---
title: "The comparison of output of SAS AND R for mixed model"
output: html_document
date: '2023-09-19'
author: Daniel He
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The comparison of output of SAS AND R for mixed model
### Data process
```{r}
library(nlme)
setwd("C:\\Users\\hed2\\OneDrive - National Institutes of Health\\Mixed model by SAS and R")
head(Orthodont)
write.csv(Orthodont,  "Orthodont.csv" ,na="",row.names = FALSE)
```

### Fit the models with unstructured covariance structure- default
```{r}
# fit the models with unstructured covariance structure of random effect
model1 <- lme(distance ~ age+Sex, data=Orthodont, random= list(Subject = pdSymm(~ 1+age+Sex ))) 
summary(model1)
# https://rpubs.com/samuelkn/CovarianceStructuresInR

model2 <- lme(distance ~ age + Sex  , random = ~1+age+Sex  | Subject, data = Orthodont) 
summary(model2)
```

### In SAS, the corresponding agument: type = un
```{r}
# PROC IMPORT DATAFILE="...\Mixed model by SAS and R\orthodont.csv" DBMS=csv
# OUT=orthodont;
# RUN;

# PROC MIXED DATA=orthodont ;
# CLASS sex  subject   ;
# MODEL distance =  sex age /solution;
# RANDOM  sex age/ subject=subject type=un ;
# RUN; QUIT;
```

### Fit the models with compound symmetric covariance structure 
```{r}
model11 <- lme(distance ~ age+Sex, data=Orthodont, random= list(Subject = pdCompSymm(~ 1+age+Sex ))) 
summary(model1)
# https://stats.stackexchange.com/questions/26556/equivalent-mixed-models-yielding-different-results-in-sas
```

### In SAS, the corresponding type = cs
```{r}
# PROC MIXED DATA=orthodont ;
# CLASS sex  subject   ;
# MODEL distance =  sex age /solution;
# RANDOM  sex age/ subject=subject type=CS ;
# RUN; QUIT;
```


### Other covariance structures

| Name|	nlme	| 
|----|-----|
| Compound symmetry	|corCompSymm	|
| AR1	|corAR1	|
| CAR1	|corCAR1	|
| Unstructured	|CorSymm|

```{r}
# Many covariance structures are reasonable for these data, such as unstructured (TYPE=UN), autoregressive (TYPE=AR(1)), compound symmetric (TYPE=CS), Toeplitz (TYPE=TOEP), etc. 
# http://support.sas.com/kb/37/107.html
```

**Variance Components (VC) is the default setting in proc mixed in SAS, but unstructured (UN) in R.**
 
