---
title: "Statistical Analysis Plan (SAP) for Epidemiologic Studies"
author: "[Author / Study Team]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
params:
  study_title: "[Study Title]"
  protocol_version: "[Protocol Version]"
  sap_version: "1.0"
  data_cutoff_date: "[YYYY-MM-DD]"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# 1. Document Information

**Study title:** `r params$study_title`  
**Protocol version/date:** `r params$protocol_version`  
**SAP version:** `r params$sap_version`  
**Data cutoff date:** `r params$data_cutoff_date`  
**Prepared by:** [Name / Role]  
**Reviewed by:** [Name / Role]  
**Approved by:** [Name / Role]

## 1.1 Purpose

This Statistical Analysis Plan (SAP) prespecifies the statistical methods for the study before final outcome analysis. The purpose is to reduce data-driven analytical decisions, improve reproducibility, and provide a transparent framework for estimating and interpreting epidemiologic associations or effects.

This template is designed primarily for observational epidemiologic studies, including:

- prospective cohort studies,
- retrospective cohort studies,
- case-control studies,
- cross-sectional studies,
- registry studies,
- electronic health record (EHR) studies,
- claims-based studies,
- survey-linked outcome studies,
- and other real-world data studies.

Sections not relevant to a specific study should be marked **Not Applicable (N/A)** rather than silently omitted.

## 1.2 SAP Scope

This SAP covers:

- study objectives and hypotheses,
- target population and analytic cohort,
- exposure, outcome, covariate, and time definitions,
- estimand / target effect,
- data quality assessment,
- descriptive analyses,
- primary and secondary statistical models,
- confounding control,
- missing data,
- model diagnostics,
- subgroup and interaction analyses,
- propensity-score and weighting analyses when applicable,
- sensitivity analyses,
- multiplicity considerations,
- robustness analyses,
- tables, figures, and listings,
- and interpretation principles.

# 2. Study Background and Rationale

Provide a concise scientific and epidemiologic rationale.

**Background:**  
[Describe the disease, exposure/intervention, public health relevance, and prior evidence.]

**Knowledge gap:**  
[Describe what remains uncertain and why the current study is needed.]

**Primary research question:**  
Among **[target population]**, is **[exposure]** associated with / causally related to **[outcome]** over **[follow-up period]**, compared with **[reference exposure]**?

If the study is explicitly causal, define the causal contrast and the assumptions required for causal interpretation. If those assumptions are not supportable, results will be described as adjusted associations rather than causal effects.

# 3. Study Objectives and Hypotheses

## 3.1 Primary Objective

To estimate the association/effect of **[primary exposure]** on **[primary outcome]** in **[target population]**.

## 3.2 Secondary Objectives

1. To estimate associations/effects for secondary outcomes.
2. To evaluate dose-response or exposure-response relationships, if applicable.
3. To examine effect modification across prespecified subgroups.
4. To evaluate robustness under alternative modeling and confounding-control assumptions.

## 3.3 Exploratory Objectives

[Specify exploratory analyses. These analyses will be clearly distinguished from confirmatory analyses.]

## 3.4 Statistical Hypotheses

For a two-sided primary comparison:

\[
H_0: \theta = \theta_0
\]

\[
H_A: \theta \neq \theta_0
\]

where \(\theta\) is the prespecified target estimand (for example, a hazard ratio, odds ratio, risk ratio, risk difference, mean difference, rate ratio, or marginal effect).

For ratio measures, \(\theta_0 = 1\). For difference measures, \(\theta_0 = 0\).

# 4. Study Design and Data Source

## 4.1 Study Design

**Design:** [Prospective cohort / retrospective cohort / case-control / cross-sectional / nested case-control / case-cohort / registry-based / other]

Describe:

- calendar study period,
- enrollment or cohort-entry period,
- duration of follow-up,
- index date / time zero,
- exposure ascertainment window,
- baseline covariate ascertainment window,
- outcome ascertainment window,
- and censoring rules.

## 4.2 Data Source(s)

Specify each source and its role:

| Data source | Population coverage | Key variables | Date range | Linkage method |
|---|---|---|---|---|
| [Source 1] | [ ] | [ ] | [ ] | [ ] |
| [Source 2] | [ ] | [ ] | [ ] | [ ] |

Potential sources include cohort databases, EHRs, claims, disease registries, surveys, laboratory systems, pharmacy data, vital records, and linked administrative data.

## 4.3 Data Freeze / Cutoff

The primary analysis will use data available through **[date]**. Any records received after this cutoff will not be incorporated into the primary analysis unless the SAP is formally amended before unblinded/final outcome analysis.

# 5. Target Population, Study Cohort, and Analysis Populations

## 5.1 Target Population

Define the population to which inference is intended to generalize.

## 5.2 Source Population

Define the population represented by the data source.

## 5.3 Inclusion Criteria

1. [Criterion 1]
2. [Criterion 2]
3. [Criterion 3]

## 5.4 Exclusion Criteria

1. [Criterion 1]
2. [Criterion 2]
3. [Criterion 3]

Eligibility will be determined without reference to post-index outcomes whenever possible.

## 5.5 Cohort Construction

Provide a reproducible flow from the source population to the final analytic cohort.

Recommended flow:

1. Source population
2. Eligible age / enrollment criteria
3. Required baseline information
4. Exposure eligibility
5. Outcome-free status at baseline, if appropriate
6. Minimum follow-up requirements, if scientifically justified
7. Exclusions for implausible or duplicate records
8. Final analytic cohort

## 5.6 Analysis Populations

Where appropriate, define:

- **Primary analysis population:** all participants meeting prespecified eligibility and variable requirements.
- **Complete-case population:** participants with complete data for variables required by a specified analysis.
- **Imputed analysis population:** participants included after multiple imputation.
- **Propensity-score overlap population:** participants meeting prespecified positivity/common-support criteria.
- **Sensitivity-analysis populations:** any alternative cohorts used to evaluate robustness.

# 6. Definition of Time, Exposure, Outcome, and Follow-up

## 6.1 Time Zero / Index Date

Time zero is **[definition]**.

Exposure classification and baseline covariates must be defined relative to this time point. Temporal alignment should minimize immortal-time bias and prevent use of post-outcome information as baseline information.

## 6.2 Exposure Definition

**Primary exposure:** [Definition]

Specify:

- source variable(s),
- coding,
- measurement window,
- reference group,
- binary/categorical/continuous representation,
- cumulative or time-varying definition,
- lag period if used,
- treatment/exposure switching rules,
- and handling of ambiguous or missing exposure status.

Example coding:

| Exposure category | Operational definition | Analysis code |
|---|---|---|
| Reference | [ ] | 0 |
| Exposed | [ ] | 1 |

For continuous exposures, specify the unit of interpretation (e.g., per 5 units, per SD, log-transformed) and whether nonlinearity will be modeled using restricted cubic splines or another prespecified method.

## 6.3 Outcome Definition

**Primary outcome:** [Definition]

Specify:

- outcome algorithm,
- source,
- first-event vs recurrent-event definition,
- validation status,
- adjudication rules if any,
- event date assignment,
- and competing events if relevant.

**Secondary outcomes:** [List]

## 6.4 Follow-up

Follow-up begins at **[time zero]** and ends at the earliest of:

1. occurrence of the outcome,
2. death, if not the outcome,
3. loss of enrollment / loss to follow-up,
4. end of data availability,
5. administrative study end,
6. or another prespecified censoring event.

Censoring assumptions and potential informative censoring will be evaluated as described in the sensitivity-analysis section.

# 7. Estimand / Target Effect

For each key objective, define the target estimand explicitly.

## 7.1 Primary Estimand

**Population:** [Target population]  
**Exposure contrast:** [Exposure A vs B / per unit increase]  
**Outcome:** [Outcome]  
**Time horizon:** [e.g., 5 years]  
**Summary measure:** [HR / RR / OR / RD / mean difference / RMST difference / incidence rate ratio]  
**Handling of competing events:** [strategy]  
**Handling of exposure changes/intercurrent events:** [strategy]  
**Covariate-adjustment target:** [conditional or marginal estimand]

The analysis should distinguish clearly between a **conditional regression coefficient** and a **marginal population-level effect** when those quantities differ.

CVD risk assuming everyone has diabetes
CVD risk assuming no one has diabetes

## 7.2 Secondary Estimands

[List each estimand using the same structure.]

# 8. Covariates and Confounding Control Strategy

## 8.1 Prespecified Covariates

Candidate confounders will be identified primarily using:

- subject-matter knowledge,
- prior epidemiologic evidence,
- temporal ordering,
- causal diagrams / DAGs,
- and the target causal contrast.

Variables will not be selected solely because their univariable p-values are statistically significant.

Create a covariate specification table:

| Covariate | Role | Definition/window | Coding | Missing-data approach | Primary model? |
|---|---|---|---|---|---|
| Age | Confounder | Baseline | Continuous | [ ] | Yes |
| Sex | Confounder | Baseline | Categorical | [ ] | Yes |
| [ ] | [ ] | [ ] | [ ] | [ ] | [ ] |

## 8.2 Variables Generally Avoided in Primary Confounding Adjustment

Unless required by the estimand, the primary adjustment set should generally avoid:

- mediators / post-exposure variables,
- post-treatment variables,
- variables measured after the outcome,
- colliders,
- instruments that add variance without controlling confounding,
- and variables whose temporal relationship to exposure/outcome makes adjustment inappropriate.

If a variable has uncertain causal status, alternative adjustment sets may be examined in sensitivity analyses.

## 8.3 Functional Form of Continuous Covariates

Continuous covariates will be modeled as [linear / spline / categorized based on clinically justified cut points].

When restricted cubic splines are used, specify the number and location of knots before outcome analysis.

# 9. Data Preparation and Quality Assessment

Before inferential modeling, perform structured data-quality checks.

## 9.1 Record-Level Checks

- duplicate participant IDs,
- duplicate encounters/events,
- inconsistent dates,
- impossible temporal ordering,
- invalid category codes,
- and unexpected repeated records.

## 9.2 Range and Plausibility Checks

For continuous variables, assess:

- minimum/maximum,
- percentiles,
- impossible values,
- units,
- digit preference/heaping,
- and extreme outliers.

Values determined to be data errors will be corrected only using documented source information. Otherwise, values will be retained, set to missing, truncated, or excluded according to prespecified rules.

## 9.3 Missingness Assessment

Summarize missingness by:

- variable,
- exposure group,
- outcome status when appropriate,
- study period,
- site/data source,
- and other relevant strata.

## 9.4 Exposure and Outcome Misclassification

Assess known limitations in measurement sensitivity, specificity, coding changes, diagnostic algorithms, or self-report. Quantitative bias analysis may be considered when suitable external information is available.

## 9.5 Follow-up Completeness

Summarize follow-up duration, censoring reasons, loss to follow-up, and person-time by exposure group.

# 10. Descriptive Analyses

## 10.1 Participant Flow

Report counts through each cohort-construction step.

## 10.2 Baseline Characteristics

Summarize baseline characteristics overall and by exposure group.

For continuous variables:

- mean and SD when approximately symmetric,
- median and IQR when skewed,
- minimum and maximum where useful.

For categorical variables:

- frequency and percentage.

## 10.3 Standardized Mean Differences

When comparing exposure groups, standardized mean differences (SMDs) will be presented when useful. SMDs are descriptive and will not determine confounder inclusion by themselves.

An absolute SMD of 0.10 may be used as a practical indicator of imbalance/balance, particularly for propensity-score diagnostics, but it is not a causal criterion for covariate selection.

## 10.4 Outcome Incidence

Depending on design and outcome type, summarize:

- number and proportion with outcome,
- incidence proportion,
- person-time,
- incidence rate,
- median follow-up,
- Kaplan-Meier estimates,
- cumulative incidence in competing-risk settings.

# 11. General Statistical Principles

## 11.1 Significance Level

Unless otherwise specified, statistical tests will be two-sided with \(\alpha = 0.05\). Estimates will be presented with 95% confidence intervals.

## 11.2 Emphasis on Estimation

Primary interpretation will emphasize:

- effect estimates,
- confidence intervals,
- clinical/public-health magnitude,
- consistency across analyses,
- and study limitations,

rather than statistical significance alone.

## 11.3 Precision and Rounding

Specify presentation precision before final output production.

Recommended defaults:

- percentages: 1 decimal place,
- effect estimates: 2 decimal places,
- confidence limits: 2 decimal places,
- p-values: 3 decimals; report `<0.001` when smaller.

## 11.4 Software

Analyses will be conducted using **R version [x.x.x]** and/or **SAS version [x.x]**. Key package/library versions will be archived for reproducibility.

# 12. Primary Statistical Analysis

The primary method will be selected according to the outcome scale and target estimand.

## 12.1 Binary Outcomes

Possible primary models:

- logistic regression for odds ratios,
- log-binomial regression for risk ratios,
- modified Poisson regression with robust variance for risk ratios,
- binomial identity-link model for risk differences when estimable.

Generic model:

\[
g\{E(Y_i)\} = \beta_0 + \beta_1 X_i + \boldsymbol{\beta}_2^T \mathbf{C}_i
\]

where \(X_i\) is the exposure and \(\mathbf{C}_i\) is the prespecified adjustment set.

## 12.2 Continuous Outcomes

Possible models:

- linear regression,
- generalized linear models,
- robust regression if prespecified and justified.

Report adjusted mean differences or marginal contrasts as appropriate.

## 12.3 Count / Rate Outcomes

Possible models:

- Poisson regression,
- negative binomial regression when overdispersion is present,
- recurrent-event models where scientifically appropriate.

Include log person-time as an offset for incidence-rate analyses.

## 12.4 Time-to-Event Outcomes

The default primary model may be Cox proportional hazards regression:

\[
h(t|X,C)=h_0(t)\exp\{\beta_1X+\boldsymbol{\beta}_2^T C\}
\]

Report hazard ratios with 95% confidence intervals.

If the proportional hazards assumption is not appropriate for the exposure effect, prespecified alternatives may include:

- time-varying coefficients,
- stratified Cox models,
- piecewise effects,
- accelerated failure-time models,
- restricted mean survival time (RMST) contrasts,
- or flexible parametric survival models.

## 12.5 Repeated / Longitudinal Outcomes

Possible models:

- linear mixed-effects models,
- generalized linear mixed models,
- generalized estimating equations (GEE),
- marginal structural models for time-varying exposure/confounding where appropriate.

Specify:

- fixed effects,
- random effects,
- covariance structure,
- time representation,
- exposure-by-time interaction,
- and target contrast.

## 12.6 Case-Control Studies

Use unconditional logistic regression unless matching requires conditional logistic regression.

Odds ratios will be interpreted in relation to the sampling design. Incidence-density sampling may permit the odds ratio to estimate an incidence rate ratio under appropriate assumptions.

## 12.7 Cross-Sectional Studies

Because temporality may be limited, results will generally be interpreted as prevalence associations rather than causal effects unless temporal ordering is otherwise established.

# 13. Crude and Adjusted Estimates

For the primary exposure-outcome relationship, report when appropriate:

1. crude/unadjusted estimate,
2. minimally adjusted estimate,
3. primary fully adjusted estimate,
4. alternative adjusted estimate(s), if prespecified.

Differences between crude and adjusted estimates may demonstrate the impact of adjustment but will not be used by themselves to identify true confounders.

# 14. Model Diagnostics and Assumption Checks

Diagnostics will be appropriate to the selected model.

## 14.1 Cox Model

Evaluate:

- proportional hazards using Schoenfeld residuals / `cox.zph`,
- functional form for continuous variables,
- influential observations,
- sparse cells / separation where relevant,
- and overall model stability.

## 14.2 Linear Models

Evaluate:

- residual distribution,
- heteroskedasticity,
- influential observations,
- functional form,
- and multicollinearity.

## 14.3 Logistic / GLM Models

Evaluate:

- separation/sparse data,
- functional form of continuous predictors,
- influence,
- calibration when prediction is a study objective,
- and overdispersion where relevant.

Model diagnostics will inform prespecified alternative models or sensitivity analyses; they will not be used to conduct unrestricted data-driven model searching.

# 15. Propensity-Score Analyses

Propensity-score methods may be used as primary or sensitivity analyses when appropriate to the study objective.

## 15.1 Propensity-Score Model

Estimate:

\[
e(C)=P(X=1|C)
\]

using prespecified baseline confounders.

Variables should be selected for confounding control, not solely based on statistical prediction of exposure.

## 15.2 Weighting Strategy

Specify the estimand:

- ATE: inverse probability of treatment weighting,
- ATT: weighting targeted to the exposed/treated,
- overlap weighting,
- standardized mortality ratio weighting,
- or another prespecified target.

For ATE IPTW:

\[
w_i = \frac{X_i}{e(C_i)} + \frac{1-X_i}{1-e(C_i)}
\]

Stabilized weights may be preferred for improved precision.

## 15.3 Positivity / Overlap

Assess:

- propensity-score distributions by exposure,
- extreme propensity scores,
- extreme weights,
- effective sample size,
- and regions of poor common support.

Prespecify truncation/winsorization thresholds, for example the 1st/99th or 0.5th/99.5th percentiles, if trimming is planned.

## 15.4 Covariate Balance

Assess weighted SMDs and graphical diagnostics (e.g., Love plots). An absolute SMD < 0.10 may be considered acceptable measured balance.

## 15.5 Weighted Outcome Model

Use weighted regression appropriate to the outcome with robust/sandwich standard errors as required.

Measured balance does not demonstrate absence of unmeasured confounding.

# 16. Time-Varying Exposure and Time-Varying Confounding

If exposure changes over time, specify:

- update frequency,
- exposure grace periods,
- cumulative exposure definitions,
- lagging rules,
- and censoring at switching if applicable.

When time-varying confounders are affected by prior exposure, conventional regression adjustment may introduce bias. Marginal structural models with inverse-probability weights may be used if prespecified and scientifically justified.

# 17. Missing Data

## 17.1 Description

Report the amount and pattern of missing data for exposure, outcome, covariates, and follow-up variables.

## 17.2 Primary Missing-Data Approach

Specify one primary strategy:

- complete-case analysis,
- multiple imputation (MI),
- inverse-probability weighting for missingness,
- maximum-likelihood methods inherent in longitudinal models,
- or another justified method.

## 17.3 Multiple Imputation

When MI is used:

- include exposure, outcome, covariates, and predictors of missingness,
- respect variable distributions and bounds,
- include transformations/interactions needed by the analysis model,
- use a sufficient number of imputations,
- combine estimates using Rubin's rules.

Imputation should be consistent with the temporal structure of the study and should not inadvertently use future information in a way incompatible with the estimand.

## 17.4 Missing-Not-at-Random Sensitivity Analyses

When clinically relevant, evaluate departures from MAR using:

- delta adjustment,
- pattern-mixture models,
- selection models,
- tipping-point analyses,
- or scenario-based analyses.

# 18. Censoring, Competing Risks, and Loss to Follow-up

## 18.1 Censoring

List all censoring mechanisms and justify non-informative censoring assumptions.

## 18.2 Informative Loss to Follow-up

Potential approaches include inverse-probability-of-censoring weights (IPCW) or sensitivity analyses under plausible missing-outcome scenarios.

## 18.3 Competing Risks

If competing events preclude the outcome, prespecify whether the scientific question targets:

- cause-specific hazards,
- subdistribution hazards,
- cumulative incidence functions,
- risk contrasts at a fixed time,
- or a composite endpoint.

Cause-specific Cox and Fine-Gray models answer different questions and should not be treated as interchangeable.

# 19. Effect Modification, Interaction, and Subgroup Analyses

Prespecified effect modifiers may include:

- age,
- sex,
- race/ethnicity when scientifically justified and appropriately interpreted,
- disease severity,
- baseline comorbidity,
- calendar period,
- or other clinically justified factors.

Effect modification will preferably be evaluated by an interaction term rather than separate within-subgroup significance tests.

Example:

\[
g\{E(Y)\}=\beta_0+\beta_1X+\beta_2Z+\beta_3(X\times Z)+\cdots
\]

Report stratum-specific estimates and the interaction p-value or confidence interval for the interaction contrast.

Subgroup analyses are generally considered supportive or exploratory unless explicitly designated as confirmatory.

# 20. Dose-Response and Nonlinear Exposure Effects

For continuous or ordinal exposures, assess the prespecified form using one or more of:

- linear trend,
- clinically defined categories,
- restricted cubic splines,
- fractional polynomials,
- or generalized additive models when exploratory.

If categories are used, cut points should preferably be prespecified or clinically motivated rather than chosen from observed outcome data.

# 21. Multiplicity

## 21.1 Primary Objective

If there is one primary exposure-outcome estimand, no multiplicity adjustment is needed for that single primary test.

## 21.2 Multiple Primary Endpoints / Exposures / Contrasts

When more than one confirmatory primary hypothesis is tested, specify a control strategy such as:

- Bonferroni,
- Holm,
- Hochberg,
- hierarchical/gatekeeping testing,
- false-discovery-rate control when appropriate for the research objective,
- or no formal adjustment with explicit exploratory interpretation.

## 21.3 Secondary and Exploratory Analyses

Unless otherwise prespecified, secondary and exploratory p-values will be interpreted descriptively with emphasis on effect sizes, uncertainty, and consistency rather than binary significance claims.

# 22. Sensitivity and Robustness Analyses

Sensitivity analyses should target specific assumptions rather than simply repeat the analysis using arbitrary alternatives.

Recommended categories include:

## 22.1 Alternative Confounder Adjustment Sets

- primary DAG-based adjustment set,
- expanded adjustment set,
- reduced set excluding variables with uncertain temporal/causal status.

## 22.2 Alternative Confounding-Control Methods

Compare, where appropriate:

- conventional multivariable regression,
- IPTW,
- overlap weighting,
- matching,
- standardization / g-computation,
- doubly robust estimators,
- TMLE or other causal estimators.

## 22.3 Propensity-Score Weight Truncation

Evaluate alternative truncation thresholds.

## 22.4 Exposure Definition

Examples:

- stricter exposure definition,
- alternative look-back window,
- cumulative exposure,
- lagged exposure,
- exclusion of uncertain exposure classification.

## 22.5 Outcome Definition

Examples:

- high-specificity outcome algorithm,
- broad/sensitive algorithm,
- adjudicated subset,
- alternative event-date rule.

## 22.6 Reverse Causation / Latency

Use lag analyses or exclude events occurring shortly after baseline when scientifically appropriate and prespecified.

## 22.7 Missing Data

Compare complete-case, MI, and MNAR scenarios as appropriate.

## 22.8 Informative Censoring

Use IPCW or scenario analyses.

## 22.9 Unmeasured Confounding

When appropriate, consider:

- E-values,
- negative-control exposures/outcomes,
- quantitative bias analysis,
- bias functions,
- or external-control calibration.

## 22.10 Model Specification

Evaluate alternative functional forms, distributions, correlation structures, or survival-model assumptions.

# 23. Negative Controls and Falsification Analyses

If available, prespecify negative-control outcomes or exposures that should not plausibly be caused by the exposure. Unexpected associations may indicate residual bias, measurement artifacts, or uncontrolled confounding.

# 24. Selection Bias and Generalizability

Evaluate whether inclusion, retention, healthcare access, testing, referral, or data capture may depend jointly on exposure and outcome determinants.

If sampling weights are available, specify whether survey/design weights will be incorporated.

Compare the analytic cohort with the target/source population when possible and describe limits to transportability/generalizability.

# 25. Survey and Complex Sampling Designs

For complex surveys, incorporate as appropriate:

- sampling weights,
- strata,
- primary sampling units,
- finite population corrections,
- and survey-specific variance estimation.

Analyses must account for the design rather than treating observations as a simple random sample.

# 26. Clustered and Multilevel Data

For clustering by site, family, physician, hospital, geographic area, or repeated person-level observations, use methods appropriate to the estimand, such as:

- cluster-robust standard errors,
- GEE,
- random-effects/multilevel models,
- frailty models,
- or fixed effects.

Specify the cluster level and correlation structure.

# 27. Small-Sample, Sparse-Data, and Separation Issues

When sparse data may cause instability, prespecify alternatives such as:

- exact methods,
- Firth penalized logistic regression,
- penalized likelihood,
- category collapsing based on clinical logic,
- or reporting descriptive estimates without overfitted adjustment.

Avoid fitting models with inadequate information relative to model complexity.

# 28. Outliers and Influential Observations

Outliers will not be removed solely because they affect statistical significance.

Potential influential observations will be:

1. verified for data quality,
2. retained if valid in the primary analysis unless a prespecified exclusion rule applies,
3. evaluated in sensitivity analyses when materially influential.

# 29. Model Selection and Data-Driven Procedures

The primary model and adjustment set will be prespecified.

Automated stepwise procedures should not be used to define the primary confounder set. If machine-learning methods are used for nuisance-model estimation, their role and tuning/cross-fitting procedures should be prespecified and separated from the causal estimand definition.

# 30. Machine Learning and High-Dimensional Adjustment (Optional)

For high-dimensional EHR/claims settings, prespecified approaches may include:

- high-dimensional propensity scores,
- penalized regression,
- Super Learner,
- double/debiased machine learning,
- targeted maximum likelihood estimation (TMLE),
- causal forests for exploratory heterogeneity.

These approaches do not replace careful specification of time zero, exposure, outcome, target population, causal structure, and estimand.

# 31. Sample Size, Precision, and Power

Observational studies may have a fixed available sample size. The SAP should state whether the study is:

- **precision-based**,
- **power-based**,
- or **database-size constrained**.

## 31.1 Precision-Based Planning

Specify the anticipated number of exposed/unexposed participants, outcome events, and expected confidence-interval width for the primary effect.

## 31.2 Power-Based Planning

Specify:

- effect size,
- event rate / variance,
- exposure prevalence or group ratio,
- alpha,
- power,
- attrition/loss to follow-up,
- and design effect if clustered.

## 31.3 Effective Sample Size for Weighted Analyses

For weighting analyses, report effective sample size and recognize that extreme weights may substantially reduce precision.

# 32. Interim Analyses and Data-Dependent Adaptation

Usually **N/A** for standard observational epidemiologic analyses.

If interim looks, sequential surveillance, or adaptive data collection are planned, specify stopping rules and error-control considerations before analysis begins.

# 33. Statistical Tables, Figures, and Listings

## 33.1 Planned Tables

**Table 1.** Cohort construction / participant flow  
**Table 2.** Baseline characteristics overall and by exposure  
**Table 3.** Missingness summary  
**Table 4.** Outcome incidence / person-time by exposure  
**Table 5.** Crude and adjusted primary effect estimates  
**Table 6.** Secondary outcome analyses  
**Table 7.** Prespecified subgroup / interaction analyses  
**Table 8.** Propensity-score balance diagnostics  
**Table 9.** Sensitivity-analysis comparison

## 33.2 Planned Figures

**Figure 1.** Study cohort flow diagram  
**Figure 2.** Exposure distribution  
**Figure 3.** Kaplan-Meier / cumulative incidence curve, if applicable  
**Figure 4.** Propensity-score overlap plot, if applicable  
**Figure 5.** Love plot / covariate balance, if applicable  
**Figure 6.** Forest plot of main and sensitivity estimates  
**Figure 7.** Restricted cubic spline exposure-response curve, if applicable

# 34. Recommended R Analysis Framework

The following code skeleton is intentionally generic and should be adapted to the specific study.

```{r packages, eval=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tableone)
library(survival)
library(cobalt)
library(survey)
library(splines)
library(broom)
```

## 34.1 Cohort Definition

```{r cohort-template, eval=FALSE}
analysis_data <- raw_data %>%
  filter(
    eligibility_flag == 1,
    !is.na(exposure),
    index_date >= study_start,
    index_date <= study_end
  )
```

## 34.2 Data Quality Checks

```{r dq-template, eval=FALSE}
# Duplicate IDs
sum(duplicated(analysis_data$id))

# Missingness
colSums(is.na(analysis_data))

# Distributions
summary(analysis_data)
```

## 34.3 Baseline Table

```{r table1-template, eval=FALSE}
vars <- c(
  "age",
  "sex",
  "bmi",
  "comorbidity_1",
  "comorbidity_2"
)

tab1 <- CreateTableOne(
  vars = vars,
  strata = "exposure",
  data = analysis_data,
  test = FALSE
)

print(tab1, smd = TRUE)
```

## 34.4 Binary Outcome Model

```{r binary-template, eval=FALSE}
fit_binary <- glm(
  outcome ~ exposure + age + sex + bmi + confounder_1,
  family = binomial(),
  data = analysis_data
)

summary(fit_binary)
```

## 34.5 Continuous Outcome Model

```{r continuous-template, eval=FALSE}
fit_continuous <- lm(
  outcome ~ exposure + age + sex + bmi + confounder_1,
  data = analysis_data
)

summary(fit_continuous)
```

## 34.6 Time-to-Event Model

```{r cox-template, eval=FALSE}
fit_cox <- coxph(
  Surv(followup_time, event) ~
    exposure + age + sex + bmi + confounder_1,
  data = analysis_data
)

summary(fit_cox)

# PH assumption
ph_test <- cox.zph(fit_cox)
print(ph_test)
plot(ph_test)
```

## 34.7 Propensity Score and IPTW

```{r iptw-template, eval=FALSE}
ps_fit <- glm(
  exposure ~ age + sex + bmi + confounder_1 + confounder_2,
  family = binomial(),
  data = analysis_data
)

analysis_data <- analysis_data %>%
  mutate(
    ps = predict(ps_fit, type = "response"),
    p_exp = mean(exposure == 1),
    sw = ifelse(
      exposure == 1,
      p_exp / ps,
      (1 - p_exp) / (1 - ps)
    )
  )

# Prespecified truncation example
limits <- quantile(analysis_data$sw, c(0.01, 0.99), na.rm = TRUE)

analysis_data <- analysis_data %>%
  mutate(
    sw_trim = pmin(pmax(sw, limits[1]), limits[2])
  )

bal.tab(
  exposure ~ age + sex + bmi + confounder_1 + confounder_2,
  data = analysis_data,
  weights = analysis_data$sw_trim,
  method = "weighting",
  un = TRUE,
  thresholds = c(m = 0.10)
)
```

## 34.8 Weighted Cox Model

```{r weighted-cox-template, eval=FALSE}
fit_iptw <- coxph(
  Surv(followup_time, event) ~ exposure,
  data = analysis_data,
  weights = sw_trim,
  robust = TRUE
)

summary(fit_iptw)
```

## 34.9 Restricted Cubic / Natural Spline Example

```{r spline-template, eval=FALSE}
fit_spline <- coxph(
  Surv(followup_time, event) ~
    ns(exposure_continuous, df = 4) +
    age + sex + bmi + confounder_1,
  data = analysis_data
)
```

# 35. Primary and Sensitivity Result Synthesis

Create one summary table containing all major estimates with the same target contrast.

Example structure:

| Analysis | Estimand/measure | Estimate | 95% CI | Key assumption changed |
|---|---|---:|---|---|
| Crude | HR | [ ] | [ ] | No adjustment |
| Primary adjusted | HR | [ ] | [ ] | Primary confounder set |
| IPTW | Marginal HR / target measure | [ ] | [ ] | PS weighting |
| Alternative adjustment | HR | [ ] | [ ] | Covariate set |
| Missing-data sensitivity | HR | [ ] | [ ] | Missingness assumption |

Consistency across analyses will be evaluated based on effect direction, magnitude, precision, and the assumptions each analysis changes—not solely on whether individual p-values cross 0.05.

# 36. Interpretation Framework

The final interpretation will distinguish among:

1. **Crude association** — observed exposure-outcome relationship without confounder adjustment.
2. **Adjusted association** — association conditional on measured covariates.
3. **Causal effect estimate** — only when design, temporal ordering, consistency, exchangeability, positivity, correct model specification, and measurement assumptions are sufficiently justified.

Results will not automatically be described as causal merely because regression, propensity scores, or weighting were used.

# 37. Key Sources of Bias and Limitations to Address

At minimum, discuss:

- unmeasured confounding,
- residual confounding,
- exposure misclassification,
- outcome misclassification,
- general measurement error,
- uncertain temporal ordering,
- reverse causation,
- selection bias,
- informative censoring,
- missing data,
- positivity / overlap violations,
- model misspecification,
- immortal-time bias where relevant,
- time-varying confounding,
- competing risks,
- sparse-data bias,
- multiple testing / selective reporting,
- data provenance and coding changes,
- and limited generalizability / transportability.

# 38. Deviations from the SAP

Any material deviation from this SAP after finalization will be documented with:

- date,
- description,
- rationale,
- whether the decision was made before or after examining outcome-related results,
- and potential impact on interpretation.

Post hoc analyses will be clearly labeled exploratory.

# 39. Reproducibility and Quality Control

The study team will maintain:

- version-controlled analysis code,
- frozen analytic datasets or reproducible data-extraction specifications,
- variable derivation documentation,
- analysis logs,
- package/software versions,
- independent or secondary QC for key endpoints and primary models where feasible,
- and reproducible generation of final tables and figures.

# 40. SAP Sign-Off

| Role | Name | Signature | Date |
|---|---|---|---|
| Lead Statistician | [ ] | [ ] | [ ] |
| Epidemiologist / PI | [ ] | [ ] | [ ] |
| Data Manager / Analyst | [ ] | [ ] | [ ] |
| Other Reviewer | [ ] | [ ] | [ ] |

# Appendix A. Study-Specific Analysis Decision Table

Complete this table before final analysis.

| Item | Prespecified decision |
|---|---|
| Primary research question | [ ] |
| Study design | [ ] |
| Target population | [ ] |
| Time zero | [ ] |
| Primary exposure | [ ] |
| Exposure reference | [ ] |
| Primary outcome | [ ] |
| Follow-up end | [ ] |
| Primary estimand | [ ] |
| Primary effect measure | [ ] |
| Primary confounder set | [ ] |
| Primary statistical model | [ ] |
| Continuous-variable functional forms | [ ] |
| Missing-data method | [ ] |
| Censoring strategy | [ ] |
| Competing-risk strategy | [ ] |
| PS method, if any | [ ] |
| Weight truncation, if any | [ ] |
| Prespecified effect modifiers | [ ] |
| Multiplicity strategy | [ ] |
| Sensitivity analyses | [ ] |
| Software/version | [ ] |
| Data cutoff | [ ] |

# Appendix B. Outcome-Specific Model Selection Guide

| Outcome type | Common primary model | Typical effect measure | Common alternatives |
|---|---|---|---|
| Binary | Logistic / modified Poisson | OR / RR / RD | GEE, log-binomial |
| Continuous | Linear regression | Mean difference | GLM, robust regression |
| Count/rate | Poisson / negative binomial | Rate ratio | GEE, recurrent-event models |
| Time-to-event | Cox PH | HR | RMST, AFT, flexible parametric |
| Repeated continuous | LMM / GEE | Mean contrast | Marginal models |
| Repeated binary | GEE / GLMM | OR / RR | Marginal standardization |
| Competing risk | Cause-specific Cox / Fine-Gray | Cause-specific HR / subdistribution HR | CIF risk contrast |

# Appendix C. Minimum Sensitivity Analysis Matrix

| Assumption | Primary approach | Suggested sensitivity analysis |
|---|---|---|
| Confounder specification | DAG/subject-matter set | Expanded/reduced set |
| Missing covariates | MI or complete case | Alternative missingness model |
| Exposure misclassification | Main algorithm | Restrictive/high-specificity definition |
| Outcome misclassification | Main algorithm | Validated/high-specificity definition |
| Positivity | Primary PS approach | Trimming/overlap weighting |
| Informative censoring | Standard censoring | IPCW |
| Reverse causation | Main lag | Longer lag/exclude early events |
| PH assumption | Cox | Time-varying effect/RMST |
| Unmeasured confounding | Standard adjustment | E-value/QBA/negative controls |

# Appendix D. Example Language for Reporting

**Adjusted association wording:**  
"After adjustment for prespecified baseline covariates, the exposure was associated with [higher/lower] [outcome] ([effect measure] = [estimate], 95% CI [lower, upper])."

**Causal wording when assumptions are explicitly supported:**  
"Under the stated assumptions of exchangeability, consistency, positivity, correct model specification, and adequate measurement of exposure, outcome, and confounders, the estimate may be interpreted as the causal effect of [exposure contrast] on [outcome]."

**Cautious observational wording:**  
"Because this is an observational study, residual and unmeasured confounding, measurement error, selection bias, and uncertainty in temporal ordering may remain; therefore, the estimate should not be interpreted as definitive proof of causality."

