---
title: "Statistical Analysis Plan (SAP) for Epidemiologic Studies"
description: "A practical R learning guide to Statistical Analysis Plan (SAP) for Epidemiologic Studies, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Resources"
category: "SAP & Protocol Resources"
topic: "Statistical Analysis Plan (SAP) for Epidemiologic Studies"
type: article
tags: ["SAP & Protocol Resources", "Resources", "Statistical", "Plan", "SAP"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
params:
  study_title: "[Study Title]"
  protocol_version: "[Protocol Version]"
  sap_version: "1.0"
  data_cutoff_date: "[YYYY-MM-DD]"
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

A statistical analysis plan translates protocol objectives into pre-specified, reproducible analytical decisions.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Statistical Analysis Plan (SAP) for Epidemiologic Studies**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. 1. Document Information
2. 1.1 Purpose
3. 1.2 SAP Scope
4. 2. Study Background and Rationale
5. 3. Study Objectives and Hypotheses
6. 3.1 Primary Objective

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Statistical Analysis Plan (SAP) for Epidemiologic Studies** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

#### 1. Document Information

<!-- generated-learning-note -->
This subsection explains **1. Document Information** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


**Study title:** `r params$study_title`  
**Protocol version/date:** `r params$protocol_version`  
**SAP version:** `r params$sap_version`  
**Data cutoff date:** `r params$data_cutoff_date`  
**Prepared by:** [Name / Role]  
**Reviewed by:** [Name / Role]  
**Approved by:** [Name / Role]

#### 1.1 Purpose

<!-- generated-learning-note -->
This subsection explains **1.1 Purpose** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


This Statistical Analysis Plan (SAP) prespecifies the statistical methods for the study before final outcome analysis. The purpose is to reduce data-driven analytical decisions, improve reproducibility, and provide a transparent framework for estimating and interpreting epidemiologic associations or effects.

This template is designed primarily for observational epidemiologic studies, including:

- prospective cohort studies,
- retrospective cohort studies,
- case-control studies,
- cross-sectional studies,
- registry studies,
- electronic health record (EHR) studies,
- claims-based studies,
- survey-linked outcome studies,
- and other real-world data studies.

Sections not relevant to a specific study should be marked **Not Applicable (N/A)** rather than silently omitted.

#### 1.2 SAP Scope

<!-- generated-learning-note -->
This subsection explains **1.2 SAP Scope** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


This SAP covers:

- study objectives and hypotheses,
- target population and analytic cohort,
- exposure, outcome, covariate, and time definitions,
- estimand / target effect,
- data quality assessment,
- descriptive analyses,
- primary and secondary statistical models,
- confounding control,
- missing data,
- model diagnostics,
- subgroup and interaction analyses,
- propensity-score and weighting analyses when applicable,
- sensitivity analyses,
- multiplicity considerations,
- robustness analyses,
- tables, figures, and listings,
- and interpretation principles.

#### 2. Study Background and Rationale

<!-- generated-learning-note -->
This subsection explains **2. Study Background and Rationale** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Provide a concise scientific and epidemiologic rationale.

**Background:**  
[Describe the disease, exposure/intervention, public health relevance, and prior evidence.]

**Knowledge gap:**  
[Describe what remains uncertain and why the current study is needed.]

**Primary research question:**  
Among **[target population]**, is **[exposure]** associated with / causally related to **[outcome]** over **[follow-up period]**, compared with **[reference exposure]**?

If the study is explicitly causal, define the causal contrast and the assumptions required for causal interpretation. If those assumptions are not supportable, results will be described as adjusted associations rather than causal effects.

#### 3. Study Objectives and Hypotheses

<!-- generated-learning-note -->
This subsection explains **3. Study Objectives and Hypotheses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


#### 3.1 Primary Objective

<!-- generated-learning-note -->
This subsection explains **3.1 Primary Objective** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


To estimate the association/effect of **[primary exposure]** on **[primary outcome]** in **[target population]**.

#### 3.2 Secondary Objectives

<!-- generated-learning-note -->
This subsection explains **3.2 Secondary Objectives** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


1. To estimate associations/effects for secondary outcomes.
2. To evaluate dose-response or exposure-response relationships, if applicable.
3. To examine effect modification across prespecified subgroups.
4. To evaluate robustness under alternative modeling and confounding-control assumptions.

#### 3.3 Exploratory Objectives

<!-- generated-learning-note -->
This subsection explains **3.3 Exploratory Objectives** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


[Specify exploratory analyses. These analyses will be clearly distinguished from confirmatory analyses.]

#### 3.4 Statistical Hypotheses

<!-- generated-learning-note -->
This subsection explains **3.4 Statistical Hypotheses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


For a two-sided primary comparison:

\[
H_0: \theta = \theta_0
\]

\[
H_A: \theta \neq \theta_0
\]

where \(\theta\) is the prespecified target estimand (for example, a hazard ratio, odds ratio, risk ratio, risk difference, mean difference, rate ratio, or marginal effect).

For ratio measures, \(\theta_0 = 1\). For difference measures, \(\theta_0 = 0\).

#### 4. Study Design and Data Source

<!-- generated-learning-note -->
This section prepares the information needed for **4. Study Design and Data Source**. Check variable definitions, units, coding, and missing values before moving to analysis.


#### 4.1 Study Design

<!-- generated-learning-note -->
This subsection explains **4.1 Study Design** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


**Design:** [Prospective cohort / retrospective cohort / case-control / cross-sectional / nested case-control / case-cohort / registry-based / other]

Describe:

- calendar study period,
- enrollment or cohort-entry period,
- duration of follow-up,
- index date / time zero,
- exposure ascertainment window,
- baseline covariate ascertainment window,
- outcome ascertainment window,
- and censoring rules.

#### 4.2 Data Source(s)

<!-- generated-learning-note -->
This section prepares the information needed for **4.2 Data Source(s)**. Check variable definitions, units, coding, and missing values before moving to analysis.


Specify each source and its role:

| Data source | Population coverage | Key variables | Date range | Linkage method |
|---|---|---|---|---|
| [Source 1] | [ ] | [ ] | [ ] | [ ] |
| [Source 2] | [ ] | [ ] | [ ] | [ ] |

Potential sources include cohort databases, EHRs, claims, disease registries, surveys, laboratory systems, pharmacy data, vital records, and linked administrative data.

#### 4.3 Data Freeze / Cutoff

<!-- generated-learning-note -->
This section prepares the information needed for **4.3 Data Freeze / Cutoff**. Check variable definitions, units, coding, and missing values before moving to analysis.


The primary analysis will use data available through **[date]**. Any records received after this cutoff will not be incorporated into the primary analysis unless the SAP is formally amended before unblinded/final outcome analysis.

#### 5. Target Population, Study Cohort, and Analysis Populations

<!-- generated-learning-note -->
This subsection explains **5. Target Population, Study Cohort, and Analysis Populations** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


#### 5.1 Target Population

<!-- generated-learning-note -->
This subsection explains **5.1 Target Population** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Define the population to which inference is intended to generalize.

#### 5.2 Source Population

<!-- generated-learning-note -->
This subsection explains **5.2 Source Population** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Define the population represented by the data source.

#### 5.3 Inclusion Criteria

<!-- generated-learning-note -->
This subsection explains **5.3 Inclusion Criteria** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


1. [Criterion 1]
2. [Criterion 2]
3. [Criterion 3]

#### 5.4 Exclusion Criteria

<!-- generated-learning-note -->
This subsection explains **5.4 Exclusion Criteria** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


1. [Criterion 1]
2. [Criterion 2]
3. [Criterion 3]

Eligibility will be determined without reference to post-index outcomes whenever possible.

#### 5.5 Cohort Construction

<!-- generated-learning-note -->
This subsection explains **5.5 Cohort Construction** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Provide a reproducible flow from the source population to the final analytic cohort.

Recommended flow:

1. Source population
2. Eligible age / enrollment criteria
3. Required baseline information
4. Exposure eligibility
5. Outcome-free status at baseline, if appropriate
6. Minimum follow-up requirements, if scientifically justified
7. Exclusions for implausible or duplicate records
8. Final analytic cohort

#### 5.6 Analysis Populations

<!-- generated-learning-note -->
This subsection explains **5.6 Analysis Populations** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Where appropriate, define:

- **Primary analysis population:** all participants meeting prespecified eligibility and variable requirements.
- **Complete-case population:** participants with complete data for variables required by a specified analysis.
- **Imputed analysis population:** participants included after multiple imputation.
- **Propensity-score overlap population:** participants meeting prespecified positivity/common-support criteria.
- **Sensitivity-analysis populations:** any alternative cohorts used to evaluate robustness.

#### 6. Definition of Time, Exposure, Outcome, and Follow-up

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **6. Definition of Time, Exposure, Outcome, and Follow-up** in the context of Statistical Analysis Plan (SAP) for Epidemiologic Studies.


#### 6.1 Time Zero / Index Date

<!-- generated-learning-note -->
This subsection explains **6.1 Time Zero / Index Date** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Time zero is **[definition]**.

Exposure classification and baseline covariates must be defined relative to this time point. Temporal alignment should minimize immortal-time bias and prevent use of post-outcome information as baseline information.

#### 6.2 Exposure Definition

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **6.2 Exposure Definition** in the context of Statistical Analysis Plan (SAP) for Epidemiologic Studies.


**Primary exposure:** [Definition]

Specify:

- source variable(s),
- coding,
- measurement window,
- reference group,
- binary/categorical/continuous representation,
- cumulative or time-varying definition,
- lag period if used,
- treatment/exposure switching rules,
- and handling of ambiguous or missing exposure status.

Example coding:

| Exposure category | Operational definition | Analysis code |
|---|---|---|
| Reference | [ ] | 0 |
| Exposed | [ ] | 1 |

For continuous exposures, specify the unit of interpretation (e.g., per 5 units, per SD, log-transformed) and whether nonlinearity will be modeled using restricted cubic splines or another prespecified method.

#### 6.3 Outcome Definition

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **6.3 Outcome Definition** in the context of Statistical Analysis Plan (SAP) for Epidemiologic Studies.


**Primary outcome:** [Definition]

Specify:

- outcome algorithm,
- source,
- first-event vs recurrent-event definition,
- validation status,
- adjudication rules if any,
- event date assignment,
- and competing events if relevant.

**Secondary outcomes:** [List]

#### 6.4 Follow-up

<!-- generated-learning-note -->
This subsection explains **6.4 Follow-up** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Follow-up begins at **[time zero]** and ends at the earliest of:

1. occurrence of the outcome,
2. death, if not the outcome,
3. loss of enrollment / loss to follow-up,
4. end of data availability,
5. administrative study end,
6. or another prespecified censoring event.

Censoring assumptions and potential informative censoring will be evaluated as described in the sensitivity-analysis section.

#### 7. Estimand / Target Effect

<!-- generated-learning-note -->
This subsection explains **7. Estimand / Target Effect** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


For each key objective, define the target estimand explicitly.

#### 7.1 Primary Estimand

<!-- generated-learning-note -->
This subsection explains **7.1 Primary Estimand** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


**Population:** [Target population]  
**Exposure contrast:** [Exposure A vs B / per unit increase]  
**Outcome:** [Outcome]  
**Time horizon:** [e.g., 5 years]  
**Summary measure:** [HR / RR / OR / RD / mean difference / RMST difference / incidence rate ratio]  
**Handling of competing events:** [strategy]  
**Handling of exposure changes/intercurrent events:** [strategy]  
**Covariate-adjustment target:** [conditional or marginal estimand]

The analysis should distinguish clearly between a **conditional regression coefficient** and a **marginal population-level effect** when those quantities differ.

CVD risk assuming everyone has diabetes
CVD risk assuming no one has diabetes

#### 7.2 Secondary Estimands

<!-- generated-learning-note -->
This subsection explains **7.2 Secondary Estimands** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


[List each estimand using the same structure.]

#### 8. Covariates and Confounding Control Strategy

<!-- generated-learning-note -->
This subsection explains **8. Covariates and Confounding Control Strategy** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


#### 8.1 Prespecified Covariates

<!-- generated-learning-note -->
This subsection explains **8.1 Prespecified Covariates** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Candidate confounders will be identified primarily using:

- subject-matter knowledge,
- prior epidemiologic evidence,
- temporal ordering,
- causal diagrams / DAGs,
- and the target causal contrast.

Variables will not be selected solely because their univariable p-values are statistically significant.

Create a covariate specification table:

| Covariate | Role | Definition/window | Coding | Missing-data approach | Primary model? |
|---|---|---|---|---|---|
| Age | Confounder | Baseline | Continuous | [ ] | Yes |
| Sex | Confounder | Baseline | Categorical | [ ] | Yes |
| [ ] | [ ] | [ ] | [ ] | [ ] | [ ] |

#### 8.2 Variables Generally Avoided in Primary Confounding Adjustment

<!-- generated-learning-note -->
This section prepares the information needed for **8.2 Variables Generally Avoided in Primary Confounding Adjustment**. Check variable definitions, units, coding, and missing values before moving to analysis.


Unless required by the estimand, the primary adjustment set should generally avoid:

- mediators / post-exposure variables,
- post-treatment variables,
- variables measured after the outcome,
- colliders,
- instruments that add variance without controlling confounding,
- and variables whose temporal relationship to exposure/outcome makes adjustment inappropriate.

If a variable has uncertain causal status, alternative adjustment sets may be examined in sensitivity analyses.

#### 8.3 Functional Form of Continuous Covariates

<!-- generated-learning-note -->
This subsection explains **8.3 Functional Form of Continuous Covariates** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Continuous covariates will be modeled as [linear / spline / categorized based on clinically justified cut points].

When restricted cubic splines are used, specify the number and location of knots before outcome analysis.

#### 9. Data Preparation and Quality Assessment

<!-- generated-learning-note -->
This section prepares the information needed for **9. Data Preparation and Quality Assessment**. Check variable definitions, units, coding, and missing values before moving to analysis.


Before inferential modeling, perform structured data-quality checks.

#### 9.1 Record-Level Checks

<!-- generated-learning-note -->
This subsection explains **9.1 Record-Level Checks** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


- duplicate participant IDs,
- duplicate encounters/events,
- inconsistent dates,
- impossible temporal ordering,
- invalid category codes,
- and unexpected repeated records.

#### 9.2 Range and Plausibility Checks

<!-- generated-learning-note -->
This subsection explains **9.2 Range and Plausibility Checks** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


For continuous variables, assess:

- minimum/maximum,
- percentiles,
- impossible values,
- units,
- digit preference/heaping,
- and extreme outliers.

Values determined to be data errors will be corrected only using documented source information. Otherwise, values will be retained, set to missing, truncated, or excluded according to prespecified rules.

#### 9.3 Missingness Assessment

<!-- generated-learning-note -->
This subsection explains **9.3 Missingness Assessment** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Summarize missingness by:

- variable,
- exposure group,
- outcome status when appropriate,
- study period,
- site/data source,
- and other relevant strata.

#### 9.4 Exposure and Outcome Misclassification

<!-- generated-learning-note -->
This subsection explains **9.4 Exposure and Outcome Misclassification** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Assess known limitations in measurement sensitivity, specificity, coding changes, diagnostic algorithms, or self-report. Quantitative bias analysis may be considered when suitable external information is available.

#### 9.5 Follow-up Completeness

<!-- generated-learning-note -->
This subsection explains **9.5 Follow-up Completeness** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Summarize follow-up duration, censoring reasons, loss to follow-up, and person-time by exposure group.

#### 10. Descriptive Analyses

<!-- generated-learning-note -->
This subsection explains **10. Descriptive Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


#### 10.1 Participant Flow

<!-- generated-learning-note -->
This subsection explains **10.1 Participant Flow** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Report counts through each cohort-construction step.

#### 10.2 Baseline Characteristics

<!-- generated-learning-note -->
This subsection explains **10.2 Baseline Characteristics** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Summarize baseline characteristics overall and by exposure group.

For continuous variables:

- mean and SD when approximately symmetric,
- median and IQR when skewed,
- minimum and maximum where useful.

For categorical variables:

- frequency and percentage.

#### 10.3 Standardized Mean Differences

<!-- generated-learning-note -->
This subsection explains **10.3 Standardized Mean Differences** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


When comparing exposure groups, standardized mean differences (SMDs) will be presented when useful. SMDs are descriptive and will not determine confounder inclusion by themselves.

An absolute SMD of 0.10 may be used as a practical indicator of imbalance/balance, particularly for propensity-score diagnostics, but it is not a causal criterion for covariate selection.

#### 10.4 Outcome Incidence

<!-- generated-learning-note -->
This subsection explains **10.4 Outcome Incidence** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Depending on design and outcome type, summarize:

- number and proportion with outcome,
- incidence proportion,
- person-time,
- incidence rate,
- median follow-up,
- Kaplan-Meier estimates,
- cumulative incidence in competing-risk settings.

#### 11. General Statistical Principles

<!-- generated-learning-note -->
This subsection explains **11. General Statistical Principles** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


#### 11.1 Significance Level

<!-- generated-learning-note -->
This subsection explains **11.1 Significance Level** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Unless otherwise specified, statistical tests will be two-sided with \(\alpha = 0.05\). Estimates will be presented with 95% confidence intervals.

#### 11.2 Emphasis on Estimation

<!-- generated-learning-note -->
This subsection explains **11.2 Emphasis on Estimation** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Primary interpretation will emphasize:

- effect estimates,
- confidence intervals,
- clinical/public-health magnitude,
- consistency across analyses,
- and study limitations,

rather than statistical significance alone.

#### 11.3 Precision and Rounding

<!-- generated-learning-note -->
This subsection explains **11.3 Precision and Rounding** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Specify presentation precision before final output production.

Recommended defaults:

- percentages: 1 decimal place,
- effect estimates: 2 decimal places,
- confidence limits: 2 decimal places,
- p-values: 3 decimals; report `<0.001` when smaller.

#### 11.4 Software

<!-- generated-learning-note -->
This subsection explains **11.4 Software** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Analyses will be conducted using **R version [x.x.x]** and/or **SAS version [x.x]**. Key package/library versions will be archived for reproducibility.

#### 12. Primary Statistical Analysis

<!-- generated-learning-note -->
This subsection explains **12. Primary Statistical Analysis** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


The primary method will be selected according to the outcome scale and target estimand.

#### 12.1 Binary Outcomes

<!-- generated-learning-note -->
This subsection explains **12.1 Binary Outcomes** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Possible primary models:

- logistic regression for odds ratios,
- log-binomial regression for risk ratios,
- modified Poisson regression with robust variance for risk ratios,
- binomial identity-link model for risk differences when estimable.

Generic model:

\[
g\{E(Y_i)\} = \beta_0 + \beta_1 X_i + \boldsymbol{\beta}_2^T \mathbf{C}_i
\]

where \(X_i\) is the exposure and \(\mathbf{C}_i\) is the prespecified adjustment set.

#### 12.2 Continuous Outcomes

<!-- generated-learning-note -->
This subsection explains **12.2 Continuous Outcomes** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Possible models:

- linear regression,
- generalized linear models,
- robust regression if prespecified and justified.

Report adjusted mean differences or marginal contrasts as appropriate.

#### 12.3 Count / Rate Outcomes

<!-- generated-learning-note -->
This subsection explains **12.3 Count / Rate Outcomes** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Possible models:

- Poisson regression,
- negative binomial regression when overdispersion is present,
- recurrent-event models where scientifically appropriate.

Include log person-time as an offset for incidence-rate analyses.

#### 12.4 Time-to-Event Outcomes

<!-- generated-learning-note -->
This subsection explains **12.4 Time-to-Event Outcomes** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


The default primary model may be Cox proportional hazards regression:

\[
h(t|X,C)=h_0(t)\exp\{\beta_1X+\boldsymbol{\beta}_2^T C\}
\]

Report hazard ratios with 95% confidence intervals.

If the proportional hazards assumption is not appropriate for the exposure effect, prespecified alternatives may include:

- time-varying coefficients,
- stratified Cox models,
- piecewise effects,
- accelerated failure-time models,
- restricted mean survival time (RMST) contrasts,
- or flexible parametric survival models.

#### 12.5 Repeated / Longitudinal Outcomes

<!-- generated-learning-note -->
This subsection explains **12.5 Repeated / Longitudinal Outcomes** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Possible models:

- linear mixed-effects models,
- generalized linear mixed models,
- generalized estimating equations (GEE),
- marginal structural models for time-varying exposure/confounding where appropriate.

Specify:

- fixed effects,
- random effects,
- covariance structure,
- time representation,
- exposure-by-time interaction,
- and target contrast.

#### 12.6 Case-Control Studies

<!-- generated-learning-note -->
This subsection explains **12.6 Case-Control Studies** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Use unconditional logistic regression unless matching requires conditional logistic regression.

Odds ratios will be interpreted in relation to the sampling design. Incidence-density sampling may permit the odds ratio to estimate an incidence rate ratio under appropriate assumptions.

#### 12.7 Cross-Sectional Studies

<!-- generated-learning-note -->
This subsection explains **12.7 Cross-Sectional Studies** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Because temporality may be limited, results will generally be interpreted as prevalence associations rather than causal effects unless temporal ordering is otherwise established.

#### 13. Crude and Adjusted Estimates

<!-- generated-learning-note -->
This section develops **13. Crude and Adjusted Estimates**. Read the model specification together with its assumptions and the scale of each reported effect.


For the primary exposure-outcome relationship, report when appropriate:

1. crude/unadjusted estimate,
2. minimally adjusted estimate,
3. primary fully adjusted estimate,
4. alternative adjusted estimate(s), if prespecified.

Differences between crude and adjusted estimates may demonstrate the impact of adjustment but will not be used by themselves to identify true confounders.

#### 14. Model Diagnostics and Assumption Checks

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **14. Model Diagnostics and Assumption Checks** in the context of Statistical Analysis Plan (SAP) for Epidemiologic Studies.


Diagnostics will be appropriate to the selected model.

#### 14.1 Cox Model

<!-- generated-learning-note -->
This section develops **14.1 Cox Model**. Read the model specification together with its assumptions and the scale of each reported effect.


Evaluate:

- proportional hazards using Schoenfeld residuals / `cox.zph`,
- functional form for continuous variables,
- influential observations,
- sparse cells / separation where relevant,
- and overall model stability.

#### 14.2 Linear Models

<!-- generated-learning-note -->
This section develops **14.2 Linear Models**. Read the model specification together with its assumptions and the scale of each reported effect.


Evaluate:

- residual distribution,
- heteroskedasticity,
- influential observations,
- functional form,
- and multicollinearity.

#### 14.3 Logistic / GLM Models

<!-- generated-learning-note -->
This section develops **14.3 Logistic / GLM Models**. Read the model specification together with its assumptions and the scale of each reported effect.


Evaluate:

- separation/sparse data,
- functional form of continuous predictors,
- influence,
- calibration when prediction is a study objective,
- and overdispersion where relevant.

Model diagnostics will inform prespecified alternative models or sensitivity analyses; they will not be used to conduct unrestricted data-driven model searching.

#### 15. Propensity-Score Analyses

<!-- generated-learning-note -->
This subsection explains **15. Propensity-Score Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Propensity-score methods may be used as primary or sensitivity analyses when appropriate to the study objective.

#### 15.1 Propensity-Score Model

<!-- generated-learning-note -->
This section develops **15.1 Propensity-Score Model**. Read the model specification together with its assumptions and the scale of each reported effect.


Estimate:

\[
e(C)=P(X=1|C)
\]

using prespecified baseline confounders.

Variables should be selected for confounding control, not solely based on statistical prediction of exposure.

#### 15.2 Weighting Strategy

<!-- generated-learning-note -->
This subsection explains **15.2 Weighting Strategy** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Specify the estimand:

- ATE: inverse probability of treatment weighting,
- ATT: weighting targeted to the exposed/treated,
- overlap weighting,
- standardized mortality ratio weighting,
- or another prespecified target.

For ATE IPTW:

\[
w_i = \frac{X_i}{e(C_i)} + \frac{1-X_i}{1-e(C_i)}
\]

Stabilized weights may be preferred for improved precision.

#### 15.3 Positivity / Overlap

<!-- generated-learning-note -->
This subsection explains **15.3 Positivity / Overlap** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Assess:

- propensity-score distributions by exposure,
- extreme propensity scores,
- extreme weights,
- effective sample size,
- and regions of poor common support.

Prespecify truncation/winsorization thresholds, for example the 1st/99th or 0.5th/99.5th percentiles, if trimming is planned.

#### 15.4 Covariate Balance

<!-- generated-learning-note -->
This subsection explains **15.4 Covariate Balance** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Assess weighted SMDs and graphical diagnostics (e.g., Love plots). An absolute SMD < 0.10 may be considered acceptable measured balance.

#### 15.5 Weighted Outcome Model

<!-- generated-learning-note -->
This section develops **15.5 Weighted Outcome Model**. Read the model specification together with its assumptions and the scale of each reported effect.


Use weighted regression appropriate to the outcome with robust/sandwich standard errors as required.

Measured balance does not demonstrate absence of unmeasured confounding.

#### 16. Time-Varying Exposure and Time-Varying Confounding

<!-- generated-learning-note -->
This subsection explains **16. Time-Varying Exposure and Time-Varying Confounding** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


If exposure changes over time, specify:

- update frequency,
- exposure grace periods,
- cumulative exposure definitions,
- lagging rules,
- and censoring at switching if applicable.

When time-varying confounders are affected by prior exposure, conventional regression adjustment may introduce bias. Marginal structural models with inverse-probability weights may be used if prespecified and scientifically justified.

#### 17. Missing Data

<!-- generated-learning-note -->
This section prepares the information needed for **17. Missing Data**. Check variable definitions, units, coding, and missing values before moving to analysis.


#### 17.1 Description

<!-- generated-learning-note -->
This subsection explains **17.1 Description** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Report the amount and pattern of missing data for exposure, outcome, covariates, and follow-up variables.

#### 17.2 Primary Missing-Data Approach

<!-- generated-learning-note -->
This section prepares the information needed for **17.2 Primary Missing-Data Approach**. Check variable definitions, units, coding, and missing values before moving to analysis.


Specify one primary strategy:

- complete-case analysis,
- multiple imputation (MI),
- inverse-probability weighting for missingness,
- maximum-likelihood methods inherent in longitudinal models,
- or another justified method.

#### 17.3 Multiple Imputation

<!-- generated-learning-note -->
This subsection explains **17.3 Multiple Imputation** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


When MI is used:

- include exposure, outcome, covariates, and predictors of missingness,
- respect variable distributions and bounds,
- include transformations/interactions needed by the analysis model,
- use a sufficient number of imputations,
- combine estimates using Rubin's rules.

Imputation should be consistent with the temporal structure of the study and should not inadvertently use future information in a way incompatible with the estimand.

#### 17.4 Missing-Not-at-Random Sensitivity Analyses

<!-- generated-learning-note -->
This subsection explains **17.4 Missing-Not-at-Random Sensitivity Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


When clinically relevant, evaluate departures from MAR using:

- delta adjustment,
- pattern-mixture models,
- selection models,
- tipping-point analyses,
- or scenario-based analyses.

#### 18. Censoring, Competing Risks, and Loss to Follow-up

<!-- generated-learning-note -->
This subsection explains **18. Censoring, Competing Risks, and Loss to Follow-up** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


#### 18.1 Censoring

<!-- generated-learning-note -->
This subsection explains **18.1 Censoring** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


List all censoring mechanisms and justify non-informative censoring assumptions.

#### 18.2 Informative Loss to Follow-up

<!-- generated-learning-note -->
This subsection explains **18.2 Informative Loss to Follow-up** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Potential approaches include inverse-probability-of-censoring weights (IPCW) or sensitivity analyses under plausible missing-outcome scenarios.

#### 18.3 Competing Risks

<!-- generated-learning-note -->
This subsection explains **18.3 Competing Risks** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


If competing events preclude the outcome, prespecify whether the scientific question targets:

- cause-specific hazards,
- subdistribution hazards,
- cumulative incidence functions,
- risk contrasts at a fixed time,
- or a composite endpoint.

Cause-specific Cox and Fine-Gray models answer different questions and should not be treated as interchangeable.

#### 19. Effect Modification, Interaction, and Subgroup Analyses

<!-- generated-learning-note -->
This subsection explains **19. Effect Modification, Interaction, and Subgroup Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Prespecified effect modifiers may include:

- age,
- sex,
- race/ethnicity when scientifically justified and appropriately interpreted,
- disease severity,
- baseline comorbidity,
- calendar period,
- or other clinically justified factors.

Effect modification will preferably be evaluated by an interaction term rather than separate within-subgroup significance tests.

Example:

\[
g\{E(Y)\}=\beta_0+\beta_1X+\beta_2Z+\beta_3(X\times Z)+\cdots
\]

Report stratum-specific estimates and the interaction p-value or confidence interval for the interaction contrast.

Subgroup analyses are generally considered supportive or exploratory unless explicitly designated as confirmatory.

#### 20. Dose-Response and Nonlinear Exposure Effects

<!-- generated-learning-note -->
This subsection explains **20. Dose-Response and Nonlinear Exposure Effects** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


For continuous or ordinal exposures, assess the prespecified form using one or more of:

- linear trend,
- clinically defined categories,
- restricted cubic splines,
- fractional polynomials,
- or generalized additive models when exploratory.

If categories are used, cut points should preferably be prespecified or clinically motivated rather than chosen from observed outcome data.

#### 21. Multiplicity

<!-- generated-learning-note -->
This subsection explains **21. Multiplicity** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


#### 21.1 Primary Objective

<!-- generated-learning-note -->
This subsection explains **21.1 Primary Objective** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


If there is one primary exposure-outcome estimand, no multiplicity adjustment is needed for that single primary test.

#### 21.2 Multiple Primary Endpoints / Exposures / Contrasts

<!-- generated-learning-note -->
This subsection explains **21.2 Multiple Primary Endpoints / Exposures / Contrasts** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


When more than one confirmatory primary hypothesis is tested, specify a control strategy such as:

- Bonferroni,
- Holm,
- Hochberg,
- hierarchical/gatekeeping testing,
- false-discovery-rate control when appropriate for the research objective,
- or no formal adjustment with explicit exploratory interpretation.

#### 21.3 Secondary and Exploratory Analyses

<!-- generated-learning-note -->
This subsection explains **21.3 Secondary and Exploratory Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Unless otherwise prespecified, secondary and exploratory p-values will be interpreted descriptively with emphasis on effect sizes, uncertainty, and consistency rather than binary significance claims.

#### 22. Sensitivity and Robustness Analyses

<!-- generated-learning-note -->
This subsection explains **22. Sensitivity and Robustness Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Sensitivity analyses should target specific assumptions rather than simply repeat the analysis using arbitrary alternatives.

Recommended categories include:

#### 22.1 Alternative Confounder Adjustment Sets

<!-- generated-learning-note -->
This subsection explains **22.1 Alternative Confounder Adjustment Sets** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


- primary DAG-based adjustment set,
- expanded adjustment set,
- reduced set excluding variables with uncertain temporal/causal status.

#### 22.2 Alternative Confounding-Control Methods

<!-- generated-learning-note -->
This section develops **22.2 Alternative Confounding-Control Methods**. Read the model specification together with its assumptions and the scale of each reported effect.


Compare, where appropriate:

- conventional multivariable regression,
- IPTW,
- overlap weighting,
- matching,
- standardization / g-computation,
- doubly robust estimators,
- TMLE or other causal estimators.

#### 22.3 Propensity-Score Weight Truncation

<!-- generated-learning-note -->
This subsection explains **22.3 Propensity-Score Weight Truncation** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Evaluate alternative truncation thresholds.

#### 22.4 Exposure Definition

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **22.4 Exposure Definition** in the context of Statistical Analysis Plan (SAP) for Epidemiologic Studies.


Examples:

- stricter exposure definition,
- alternative look-back window,
- cumulative exposure,
- lagged exposure,
- exclusion of uncertain exposure classification.

#### 22.5 Outcome Definition

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **22.5 Outcome Definition** in the context of Statistical Analysis Plan (SAP) for Epidemiologic Studies.


Examples:

- high-specificity outcome algorithm,
- broad/sensitive algorithm,
- adjudicated subset,
- alternative event-date rule.

#### 22.6 Reverse Causation / Latency

<!-- generated-learning-note -->
This subsection explains **22.6 Reverse Causation / Latency** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Use lag analyses or exclude events occurring shortly after baseline when scientifically appropriate and prespecified.

#### 22.7 Missing Data

<!-- generated-learning-note -->
This section prepares the information needed for **22.7 Missing Data**. Check variable definitions, units, coding, and missing values before moving to analysis.


Compare complete-case, MI, and MNAR scenarios as appropriate.

#### 22.8 Informative Censoring

<!-- generated-learning-note -->
This subsection explains **22.8 Informative Censoring** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Use IPCW or scenario analyses.

#### 22.9 Unmeasured Confounding

<!-- generated-learning-note -->
This subsection explains **22.9 Unmeasured Confounding** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


When appropriate, consider:

- E-values,
- negative-control exposures/outcomes,
- quantitative bias analysis,
- bias functions,
- or external-control calibration.

#### 22.10 Model Specification

<!-- generated-learning-note -->
This section develops **22.10 Model Specification**. Read the model specification together with its assumptions and the scale of each reported effect.


Evaluate alternative functional forms, distributions, correlation structures, or survival-model assumptions.

#### 23. Negative Controls and Falsification Analyses

<!-- generated-learning-note -->
This subsection explains **23. Negative Controls and Falsification Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


If available, prespecify negative-control outcomes or exposures that should not plausibly be caused by the exposure. Unexpected associations may indicate residual bias, measurement artifacts, or uncontrolled confounding.

#### 24. Selection Bias and Generalizability

<!-- generated-learning-note -->
This subsection explains **24. Selection Bias and Generalizability** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Evaluate whether inclusion, retention, healthcare access, testing, referral, or data capture may depend jointly on exposure and outcome determinants.

If sampling weights are available, specify whether survey/design weights will be incorporated.

Compare the analytic cohort with the target/source population when possible and describe limits to transportability/generalizability.

#### 25. Survey and Complex Sampling Designs

<!-- generated-learning-note -->
This subsection explains **25. Survey and Complex Sampling Designs** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


For complex surveys, incorporate as appropriate:

- sampling weights,
- strata,
- primary sampling units,
- finite population corrections,
- and survey-specific variance estimation.

Analyses must account for the design rather than treating observations as a simple random sample.

#### 26. Clustered and Multilevel Data

<!-- generated-learning-note -->
This section prepares the information needed for **26. Clustered and Multilevel Data**. Check variable definitions, units, coding, and missing values before moving to analysis.


For clustering by site, family, physician, hospital, geographic area, or repeated person-level observations, use methods appropriate to the estimand, such as:

- cluster-robust standard errors,
- GEE,
- random-effects/multilevel models,
- frailty models,
- or fixed effects.

Specify the cluster level and correlation structure.

#### 27. Small-Sample, Sparse-Data, and Separation Issues

<!-- generated-learning-note -->
This section prepares the information needed for **27. Small-Sample, Sparse-Data, and Separation Issues**. Check variable definitions, units, coding, and missing values before moving to analysis.


When sparse data may cause instability, prespecify alternatives such as:

- exact methods,
- Firth penalized logistic regression,
- penalized likelihood,
- category collapsing based on clinical logic,
- or reporting descriptive estimates without overfitted adjustment.

Avoid fitting models with inadequate information relative to model complexity.

#### 28. Outliers and Influential Observations

<!-- generated-learning-note -->
This subsection explains **28. Outliers and Influential Observations** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Outliers will not be removed solely because they affect statistical significance.

Potential influential observations will be:

1. verified for data quality,
2. retained if valid in the primary analysis unless a prespecified exclusion rule applies,
3. evaluated in sensitivity analyses when materially influential.

#### 29. Model Selection and Data-Driven Procedures

<!-- generated-learning-note -->
This section prepares the information needed for **29. Model Selection and Data-Driven Procedures**. Check variable definitions, units, coding, and missing values before moving to analysis.


The primary model and adjustment set will be prespecified.

Automated stepwise procedures should not be used to define the primary confounder set. If machine-learning methods are used for nuisance-model estimation, their role and tuning/cross-fitting procedures should be prespecified and separated from the causal estimand definition.

#### 30. Machine Learning and High-Dimensional Adjustment (Optional)

<!-- generated-learning-note -->
This subsection explains **30. Machine Learning and High-Dimensional Adjustment (Optional)** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


For high-dimensional EHR/claims settings, prespecified approaches may include:

- high-dimensional propensity scores,
- penalized regression,
- Super Learner,
- double/debiased machine learning,
- targeted maximum likelihood estimation (TMLE),
- causal forests for exploratory heterogeneity.

These approaches do not replace careful specification of time zero, exposure, outcome, target population, causal structure, and estimand.

#### 31. Sample Size, Precision, and Power

<!-- generated-learning-note -->
This subsection explains **31. Sample Size, Precision, and Power** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Observational studies may have a fixed available sample size. The SAP should state whether the study is:

- **precision-based**,
- **power-based**,
- or **database-size constrained**.

#### 31.1 Precision-Based Planning

<!-- generated-learning-note -->
This subsection explains **31.1 Precision-Based Planning** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Specify the anticipated number of exposed/unexposed participants, outcome events, and expected confidence-interval width for the primary effect.

#### 31.2 Power-Based Planning

<!-- generated-learning-note -->
This subsection explains **31.2 Power-Based Planning** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Specify:

- effect size,
- event rate / variance,
- exposure prevalence or group ratio,
- alpha,
- power,
- attrition/loss to follow-up,
- and design effect if clustered.

#### 31.3 Effective Sample Size for Weighted Analyses

<!-- generated-learning-note -->
This subsection explains **31.3 Effective Sample Size for Weighted Analyses** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


For weighting analyses, report effective sample size and recognize that extreme weights may substantially reduce precision.

#### 32. Interim Analyses and Data-Dependent Adaptation

<!-- generated-learning-note -->
This section prepares the information needed for **32. Interim Analyses and Data-Dependent Adaptation**. Check variable definitions, units, coding, and missing values before moving to analysis.


Usually **N/A** for standard observational epidemiologic analyses.

If interim looks, sequential surveillance, or adaptive data collection are planned, specify stopping rules and error-control considerations before analysis begins.

#### 33. Statistical Tables, Figures, and Listings

<!-- generated-learning-note -->
This section uses **33. Statistical Tables, Figures, and Listings** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.


#### 33.1 Planned Tables

<!-- generated-learning-note -->
This subsection explains **33.1 Planned Tables** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


**Table 1.** Cohort construction / participant flow  
**Table 2.** Baseline characteristics overall and by exposure  
**Table 3.** Missingness summary  
**Table 4.** Outcome incidence / person-time by exposure  
**Table 5.** Crude and adjusted primary effect estimates  
**Table 6.** Secondary outcome analyses  
**Table 7.** Prespecified subgroup / interaction analyses  
**Table 8.** Propensity-score balance diagnostics  
**Table 9.** Sensitivity-analysis comparison

#### 33.2 Planned Figures

<!-- generated-learning-note -->
This section uses **33.2 Planned Figures** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.


**Figure 1.** Study cohort flow diagram  
**Figure 2.** Exposure distribution  
**Figure 3.** Kaplan-Meier / cumulative incidence curve, if applicable  
**Figure 4.** Propensity-score overlap plot, if applicable  
**Figure 5.** Love plot / covariate balance, if applicable  
**Figure 6.** Forest plot of main and sensitivity estimates  
**Figure 7.** Restricted cubic spline exposure-response curve, if applicable

#### 34. Recommended R Analysis Framework

<!-- generated-learning-note -->
This subsection explains **34. Recommended R Analysis Framework** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


The following code skeleton is intentionally generic and should be adapted to the specific study.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r packages, eval=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tableone)
library(survival)
library(cobalt)
library(survey)
library(splines)
library(broom)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### 34.1 Cohort Definition

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **34.1 Cohort Definition** in the context of Statistical Analysis Plan (SAP) for Epidemiologic Studies.



<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r cohort-template, eval=FALSE}
analysis_data <- raw_data %>%
  filter(
    eligibility_flag == 1,
    !is.na(exposure),
    index_date >= study_start,
    index_date <= study_end
  )
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.


#### 34.2 Data Quality Checks

<!-- generated-learning-note -->
This section prepares the information needed for **34.2 Data Quality Checks**. Check variable definitions, units, coding, and missing values before moving to analysis.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r dq-template, eval=FALSE}
# Duplicate IDs
sum(duplicated(analysis_data$id))

# Missingness
colSums(is.na(analysis_data))

# Distributions
summary(analysis_data)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


#### 34.3 Baseline Table

<!-- generated-learning-note -->
This subsection explains **34.3 Baseline Table** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r table1-template, eval=FALSE}
vars <- c(
  "age",
  "sex",
  "bmi",
  "comorbidity_1",
  "comorbidity_2"
)

tab1 <- CreateTableOne(
  vars = vars,
  strata = "exposure",
  data = analysis_data,
  test = FALSE
)

print(tab1, smd = TRUE)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### 34.4 Binary Outcome Model

<!-- generated-learning-note -->
This section develops **34.4 Binary Outcome Model**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r binary-template, eval=FALSE}
fit_binary <- glm(
  outcome ~ exposure + age + sex + bmi + confounder_1,
  family = binomial(),
  data = analysis_data
)

summary(fit_binary)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 34.5 Continuous Outcome Model

<!-- generated-learning-note -->
This section develops **34.5 Continuous Outcome Model**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r continuous-template, eval=FALSE}
fit_continuous <- lm(
  outcome ~ exposure + age + sex + bmi + confounder_1,
  data = analysis_data
)

summary(fit_continuous)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 34.6 Time-to-Event Model

<!-- generated-learning-note -->
This section develops **34.6 Time-to-Event Model**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r cox-template, eval=FALSE}
fit_cox <- coxph(
  Surv(followup_time, event) ~
    exposure + age + sex + bmi + confounder_1,
  data = analysis_data
)

summary(fit_cox)

# PH assumption
ph_test <- cox.zph(fit_cox)
print(ph_test)
plot(ph_test)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 34.7 Propensity Score and IPTW

<!-- generated-learning-note -->
This subsection explains **34.7 Propensity Score and IPTW** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.



<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r iptw-template, eval=FALSE}
ps_fit <- glm(
  exposure ~ age + sex + bmi + confounder_1 + confounder_2,
  family = binomial(),
  data = analysis_data
)

analysis_data <- analysis_data %>%
  mutate(
    ps = predict(ps_fit, type = "response"),
    p_exp = mean(exposure == 1),
    sw = ifelse(
      exposure == 1,
      p_exp / ps,
      (1 - p_exp) / (1 - ps)
    )
  )

# Prespecified truncation example
limits <- quantile(analysis_data$sw, c(0.01, 0.99), na.rm = TRUE)

analysis_data <- analysis_data %>%
  mutate(
    sw_trim = pmin(pmax(sw, limits[1]), limits[2])
  )

bal.tab(
  exposure ~ age + sex + bmi + confounder_1 + confounder_2,
  data = analysis_data,
  weights = analysis_data$sw_trim,
  method = "weighting",
  un = TRUE,
  thresholds = c(m = 0.10)
)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.


#### 34.8 Weighted Cox Model

<!-- generated-learning-note -->
This section develops **34.8 Weighted Cox Model**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r weighted-cox-template, eval=FALSE}
fit_iptw <- coxph(
  Surv(followup_time, event) ~ exposure,
  data = analysis_data,
  weights = sw_trim,
  robust = TRUE
)

summary(fit_iptw)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 34.9 Restricted Cubic / Natural Spline Example

<!-- generated-learning-note -->
This subsection explains **34.9 Restricted Cubic / Natural Spline Example** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r spline-template, eval=FALSE}
fit_spline <- coxph(
  Surv(followup_time, event) ~
    ns(exposure_continuous, df = 4) +
    age + sex + bmi + confounder_1,
  data = analysis_data
)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 35. Primary and Sensitivity Result Synthesis

<!-- generated-learning-note -->
This section focuses on **35. Primary and Sensitivity Result Synthesis**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.


Create one summary table containing all major estimates with the same target contrast.

Example structure:

| Analysis | Estimand/measure | Estimate | 95% CI | Key assumption changed |
|---|---|---:|---|---|
| Crude | HR | [ ] | [ ] | No adjustment |
| Primary adjusted | HR | [ ] | [ ] | Primary confounder set |
| IPTW | Marginal HR / target measure | [ ] | [ ] | PS weighting |
| Alternative adjustment | HR | [ ] | [ ] | Covariate set |
| Missing-data sensitivity | HR | [ ] | [ ] | Missingness assumption |

Consistency across analyses will be evaluated based on effect direction, magnitude, precision, and the assumptions each analysis changes—not solely on whether individual p-values cross 0.05.

#### 36. Interpretation Framework

<!-- generated-learning-note -->
This section focuses on **36. Interpretation Framework**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.


The final interpretation will distinguish among:

1. **Crude association** — observed exposure-outcome relationship without confounder adjustment.
2. **Adjusted association** — association conditional on measured covariates.
3. **Causal effect estimate** — only when design, temporal ordering, consistency, exchangeability, positivity, correct model specification, and measurement assumptions are sufficiently justified.

Results will not automatically be described as causal merely because regression, propensity scores, or weighting were used.

#### 37. Key Sources of Bias and Limitations to Address

<!-- generated-learning-note -->
This subsection explains **37. Key Sources of Bias and Limitations to Address** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


At minimum, discuss:

- unmeasured confounding,
- residual confounding,
- exposure misclassification,
- outcome misclassification,
- general measurement error,
- uncertain temporal ordering,
- reverse causation,
- selection bias,
- informative censoring,
- missing data,
- positivity / overlap violations,
- model misspecification,
- immortal-time bias where relevant,
- time-varying confounding,
- competing risks,
- sparse-data bias,
- multiple testing / selective reporting,
- data provenance and coding changes,
- and limited generalizability / transportability.

#### 38. Deviations from the SAP

<!-- generated-learning-note -->
This subsection explains **38. Deviations from the SAP** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Any material deviation from this SAP after finalization will be documented with:

- date,
- description,
- rationale,
- whether the decision was made before or after examining outcome-related results,
- and potential impact on interpretation.

Post hoc analyses will be clearly labeled exploratory.

#### 39. Reproducibility and Quality Control

<!-- generated-learning-note -->
This subsection explains **39. Reproducibility and Quality Control** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


The study team will maintain:

- version-controlled analysis code,
- frozen analytic datasets or reproducible data-extraction specifications,
- variable derivation documentation,
- analysis logs,
- package/software versions,
- independent or secondary QC for key endpoints and primary models where feasible,
- and reproducible generation of final tables and figures.

#### 40. SAP Sign-Off

<!-- generated-learning-note -->
This subsection explains **40. SAP Sign-Off** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


| Role | Name | Signature | Date |
|---|---|---|---|
| Lead Statistician | [ ] | [ ] | [ ] |
| Epidemiologist / PI | [ ] | [ ] | [ ] |
| Data Manager / Analyst | [ ] | [ ] | [ ] |
| Other Reviewer | [ ] | [ ] | [ ] |

#### Appendix A. Study-Specific Analysis Decision Table

<!-- generated-learning-note -->
This subsection explains **Appendix A. Study-Specific Analysis Decision Table** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


Complete this table before final analysis.

| Item | Prespecified decision |
|---|---|
| Primary research question | [ ] |
| Study design | [ ] |
| Target population | [ ] |
| Time zero | [ ] |
| Primary exposure | [ ] |
| Exposure reference | [ ] |
| Primary outcome | [ ] |
| Follow-up end | [ ] |
| Primary estimand | [ ] |
| Primary effect measure | [ ] |
| Primary confounder set | [ ] |
| Primary statistical model | [ ] |
| Continuous-variable functional forms | [ ] |
| Missing-data method | [ ] |
| Censoring strategy | [ ] |
| Competing-risk strategy | [ ] |
| PS method, if any | [ ] |
| Weight truncation, if any | [ ] |
| Prespecified effect modifiers | [ ] |
| Multiplicity strategy | [ ] |
| Sensitivity analyses | [ ] |
| Software/version | [ ] |
| Data cutoff | [ ] |

#### Appendix B. Outcome-Specific Model Selection Guide

<!-- generated-learning-note -->
This section develops **Appendix B. Outcome-Specific Model Selection Guide**. Read the model specification together with its assumptions and the scale of each reported effect.


| Outcome type | Common primary model | Typical effect measure | Common alternatives |
|---|---|---|---|
| Binary | Logistic / modified Poisson | OR / RR / RD | GEE, log-binomial |
| Continuous | Linear regression | Mean difference | GLM, robust regression |
| Count/rate | Poisson / negative binomial | Rate ratio | GEE, recurrent-event models |
| Time-to-event | Cox PH | HR | RMST, AFT, flexible parametric |
| Repeated continuous | LMM / GEE | Mean contrast | Marginal models |
| Repeated binary | GEE / GLMM | OR / RR | Marginal standardization |
| Competing risk | Cause-specific Cox / Fine-Gray | Cause-specific HR / subdistribution HR | CIF risk contrast |

#### Appendix C. Minimum Sensitivity Analysis Matrix

<!-- generated-learning-note -->
This subsection explains **Appendix C. Minimum Sensitivity Analysis Matrix** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


| Assumption | Primary approach | Suggested sensitivity analysis |
|---|---|---|
| Confounder specification | DAG/subject-matter set | Expanded/reduced set |
| Missing covariates | MI or complete case | Alternative missingness model |
| Exposure misclassification | Main algorithm | Restrictive/high-specificity definition |
| Outcome misclassification | Main algorithm | Validated/high-specificity definition |
| Positivity | Primary PS approach | Trimming/overlap weighting |
| Informative censoring | Standard censoring | IPCW |
| Reverse causation | Main lag | Longer lag/exclude early events |
| PH assumption | Cox | Time-varying effect/RMST |
| Unmeasured confounding | Standard adjustment | E-value/QBA/negative controls |

#### Appendix D. Example Language for Reporting

<!-- generated-learning-note -->
This subsection explains **Appendix D. Example Language for Reporting** and shows how it contributes to the overall Statistical Analysis Plan (SAP) for Epidemiologic Studies workflow.


**Adjusted association wording:**  
"After adjustment for prespecified baseline covariates, the exposure was associated with [higher/lower] [outcome] ([effect measure] = [estimate], 95% CI [lower, upper])."

**Causal wording when assumptions are explicitly supported:**  
"Under the stated assumptions of exchangeability, consistency, positivity, correct model specification, and adequate measurement of exposure, outcome, and confounders, the estimate may be interpreted as the causal effect of [exposure contrast] on [outcome]."

**Cautious observational wording:**  
"Because this is an observational study, residual and unmeasured confounding, measurement error, selection bias, and uncertainty in temporal ordering may remain; therefore, the estimate should not be interpreted as definitive proof of causality."

## Interpreting the Results

Use the material as a checklist and adapt each element to the study design, estimands, data structure, and reporting requirements. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
