---
title: "Mixed model and linear model, Join model and Two stages model, MCMC"
description: "A practical R learning guide to Mixed model and linear model, Join model and Two stages model, MCMC, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Adaptive & Bayesian Methods"
topic: "Mixed model and linear model, Join model and Two stages model, MCMC"
type: article
tags: ["Adaptive & Bayesian Methods", "Biostatistics", "Mixed", "linear", "Join"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  enabled: false
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Adaptive and Bayesian methods update evidence as data accumulate and express uncertainty through probability distributions or decision rules.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Mixed model and linear model, Join model and Two stages model, MCMC**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Join model, Mixed model and linear model
2. Data manipulation
3. Prepare the data input (list() format)
4. Model fitting
5. Create a JAGS model
6. Prediction

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Adaptive & Bayesian Methods",
  topic = "Mixed model and linear model, Join model and Two stages model, MCMC"
)
learning_overview
```

## Learning Material

The following material develops **Mixed model and linear model, Join model and Two stages model, MCMC** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Join model, Mixed model and linear model

<!-- generated-learning-note -->
This section develops **Join model, Mixed model and linear model**. Read the model specification together with its assumptions and the scale of each reported effect.


Stage 1: A mixed model: Random intercept and slope.

Stage.1  = lmer(EFW3 ~ 1+t1 + t2 + t3 + tt1 + tt2 + tt3+(1+t1| Subject_ID)   , data = train.data, REML = TRUE)

Stage 2: linear regression model

Stage.2 =lm(data=data_lm,NDSBWGT_fm024~alpha+beta+EFW35+diff)

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# Clean 'Global Environment'  
rm(list = ls())
# Download the needed libraries
library(tidyverse)
library(caret)
library(ggplot2)
library(rms)
library(reshape) 
library(nlme)
library(mvtnorm)
library(lme4)
library("rms")
library(ROCR)
library(magicfor)
library(coda)
library(rjags)
library(ROCR)
library(boot)
library(data.table)
library("gridExtra")
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Data manipulation

<!-- generated-learning-note -->
This section prepares the information needed for **Data manipulation**. Check variable definitions, units, coding, and missing values before moving to analysis.


<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r, echo=FALSE}
#| eval: false
#| code-fold: true
# load data
setwd("N:\\DIPHRData\\Fetal 3D Study\\Projects\\BW prediction with Velocity\\output\\two stage model")
DF <- read.csv("Analytic_1719 - v1.081618.csv")
# length(unique(DF$Subject_ID))

# manage the data
# create diff and spline terms
DF_1 = DF[  c(  "Subject_ID" ,   "GA_WKS" ,      "EFW3"      ,"NDSBWGT_fm024",   "VISNO","GADel","GA_WKS" )]
DF_1$diff=round((DF_1$GADel-DF_1$GA_WKS)*7,0)
nDF = na.omit(DF_1) #or imputation
Zeta = quantile(nDF$GA_WKS, probs = c(25, 50, 75)/100, na.rm = TRUE)
newDF = data.frame(Subject_ID = nDF$Subject_ID, 
                   EFW3 =  (nDF$EFW3), 
                   t1 = nDF$GA_WKS, 
                   t2 = nDF$GA_WKS**2, 
                   t3 = nDF$GA_WKS**3,
                   tt1 = (nDF$GA_WKS - Zeta[1])**3 * (nDF$GA_WKS > Zeta[1]),
                   tt2 = (nDF$GA_WKS - Zeta[2])**3 * (nDF$GA_WKS > Zeta[2]),
                   tt3 = (nDF$GA_WKS - Zeta[3])**3 * (nDF$GA_WKS > Zeta[3]),
                   NDSBWGT_fm024=nDF$NDSBWGT_fm024,
                   VISNO=nDF$VISNO,
                   GADel=nDF$GADel,
                   GA_WKS=nDF$GA_WKS,
                   diff=nDF$diff)
# nrow(newDF)

ts1=newDF[,c(1,3,10)]
ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")
ts22=na.omit(ts2)
index=ts22$Subject_ID
# length(index)
newDF=newDF[newDF$Subject_ID %in% index, ]
# table(newDF$VISNO)

 
```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# colnames(train.data)
# ncol(train.data)
# nrow(train.data)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Prepare the data input (list() format)

<!-- generated-learning-note -->
This section prepares the information needed for **Prepare the data input (list() format)**. Check variable definitions, units, coding, and missing values before moving to analysis.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}


```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Model fitting

<!-- generated-learning-note -->
This section develops **Model fitting**. Read the model specification together with its assumptions and the scale of each reported effect.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
 
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Create a JAGS model

<!-- generated-learning-note -->
This section develops **Create a JAGS model**. Read the model specification together with its assumptions and the scale of each reported effect.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}

```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`can not have missing for the predictors.`



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# quantile.model.JMB
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Prediction

<!-- generated-learning-note -->
This subsection explains **Prediction** and shows how it contributes to the overall Mixed model and linear model, Join model and Two stages model, MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
out_rmse=NULL #for RMSE ci calculation
out_para=NULL #two stage coefficients, mixed and linear regression
for (B in c(1:1)){ # calculate rmse- root mean square error 

# split the data
  set.seed(NULL)
  index=sample(unique(newDF$Subject_ID),length(unique(newDF$Subject_ID))*0.9)
  train.data  <- newDF[newDF$Subject_ID %in% index, ]
  test.data <- newDF[!(newDF$Subject_ID %in% index), ]
  
# Prepare the data input needed for fitting the Bayesian model-------------------------------------------------------------------------------- 
  ## order by VISNO and remove duplicated Subject_ID
  data= train.data
  M <- as.numeric (length(unique(train.data$VISNO))) # Number of weeks 6
  M
  N <- as.numeric (length(unique(train.data$Subject_ID))) # Number of patients , 1545
  N
  
  ts1=train.data[,c(1,3,10)]
  ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")[-1]
  t1 = as.matrix(ts2)
  dim(t1)
  
  ts1=train.data[,c(1,4,10)]
  ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")[-1]
  t2 = as.matrix(ts2)
  dim(t2)
  
  ts1=train.data[,c(1,5,10)]
  ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")[-1]
  t3 = as.matrix(ts2)
  dim(t3)
  
  ts1=train.data[,c(1,6,10)]
  ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")[-1]
  tt1 = as.matrix(ts2)
  dim(tt1)
  
  ts1=train.data[,c(1,7,10)]
  ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")[-1]
  tt2 = as.matrix(ts2)
  dim(tt2)
  
  ts1=train.data[,c(1,8,10)]
  ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")[-1]
  tt3 = as.matrix(ts2)
  dim(tt3)


  data_merge2 <- train.data[order(train.data$VISNO, decreasing=TRUE),]
  data_lm <- data_merge2[!duplicated(data_merge2$Subject_ID),]
  # nrow(data_lm)
  
  NDSBWGT_fm024=data_lm$NDSBWGT_fm024 # the GTN status: Binary outcome  
  length(NDSBWGT_fm024) #1545
  
  ts1=train.data[,c(1,2,10)]
  ts2=reshape(ts1, idvar = "Subject_ID", timevar = "VISNO", direction = "wide")[-1]
  EFW3 = as.matrix(ts2) # The matrix of the repeated log(hCG) measurements 
  dim(EFW3)
  
  diff=data_lm$diff # The age at the time of evacuation
  length(diff)
  
  EFW35=data_lm$EFW3
  length(EFW35)

 # Model fitting: The bayesian joint model (JMBB)----------------------------------------------------------------------------------------------
  
  Model_string.JMB = "model {
for (i in 1:N) {
## random effects 
## mean is 0
b0[i,1] = 0   
b0[i,2] = 0
b[i,1:2] ~ dmnorm(b0[i,1:2],ISigma[,]) ## MV normal

for (j in 1:M) {
## The mixed model
EFW3[i,j] ~ dnorm(mu[i,j], tau)
mu[i,j] = beta[1]+b[i,1] + (beta[2]+b[i,2])*t1[i,j] + beta[3]*t2[i,j]   + beta[4]*t3[i,j] + beta[5]*tt1[i,j] + beta[6]*tt2[i,j] + beta[7]*tt3[i,j]
}

## linear regression model 
NDSBWGT_fm024[i] ~ dnorm(m[i],1)
m[i] = a[1] + a[2]*b[i,1] + a[3]*b[i,2]+a[4]*EFW35[i]+a[5]*diff[i]
}

## prior distribution of the parameters

#(1) Coefficients
for (l in 1:7) { beta[l] ~dnorm(0, .01) }
for (ll in 1:5) { a[ll] ~ dnorm(0, .01) }

#(2) Precision parameters
tau ~dgamma(.01,.01)
sigma.tau = 1/tau

#(3) Variance-covariance matrix, prior 
ISigma[1:2,1:2] ~dwish(R[,],3)  
#Evaluate the Density of the Wishart Distribution, inverse of co variance distribution
Sigma[1:2,1:2] = inverse(ISigma[,]) 
R[1,1] =1
R[1,2] =0
R[2,2] =1
R[2,1] =0
}"


# In Bayesian statistics, the Wishart distribution is (the conjugate prior of ) the inverse covariance-matrix of a multivariate-normal random-vector

# library(mnormt )
# library(MCMCpack)
# 
# mu <- c(0,0,0)
# Sigma <- matrix(c(1,2,0,2,5,0.5,0,0.5,3), 3, 3)
# f <- dmnorm( (c(1,.3,.3 ) ), mu, Sigma)
# f #mean distribution
# density <- dwish(matrix(c(1,.3,.3,1),2,2), mu, Sigma)
# density #co variance distribution

# Create a JAGS model object
Model.JMB = jags.model(textConnection(Model_string.JMB), 
                       data = list(EFW3=EFW3,N=N,M=M,NDSBWGT_fm024=NDSBWGT_fm024,t1=t1,t2=t2, t3=t3,tt1=tt1,tt2=tt2,tt3=tt3,diff=diff,EFW35=EFW35 ),
                       n.chains=3)

update(Model.JMB, 10000, progress.bar="none") #Updating the model

params=c('beta', 'a','Sigma','sigma.tau') # Generate posterior samples in mcmc.list format
samps = coda.samples(Model.JMB, params, n.iter = 10000) 

burn.in = 1000  # burn in period 
summary.model.JMB=summary(window(samps, start = burn.in))  # the posterior means of the joint model coefficients 

quantile.model.JMB=as.data.frame(summary.model.JMB$quantiles)

# B=99

Params.boot=as.data.frame(summary.model.JMB[["quantiles"]][,3])
Params.boot.new=as.data.frame(transpose(Params.boot))
Params.boot.new=cbind(B,Params.boot.new)
names(Params.boot.new)=c("B","d11","d12","d21","d22","Alpha0", "Alpha1","Alpha2","Alpha3","Alpha4","Beta0","Beta1","Beta2","Beta3","Beta4","Beta5","Beta6","Sigma.res")
head(Params.boot.new)

outcome1=NULL #prediction vs bw
outcome2=NULL # coefficients
for (BB in 1: length(unique (test.data$Subject_ID ))) {
# new data, internal cross validation
new.patient=test.data[test.data$Subject_ID %in% c(unique (test.data$Subject_ID )[BB]),] #calculate one subject prediction

Xbeta=Params.boot.new$Beta0+Params.boot.new$Beta1*new.patient$t1 +
  Params.boot.new$Beta2*new.patient$t2 +
  Params.boot.new$Beta3*new.patient$t3 +
  Params.boot.new$Beta4*new.patient$tt1 +
  Params.boot.new$Beta5*new.patient$tt2 +
  Params.boot.new$Beta6*new.patient$tt3 #calcualte fixed effect
 
R=diag(length(new.patient$EFW3))*(Params.boot.new$Sigma.res **2) # The residuals variance matrix

Z=matrix(c(rep(1,length(new.patient$EFW3)),new.patient$t1),nrow=length(new.patient$EFW3),ncol=2) # The random effects design matrix 

kk=new.patient$EFW3-Xbeta #diff between bw with fixed effect

G=matrix(c(Params.boot.new$d11**2,Params.boot.new$d12,Params.boot.new$d12,Params.boot.new$d22**2),nrow=2,ncol=2) #G matrix, random effect covariance

SigmaMatrix=Z%*%G%*%t(Z)+R #y covariance matrix= fixed effect covariance

Inversigma=solve(SigmaMatrix) #inversigma? 

REPredicted=G%*%t(Z)%*%Inversigma%*%  kk # random effect coefficients

RandomIntercept=REPredicted[1,1]
RandmSlope=REPredicted[2,1]

# caculate bw prediction
bw=Params.boot.new$Alpha0+Params.boot.new$Alpha1*RandomIntercept +Params.boot.new$Alpha2*RandmSlope+ 
  Params.boot.new$Alpha3*tail(new.patient$EFW3,1) +Params.boot.new$Alpha4*tail(new.patient$diff,1)
  
# output bw prediction and observation  
out=cbind(BB,unique (test.data[,c(1,9)])$NDSBWGT_fm024[BB],bw)
outcome1=rbind(outcome1,out)
}

RMSE = round(RMSE(outcome1[,3], outcome1[,2]),3)
Rsquare = round(R2(outcome1[,3], outcome1[,2]),3)

result=cbind(RMSE,Rsquare )

out_rmse=rbind(out_rmse,result)
}

# output results
out_rmse=as.data.frame(out_rmse)
quants <- c(0.025, 0.50, 0.975 )
ci=round(apply( out_rmse[1:2] , 2 , quantile , probs = quants , na.rm = TRUE ),3)
colnames(ci)=c("join_model_RMSE"  ,  "join_model_Rsquare")
ci

# > ci
#          RMSE Rsquare
# 2.5%  241.368   0.456
# 50%   282.596   0.591
# 97.5% 442.610   0.701
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


`give random effect coefficients based on the new data then predict`

#### Perfermance measurement

<!-- generated-learning-note -->
This subsection explains **Perfermance measurement** and shows how it contributes to the overall Mixed model and linear model, Join model and Two stages model, MCMC workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}

```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



#### Two stages model

<!-- generated-learning-note -->
This section develops **Two stages model**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
out_rmse=NULL #for RMSE ci calculation
out_para=NULL #two stage coefficients, mixed and linear regression
for (j in c(1:50)){ # calculate rmse- root mean square error 
set.seed(NULL)
index=sample(unique(newDF$Subject_ID),length(unique(newDF$Subject_ID))*0.9)
train.data  <- newDF[newDF$Subject_ID %in% index, ]
test.data <- newDF[!(newDF$Subject_ID %in% index), ]
# j=1
# Stage 1: A mixed model: Random intercept and slope.
Stage.1  = lmer(EFW3 ~ 1+t1 + t2 + t3 + tt1 + tt2 + tt3+(1+t1| Subject_ID)   , data = train.data, REML = TRUE) 
# summary(Stage.1)
RE = ranef(Stage.1)[["Subject_ID"]]  # The random effects coefficients
df1 = cbind(Subject_ID = rownames(RE), RE)
# head(df1)
colnames(df1)=c("Subject_ID"    ,"alpha",  "beta")
df1$Subject_ID=as.integer(df1$Subject_ID)
# summary(df1)
# length(unique(df1$Subject_ID))

# Stage 2: linear regression for bw using the random effects coefficients and other predictors
df2= train.data
# merge above two datasets
data_merge=full_join(df1, df2, by = "Subject_ID") 
# subset/filter a dataset with last visit EFW3
## order by VISNO and remove duplicated Subject_ID
data_merge2 <- data_merge[order(data_merge$VISNO, decreasing=TRUE),]
data_merge3 <- data_merge2[!duplicated(data_merge2$Subject_ID),]
data_lm=data_merge3 %>% filter(VISNO>3) #keep after 3th visit

# linear regression model
M=lm(data=data_lm,NDSBWGT_fm024~alpha+beta+EFW3+diff)
# summary(M)

# calculate the prediction of bw using two stage model
# fixed effect coefficients
Beta0= round(as.numeric(fixef(Stage.1)[1]), 2)
Beta1= round(as.numeric(fixef(Stage.1)[2]), 2)
Beta2= round(as.numeric(fixef(Stage.1)[3]), 2)
Beta3= round(as.numeric(fixef(Stage.1)[4]), 2)
Beta4= round(as.numeric(fixef(Stage.1)[5]), 2)
Beta5= round(as.numeric(fixef(Stage.1)[6]), 2)
Beta6= round(as.numeric(fixef(Stage.1)[7]), 2)
# random effect covariance
vcov = as.data.frame(VarCorr(Stage.1))
d11= round(vcov$sdcor[1], 2)
d22= round(vcov$sdcor[2], 2)
d12= round(vcov$sdcor[3] * d11* d22, 2)
Sigma.res= round(vcov$sdcor[4], 2)
# linear regression coefficients
Alpha0= round(as.numeric(coefficients(M)[1]), 2)
Alpha1= round(as.numeric(coefficients(M)[2] ), 2)
Alpha2= round(as.numeric(coefficients(M)[3] ), 2)
Alpha3= round(as.numeric(coefficients(M)[4]), 2)
Alpha4= round(as.numeric(coefficients(M)[5]), 2)
# organize these coefficients
B=j
Param=rbind(B,Beta0,Beta1,Beta2,Beta3,Beta4,Beta5,Beta6,Sigma.res,
            d11,d12,d22, Alpha0, Alpha1,Alpha2,Alpha3,Alpha4)
Params.boot=as.data.frame(Param)
Params.boot.new=as.data.frame(transpose(Params.boot))
names(Params.boot.new)=c("B","Beta0","Beta1","Beta2","Beta3","Beta4","Beta5","Beta6","Sigma.res","d11","d12","d22","Alpha0", "Alpha1","Alpha2","Alpha3","Alpha4")
head(Params.boot.new)


# using new data to do internal cross validation
out_bw_pred=NULL #for prediction and birthweight
s_test=unique(test.data$Subject_ID)
len=length(s_test)
for (i in c(1:len)) {
  # i=1
new.patient=test.data[test.data$Subject_ID %in% s_test[i],] #calculate prediction subject by subject

Xbeta=Params.boot.new$Beta0+Params.boot.new$Beta1*new.patient$t1 +
  Params.boot.new$Beta2*new.patient$t2 +
  Params.boot.new$Beta3*new.patient$t3 +
  Params.boot.new$Beta4*new.patient$tt1 +
  Params.boot.new$Beta5*new.patient$tt2 +
  Params.boot.new$Beta6*new.patient$tt3 #calculate fixed effect
  
R=diag(length(new.patient$EFW3))*(Params.boot.new$Sigma.res **2) # The residuals variance matrix

Z=matrix(c(rep(1,length(new.patient$EFW3)),new.patient$t1),nrow=length(new.patient$EFW3),ncol=2) # The random effects design matrix 

kk=new.patient$EFW3-Xbeta #diff between bw with fixed effect

G=matrix(c(Params.boot.new$d11**2,Params.boot.new$d12,Params.boot.new$d12,Params.boot.new$d22**2),nrow=2,ncol=2) #G matrix, random effect covariance matrix

SigmaMatrix=Z%*%G%*%t(Z)+R #y covariance matrix= fixed effect covariance matrix

Inversigma=solve(SigmaMatrix) #inversigma? 

REPredicted=G%*%t(Z)%*%Inversigma%*%  kk # random effect coefficients

RandomIntercept=REPredicted[1,1]
RandmSlope=REPredicted[2,1]

# using 0 instead of NA? 
Params.boot.new$Alpha1= ifelse((is.na(Params.boot.new$Alpha1 )==T | Params.boot.new$Alpha1>10000),0,Params.boot.new$Alpha1 ) 
Params.boot.new$Alpha2= ifelse((is.na(Params.boot.new$Alpha2 )==T | Params.boot.new$Alpha2>10000),0,Params.boot.new$Alpha2 ) 

# calculate bw prediction
bw=Params.boot.new$Alpha0+Params.boot.new$Alpha1*RandomIntercept +Params.boot.new$Alpha2*RandmSlope+ 
  Params.boot.new$Alpha3*tail(new.patient$EFW3,1) +Params.boot.new$Alpha4*tail(new.patient$diff,1)

# output bw prediction and observation  
out=cbind(i,new.patient$NDSBWGT_fm024[1],bw)
out_bw_pred=rbind(out_bw_pred,out)
# print(i)
}

# calculate RMSE,Rsquare
RMSE = round(RMSE(out_bw_pred[,3], out_bw_pred[,2]),3)
Rsquare = round(R2(out_bw_pred[,3], out_bw_pred[,2]),3)

# output RMSE,Rsquare  
out3=cbind(j,RMSE,Rsquare)
out_rmse=rbind(out_rmse,out3)
# print(j)
}

# output results
out_rmse=as.data.frame(out_rmse)
quants <- c(0.025, 0.50, 0.975 )
ci=round(apply( out_rmse[2:3] , 2 , quantile , probs = quants , na.rm = TRUE ),3)
colnames(ci)=c("two_stage_RMSE"  ,  "two_stage_Rsquare")
ci


```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

## Interpreting the Results

Inspect priors, likelihoods, posterior summaries, convergence, simulations, and operating characteristics. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
