---
title: "Bayesian Linear Regression Using brms"
output: html_document
---

## Introduction

This example fits a Bayesian linear regression model using the `brms` package.

The model is

$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i,
$$

where

$$
\epsilon_i \sim N(0,\sigma^2).
$$

## Load Package and Simulate Data

```{r setup, message=FALSE, warning=FALSE}
library(brms)

set.seed(123)

n <- 31

dat <- data.frame(
  x = seq(-15, 15, length.out = n)
)

dat$y <- 5 * dat$x + rnorm(
  n,
  mean = 0,
  sd = 10
)
```

## Plot the Simulated Data

```{r data-plot}
plot(
  dat$x,
  dat$y,
  pch = 19,
  xlab = "x",
  ylab = "y",
  main = "Simulated Data"
)
```

## Define Prior Distributions

The prior distributions are

$$
\beta_1 \sim N(5,2^2),
$$

$$
\beta_0 \sim N(0,5^2),
$$

and

$$
\sigma \sim \text{Exponential}(0.1).
$$

```{r priors}
priors <- c(
  brms::set_prior(
    "normal(5, 2)",
    class = "b"
  ),
  brms::set_prior(
    "normal(0, 5)",
    class = "Intercept"
  ),
  brms::set_prior(
    "exponential(0.1)",
    class = "sigma"
  )
)
```

## Fit the Bayesian Linear Regression Model

```{r fit-model, results='hide'}
fit <- brms::brm(
  y ~ x,
  data = dat,
  family = gaussian(),
  prior = priors,
  chains = 4,
  iter = 4000,
  warmup = 2000,
  cores = 4,
  seed = 123,
  refresh = 0
)
```

## Model Summary

```{r model-summary}
summary(fit)
```

## MCMC Diagnostic Plots

```{r diagnostic-plots}
# Close all active graphics devices
while (dev.cur() > 1) {
  dev.off()
}

# Draw brms diagnostic plots
plot(fit)
```

## Interpretation

The posterior estimate of `b_x` represents the slope of the relationship between `x` and `y`.

The posterior estimate of `Intercept` represents the expected value of `y` when (x=0).

The posterior estimate of `sigma` represents the residual standard deviation.
