---
title: "Epidemiology"
description: "A practical R learning guide to Epidemiology, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Epidemiology & Real-World Evidence"
category: "Epidemiologic Study Designs"
topic: "Epidemiology"
type: article
tags: ["Epidemiologic Study Designs", "Epidemiology & Real-World Evidence", "Epidemiology"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Epidemiologic design determines how populations, exposures, outcomes, follow-up, and comparisons support a target scientific question.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Epidemiology**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Epidemiology
2. What is epidemiology?
3. Framingham Heart Study
4. Sources of Public Health Data
5. Measures of Association
6. Confidence interval and p value

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Epidemiology & Real-World Evidence",
  category = "Epidemiologic Study Designs",
  topic = "Epidemiology"
)
learning_overview
```

## Learning Material

The following material develops **Epidemiology** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Epidemiology

<!-- generated-learning-note -->
This subsection explains **Epidemiology** and shows how it contributes to the overall Epidemiology workflow.


Brief summary from J. Craig Longenecker...  

#### What is epidemiology?

<!-- generated-learning-note -->
This subsection explains **What is epidemiology?** and shows how it contributes to the overall Epidemiology workflow.


"Epidemiology is the study of 
the distribution and 
determinants of health related 
states or events in 
specified populations, and 
the application of this study 
to control of health 
problems."


#### Framingham Heart Study

<!-- generated-learning-note -->
This subsection explains **Framingham Heart Study** and shows how it contributes to the overall Epidemiology workflow.



Highlights: Some of the Most Significant Milestones Include:

1960: Cigarette smoking found to increase the risk of heart disease

1961: Cholesterol level, blood pressure, and electrocardiogram 
abnormalities found to increase the risk of heart disease

1967: Physical activity found to reduce the risk of heart disease and
obesity to increase the risk of heart disease

1970: High blood pressure found to increase the risk of stroke

1978: Psychosocial factors found to affect heart disease

1988: High levels of HDL cholesterol found to reduce risk of death

1994: Enlarged left ventricle (one of two lower chambers of the heart)
shown to increase the risk of stroke

1996: Progression from hypertension to heart failure described

#### Sources of Public Health Data

<!-- generated-learning-note -->
This section prepares the information needed for **Sources of Public Health Data**. Check variable definitions, units, coding, and missing values before moving to analysis.


Epidemiologic Followup Study  [(NHEFS)](https://wwwn.cdc.gov/Nchs/Nhanes/Nhefs/Default.aspx)

National Health and Nutrition 
Examination Survey [(NHANES)](https://www.cdc.gov/nchs/nhanes/index.htm)

`Family questionnaire
- Demographics, housing, smoking,
income, food security

Computer-assisted personal interview
- Current health status
- Alcohol use, drug use
- Sexual history
- Depression screener
- Kidney function
- Pesticide use
- Physical activity
24-hour dietary recall

Examination Components
- Arthritis
- Audiometry
- Bone Density Dual-Energy X-Ray
Absorptiometry {DXA)

Examination Components (cont.)
- Body Measurements - Anthropometry
- Oral Glucose Tolerance Test (OGTT)
- Oral Health
- Physician's Exam

Laboratory Components:
- Venipuncture
Urine Collection
- Bone Mineral Status Markers
- Diabetes Proftle
- Infectious Disease Profile (inc. STDs)
- C-reactive Protein
- Kidney Disease Profile
- Pregnancy Test
- Prostate Specific Antigen
- Blood Lipids
- Environmental Health Profile`

#### Measures of Association

<!-- generated-learning-note -->
This subsection explains **Measures of Association** and shows how it contributes to the overall Epidemiology workflow.


#### Confidence interval and p value

<!-- generated-learning-note -->
This subsection explains **Confidence interval and p value** and shows how it contributes to the overall Epidemiology workflow.


In statistical estimate, this “confidence level” should not be confused with the 95% Confidence Interval, which is not based on hypothesis testing, but 
on estimation (two different types of inferential statistics) based on CLT. 

In hypothesis testing, assuming that no association exists IN THE TARGET 
POPULATION (i.e. under the null hypothesis), The p-value is the probability that 
THE STUDY found THIS observed (or greater) difference 
(i.e., this specific estimate in my study) 
by chance alone.

For a statistic, confidence intervals are calculated from the same equations that generate p-values. 



#### “Epidemiologic” Measures of Association

<!-- generated-learning-note -->
This subsection explains **“Epidemiologic” Measures of Association** and shows how it contributes to the overall Epidemiology workflow.


• Differences in:
– Means
– Medians
– Proportions

• The slope of a regression line



• Relative Risk (and Relative Rate)
• Attributable Risk
• Odds Ratio
• Number Needed to Treat

 
#### PECO

<!-- generated-learning-note -->
This subsection explains **PECO** and shows how it contributes to the overall Epidemiology workflow.


Frame a research question into four parts: “Population of interest,
Exposure, Comparison and Outcome” (PECO). 

Everything in PH or clinical research flows from PECO
– Which variables should be measured, and how
– The population to generalize from
– The research study design and research methods needed
– The outcomes you are concerned about
– The inferences that you can make from the study
– Search terms you use in searching the literature

you can frame the question into a PECO statement does not
mean that a study has the right to make causal statements.

#### Relative risks and relative rates

<!-- generated-learning-note -->
This subsection explains **Relative risks and relative rates** and shows how it contributes to the overall Epidemiology workflow.


• Relative risk
• Relative rate
• Attributable risk
• Odds ratio

The OR is always farther away from 1.0 than is the RR. The higher the incidence
and the higher the RR, the less the OR can be used as an estimate of the RR. 

The choice of statistical tests depends on the NATURE of the two variables. 


#### Measures of Disease Frequency

<!-- generated-learning-note -->
This subsection explains **Measures of Disease Frequency** and shows how it contributes to the overall Epidemiology workflow.


#### 8 Basic measures of the burden of disease

<!-- generated-learning-note -->
This subsection explains **8 Basic measures of the burden of disease** and shows how it contributes to the overall Epidemiology workflow.


• How common is it?

– Prevalence of disease/risk factors

– Incidence of disease/comorbidity

• How severe is it?

– Mortality rates (incidence of death)

– Median Survival

– 5‐year (or other time‐) survival

– Fatality

– YPLL: “Years of Potential Life Lost” due to early death

– DALY: “Disability‐Adjusted Life‐Years


#### Proportion, rate and ratio

<!-- generated-learning-note -->
This subsection explains **Proportion, rate and ratio** and shows how it contributes to the overall Epidemiology workflow.


The numerator: How do we define Health/Disease‐Related “Events". 



In incidence rate, the denominator is the total disease-free observation time in
each group. Person-time is only accrued while the subject is
being followed.


CRUDE, SPECIFIC, AND ADJUSTED RATES

Direct adjustment: apply observed rate of disease/
mortality in populations of interest to the population
structure of a standard population to derive expected #
cases. Then compare adjusted rates of the populations of
interest.

Median Survival

Length of time to 
which half the study 
population survives. Not affected by extremes. 

5‐year  survival
Number of people alive 5 years after diagnosis. Note artifactual increase in survival due to 
earlier detection.



#### Disease transmission

<!-- generated-learning-note -->
This subsection explains **Disease transmission** and shows how it contributes to the overall Epidemiology workflow.


What Factors Affect the Reproductive Number?

Contacts per unit time X  Infections per contact X d Duration of infectivity X  Susceptible Fraction

Re=R0XS

Major unknowns preventing precise predictions
1. Immunity/vaccine?
2. Treatment?
3. Mutation?
4. Human behavior?
5. Widespread testing availability?
6. Testing accuracy?
7. Asymptomatic cases?
8. Seasonal pattern?





 
#### Observational Studies

<!-- generated-learning-note -->
This subsection explains **Observational Studies** and shows how it contributes to the overall Epidemiology workflow.


• Descriptive epidemiological studies investigate the
distribution of diseases and risk factors (exposures) by
frequency in terms of person, place, and time.

• Analytical epidemiological studies are conducted to (attemptto)
determine cause and, sometimes, prevention, of disease
based upon comparison of populations in relation to their
exposure status.

NHANES is a cross sectional study. From these separate cross-sectional studies done between 1960 and 2000 we can see that prevalence of obesity in US adults is increasing. Different groups 
were used for each study, but the samples are meant to be rep representative of the
US population so we can look at trends over time. Cross section study can not determine the temporality. 

#### Case control study

<!-- generated-learning-note -->
This subsection explains **Case control study** and shows how it contributes to the overall Epidemiology workflow.

The goal is to estimate the frequency of exposure in
cases relative to controls.


How to Combat Some of the Weaknesses in a Case-Control Study:Matching Multiple Controls
Blinding of Investigators

#### Cohort study

<!-- generated-learning-note -->
This subsection explains **Cohort study** and shows how it contributes to the overall Epidemiology workflow.

Incidence of a disease (or outcome) is compared
among exposed and unexposed individuals.

#### Matched design

<!-- generated-learning-note -->
This subsection explains **Matched design** and shows how it contributes to the overall Epidemiology workflow.

The cases and controls are not independent of each other. This provides greater efficiency and statistical power. Choose the appropriate statistical test for the association of a paired (matched) binary outcome variable:

- With a continuous, normal variable: Paired t-test (instead of t-test)
- With a continuous, non-normal variable: Wilcoxon signed rank test (instead of
Mann-Whitney-U)
- With a binary variable: McNemar's test (instead of chi-square)


#### Experimental Studies

<!-- generated-learning-note -->
This subsection explains **Experimental Studies** and shows how it contributes to the overall Epidemiology workflow.

 

In an experiment investigators apply
treatments to experimental units
(people, animals, plots of land, etc.)
and then proceed to observe the effect
of the treatments on the experimental
units.

Types of Trials

• Devices (prosthesis, heart valve, joint replacement)
• Procedure (surgery)
• Behavioral change (smoking cessation, dietary change, exercise)
• Pharmaceutical (prevention or treatment)


Masking

Placebo or fake procedure 

Double blind 



Random Assignment

Therefore, any differences in outcome can be attributed to the treatment
and not differences in:

age;
sex;
smoking;
alcohol;
education;
stage/severity of disease;
hospital;
physician;
previous treatments;

#### Phases of Clinical Trials

<!-- generated-learning-note -->
This subsection explains **Phases of Clinical Trials** and shows how it contributes to the overall Epidemiology workflow.



- phase 1 (randomized placebo trial), 3.4m

participants: 20-100 healthy volunteers 

length: several months

purpose: safety and dosage

answer: 
how drug works in the body;
side effect


70% of drugs move to the next phase.




- phase 2 (randomized control trial), 8.6m

participants: several hundred with disease  

length: several months to  years 

purpose: efficacy and side effects 

answer: 
efficacy- how well does treatment perform in idea condition;
side effects

33% of drugs move to the next phase.


- phase 3 (randomized control trial), 21.4m

participants: 300- 3000 with disease  

length: 1-4 years 

purpose: efficacy and monitor adverse reactions 

answer: 
efficacy- how well does treatment perform to a specific population;
side effects- less common side effects are more likely to be detected in these 
larger, longer studies. 

25-33% of drugs move to the next phase.


- phase 4 

participants: several thousand with disease  

time: after fda approval

purpose: efficacy and monitor safety 


#### Sample size, Errors, Power

<!-- generated-learning-note -->
This subsection explains **Sample size, Errors, Power** and shows how it contributes to the overall Epidemiology workflow.


Need enough subjects to see the effect of treatment (if treatment is indeed effective); that is statistical power. Need to balance this with financial and time constraints. Also need to balance the chance of making Type I and II errors.

Determining an Appropriate Sample Size

- difference between groups;
- level of sig;
- desired power;
- two side or one side;
- response/ loss follow- up rate;

#### Randomization

<!-- generated-learning-note -->
This subsection explains **Randomization** and shows how it contributes to the overall Epidemiology workflow.


Study participants are assigned to one of two ( or more)
interventions using an explicit method that assures the
assignment will be random, or by chance.

Any differences in baseline
characteristics of the study
groups indicates breakdown
in the randomization process.

Blocked Randomization: blocks of 10 participants may be randomized at one time. 

#### ITT, PPT

<!-- generated-learning-note -->
This subsection explains **ITT, PPT** and shows how it contributes to the overall Epidemiology workflow.

An “intention‐to‐treat” analysis keeps the randomization
assignment intact at all costs. In an ITT analysis, cross‐over will bias the RR toward the
null, but will NOT bias the RR away from the null. 

Per-protocol analysis is tempting to want to analyze cross‐overs with the
actual treatment they received. Can bias the results toward the null OR away from the null. 

#### Surrogate

<!-- generated-learning-note -->
This subsection explains **Surrogate** and shows how it contributes to the overall Epidemiology workflow.


Measure change in a continuous outcome from baseline. Strength: Intuitive, and results expressed in absolute terms. Limitation: Can not calculate RR, RRR, ARR, NNT, K-M. 

Surrogate: often occur earlier than the clinical outcomes, this:
− reduces cost
− reduces study duration
− reduces study size

#### Bias and Confounding

<!-- generated-learning-note -->
This subsection explains **Bias and Confounding** and shows how it contributes to the overall Epidemiology workflow.


Measured value= true value+ bias+ random

Bias (systematic error: selection+ information+ confounding)+ random= error 



Bias can only be fixed by doing things in an unbiased manner (there is
no other way). However, in some cases, you can estimate the magnitude of the
effect of bias on the study estimates.

#### Selection bias

<!-- generated-learning-note -->
This subsection explains **Selection bias** and shows how it contributes to the overall Epidemiology workflow.


#### Misclassification- information bias

<!-- generated-learning-note -->
This subsection explains **Misclassification- information bias** and shows how it contributes to the overall Epidemiology workflow.

Resulting from non-differential misclassification
– Exposure is under- or over-estimated similarly in cases and controls.
– Cases/non-cases are misclassified similarly in the exposed and
unexposed.
– In this case, associations tend to be biased towards the null.

Resulting from differential misclassification
– Exposure is under- or over-estimated in cases and the opposite in the
in the controls.
– Cases/non-cases are misclassified differently in the exposed and
unexposed
– In this case, associations can be biased towards or away from the null
Slide

#### Confounding

<!-- generated-learning-note -->
This subsection explains **Confounding** and shows how it contributes to the overall Epidemiology workflow.

Confounding: Distortion of an Association
Confounding is not an error in selection or measurement (as
selection and information bias are). Therefore some epidemiologists call confounding a bias, while
others do not.

How to control for confounding: 3 levels
• In selection of participants
– Matching (most commonly in a case-control study):
– Restriction (any study design)
• In assigning the exposure (intervention)
– Randomization (only done in RCT’s)
• In the analysis (can be used in ANY study design)
– Stratification
– Adjustment (multivariable regression)
– Direct/Indirect Standardization
  • When individual data on exposure and outcome are not available
  
  If the adjusted estimate (b) for a given X differs from the unadjusted
estimate, then there was confounding by the other covariates in the
model.

#### Accuracy and precision

<!-- generated-learning-note -->
This subsection explains **Accuracy and precision** and shows how it contributes to the overall Epidemiology workflow.

Validity = Accuracy: How close is the result to the truth?

Reliability= Precision: How close are repeated measures close to one
another?



#### Regression model

<!-- generated-learning-note -->
This section develops **Regression model**. Read the model specification together with its assumptions and the scale of each reported effect.





Regression coefficients in a regression model can be interpreted as “For a 1‐unit increase in X (the exposure variable), the outcome increases by b.”

1. What type of regression model do I see?
2. Are these crude or adjusted estimates, or both?
3. Is this table showing:

– Associations of many exposures (X’s) with the outcome (Y)?
• Does each row represent a separate model with DIFFERENT X’s adjusted for the SAME covariates?
• Does each row represent a separate model with the SAME X adjusted for DIFFERENT covariates?
• The primary X of interest might be presented several different ways (continuous, binary, quartiles)

4. Are the exposures (X’s) all categorical, or are any continuous?
– It is usually quite easy to find the reference group for categorical X’s
– If a continuous variable is there, figure out how it was put into the model
• IF “per X units” is specified, then it is +X units increase
5. How many models are there in the table??


Causation 

mycobacterium tuberculosis is the necessary
cause of tuberculosis but often is not a sufficient cause without poverty,
poor nutrition, overcrowding, etc. Smoking alone can cause lung cancer, but other factors can cause it as well, without smoking being present.


#### Diagnostic Tests

<!-- generated-learning-note -->
This subsection explains **Diagnostic Tests** and shows how it contributes to the overall Epidemiology workflow.


Goal: to distinguish accurately between diseased
and non‐diseased individuals

ALL diagnostic tests suffer from some level of
inaccuracy. We need a way to quantify the accuracy of a
given test. Usually: overlap between healthy and diseased states
 makes a “cut‐point” difficult to define.

#### Bayes' theorem

<!-- generated-learning-note -->
This subsection explains **Bayes' theorem** and shows how it contributes to the overall Epidemiology workflow.

Does my trust of the same test result differ from
patient to patient or setting to setting? If I receive a “negative” test result, how much can I
trust it?Prevalence can affect PPV and NPV dramatically.Prevalence itself does NOT affect sensitivity and
specificity.   

$$
P(B|A) = \frac{P(A|B)P(B)}{P(A)}
$$

#### Sensitivity and specificity

<!-- generated-learning-note -->
This subsection explains **Sensitivity and specificity** and shows how it contributes to the overall Epidemiology workflow.


• PPV is maximized when re‐test probability is high

• NPV is maximized when Pre‐test probability is low


• Reliability (Relates to Precision)
– Many measures (% Agreement, Kappa, Intraclass correlation
coefficient, Crohnbach’s alpha)

• Validity (Relates to Accuracy)
– Many measures (Youden’s J‐statistic, construct validity,
content validity, etc)
– Sensitivity
– Specificity
– Positive Predictive Value
– Negative Predictive Value

Sensitivity is increased at the expense of specificity.

Screening: Does finding and treating the disease EARLIER
improve the OUTCOME? Screening tests tend to maximize Sensitivity at
expense of PPV. diagnosis have post test probabilities near 0 or 1. Screening is only part of a preventive medicine and health
maintenance program.

### Applying the Material

This source is primarily conceptual or procedural. Use its definitions and decision points as a checklist, and document how each item is implemented in the intended study or analysis.

## Interpreting the Results

Identify the source population, time zero, eligibility criteria, measurement process, and target effect before interpreting estimates. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
