---
title: "Simulations and Bootstrapping"
output: html_document
date: '2023-12-10'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Simulations and Bootstrapping

The bootstrapping approach will always work because it doesn't assume any underlying distribution of the data. 

- Test the robust of t test what if violation of the normality assumption 

`How sensitive is the two-sample t-test to violations of the normality assumption? Test this by using the rpois() function, and try different lambda values`

```{r}
set.seed(123)
hist(rpois(100, 10))
hist(rpois(100, 5))
hist(rpois(100, 2)) #not normality intuition

psum=c()
r=1
while (r<=20) {
  r=r+1
  count=1
  pval=c()
while ( count<=1000) {
  sample1=rpois(100,r) #re sample
  sample2=rpois(100,r)
  pval=c(pval,t.test(sample1,sample2)$p.value)
  count=count+1
}
  psum=c(psum,sum(pval<0.05)/1000)
}
psum

```


`two-sample t-test is robust even violation of the normality assumption but it will be not applicable under some extrem conditions.  `


`The purpose of the t-test is to compare certain characteristics representing groups, and the mean values become representative when the population has a normal distribution. This is the reason why satisfaction of the normality assumption is essential in the t-test.`

- ANOVA by bootstrapping


`Consider what null hypothesis (Sample with replacement to create empirical sampling distributions); After creating bootstrap samples, run statistical test; p-value= proportion of times bootstrap data show statistics more extreme than the observed statistic (and CI); `


```{r}
# example 1
hist(iris$Petal.Width) #violation of normal distribution
one.way <- aov(Petal.Width ~ Species, data = iris)
anov=summary(one.way)[[1]]$`F value`[1]
summary(one.way)

set.seed(123)
count=1
boot=c()
while ( count<=1000) {
  sample1=iris[sample(1:nrow(iris),replace=T,size =  table(iris$Species)[1]),]
  sample1$group=1
  sample2=iris[sample(1:nrow(iris),replace=T,size =  table(iris$Species)[2]),]
  sample2$group=2
  sample3=iris[sample(1:nrow(iris),replace=T,size =  table(iris$Species)[3]),]
  sample3$group=3
  sample=rbind(sample1,sample2,sample3)
  
  test=aov(Petal.Width ~ as.factor(group), data = sample)
  boot =c(boot,summary(test)[[1]]$`F value`[1]) 
  
  count=count+1
}
(sum(abs(boot)>abs(anov))/1000) # Equivalent to p value

hist(boot)

quantile(boot,.025)
quantile(boot,.975)

```

- The Central Limit Theorem

`sampling distribution of means from a population will be approximately normally distributed regardless of the underlying population distribution.`

```{r}
set.seed(123)
hist(rpois(1000, 2))

r=1
while (r<=10) {
  r=r+1
  count=1
  pval=c()
  while ( count<=100) {
    sample1=rpois(30,r)
    pval=c(pval,mean(sample1))
    count=count+1
  }
  hist(pval, prob = TRUE)
  lines(density(pval), col = 4, lwd = 2)
}
```

`the means (not other statistics, like median) follow the normal distribution but it also depends on the sample size and the original distribution. `

```{r}
# if median
r=1
while (r<=10) {
  r=r+1
  count=1
  pval=c()
  while ( count<=1000) {
    sample1=rpois(30,r)
    pval=c(pval,median(sample1))
    count=count+1
  }
  hist(pval, prob = TRUE)
  lines(density(pval), col = 4, lwd = 2)
}
```



- Test of Normality

`The critical assumption for this type of test concerns the normality of residuals, not the dependent variable. Also, with a large sample size the null of the Shapiro-Wilk test will often be rejected.` 


```{r}
hist(residuals(one.way)) 
#Create a Normal Q-Q Plot of the residuals for "model_job_loss".
qqnorm(residuals(one.way), main = "Normal Q-Q Plot of Residuals")
qqline(residuals(one.way))

shapiro.test(residuals(one.way))
```


- Kruskal-Wallis Rank Sum Test (nonparametric)
```{r}
kruskal.test(iris$Petal.Width ,iris$Species)
```










