---
title: "Data visualization"
description: "A practical R learning guide to Data visualization, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Data Management & Programming"
category: "Data Cleaning & Quality Control"
topic: "Data visualization"
type: article
tags: ["Data Cleaning & Quality Control", "Data Management & Programming", "Data", "visualization"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: false
---

## Why This Topic Matters

Data cleaning and quality control make analytical results traceable and reduce errors caused by inconsistent or implausible values.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Data visualization**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Scatter plot
2. Create a empty canvas
3. Add a layer/geom of points to the canvas
4. Add another aesthetic
5. Add other aesthetic
6. Bar chart

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Data Management & Programming",
  category = "Data Cleaning & Quality Control",
  topic = "Data visualization"
)
learning_overview
```

## Learning Material

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
#| eval: false
#| code-fold: true
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(echo = TRUE)

```

- Loading data set
```{r}
library(printr)
library(tidyverse)
head(iris)
```
#### Scatter plot
##### Create a empty canvas
- then create `aesthetic mapping`  
- tell the function which dataset and variables to use
```{r}
ggplot(data = iris,        # which data set? canvas? 
       aes(x=Sepal.Length , y=Petal.Length   ))  # which variables as aesthetics? x and y are mapped to columns of the data; different geoms can have different aesthetics (different variables). 
```

 
##### Add a layer/geom of `points` to the canvas
```{r}
ggplot(data = iris,
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point()   # adding the geometrical representation 
```
 
```{r,eval=F}
# same plot as above
ggplot(data = iris) + 
  geom_point(  aes(x=Sepal.Length , y=Petal.Length  )) 
```
##### Add another aesthetic
- add a curve/straight line to fit these points
- geom provides the aesthetic to ggplot
```{r}
# Loess curve
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point() +
  geom_smooth() 

```

```{r}
# Linear regression line
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

##### Add other aesthetic
- set other aesthetics `colour, alpha (transparency), and size of points`

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width     ), 
             alpha = .5, 
             colour = "red")
```

 

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=Species), #white is a variable here 
             alpha=.9)
```

 
- categorize `Petal.Width` then map colour to this new variable
```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(data=iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=growth), 
             alpha=.9)
```

 
#### Bar chart
```{r}
ggplot(data = iris) + 
  geom_bar(aes(x = growth))
```

- bar chart after `group_by` 
- then use `stat='identity'` 
```{r}
results  <- iris %>% 
  group_by(Species, growth) %>% 
  summarize(Sepal.Length.mean=mean (Sepal.Length ))

 gop <- results  %>% 
  filter(Species != "setosa_null"  )
 gop
```
- though meaningless below until line chart (just use the mean as the sum for demonstration)
```{r}
# We can also store parts of a plot in an object
plot1 <- ggplot(gop) + 
  geom_bar(aes(x=growth , y=Sepal.Length.mean), 
           
           stat='identity')
plot1
```
##### Add some options for the whole ggplot rather than layers
- `switch` the x and y axes 
```{r}
plot1 + 
  coord_flip()
```

- `reorder` x categories (`-`means descending)
```{r}
ggplot( gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip()
```

- add x axis `label` and a `theme` 
```{r}
ggplot(gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip() + 
  xlab("Growth categories") + 
  guides(fill=F) +
  theme_minimal()
```

- set theme
```{r}
library(ggthemes)
ggplot(data = iris) + 
  geom_bar(aes(x = growth)) + 
  theme_economist()
```

##### Grouped bar chart 
-bar chart with different panels
```{r}
ggplot(mpg, aes(x = class)) + 
  geom_bar() + 
  facet_wrap( ~ year)
```

- actual number (groups are stacked by default)
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species)
           , stat='identity'
           )
```

```{r}
ggplot(mpg, aes(x = class )) + 
  geom_bar(aes(group = year, fill = year), position = "stack")
```

- percentage
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='fill')
```
 
- groups are dodge with `actual number `
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='dodge')

```
- groups are dodge with `percentage`
```{r}
gop2 <- gop %>% 
  group_by(growth ) %>% 
  mutate(Sepal.Length.prop=Sepal.Length.mean/sum(Sepal.Length.mean))

ggplot(gop2) + 
  geom_bar(aes(x=growth, y=Sepal.Length.prop, fill=Species), 
           stat='identity', 
           position='dodge') + 
  ylab("Votes (%)")
```

 
#### Line charts
 
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length))
```
 
##### Grouped by `colour variable`
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length, colour = Species))
```

 
- grouped by state then set how many rows or columns
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length) ) + 
  
  facet_wrap(~Species, nrow = 1) +      #set how many rows
  coord_flip()  
```

##### Multiple aesthetics
```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) + 
  geom_line(size=2,color="purple")+  
  # number format
  scale_x_log10(labels = scales::label_number())+
  geom_point( aes(size =  Sepal.Length,colour = as.factor(growth)),show.legend = F)+
  facet_wrap(~ Species) 

```

## Interpreting the Results

Use summaries and visual checks to confirm ranges, distributions, duplicates, and missing-data patterns. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
