---
title: "BOIN, CRM and 3+3"
output: html_document
date: "2026-06-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### BOIN, CRM and 3+3  

#### BOIN    
 
- Define trial parameters (target DLT, cohort size, number of cohorts)
- Generate BOIN decision boundaries
- Extract escalation and de-escalation cutoffs
- Input current cohort toxicity data
- Calculate observed toxicity rate
- Compare toxicity rate with BOIN boundaries
- Make dose decision: escalate, stay, or de-escalate
 
```{r  }
# =========================================================
# Simple BOIN Example in R
# Single-Agent Phase I Dose Escalation Trial
# =========================================================

library(BOIN)

# =========================================================
# Step 1: Define Trial Design Parameters
# =========================================================

target <- 0.30
# Target DLT (Dose Limiting Toxicity) rate = 30%

ncohort <- 10
# Maximum number of cohorts planned in the trial

cohortsize <- 3
# Number of patients enrolled per cohort

# =========================================================
# Step 2: Generate BOIN Decision Boundaries
# =========================================================

bound <- get.boundary(
  target = target,
  ncohort = ncohort,
  cohortsize = cohortsize
)

# Print BOIN design summary and boundaries
print(bound)

# =========================================================
# Step 3: Extract Escalation / De-escalation Boundaries
# =========================================================

lambda_e <- bound$lambda_e
# Escalation boundary
# If observed toxicity rate is BELOW this value,
# escalate to the next higher dose

lambda_d <- bound$lambda_d
# De-escalation boundary
# If observed toxicity rate is ABOVE this value,
# de-escalate to the next lower dose

# Print boundaries
cat("Escalation Boundary (lambda_e):", lambda_e, "\n")
cat("De-escalation Boundary (lambda_d):", lambda_d, "\n")

# =========================================================
# Step 4: Current Cohort Data
# =========================================================

current_n <- 3
# Number of patients treated at current dose

current_tox <- 0
# Number of DLTs observed at current dose

# Calculate observed toxicity rate
obs_rate <- current_tox / current_n

cat("Observed Toxicity Rate:", obs_rate, "\n")

# =========================================================
# Step 5: BOIN Dose Escalation Decision
# =========================================================

if (obs_rate < lambda_e) {
  
  decision <- "Escalate"
  
} else if (obs_rate > lambda_d) {
  
  decision <- "De-escalate"
  
} else {
  
  decision <- "Stay at Current Dose"
  
}

# Print final decision
cat("BOIN Decision:", decision, "\n")
``` 
#### CRM    
 
- Define CRM design parameters. 
Set the target DLT rate, dose levels, and prior toxicity skeleton.
- Create observed trial data. 
Enter patient-level dose assignment and DLT outcomes.
- Fit the CRM model. 
Use the observed data to update toxicity probabilities for all dose levels.
- Summarize estimated toxicity. 
Compare prior skeleton values with posterior estimated toxicity probabilities.
- Select next recommended dose. 
Choose the dose with estimated toxicity closest to the target DLT rate.
- Output recommendation. 
Print the recommended dose level for the next cohort.

 
```{r  }
# =========================================================
# Simple CRM Example in R
# Single-Agent Phase I Dose Escalation Trial
# =========================================================

# install.packages("dfcrm")
library(dfcrm)

# =========================================================
# Step 1: Define CRM Design Parameters
# =========================================================

target <- 0.30
# Target DLT rate = 30%

dose_level <- 1:5
# Five dose levels

prior <- c(0.05, 0.10, 0.20, 0.35, 0.50)
# CRM skeleton:
# Initial guesses of DLT probability for each dose level
# Must be increasing with dose

# =========================================================
# Step 2: Create Observed Trial Data
# =========================================================
# Example:
# Cohort 1: Dose 1, 0/3 DLT
# Cohort 2: Dose 2, 0/3 DLT
# Cohort 3: Dose 3, 1/3 DLT
# Cohort 4: Dose 3, 1/3 DLT

level <- c(
  rep(1, 3),
  rep(2, 3),
  rep(3, 3),
  rep(3, 3)
)

tox <- c(
  0, 0, 0,   # Dose 1: 0/3 DLT
  0, 0, 0,   # Dose 2: 0/3 DLT
  0, 0, 1,   # Dose 3: 1/3 DLT
  0, 1, 0    # Dose 3: 1/3 DLT
)

trial_data <- data.frame(
  patient_id = 1:length(tox),
  dose_level = level,
  dlt = tox
)

print(trial_data)

# =========================================================
# Step 3: Fit CRM Model
# =========================================================

fit <- crm(
  prior = prior,
  target = target,
  tox = tox,
  level = level,
  model = "empiric",
  method = "bayes"
)

print(fit)

# =========================================================
# Step 4: Extract CRM Estimated Toxicity Probabilities
# =========================================================

estimated_tox <- fit$ptox

crm_summary <- data.frame(
  dose_level = dose_level,
  skeleton = prior,
  estimated_toxicity = estimated_tox
)

print(crm_summary)

# =========================================================
# Step 5: Recommended Dose for Next Cohort
# =========================================================

recommended_dose <- fit$mtd

cat("Recommended dose level for next cohort:", recommended_dose, "\n")
```




#### 3+3     
 
- Define patient dose levels and DLT outcomes
- Create trial dataset
- Summarize number of patients and DLTs at each dose
- Apply 3+3 rules sequentially:  
0/3 DLT → escalate
1/3 DLT → expand to 6 patients
≤1/6 DLT → escalate
≥2/3 or ≥2/6 DLT → de-escalate/stop
-  Select the previous safe dose as the MTD 

 
```{r  }
# =========================================================
# Simple 3+3 Design Example in R
# Single-Agent Phase I Dose Escalation Trial
# =========================================================

# =========================================================
# Step 1: Define Observed Trial Data
# =========================================================

# Example:
# Cohort 1: Dose 1, 0/3 DLT  -> escalate
# Cohort 2: Dose 2, 0/3 DLT  -> escalate
# Cohort 3: Dose 3, 1/3 DLT  -> expand dose 3
# Cohort 4: Dose 3, 0/3 DLT  -> dose 3 has 1/6 DLT, escalate
# Cohort 5: Dose 4, 2/3 DLT  -> de-escalate, stop
# Final MTD = Dose 3

dose_level <- c(
  rep(1, 3),
  rep(2, 3),
  rep(3, 3),
  rep(3, 3),
  rep(4, 3)
)

dlt <- c(
  0, 0, 0,   # Dose 1: 0/3 DLT
  0, 0, 0,   # Dose 2: 0/3 DLT
  0, 0, 1,   # Dose 3: 1/3 DLT
  0, 0, 0,   # Dose 3: total 1/6 DLT
  1, 1, 0    # Dose 4: 2/3 DLT
)

trial_data <- data.frame(
  patient_id = 1:length(dlt),
  dose_level = dose_level,
  dlt = dlt
)

print(trial_data)

# =========================================================
# Step 2: Summarize DLT by Dose
# =========================================================

summary_by_dose <- aggregate(
  dlt ~ dose_level,
  data = trial_data,
  FUN = function(x) c(n = length(x), tox = sum(x), rate = mean(x))
)

summary_by_dose <- do.call(data.frame, summary_by_dose)
colnames(summary_by_dose) <- c("dose_level", "n", "dlt", "dlt_rate")

print(summary_by_dose)

# =========================================================
# Step 3: Apply Simple 3+3 Rule Manually
# =========================================================

cat("3+3 decision path:\n")
cat("Dose 1: 0/3 DLT -> Escalate to Dose 2\n")
cat("Dose 2: 0/3 DLT -> Escalate to Dose 3\n")
cat("Dose 3: 1/3 DLT -> Expand Dose 3 to 6 patients\n")
cat("Dose 3: 1/6 DLT -> Escalate to Dose 4\n")
cat("Dose 4: 2/3 DLT -> Stop and de-escalate\n")
cat("Final MTD: Dose 3\n")
```


