---
title: "BOIN, CRM and 3+3"
description: "A practical R learning guide to BOIN, CRM and 3+3, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Clinical Trial Process & Design"
category: "Trial Designs"
topic: "BOIN, CRM and 3+3"
type: article
tags: ["Trial Designs", "Clinical Trial Process & Design", "BOIN", "CRM"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Clinical trial design connects the scientific question to an ethical and operational plan for generating evidence.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **BOIN, CRM and 3+3**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. BOIN, CRM and 3+3
2. BOIN
3. CRM
4. 3+3

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Clinical Trial Process & Design",
  category = "Trial Designs",
  topic = "BOIN, CRM and 3+3"
)
learning_overview
```

## Learning Material

The following material develops **BOIN, CRM and 3+3** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### BOIN, CRM and 3+3

<!-- generated-learning-note -->
This subsection explains **BOIN, CRM and 3+3** and shows how it contributes to the overall BOIN, CRM and 3+3 workflow.


#### BOIN

<!-- generated-learning-note -->
This subsection explains **BOIN** and shows how it contributes to the overall BOIN, CRM and 3+3 workflow.

 
- Define trial parameters (target DLT, cohort size, number of cohorts)
- Generate BOIN decision boundaries
- Extract escalation and de-escalation cutoffs
- Input current cohort toxicity data
- Calculate observed toxicity rate
- Compare toxicity rate with BOIN boundaries
- Make dose decision: escalate, stay, or de-escalate

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r  }
# =========================================================
# Simple BOIN Example in R
# Single-Agent Phase I Dose Escalation Trial
# =========================================================

library(BOIN)

# =========================================================
# Step 1: Define Trial Design Parameters
# =========================================================

target <- 0.30
# Target DLT (Dose Limiting Toxicity) rate = 30%

ncohort <- 10
# Maximum number of cohorts planned in the trial

cohortsize <- 3
# Number of patients enrolled per cohort

# =========================================================
# Step 2: Generate BOIN Decision Boundaries
# =========================================================

bound <- get.boundary(
  target = target,
  ncohort = ncohort,
  cohortsize = cohortsize
)

# Print BOIN design summary and boundaries
print(bound)

# =========================================================
# Step 3: Extract Escalation / De-escalation Boundaries
# =========================================================

lambda_e <- bound$lambda_e
# Escalation boundary
# If observed toxicity rate is BELOW this value,
# escalate to the next higher dose

lambda_d <- bound$lambda_d
# De-escalation boundary
# If observed toxicity rate is ABOVE this value,
# de-escalate to the next lower dose

# Print boundaries
cat("Escalation Boundary (lambda_e):", lambda_e, "\n")
cat("De-escalation Boundary (lambda_d):", lambda_d, "\n")

# =========================================================
# Step 4: Current Cohort Data
# =========================================================

current_n <- 3
# Number of patients treated at current dose

current_tox <- 0
# Number of DLTs observed at current dose

# Calculate observed toxicity rate
obs_rate <- current_tox / current_n

cat("Observed Toxicity Rate:", obs_rate, "\n")

# =========================================================
# Step 5: BOIN Dose Escalation Decision
# =========================================================

if (obs_rate < lambda_e) {
  
  decision <- "Escalate"
  
} else if (obs_rate > lambda_d) {
  
  decision <- "De-escalate"
  
} else {
  
  decision <- "Stay at Current Dose"
  
}

# Print final decision
cat("BOIN Decision:", decision, "\n")
``` 

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

#### CRM

<!-- generated-learning-note -->
This subsection explains **CRM** and shows how it contributes to the overall BOIN, CRM and 3+3 workflow.

 
- Define CRM design parameters. 
Set the target DLT rate, dose levels, and prior toxicity skeleton.
- Create observed trial data. 
Enter patient-level dose assignment and DLT outcomes.
- Fit the CRM model. 
Use the observed data to update toxicity probabilities for all dose levels.
- Summarize estimated toxicity. 
Compare prior skeleton values with posterior estimated toxicity probabilities.
- Select next recommended dose. 
Choose the dose with estimated toxicity closest to the target DLT rate.
- Output recommendation. 
Print the recommended dose level for the next cohort.

 

<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r  }
#| eval: false
#| code-fold: true
# =========================================================
# Simple CRM Example in R
# Single-Agent Phase I Dose Escalation Trial
# =========================================================

# install.packages("dfcrm")
library(dfcrm)

# =========================================================
# Step 1: Define CRM Design Parameters
# =========================================================

target <- 0.30
# Target DLT rate = 30%

dose_level <- 1:5
# Five dose levels

prior <- c(0.05, 0.10, 0.20, 0.35, 0.50)
# CRM skeleton:
# Initial guesses of DLT probability for each dose level
# Must be increasing with dose

# =========================================================
# Step 2: Create Observed Trial Data
# =========================================================
# Example:
# Cohort 1: Dose 1, 0/3 DLT
# Cohort 2: Dose 2, 0/3 DLT
# Cohort 3: Dose 3, 1/3 DLT
# Cohort 4: Dose 3, 1/3 DLT

level <- c(
  rep(1, 3),
  rep(2, 3),
  rep(3, 3),
  rep(3, 3)
)

tox <- c(
  0, 0, 0,   # Dose 1: 0/3 DLT
  0, 0, 0,   # Dose 2: 0/3 DLT
  0, 0, 1,   # Dose 3: 1/3 DLT
  0, 1, 0    # Dose 3: 1/3 DLT
)

trial_data <- data.frame(
  patient_id = 1:length(tox),
  dose_level = level,
  dlt = tox
)

print(trial_data)

# =========================================================
# Step 3: Fit CRM Model
# =========================================================

fit <- crm(
  prior = prior,
  target = target,
  tox = tox,
  level = level,
  model = "empiric",
  method = "bayes"
)

print(fit)

# =========================================================
# Step 4: Extract CRM Estimated Toxicity Probabilities
# =========================================================

estimated_tox <- fit$ptox

crm_summary <- data.frame(
  dose_level = dose_level,
  skeleton = prior,
  estimated_toxicity = estimated_tox
)

print(crm_summary)

# =========================================================
# Step 5: Recommended Dose for Next Cohort
# =========================================================

recommended_dose <- fit$mtd

cat("Recommended dose level for next cohort:", recommended_dose, "\n")
```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.





#### 3+3

<!-- generated-learning-note -->
This subsection explains **3+3** and shows how it contributes to the overall BOIN, CRM and 3+3 workflow.

 
- Define patient dose levels and DLT outcomes
- Create trial dataset
- Summarize number of patients and DLTs at each dose
- Apply 3+3 rules sequentially:  
0/3 DLT → escalate
1/3 DLT → expand to 6 patients
≤1/6 DLT → escalate
≥2/3 or ≥2/6 DLT → de-escalate/stop
-  Select the previous safe dose as the MTD 

 

<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r  }
#| eval: false
#| code-fold: true
# =========================================================
# Simple 3+3 Design Example in R
# Single-Agent Phase I Dose Escalation Trial
# =========================================================

# =========================================================
# Step 1: Define Observed Trial Data
# =========================================================

# Example:
# Cohort 1: Dose 1, 0/3 DLT  -> escalate
# Cohort 2: Dose 2, 0/3 DLT  -> escalate
# Cohort 3: Dose 3, 1/3 DLT  -> expand dose 3
# Cohort 4: Dose 3, 0/3 DLT  -> dose 3 has 1/6 DLT, escalate
# Cohort 5: Dose 4, 2/3 DLT  -> de-escalate, stop
# Final MTD = Dose 3

dose_level <- c(
  rep(1, 3),
  rep(2, 3),
  rep(3, 3),
  rep(3, 3),
  rep(4, 3)
)

dlt <- c(
  0, 0, 0,   # Dose 1: 0/3 DLT
  0, 0, 0,   # Dose 2: 0/3 DLT
  0, 0, 1,   # Dose 3: 1/3 DLT
  0, 0, 0,   # Dose 3: total 1/6 DLT
  1, 1, 0    # Dose 4: 2/3 DLT
)

trial_data <- data.frame(
  patient_id = 1:length(dlt),
  dose_level = dose_level,
  dlt = dlt
)

print(trial_data)

# =========================================================
# Step 2: Summarize DLT by Dose
# =========================================================

summary_by_dose <- aggregate(
  dlt ~ dose_level,
  data = trial_data,
  FUN = function(x) c(n = length(x), tox = sum(x), rate = mean(x))
)

summary_by_dose <- do.call(data.frame, summary_by_dose)
colnames(summary_by_dose) <- c("dose_level", "n", "dlt", "dlt_rate")

print(summary_by_dose)

# =========================================================
# Step 3: Apply Simple 3+3 Rule Manually
# =========================================================

cat("3+3 decision path:\n")
cat("Dose 1: 0/3 DLT -> Escalate to Dose 2\n")
cat("Dose 2: 0/3 DLT -> Escalate to Dose 3\n")
cat("Dose 3: 1/3 DLT -> Expand Dose 3 to 6 patients\n")
cat("Dose 3: 1/6 DLT -> Escalate to Dose 4\n")
cat("Dose 4: 2/3 DLT -> Stop and de-escalate\n")
cat("Final MTD: Dose 3\n")
```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.

## Interpreting the Results

Focus on the design assumptions, decision rules, operating characteristics, and how they affect interpretation. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
