---
title: "Bias analysis and control"
output: html_document
date: '2024-01-22'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Refrerence](https://cran.r-project.org/web/packages/episensr/vignettes/episensr.html)



### Selection Bias

Selection Bias, occurs when the study selected population doesn’t `represent` the target
population (like refuse answer and leave study).

#### No selection bias


- if the response rate is 100%. 

bias_parms is 4 probabilities: among cases exposed (1), among cases unexposed(2), among noncases exposed(3), and among noncases unexposed(4). 

```{r}
library(episensr)

stang <- selection(matrix(c(136, 107, 297, 165),
                          dimnames = list(c("UM+", "UM-"), c("Mobile+", "Mobile-")),
                          nrow = 2, byrow = TRUE),
                   bias_parms = c(1, 1, 1, 1))
stang
```

or 

```{r}
library(episensr)

stang <- selection(matrix(c(136, 107, 297, 165),
                          dimnames = list(c("UM+", "UM-"), c("Mobile+", "Mobile-")),
                          nrow = 2, byrow = TRUE),
                   bias_parms = 1)
stang
```



####  Corrected selection bias

by 4 probabilities: among cases exposed (1), among cases unexposed(2), among noncases exposed(3), and among noncases unexposed(4). 

or a single positive selection-bias factor which is **the ratio of the exposed versus
unexposed selection probabilities (important)** comparing cases and noncases [(1/2)/(3/4)=0.8]. 

- If there is 0.8 response rate in nocase exposed, there is a selection bias in the control group because not good representative. The association is toward null. 

```{r}
library(episensr)

stang <- selection(matrix(c(136, 107, 297, 165),
                          dimnames = list(c("UM+", "UM-"), c("Mobile+", "Mobile-")),
                          nrow = 2, byrow = TRUE),
                   bias_parms = 1/0.8)
stang
```

or



```{r}
library(episensr)

stang <- selection(matrix(c(136, 107, 297, 165),
                          dimnames = list(c("UM+", "UM-"), c("Mobile+", "Mobile-")),
                          nrow = 2, byrow = TRUE),
                   bias_parms = c(1, 1, 0.8, 1))
stang
```


- If the response rate is 0.45-0.9 in case group, can not correct selection bias by this approach. you have to compare the characters of response population to no response population.


```{r}
library(episensr)

stang <- selection(matrix(c(136, 107, 297, 165),
                          dimnames = list(c("UM+", "UM-"), c("Mobile+", "Mobile-")),
                          nrow = 2, byrow = TRUE),
                   bias_parms = c(0.45, 0.9, 0.5, 1))
stang
```

### Information Bias (Misclassification)

Highly sensitive tests (>90%) yield few false negatives, so a negative result constitutes strong evidence to rule out or exclude the target disease. Conversely, highly specific tests (>90%) generate a very low number of false positives, so a positive result strongly implies that the target disease/condition is present. Note the prevalence (low) also affect the `diagnosis` test result. The association is toward to null. 

```{r}
misclassification(matrix(c(126, 92, 71, 224),
                         dimnames = list(c("Case", "Control"),
                                         c("Smoking +", "Smoking - ")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure",
                  bias_parms =c(0.90, 0.90, 0.90, 0.90))
```

#### No differential miscllasification 

non-differential misclassification will always bias towards the null.

- Let’s say the sensitivity of self-reported smoking is 94% and specificity is 97%, for both the case and control groups. 

The argument bias_parms should be made of the following components: (1) Sensitivity of classification among those with the outcome, (2) Sensitivity of classification among those without the outcome, (3) Specificity of classification among those with the outcome, and (4) Specificity of classification among those without the outcome.

```{r}
misclassification(matrix(c(126, 92, 71, 224),
                         dimnames = list(c("Case", "Control"),
                                         c("Smoking +", "Smoking - ")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure",
                  bias_parms = c(0.94, 0.94, 0.97, 0.97))
```


#### Differential miscllasification 

The effects of differential misclassification and selection bias will depend on the details.

Let’s say the sensitivity of self-reported smoking and specificity are different in the case and control groups. 

- Higher sensitivity (false positive) are in the case group. The association is away null. 


```{r}
misclassification(matrix(c(126, 92, 71, 224),
                         dimnames = list(c("Case", "Control"),
                                         c("Smoking +", "Smoking - ")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure",
                  bias_parms = c(0.95, 1, 0.80, 1))
```

- Higher sensitivity (false positive) are in the control group. The association is toward null. 


```{r}
misclassification(matrix(c(126, 92, 71, 224),
                         dimnames = list(c("Case", "Control"),
                                         c("Smoking +", "Smoking - ")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure",
                  bias_parms = c(1, 0.95, 1, 0.80))

```

- Higher sensitivity (false positive) are in the case group with low sensitivity and specificity in the control group. The association is toward null. 


```{r}
misclassification(matrix(c(126, 92, 71, 224),
                         dimnames = list(c("Case", "Control"),
                                         c("Smoking +", "Smoking - ")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure",
                  bias_parms = c(0.95, 0.80, 0.80, 0.80))
```

- Higher sensitivity (false positive) are in the control group with low sensitivity and specificity in the case group. The association is toward null. 


```{r}
misclassification(matrix(c(126, 92, 71, 224),
                         dimnames = list(c("Case", "Control"),
                                         c("Smoking +", "Smoking - ")),
                         nrow = 2, byrow = TRUE),
                  type = "exposure",
                  bias_parms = c(0.80, 0.95, 0.80, 0.80))

```

#### Covariate misclassification

- No covariate misclassification

```{r}
misclassification.cov(array(c(1319, 38054, 5641, 405546, 565, 3583,
                              781, 21958, 754, 34471, 4860, 383588),
                            dimnames = list(c("Twins+", "Twins-"),
                                            c("Folic acid+", "Folic acid-"),
                                            c("Total", "IVF+", "IVF-")),
                            dim = c(2, 2, 3)),
                      bias_parms = NULL)
```

- No differential misclassifcation

Data on IVF were not available and a proxy for it was used, period of involuntary childlessness. However, it was a poor proxy for IVF, with a sensitivity of 60% (here use 0.9) and a specificity of 95%. These bias parameters were assumed to be nondifferential.

```{r}
misclassification.cov(array(c(1319, 38054, 5641, 405546, 565, 3583,
                              781, 21958, 754, 34471, 4860, 383588),
                            dimnames = list(c("Twins+", "Twins-"),
                                            c("Folic acid+", "Folic acid-"),
                                            c("Total", "IVF+", "IVF-")),
                            dim = c(2, 2, 3)),
                      bias_parms = c(.9, .9, .95, .95))
```
- Differential misclassifcation

```{r}
misclassification.cov(array(c(1319, 38054, 5641, 405546, 565, 3583,
                              781, 21958, 754, 34471, 4860, 383588),
                            dimnames = list(c("Twins+", "Twins-"),
                                            c("Folic acid+", "Folic acid-"),
                                            c("Total", "IVF+", "IVF-")),
                            dim = c(2, 2, 3)),
                      bias_parms = c(0.6, 0.95, 0.95, 0.95))
```

The mantel haenszel association with missclassifiation is away null.


### Confounding bias

#### unmeasured or unknown confounders

The frequency of drug-attributed rash in relation to allopurinol exposure with sex treated as a potential confounding factor (`stratification`) 


```{r}
library(tidyverse)
RR_no=0.5249886 # RR between confounder and outcome among non-exposed

p1=0.5671642  # prevalence of confounder among exposed

p2=0.5417661  # prevalence of confounder among unexposed 

rash <- matrix(c(15, 94, 52, 1163),
               dimnames = list(c("Rash +", "Rash -"),
                               c("Allopurinol +", "Allopurinol -")),
               nrow = 2, byrow = TRUE)

rash %>% confounders(., type = "RR", bias_parms = c(RR_no, p1, p2))
```

- Generate three matrix by confounders or get these parameters from other researches

```{r}
## Outcome by confounders
rash_conf <- matrix(c(36, 58, 645, 518),
                    dimnames = list(c("Rash +", "Rash -"),
                                    c("Males", "Females")),
                    nrow = 2, byrow = TRUE)
rash_conf

## By confounders: among males
rash_males <- matrix(c(5, 36, 33, 645),
                     dimnames = list(c("Rash +", "Rash -"),
                                     c("Allopurinol +", "Allopurinol -")),
                     nrow = 2, byrow = TRUE)
rash_males

## By confounders: among females
rash_females <- matrix(c(10, 58, 19, 518),
                       dimnames = list(c("Rash +", "Rash -"),
                                       c("Allopurinol +", "Allopurinol -")),
                       nrow = 2, byrow = TRUE)
rash_females
```

```{r}
(RR_no <- (36/(36+645))/(58/(58+518))) # RR between confounder and outcome among non-exposed
## [1] 0.5249886
(p1 <- (5+33)/(15+52)) # prevalence of confounder among exposed
## [1] 0.5671642
(p2 <- (36+645)/(94+1163)) # prevalence of confounder among unexposed
## [1] 0.5417661
```



- Adjust RR directly 


```{r}
confounders.ext(RR = 2, bias_parms = c(0.1, 0.9, 0.1, 0.4))
```



### Summary
 

* There are several mechanisms that can produce selectinon bias:

  - Selection of a comparison group ("controls") that is not representative of the population that produced the cases in a case-control study. (Control selection bias)
  - Differential loss to follow up in a cohort study, such that the likelihood of being lost to follow up is related to outcome status and exposure status. (Loss to follow-up bias)
  - Refusal, non-response, or agreement to participate that is related to the exposure and disease (Self-selection bias)
  - Using the general population as a comparison group for an occupational cohort study ("Healthy worker effect")
  - Differential referral or diagnosis of subjects
  
* There are several mechanisms that can produce misclassification bias:  
  - Recall bias
  - Interviewer Bias
  - Differences in the Quality of Information

* Unlike selection and information bias, which can be introduced by the investigator or by the subjects, confounding is a type of bias that can be adjusted for in the analysis.


* Confounding bias can be analyzed by regression and the association can be adjusted by data from other studies (instrument variable, Bayesian).

* Selection bias and misclassification bias can be adjusted. selection bias can be identified by no response rate (or comparing the characteristics) and misclassification bias can be identified by validity diagnostic test. 











