---
title: "Data summarization"
description: "A practical R learning guide to Data summarization, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Data Management & Programming"
category: "Data Cleaning & Quality Control"
topic: "Data summarization"
type: article
tags: ["Data Cleaning & Quality Control", "Data Management & Programming", "Data", "summarization"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Data cleaning and quality control make analytical results traceable and reduce errors caused by inconsistent or implausible values.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Data summarization**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. How to do aggregation/ summarization
2. Summarization after grouping
3. Summarization with upgroup
4. Mutate new variables after grouping
5. Recode and generate new variables, then value label

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Data Management & Programming",
  category = "Data Cleaning & Quality Control",
  topic = "Data summarization"
)
learning_overview
```

## Learning Material

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### How to do aggregation/ summarization
##### Summarization after grouping
```{r echo=TRUE,  message=FALSE}
library(tidyverse)
```

```{r echo=TRUE-2,  include=FALSE}
library(tidyverse)
```

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(Support = mean(Sepal.Length)) %>%    # average
  arrange(-Support)                         # sort
```

```{r}
iris %>% 
  group_by(Species) %>% 
summarize(mean_s = mean(Sepal.Width), 
            meas_p = mean(Petal.Length), 
            diff = mean(Sepal.Width-Petal.Length)) %>% 
  arrange(-diff) 
```

```{r}
iris %>% 
  group_by(Species) %>% 
summarize(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```
##### Summarization with upgroup
```{r}
iris %>% 
  ungroup( ) %>% 
summarize(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```
##### Mutate new variables after grouping
```{r}
iris %>% 
  group_by(Species) %>% 
mutate(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```


```{r}
iris %>% 
  group_by(Species) %>% 
mutate(n = n(), 
            meas_p = mean(Petal.Length, na.rm = T), 
            sd = sd(Petal.Length))  %>% 
  summarize (n_mean = paste ("sample size:",mean(n)), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))
```

##### Recode and generate new variables, then value label
```{r}
irisifelse <- iris %>% 
mutate(Species2 = ifelse(Species == "setosa", NA, Species))
# relabel values
irisifelse$Species2 <- factor(irisifelse$Species2,labels = c( "versi","virg"))
irisifelse
```
```{r}
str(irisifelse)
```

## Interpreting the Results

Use summaries and visual checks to confirm ranges, distributions, duplicates, and missing-data patterns. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
