---
title: "Bayesian Inference in R"
output: html_document
date: '2024-01-26'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Reference](https://www.boelstad.net/post/bayesian_statistics_introduction/) [etc](https://www.statlect.com/fundamentals-of-statistics/normal-distribution-Bayesian-estimation)

### Bayesian inference 

Let Y be a random variable representing the data. We have a (parametric) model for Y  with Y∼f(y∣θ),θ∈Θ, Θ  the parameter space. That is probability density function and can result in the likelihood function. Then we have a prior distribution represented by π(θ). Given an observation of Y, the posterior distribution of θ is

$$
f(\theta \mid y) =\frac{f(y\mid\theta) \pi(\theta)}{\int_\Theta f(y\mid\theta) \pi(\theta)\; d\theta}
$$

the posterior is proportional to the likelihood times the prior ($f(y)$is a constant).



### Continuous distributions, estimating one parameter

To find the max probability of mu given some observations. 


#### The prior

Assign mu a arbitrary prior probability (or estimated by maximum likelihood method using a sample). To keep things simple and intuitive, we will assess our distributions (i.e. prior, likelihood, posterior) at equally spaced points . for each specific value approximate the probability $\pi$ that μ actually lies within a narrow interval (0.01) around this point (treating them as categorical distributions without gap).

$$
\int{f(y\mid\theta) }= \int_{-\infty}^\infty \frac1{\sqrt{2\pi}} e^{-\frac12 (y-\theta)^2}\; d\theta = 1
$$

```{r}
res <- 1 / 100              # Set resolution for the sequence of potential mu-values, 0.01 interval
mu.h <- seq(-10, 10, res)   # Define values at which to assess the probability density

density <- dnorm(mu.h, mean = -1, sd = 1.5) # Get the density for each value in mu.h
prior <- density * res      # Multiply with resolution to get probability per interval

length(mu.h)
head(mu.h)

sum(prior)
length(prior)
```



#### Individual observation

```{r}
set.seed(4)                 # Set seed for reproducibility
n <- 100                      # Set number of observations
y <- rnorm(n, mean = 1, sd = 2) # Generate data
y                           # Take a look at the data
```



#### The likelihood

Suppose we know the standard deviation is 2. For a single observation, $y_i$, the likelihood is then: $p\left(y_i \mid \mu\right)=N\left(y_i \mid \mu, 2^2\right)$. The likelihood for the whole dataset is the product of all the individual likelihoods:

$$
p(y \mid \mu)=\prod_{i=1}^n N\left(y_i \mid \mu, 2^2\right)
$$

we can calculate the likelihood across the set of potential values for $\mu$. 



```{r}
likelihood <- vector()      # Create an empty vector

for (s in mu.h) {           # Loop over each value in mu.h , 2001 MU
  temp.lik <- 1             # Create a temporary likelihood scalar
  
  for (i in 1:n) {          # Loop over each individual data point, 30 observations
    temp.lik <- temp.lik * dnorm(y[i], mean = s, sd = 2) # Multiply individual likelihoods
  }
  likelihood <- c(likelihood, temp.lik) # Store the likelihood for each value in mu.h
}

# output likelihood value
head(likelihood)
length(likelihood)
```


```{r}
# normalized.likelihood <- likelihood / sum(likelihood)
```

#### The posterior

As we know from Bayes' theorem, we can obtain a distribution that is proportional (" $\propto$ ") to the posterior by multiplying the likelihood and the prior:

$$
p(\mu \mid y) \propto p(y \mid \mu) p(\mu)
$$



```{r}
unnormalized.posterior <- likelihood * prior
```

- Then divided by p(y)

```{r}
normalizing.constant <-  sum(likelihood * prior) #p(y)
normalizing.constant #integrate

# head(unnormalized.posterior) #no integrate

posterior <- unnormalized.posterior / sum(likelihood * prior)
```




#### Determine which one having max posterior of mu

```{r}
which.max(posterior)
mu.h[which.max(posterior)]
```


#### Plot

prior and posterior probability
```{r}
plot(mu.h, posterior / res, type = "l", ylab = "")
# lines(mu.h, normalized.likelihood / res, col = "red")
lines(mu.h, prior / res, col = "green")

#having a corresponding relationship for the two normal distributions 
```


### Continuous distributions, estimating two parameters

let us try to estimate both the mean, μ, and the standard deviation, σ. The standard deviation must be positive, let us give σ a flat prior over the range from 0 to 10, using a uniform distribution: p(σ)∼U(0,10).


#### The joint prior

$$
p(\mu, \sigma)=p(\mu) p(\sigma)
$$

mu prior

```{r}
res <- 1 / 100              # Set resolution for the sequence of potential mu-values, 0.01 interval
mu.h <- seq(-10, 10, res)   # Define values at which to assess the probability density

density <- dnorm(mu.h, mean = -1, sd = 1.5) # Get the density for each value in mu.h
prior <- density * res      # Multiply with resolution to get probability per interval

length(mu.h)
head(mu.h)

sum(prior)
length(prior)

mu.prior <- prior           
```

σ prior

```{r}
sigma.h <- seq(0, 10, res)    # Define values at which to assess the probabilities for sigma
sigma.density <- dunif(sigma.h, min(sigma.h), max(sigma.h)) # Density-values for a flat prior
sigma.prior <- sigma.density * res # Multiply with res to get probability per interval


length(sigma.h)
head(sigma.h)

sum(sigma.prior)
length(sigma.prior)
```

Now that we have specified priors for μ and σ, we can also specify their joint prior distribution, p(μ,σ).

```{r}
joint.prior <- sapply(mu.prior, function(mu.s) { # Using sapply's instead of loops
                sapply(sigma.prior, function(sigma.s) { mu.s * sigma.s })}) 
```

This gives us a two-dimensional grid of probabilities, representing all possible combinations of μ and σ values that we are considering.

```{r}
joint.prior[1:5,1:5]  
```

#### Individual observation

```{r}
set.seed(4)                 # Set seed for reproducibility
n <- 200                      # Set number of observations
y <- rnorm(n, mean = 1, sd = 2) # Generate data                           # Take a look at the data
```


#### The joint likelihood

If we calculate the likelihood across the possible combinations of values for μ and σ,

$$
p(y \mid \mu,\sigma)=\prod_{i=1}^n N\left(y_i \mid \mu, \sigma^2\right)
$$

```{r}
joint.likelihood <- sapply(mu.h, function(mu.s) { # Sapply's instead of loops
                      sapply(sigma.h, function(sigma.s) { prod(dnorm(y, mu.s, sigma.s)) })})
# normalized.joint.likelihood <- joint.likelihood / sum(joint.likelihood) # Normalize 
```

#### The joint posterior

The posterior distribution is now a joint distribution for our two parameters. 

$$
p(\mu, \sigma \mid y) \propto p(y \mid \mu, \sigma) p(\mu, \sigma)
$$

```{r}
unnormalized.joint.posterior <- joint.likelihood * joint.prior
normalizing.constant <- 1 / sum(joint.likelihood * joint.prior)

joint.posterior <- unnormalized.joint.posterior * normalizing.constant  #have the same max as unnormalized.joint.posterior
```


#### Determine which one having max posterior of mu, sigma

```{r}
# find the position in the matrix
df<-matrix(joint.posterior,nrow = 1001, byrow = F)
dim(df)

position=which(df==max(joint.posterior),arr.ind = T)
position

# then detect the ordinal mu and sigma by position
mu.h[position[2]]

sigma.h[position[1]]

```

Plot
```{r}
contour(mu.h,sigma.h,t(joint.posterior))
```

### Marginal distributions

we often want to draw conclusions about the values of a given parameter without reference to the other parameter(s) in the model. a marginal posterior distribution is to average the joint distribution over the parameter(s).

to get the marginal posterior for p(μ∣y), we integrate the joint posterior over σ: by summing the posterior probabilities in our grid for each potential value of σ.

$$
p(\mu \mid y)=\int p(\mu, \sigma \mid y) d(\sigma)
$$

#### marginal posterior for mu

```{r}
mu.marginal.posterior <- colSums(df) #integrate or sum

which.max(mu.marginal.posterior)

mu.h[which.max(mu.marginal.posterior)]
```

- Plot

```{r}
plot(mu.h,mu.marginal.posterior/res,type = "l"); lines(mu.h,mu.prior/res,col=2)
```

#### marginal posterior for sigma


```{r}
sigma.marginal.posterior <- rowSums(df) #integrate or sum

which.max(sigma.marginal.posterior )

sigma.h[which.max(sigma.marginal.posterior )]
```

- Plot

```{r}
plot(sigma.h,sigma.marginal.posterior/res,type = "l"); lines(sigma.h,sigma.prior/res,col=2)
```

 the number of elements in the grid increases exponentially with the number of parameters. For such problems, we would normally sample from the posterior distribution using a form of Markov Chain Monte Carlo (MCMC) or solved analytically.
 
 
### Beyes' theory 

The prevalence is 2.28%. Assume the test (94% and 98%, correspondingly for sensitivity and specificity). What is the probability with positive test result has the disease?

```{r}
# d is disease and t is test

#Prior p(theta1)
p.d1 <- .0228
p.d0 <- 1-p.d1

#Sensitivity: p(tau1|theta1)
p.t1.d1 <- .94

#Specificity: p(tau0|theta0)
p.t0.d0 <- .98
p.t1.d0 <- 1-p.t0.d0

#Bayes prob
(p.d1.t1 <- p.t1.d1* p.d1 /( p.t1.d1 *p.d1+ p.t1.d0 *p.d0))

# PPV = (sensitivity x prevalence) / [ (sensitivity x prevalence) + ((1 – specificity) x (1 – prevalence)) ]
(p.t1.d1 * p.d1) / ( (p.t1.d1 * p.d1) + (1 - p.t0.d0) * (1 - p.d1) )

```

### Binomilal distributions 

 
$$
P(x ; p, n)=\left(\begin{array}{c}
n \\
x
\end{array}\right)(p)^x(1-p)^{(n-x)} \quad \text { for } x=0,1,2, \cdots, n
$$

#### The prior

use the normal distribution as prior distribution 

$$
\int{f(y\mid\theta) }= \int_{-\infty}^\infty \frac1{\sqrt{2\pi}} e^{-\frac12 (y-\theta)^2}\; d\theta = 1
$$

```{r}
res <- 1 / 100              # Set resolution for the sequence of potential mu-values, 0.01 interval
mu.h <- seq(0, 1, res)   # Define values at which to assess the probability density

density <- dnorm(mu.h, mean = -1, sd = 1.5) # Get the density for each value in mu.h
prior <- density * res      # Multiply with resolution to get probability per interval

length(mu.h)
head(mu.h)

sum(prior)
length(prior)
        
```



```{r}
# sigma.h <- seq(0, 10, res)    # Define values at which to assess the probabilities for sigma
# sigma.density <- dunif(sigma.h, min(sigma.h), max(sigma.h)) # Density-values for a flat prior
# sigma.prior <- sigma.density * res # Multiply with res to get probability per interval
# 
# 
# length(sigma.h)
# head(sigma.h)
# 
# sum(sigma.prior)
# length(sigma.prior)
```


#### Individual observation

```{r}
set.seed(4)                 # Set seed for reproducibility
n <- 500                      # Set number of observations
y <- rbinom(n, size=1, p = 0.26) # Generate data                           # Take a look at the data
y
```


#### The likelihood

use binomial distribution likelihood function. 

$$
p(y \mid p)=\prod_{i=1}^n B\left(y_i \mid p\right)
$$

Where n=1. 

```{r}
joint.likelihood <- sapply(mu.h, function(s) prod(dbinom(y, 1, s)) )
# normalized.joint.likelihood <- joint.likelihood / sum(joint.likelihood) # Normalize 

```

#### The posterior

$$
p(p \mid y) \propto p(y \mid p) p(p)
$$

```{r}
unnormalized.joint.posterior <- joint.likelihood * prior
normalizing.constant <- 1 / sum(joint.likelihood * prior)

joint.posterior <- unnormalized.joint.posterior * normalizing.constant  #have the same max as unnormalized.joint.posterior
```


#### Determine which one having max posterior of p

```{r}
# find the position in the matrix
mu.h[which.max(joint.posterior)]

```

Plot
```{r}
plot(mu.h, joint.posterior / res, type = "l", ylab = "")
# lines(mu.h, normalized.likelihood / res, col = "red")
lines(mu.h, prior / res, col = "green")
 
```












