---
title: "Linear mixed model covariance decomposition (nlme)"
description: "A practical R learning guide to Linear mixed model covariance decomposition (nlme), with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Longitudinal & Repeated Measures"
topic: "Linear mixed model covariance decomposition (nlme)"
type: article
tags: ["Longitudinal & Repeated Measures", "Biostatistics", "Linear", "mixed", "covariance"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Repeated observations from the same unit are correlated, so longitudinal models must represent both mean change and within-unit dependence.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Linear mixed model covariance decomposition (nlme)**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Linear mixed model covariance decomposition (nlme)
2. Load data
3. Using lme (nlme)
4. Model diagnosing
5. Get x, y, z matrixs (using lme4)
6. Fixed effect coefficient

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Linear mixed model covariance decomposition (nlme)** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Linear mixed model covariance decomposition (nlme)

<!-- generated-learning-note -->
This section develops **Linear mixed model covariance decomposition (nlme)**. Read the model specification together with its assumptions and the scale of each reported effect.


#### Load data

<!-- generated-learning-note -->
This section prepares the information needed for **Load data**. Check variable definitions, units, coding, and missing values before moving to analysis.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block imports or selects the data used in the example. Verify the file location, column types, coding, and missing values when substituting your own data.

```{r}
data(Orthodont, package="nlme")
Data <- Orthodont
```

<!-- generated-learning-note -->
**How to read the output.** Inspect dimensions, names, types, and several rows of the imported data. Unexpected structure at this stage will propagate into every later result.




<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(ggplot2)
ggplot(Data, aes(y = distance, x = age,  color = Sex)) +
    geom_smooth(method = "lm",se=F) + geom_point() + theme_classic()
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(ggplot2)
ggplot(Data, aes(y = distance, x = age, fill = Subject, color = Sex)) +
    geom_smooth(method = "lm",se=F) + geom_point() + theme_classic()
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Using lme (nlme)

<!-- generated-learning-note -->
This subsection explains **Using lme (nlme)** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(nlme)
fit.lmer2 <- lme(distance ~ age + Sex  , random = ~1   | Subject, data = Data) 
summary(fit.lmer2)
library(lme4)
# lme4
fit.lmer <- lmer(distance ~ age + Sex  +(1 | Subject), data = Data) 
# compound symmetry
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


- t test

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
tstat <- fixef(fit.lmer2)/sqrt(diag(vcov(fit.lmer2)))
pval <- 2*pt(-abs(tstat), df = fit.lmer2$fixDF$X)
Tresult <- data.frame(t = tstat, p = round(pval, 4))
print(Tresult)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


- F test

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
anova(fit.lmer2 ) 
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


- restricted model test

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
anova(fit.lmer2, L = c(0, 1, 0)) 
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


- likelihood test

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
fit.lmer2.2 <- lme(distance ~   Sex  , random = ~1   | Subject, data = Data) 
anova(fit.lmer2, fit.lmer2.2)

```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.

`And now by hand` 

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
logLik  <- logLik(fit.lmer2)
logLik0 <- logLik(fit.lmer2.2)

LR <- 2 * (logLik - logLik0)
pval <- pchisq(LR, df = 2, lower.tail = FALSE)
LRresult <- data.frame(LR = LR, p = round(pval, 4), row.names = "age")
print(LRresult)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

<!-- https://www2.karlin.mff.cuni.cz/~komarek/vyuka/2020_21/nmst432/lmm_lme.html -->

#### Model diagnosing

<!-- generated-learning-note -->
This section develops **Model diagnosing**. Read the model specification together with its assumptions and the scale of each reported effect.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(nlme)
plot.lme(fit.lmer2)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Get x, y, z matrixs (using lme4)

<!-- generated-learning-note -->
This subsection explains **Get x, y, z matrixs (using lme4)** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
X=(getME(fit.lmer, "X"))
y=getME(fit.lmer, "y")
Z <- getME(fit.lmer, "Z")

Z2 <- model.matrix(~Subject-1,data=Data)

```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`check z matrix`

<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
dummyz<- as.matrix (Z)==Z2
table(dummyz)
dim(Z)
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.

`check y and X matrixs`

<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
dummyx<- Data[,c(2)]==X[,2]
table(dummyx)
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.


#### Fixed effect coefficient

<!-- generated-learning-note -->
This subsection explains **Fixed effect coefficient** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
bhat <- fit.lmer2$coef$fixed
bhat
# fixef()
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

- fixed effect coefficient confidence intervals 

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
intervals(fit.lmer2)
 
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Plot fixed effect coefficients and theri ci

<!-- generated-learning-note -->
This section uses **Plot fixed effect coefficients and theri ci** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
tab <- cbind(Est = intervals(fit.lmer2)[["fixed"]][,2], LL = intervals(fit.lmer2)[["fixed"]][,1], UL=
intervals(fit.lmer2)[["fixed"]][,3])
#Extracting the fixed effect estimates and manually calculating the 95% confidence limits
#qnorm extracts the standard normal critical value for 1-alpha/2= 1-0.05/2=0.975
Odds= (tab)
round(Odds,4)

Odds=as.data.frame(Odds)
Odds$Label=rownames(Odds)
 
ggplot(Odds[-1,],aes(x=Est,y=Label))+geom_point()+
geom_errorbarh(aes(xmax = UL, xmin = LL))+
  
theme_bw()+geom_vline(xintercept=0,col="darkgray",size=1.2,linetype=2)+
theme(axis.title.y = element_blank(),axis.text = element_text(size=10),
axis.title.x = element_text(size=12))+labs(x="Beta coefficients")
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



`And now by hand` 

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
print(tquants <- qt(0.975, df = fit.lmer2$fixDF$X))

Low <- fixef(fit.lmer2) - tquants * sqrt(diag(vcov(fit.lmer2)))
Upp <- fixef(fit.lmer2) + tquants * sqrt(diag(vcov(fit.lmer2)))
EstInt <- data.frame(Lower = Low, Estimate = fixef(fit.lmer2), Upper = Upp)
print(EstInt)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



#### Random effect coefficient

<!-- generated-learning-note -->
This subsection explains **Random effect coefficient** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
fit.lmer2$coef$random
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Fixed effect coefficients covariance

<!-- generated-learning-note -->
This subsection explains **Fixed effect coefficients covariance** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
fit.lmer2$varFix
# vcov(fit.lmer2)
sqrt(diag(vcov(fit.lmer2)))
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`fixed effect coefficients correlation `

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
-0.04174818/prod(sqrt( diag(fit.lmer2$varFix)[-3] ))
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



#### Random effect covariance and correlation

<!-- generated-learning-note -->
This subsection explains **Random effect covariance and correlation** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
pdMatrix(fit.lmer2$modelStruct$reStruct[[1]])*fit.lmer2$sigma**2

getVarCov(fit.lmer2)  #using default 
getVarCov(fit.lmer2, individual = "F01", type = "marginal")
getVarCov(fit.lmer2, type = "conditional")
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
vc <- as.numeric(as.matrix(VarCorr(fit.lmer2))[,1])
vc/sum(vc)
VarCorr(fit.lmer2)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Get y covariance directly

<!-- generated-learning-note -->
This subsection explains **Get y covariance directly** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.

`= Z REcov Zt + sigma**2`

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
require(mgcv)
library(nlme)
 
cov <- extract.lme.cov(fit.lmer2,Data)
head(cov)
dim(cov)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Residual variance

<!-- generated-learning-note -->
This subsection explains **Residual variance** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
fit.lmer2$sigma**2
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Compute fixed effect coefficients

<!-- generated-learning-note -->
This subsection explains **Compute fixed effect coefficients** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(matlib)
inv(t(as.matrix(X))%*%inv(as.matrix(cov))%*%as.matrix(X))%*%t(as.matrix(X))%*%inv(as.matrix(cov))%*%y
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

#### Compute covariance of fixed efffect coefficients

<!-- generated-learning-note -->
This subsection explains **Compute covariance of fixed efffect coefficients** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
inv(t(as.matrix(X))%*%inv(as.matrix(cov))%*%as.matrix(X))
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Compute random effect coefficients

<!-- generated-learning-note -->
This subsection explains **Compute random effect coefficients** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
Lambda_new <-as.numeric(VarCorr(fit.lmer2)[1])*diag(length(levels(Data$Subject)))
# head(Lambda_new)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.




<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
uhat <- fit.lmer2$coef$random 
comput_uhat <- (as.matrix(Lambda_new))%*%t(Z)%*%inv(as.matrix(cov))%*%(y-as.matrix(X)%*%(bhat))
cbind((comput_uhat@x),(uhat[["Subject"]]))
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Compute predicted values

<!-- generated-learning-note -->
This subsection explains **Compute predicted values** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.



`usually computing mean and  prediction ci by using bootstrapping and simulated methods` 

approximate at below:  


*This [webpage](<!-- https://stackoverflow.com/questions/14358811/extract-prediction-band-from-lme-fit -->) calculates the sd like above, but I think the following is also correct. *

#### the conditional distribution

<!-- generated-learning-note -->
This subsection explains **the conditional distribution** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


$$
\mathbf{y}_{} \mid \mathbf{b}_{} \sim \mathbf{N}\left(\mathbf{X}_{} \boldsymbol{\beta}+\mathbf{Z}_{} \mathbf{b}_{}, \boldsymbol{\Sigma}_{}\right)
$$
therefore

- `conditional mean prediction`
$$
\begin{align}
E(\hat{Y}_0)&=E\mathbf{(X_0\hat{\beta} + Z_0\hat{u})}=\mathbf{X_0\beta + Z_0{u}}=E\mathbf{(Y_0)}\\
var(\hat{Y}_0)&=E\mathbf{((X_0\hat{\beta}-X_0\beta)+ (Z_0\hat{u} - Z_0{u}))}^2\\&=E\mathbf{((X_0\hat{\beta}-X_0\beta)^2+ (Z_0\hat{u} - Z_0{u})^2+ 2 ((X_0\hat{\beta}-X_0\beta)* (Z_0\hat{u} - Z_0{u}))  )}
\\
              &=E\mathbf{\left(
X_0(\hat{\beta}-\beta)(\hat{\beta}-\beta)'X_0' \right)}+
E\mathbf{\left(
Z_0(\hat{u}-u)(\hat{u}-u)'Z_0' \right)}+2E ((X_0(\hat{\beta}-\beta))* (Z_0(\hat{u} - {u}))) ?
\\
              &=\sigma^2\mathbf{X_0\left( X'X
\right)^{-1}X_0'}+ \sigma^2\mathbf{Z_0\left( Z'Z
\right)^{-1}Z_0'} + 2 *0*0 \quad \text {if independent}\\
&= \mathbf{X_0 cov (X) X_0'}+  \mathbf{Z_0 cov (Z) Z_0'}  \\
\end{align}
$$

    `generally`
$$
\begin{align}
Var(X+Y) &= Cov(X+Y,X+Y)  \\
 &= E((X+Y)^2)-E(X+Y)E(X+Y)   \\
&\text{by expanding,}  \\
 &= E(X^2) - (E(X))^2 + E(Y^2) - (E(Y))^2 + 2(E(XY) - E(X)E(Y))  \\
 &= Var(X) + Var(Y) + 2(E(XY)) - E(X)E(Y))  \\
\end{align}
$$


- `conditional individual prediction`
$$
\begin{align}
E(\hat{Y}_0)&= E\mathbf{(Y_0)}\\
var(\hat{Y}_0) 
&= \mathbf{X_0 cov (X) X_0'}+  \mathbf{Z_0 cov (Z) Z_0'} + \sigma^2 \\
\end{align}
$$

$$
\begin{align}
\hat{Y}_0& \sim N(\mu_{\hat{Y}_0},\sigma^2_{\hat{Y}_0})\\
\end{align}
$$

- `lower level (conditional prediction) in R` 

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# how to calculate predicted values
yhat <- X%*%(fit.lmer2$coef$fixed)+Z%*% as.numeric ( uhat[["Subject"]])

head(cbind (yhat,predict(fit.lmer2),y))  #create individual trajectory curve   

```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
#compute standard error for marginal predictions
predvar_rand <- diag(X %*% fit.lmer2$varFix %*% t(X)) + diag(Z %*%  diag(getVarCov(fit.lmer2)[1]  ,27) %*% t(Z))

SE_rand <- sqrt (predvar_rand)  #mean prediction 
SE_rand2 <- sqrt(predvar_rand+fit.lmer2$sigma^2)  #individual prediction 
head(SE_rand,20)
head(SE_rand2,20)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`see below `


#### The marginal distribution

<!-- generated-learning-note -->
This subsection explains **The marginal distribution** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.

$$
\mathbf{y}_{} \sim \mathrm{N}\left(\mathbf{X}_{} \boldsymbol{\beta}, \mathbf{V}_{}\right)
$$

$$
\operatorname{Cov}[\mathbf{y}]=\mathbf{V}=\mathbf{Z D Z}^{\prime}+\Sigma
$$

*$\mathbf{Z}_{} \mathbf{D Z}_{}^{\prime}$ represents the random-effects structure*

therefore

- `marginal mean prediction`

$$
\begin{align}
E(\hat{Y}_0)&=E\mathbf{(X_0\hat{\beta})}=\mathbf{X_0\beta}=E\mathbf{(Y_0)}\\
var(\hat{Y}_0)&=E\mathbf{(X_0\hat{\beta}-X_0\beta)}^2\\
              &=E\mathbf{\left(
X_0(\hat{\beta}-\beta)(\hat{\beta}-\beta)'X_0' \right)}\\
              &=E\mathbf{X_0\left(
(\hat{\beta}-\beta)(\hat{\beta}-\beta)' \right)X_0'}\\
              &=\mathbf{X_0Cov(X)X_0'}\\
\end{align}
$$

- `marginal individual prediction`
$$
\begin{align}
E(\hat{Y}_0)&= E\mathbf{(Y_0)}\\
var(\hat{Y}_0) 
&= \mathbf{X_0 cov (X) X_0'}+    \sigma^2 \\
\end{align}
$$

- `higher level (marginal)` in R
 

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
#compute standard error for marginal predictions
predvar <- diag(X %*% fit.lmer2$varFix %*% t(X))

SE <- sqrt (predvar)  #mean prediction 
SE2 <- sqrt(predvar+fit.lmer2$sigma^2)  #individual prediction
head(SE,10)
head(SE2,10)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
up=predict(fit.lmer2, newdata=Data, level=0) +1.96 *SE  #mean prediction 
up2=predict(fit.lmer2, newdata=Data, level=0) +1.96 *SE2  #individual prediction 
head(up)
head(up2)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r,message=F}
library(tidyverse)
library(ggeffects)
ggpredict(fit.lmer2,terms=c("age" ))
ggpredict(fit.lmer2,terms=c("age" ))  %>% plot(rawdata = T, dot.alpha = 0.2)
# ggpredict(fit.lmer2,"age",   type = "re" )  %>% plot(rawdata = T, dot.alpha = 0.2)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


**Compute SE by hand and compare with above SE: **

0.520408163
0.489795918
0.489795918
0.520408163


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
head(SE,10)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.




<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# head(predict(fit.lmer2))
# head (predict(fit.lmer2, newdata=Data, level=1)) #predicted value
# head (predict(fit.lmer2, newdata=Data, level=0)) #mean and ci   
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.

<!-- `fixed effect variance` -->
<!-- ```{r} -->
<!-- diag (X%*%(vcov(fit.lmer2))%*%t(X)) -->
<!-- ``` -->

<!-- `marginal y mean and variance directly` -->
<!-- # ```{r} -->
<!-- # ymean <- X%*%(fit.lmer2$coef$fixed)  # multiple comparisons (combine clusters) -->
<!-- # ysd  <- sqrt (diag (cov)) -->
<!-- # head(cbind (ymean,ysd )) -->
<!-- # ``` -->


#### Using lme with Gaussian

<!-- generated-learning-note -->
This subsection explains **Using lme with Gaussian** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.

`it means the correlation in a group but not correlation in variables (different covariances)`

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(nlme)

fit.lmer.gaus <- lme(distance ~ age + Sex  , random = ~1 | Subject, correlation=corGaus(form= ~ age|Subject, nugget=TRUE), data = Data) 
summary(fit.lmer.gaus)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

- Plotting of the empirical covariation (of residuals) versus "age"

`when describing the spatial pattern of a measured variable, a common way of visualizing the spatial autocorrelation of a variable is a variogram plot.`

`checking the semi-variance change by the distance of spatial autocorrelation. `


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r,warning=FALSE}
print(plot(Variogram(fit.lmer.gaus, form =~as.numeric(age)|Subject,
data = Data)))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.




<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
require(mgcv)
library(nlme)
 
cov.gaus <- extract.lme.cov(fit.lmer.gaus,Data)
head(cov.gaus)
dim(cov.gaus)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

- Compute fixed effect coefficients using Gaussian

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(matlib)
inv(t(as.matrix(X))%*%inv(as.matrix(cov.gaus))%*%as.matrix(X))%*%t(as.matrix(X))%*%inv(as.matrix(cov.gaus))%*%y
fit.lmer.gaus$coef$fixed
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


*the same below*

#### Using lme with autoregressive

<!-- generated-learning-note -->
This subsection explains **Using lme with autoregressive** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
fit.lmer.AR1 <- lme(distance ~ age + Sex  , random = ~1 | Subject, correlation=corAR1(form= ~ as.numeric(age)|Subject ), data = Data) 
summary(fit.lmer.AR1)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.

- Plotting of the empirical covariation (of residuals) versus "age" 

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r,warning=FALSE}
print(plot(Variogram(fit.lmer.AR1, form =~as.numeric(age)|Subject,
data = Data)))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
require(mgcv)
library(nlme)
 
cov.AR1 <- extract.lme.cov(fit.lmer.AR1,Data)
head(cov.AR1)
dim(cov.AR1)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Using lme with exponential

<!-- generated-learning-note -->
This subsection explains **Using lme with exponential** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
fit.lmer.Exp <- lme(distance ~ age + Sex  , random = ~1 | Subject, correlation=corExp(form= ~ as.numeric(age)|Subject ), data = Data) 
summary(fit.lmer.Exp)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.

- Plotting of the empirical covariation (of residuals) versus "age"

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r,warning=FALSE}
print(plot(Variogram(fit.lmer.Exp, form =~as.numeric(age)|Subject,
data = Data)))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
require(mgcv)
library(nlme)
 
cov.Exp <- extract.lme.cov(fit.lmer.Exp,Data)
head(cov.Exp)
dim(cov.Exp)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.




#### Using lme with unstructured

<!-- generated-learning-note -->
This subsection explains **Using lme with unstructured** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
fit.lmer.corSymm <- lme(distance ~ age + Sex  , random = ~1 |  (Subject), correlation=corSymm(form= ~ 1| (Subject) ),
                     weights = varIdent(form=~1|age),
                     data = Data)
summary(fit.lmer.corSymm)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.

- Plotting of the empirical covariation (of residuals) versus "age"

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r,warning=FALSE}
print(plot(Variogram(fit.lmer.corSymm, form =~as.numeric(age)|Subject,
data = Data)))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
require(mgcv)
library(nlme)
 
cov.corSymm <- extract.lme.cov(fit.lmer.corSymm,Data)
head(cov.corSymm)
dim(cov.corSymm)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


<!-- - Using gls with unstructured  -->
<!-- ```{r} -->
<!-- gls.corSymm <- gls(distance ~ age + Sex  ,    -->
<!--                      correlation=corSymm(form= ~ 1| (Subject) ), -->
<!--                      weights = varIdent(form=~1|age), -->
<!--                      data = Data) -->
<!-- summary(gls.corSymm) -->
<!-- ``` -->

#### Using lme with compound symm

<!-- generated-learning-note -->
This subsection explains **Using lme with compound symm** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
fit.lmer.Symm <- lme(distance ~ age + Sex  , random = ~1 |  (Subject), correlation=corCompSymm(form= ~ as.numeric(age)| (Subject) ), data = Data) 
summary(fit.lmer.Symm)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.

- Plotting of the empirical covariation (of residuals) versus "age"

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r,warning=FALSE}
print(plot(Variogram(fit.lmer.Symm, form =~as.numeric(age)|Subject,
data = Data)))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
require(mgcv)
library(nlme)
 
cov.Symm <- extract.lme.cov(fit.lmer.Symm,Data)
head(cov.Symm)
dim(cov.Symm)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Using gls with unstructured, corSymm

<!-- generated-learning-note -->
This subsection explains **Using gls with unstructured, corSymm** and shows how it contributes to the overall Linear mixed model covariance decomposition (nlme) workflow.


<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
gls.corsymm <- gls(distance ~ Sex +age, Orthodont,
                   correlation = corSymm(form = ~ 1 | Subject),
                   weights = varIdent(form=~1|age))
summary(gls.corsymm)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.




#### Using gls with compound symm, which = lmer model default = lme default = lme compound symm =lme auto reg in this example

<!-- generated-learning-note -->
This section develops **Using gls with compound symm, which = lmer model default = lme default = lme compound symm =lme auto reg in this example**. Read the model specification together with its assumptions and the scale of each reported effect.


<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
gls.comsys <- gls(distance ~ Sex +age, Orthodont,
                   correlation = corCompSymm(form = ~ 1 | Subject))
summary(gls.comsys)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


**lme4 does not allow to specify it, and it can only fit LMMs with independent residual errors. However, the range of available variance-covariance matrices for the random effects are restricted to diagonal or general matrices. **

## Interpreting the Results

Check fixed effects, random effects, covariance assumptions, and subject-specific versus population-level interpretation. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
