---
title: "Joint model with longitudinal and survival data"
description: "A practical R learning guide to Joint model with longitudinal and survival data, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Longitudinal & Repeated Measures"
topic: "Joint model with longitudinal and survival data"
type: article
tags: ["Longitudinal & Repeated Measures", "Biostatistics", "Joint", "longitudinal", "survival"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  enabled: false
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Repeated observations from the same unit are correlated, so longitudinal models must represent both mean change and within-unit dependence.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Joint model with longitudinal and survival data**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Joint model with longitudinal and survival data
2. Lattice plot for data
3. Cox model with longitudinal data
4. Perfromance of model
5. Predict and plot
6. Competing Risks joint model

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Longitudinal & Repeated Measures",
  topic = "Joint model with longitudinal and survival data"
)
learning_overview
```

## Learning Material

The following material develops **Joint model with longitudinal and survival data** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Joint model with longitudinal and survival data

<!-- generated-learning-note -->
This section prepares the information needed for **Joint model with longitudinal and survival data**. Check variable definitions, units, coding, and missing values before moving to analysis.


####### Lattice plot for data

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(JM)
lattice::xyplot(log(serBilir) ~ year | id, group = id, 
       data = pbc2[pbc2$id %in% c(1:10),], 
       xlab = "years", ylab = expression(log("serBilir")), col = 1, type = "b")
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.



####### Cox model with longitudinal data

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(tidyverse)
# data transform
pbc2_new=pbc2
pbc2_new$start=pbc2_new$year
pbc2_new=pbc2_new %>%
  arrange(id, year) %>%   
  group_by(id) %>%
  mutate(stop = lead(year, n = 1))  
pbc2_new=pbc2_new %>% mutate(event=ifelse(is.na(stop), status2,0 ), event_cr=ifelse(is.na(stop), status,1 ),stop = ifelse(is.na(stop), years, stop))
pbc2_new$event_cr=pbc2_new$event_cr-1
# cox
coxph=coxph(Surv(start,stop,event) ~ drug+ log(serBilir) +sex, data = pbc2_new, x = TRUE,model = TRUE) # event is death
coxph
table(pbc2_new$event_cr)
# FGR (Hist(start,stop, event_cr)~drug+ log(serBilir) +sex, data = pbc2_new)   #competing risk model doesn't work in this form
# Error in Hist(start, stop, event_cr) : 
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


Proportional-hazards (PH) assumption


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
test.ph <- cox.zph(coxph)
test.ph
# log(serBilir) not meet the ph assumption
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Updated cox model time dependent 

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}

fit.ph<-coxph(Surv(start,stop,event) ~ drug+ log(serBilir) +sex +tt(log(serBilir)),
              tt=function(x,t,...) x*log(t+20 ),data=pbc2_new)
summary(fit.ph)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.





<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# for sex
plot(survfit(Surv(start, stop, event) ~ sex, data = pbc2_new),
     xlab = "Time",
     ylab = "Survival probability")
# for drug
plot(survfit(Surv(start, stop, event) ~ drug, data = pbc2_new),
     xlab = "Time",
     ylab = "Survival probability")
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.




####### Joint model with longitudinal and survival data

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# two datasets
# View(pbc2)  #long format , outcome is log serum bilirubin 
# View(pbc2.id) #baseline data, time to event data , event is death
lmeFit <-   
  # lme(log(serBilir) ~ drug * ns(year, 3), 
  #                           random = list(id = pdDiag(form = ~ ns(year, 3))), data = pbc2)
  lme(log(serBilir) ~ drug * year, random = ~ year | id, data = pbc2)
coxFit <- coxph(Surv(years, status2) ~ drug+sex, data = pbc2.id, x = TRUE)
# to fit this model we need to specify the 'derivForm' argument, which is a list
# with first component the derivative of the fixed-effects formula of 'lmeFit' with
# respect to 'obstime', second component the indicator of which fixed-effects 
# coefficients correspond to the previous defined formula, third component the 
# derivative of the random-effects formula of 'lmeFit' with respect to 'obstime', 
# and fourth component the indicator of which random-effects correspond to the 
# previous defined formula
dForm <- list(fixed = ~ 1 + drug, indFixed = c(3, 4), random = ~ 1, indRandom = 2)
jointModel=jointModel(lmeFit, coxFit, timeVar = "year", method = "spline-PH-aGH",
           parameterization = "both", derivForm = dForm)
# jointModel=jointModel(lmeFit, coxFit, timeVar = "year", method = "spline-PH-aGH")
jointModel
exp(confint(jointModel, parm = "Event"))


```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


####### Perfromance of model

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ND <- pbc2[pbc2$id == 2, ] # the data of Subject 2
head(ND,20)
plot(rocJM(jointModel, dt = c(1, 9,  10), ND, idVar = "id"))
# dt: a numeric vector indicating the lengths of the time intervals of primary interest
# within which we want to distinguish between subjects who died within the
# intervals from subjects who survived longer than that.
aucJM(jointModel, pbc2, Tstart = 1, Thoriz = 8)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



####### Predict and plot

Prediction

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
set.seed(300716) # it uses Monte Carlo samples
preds <- survfitJM(jointModel, newdata = pbc2[pbc2$id %in% c("2"), ],
                   idVar = "id")  # last.time = "Time"
preds
# 
survfitJM(jointModel, newdata = pbc2[pbc2$id %in% c("2"), ],
          idVar = "id",
          survTimes = c(8, 10, 12))  # you can specify the times

```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Plot

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
par(mfrow=c(1,2))
plot(preds, which = "2", conf.int = TRUE)

plot(preds, which = "2", conf.int = TRUE, 
     fun = function (x) -log(x), ylab = "Cumulative Risk")

# overall true plot
sfit2 <- survfitJM(jointModel, newdata = pbc2[pbc2$id == "2", ], idVar = "id")
plot(sfit2, include.y = TRUE, conf.int = TRUE)

sfit7 <- survfitJM(jointModel, newdata = pbc2[pbc2$id == "7", ], idVar = "id")
plot(sfit7, include.y = TRUE, conf.int = TRUE)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


Predicted plot using the first n observations 

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
survPreds <- vector("list", nrow(ND))
for (i in 1:nrow(ND)) {
  survPreds[[i]] <- survfitJM(jointModel, newdata = ND[1:3, ])
}
ND[1:3, ]
survPreds[[3]]

par(mfrow = c(2, 2), oma = c(0, 2, 0, 2))
for (i in c(1,3,5,7)) {
  plot(survPreds[[i]], estimator = "median", conf.int = TRUE,
       include.y = TRUE, main = paste("Follow-up time:",
                                      round(survPreds[[i]]$last.time, 1)), ylab = "", ylab2 = "")
}
mtext("log serum bilirubin", side = 2, line = -1, outer = TRUE)
mtext("Survival Probability", side = 4, line = -1, outer = TRUE)
par(mfrow=c(1,1))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.




####### Competing Risks joint model


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# data transform
# we first expand the PBC dataset in the competing risks long format
# with two competing risks being death and transplantation
pbc2.idCR <- crLong(pbc2.id, "status", "alive")  #alive is censoring, others are competing outcomes 
 # 1 dead
 # 0 crisk
# 0 dead
# 0 crisk
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# we fit the linear mixed model as before
lmeFit.pbc <- lme(log(serBilir) ~ drug * ns(year, 3), 
                  random = list(id = pdDiag(form = ~ ns(year, 3))), data = pbc2)
 # lme(log(serBilir) ~ drug * year, random = ~ year | id, data = pbc2)
 
# however, for the survival model we need to use the data in the long
# format, and include the competing risks indicator as a stratification
# factor. We also take interactions of the baseline covariates with the
# stratification factor in order to allow the effect of these covariates
# to be different for each competing risk
coxCRFit.pbc <- coxph(Surv(years, status2) ~ (drug+sex  )*strata + strata(strata), 
                      data = pbc2.idCR, x = TRUE)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.



<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r}
#| eval: false
#| code-fold: true
# the corresponding joint model is fitted simply by including the above
# two submodels as main arguments, setting argument CompRisk to TRUE, 
# and choosing as method = "spline-PH-aGH". Similarly as above, we also 
# include strata as an interaction factor to allow serum bilirubin to 
# have a different effect for each of the two competing risks
jmCRFit.pbc <- jointModel(lmeFit.pbc, coxCRFit.pbc, timeVar = "year", 
                          method = "spline-PH-aGH", 
                          interFact = list(value = ~ strata, data = pbc2.idCR), 
                          CompRisk = TRUE)
summary(jmCRFit.pbc)

exp(confint(jmCRFit.pbc, parm = "Event"))

# https://stats.stackexchange.com/questions/427547/jointmodelbayes-output
# survfitJM() is not currently implemented for competing risks joint models.
```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.


[Reference1](https://www.rdocumentation.org/packages/JM/versions/1.5-2/topics/jointModel) 
[Reference2](https://bookdown.org/sestelo/sa_financial/the-jm-package.html) 
[Reference3](http://jmr.r-forge.r-project.org/)

## Interpreting the Results

Check fixed effects, random effects, covariance assumptions, and subject-specific versus population-level interpretation. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
