---
title: "Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin"
description: "A practical R learning guide to Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Epidemiology & Real-World Evidence"
category: "Real-World Data & Evidence"
topic: "Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin"
type: article
tags: ["Real-World Data & Evidence", "Epidemiology & Real-World Evidence", "Real-World", "Evidence", "Study"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Real-world evidence depends on whether routinely collected data can support a clearly specified causal or descriptive objective.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. 1. Background and Research Question
2. 2. Create / Import Real-World Data
3. 3. Define the Study Cohort
4. 4. Define Exposure, Outcome, and Baseline Covariates
5. 5. Data Quality Checks
6. 6. Examine Baseline Imbalance Before Adjustment

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

#### 1. Background and Research Question

<!-- generated-learning-note -->
This subsection explains **1. Background and Research Question** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


This educational example mirrors the design of a published real-world retrospective cohort study among patients with type 2 diabetes who were initially treated with metformin and subsequently added either insulin or a sulfonylurea.

**Research question:**  
Among eligible patients with type 2 diabetes receiving metformin, is adding insulin associated with a different risk of myocardial infarction, stroke, or all-cause death compared with adding a sulfonylurea?

Because treatment is not randomized, baseline confounding must be addressed.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(MatchIt)
library(cobalt)
library(survival)
library(tableone)

set.seed(2026)
```

#### 2. Create / Import Real-World Data

<!-- generated-learning-note -->
This section prepares the information needed for **2. Create / Import Real-World Data**. Check variable definitions, units, coding, and missing values before moving to analysis.


In a real study, this step would import EHR, claims, registry, pharmacy, laboratory, or linked administrative data.

Here we simulate a dataset with the same basic structure.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r simulate-data}
n <- 5000

rwd <- data.frame(
  id = 1:n,
  age = round(rnorm(n, 62, 9)),
  male = rbinom(n, 1, 0.94),
  bmi = round(rnorm(n, 31, 5), 1),
  hba1c = round(rnorm(n, 8.1, 1.3), 1),
  creatinine = round(pmax(rnorm(n, 1.1, 0.35), 0.4), 2),
  systolic_bp = round(rnorm(n, 132, 15)),
  smoking = rbinom(n, 1, 0.25),
  prior_cvd = rbinom(n, 1, 0.25),
  hypertension = rbinom(n, 1, 0.65),
  ckd = rbinom(n, 1, 0.18)
)

# Treatment selection depends on baseline patient characteristics.
# This intentionally creates confounding by indication.
lp_treat <-
  -3.0 +
  0.02 * (rwd$age - 60) +
  0.55 * (rwd$hba1c - 8) +
  0.35 * rwd$prior_cvd +
  0.45 * rwd$ckd +
  0.02 * (rwd$bmi - 30)

p_insulin <- plogis(lp_treat)

rwd$treatment <- ifelse(
  rbinom(n, 1, p_insulin) == 1,
  "Insulin",
  "Sulfonylurea"
)

rwd$treatment <- factor(
  rwd$treatment,
  levels = c("Sulfonylurea", "Insulin")
)

table(rwd$treatment)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### 3. Define the Study Cohort

<!-- generated-learning-note -->
This subsection explains **3. Define the Study Cohort** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


A real-world study should prespecify eligibility criteria, index date, baseline period, treatment strategy, follow-up, and outcome.

For demonstration, we restrict the cohort to adults with plausible baseline values.


<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r cohort-definition}
cohort <- rwd %>%
  filter(
    age >= 40,
    age <= 85,
    hba1c >= 5,
    hba1c <= 14,
    creatinine > 0
  )

nrow(cohort)
table(cohort$treatment)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.


#### 4. Define Exposure, Outcome, and Baseline Covariates

<!-- generated-learning-note -->
This subsection explains **4. Define Exposure, Outcome, and Baseline Covariates** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


The exposure is treatment intensification:

- Insulin
- Sulfonylurea

Potential baseline confounders include age, sex, BMI, HbA1c, creatinine, blood pressure, smoking, prior cardiovascular disease, hypertension, and chronic kidney disease.

We now simulate a time-to-event outcome. In a real study, outcome events would be identified from diagnosis/procedure codes, death records, or adjudication.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r outcome-generation}
lp_event <-
  -4.3 +
  0.22 * (cohort$treatment == "Insulin") +
  0.025 * (cohort$age - 60) +
  0.30 * cohort$prior_cvd +
  0.20 * cohort$ckd +
  0.15 * cohort$smoking +
  0.08 * (cohort$hba1c - 8)

event_rate <- exp(lp_event)

true_event_time <- rexp(
  nrow(cohort),
  rate = event_rate
)

censor_time <- runif(
  nrow(cohort),
  min = 1,
  max = 5
)

cohort$followup_time <- pmin(true_event_time, censor_time)

cohort$event <- as.integer(
  true_event_time <= censor_time
)

table(cohort$event)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### 5. Data Quality Checks

<!-- generated-learning-note -->
This section prepares the information needed for **5. Data Quality Checks**. Check variable definitions, units, coding, and missing values before moving to analysis.


Before analysis, check missingness, implausible values, duplicates, coding consistency, and follow-up.


<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r data-quality}
# Duplicate patient IDs
sum(duplicated(cohort$id))

# Missingness
colSums(is.na(cohort))

# Basic distributions
summary(
  cohort[, c(
    "age",
    "bmi",
    "hba1c",
    "creatinine",
    "systolic_bp",
    "followup_time"
  )]
)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


#### 6. Examine Baseline Imbalance Before Adjustment

<!-- generated-learning-note -->
This subsection explains **6. Examine Baseline Imbalance Before Adjustment** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


Because treatment was not randomized, the treatment groups may differ substantially at baseline.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r baseline-before}
vars <- c(
  "age",
  "male",
  "bmi",
  "hba1c",
  "creatinine",
  "systolic_bp",
  "smoking",
  "prior_cvd",
  "hypertension",
  "ckd"
)

table1_before <- CreateTableOne(
  vars = vars,
  strata = "treatment",
  data = cohort,
  test = FALSE
)

print(
  table1_before,
  smd = TRUE
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### 7. Estimate the Propensity Score

<!-- generated-learning-note -->
This section develops **7. Estimate the Propensity Score**. Read the model specification together with its assumptions and the scale of each reported effect.


The propensity score estimates each patient's probability of receiving insulin conditional on measured baseline covariates.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r propensity-score}
ps_model <- glm(
  I(treatment == "Insulin") ~
    age +
    male +
    bmi +
    hba1c +
    creatinine +
    systolic_bp +
    smoking +
    prior_cvd +
    hypertension +
    ckd,
  family = binomial(),
  data = cohort
)

cohort$ps <- predict(
  ps_model,
  type = "response"
)

summary(cohort$ps)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 8. Propensity Score Matching

<!-- generated-learning-note -->
This subsection explains **8. Propensity Score Matching** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


Perform 1:1 nearest-neighbor propensity-score matching.


<!-- generated-learning-note -->
**Purpose.** This block imports or selects the data used in the example. Verify the file location, column types, coding, and missing values when substituting your own data.

```{r matching}
m.out <- matchit(
  treatment ~
    age +
    male +
    bmi +
    hba1c +
    creatinine +
    systolic_bp +
    smoking +
    prior_cvd +
    hypertension +
    ckd,
  data = cohort,
  method = "nearest",
  distance = "glm",
  ratio = 1,
  caliper = 0.2
)

summary(m.out)

matched <- match.data(m.out)

table(matched$treatment)
```

<!-- generated-learning-note -->
**How to read the output.** Inspect dimensions, names, types, and several rows of the imported data. Unexpected structure at this stage will propagate into every later result.


#### 9. Check Covariate Balance

<!-- generated-learning-note -->
This subsection explains **9. Check Covariate Balance** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


A common rule of thumb is an absolute standardized mean difference (SMD) below 0.10.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r balance}
bal.tab(
  m.out,
  un = TRUE,
  thresholds = c(m = 0.10)
)

love.plot(
  m.out,
  threshold = 0.10,
  abs = TRUE
)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


Good balance among measured covariates does **not** prove that unmeasured confounding has been eliminated.

#### 10. Primary Outcome Analysis

<!-- generated-learning-note -->
This subsection explains **10. Primary Outcome Analysis** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


Use a Cox proportional hazards model to compare time to the composite event between treatment groups.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r primary-analysis}
cox_fit <- coxph(
  Surv(followup_time, event) ~ treatment,
  data = matched,
  weights = weights,
  robust = TRUE
)

summary(cox_fit)

exp(
  cbind(
    HR = coef(cox_fit),
    confint(cox_fit)
  )
)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


Interpretation:

- HR > 1: higher event hazard in the insulin group.
- HR < 1: lower event hazard in the insulin group.
- A 95% confidence interval containing 1 indicates that the difference is not statistically significant at the conventional 0.05 level.

Because this is an observational study, the estimate should be interpreted as an adjusted association / causal estimate under the required causal assumptions rather than proof of causality.

#### 11. Sensitivity Analysis 1: IPTW

<!-- generated-learning-note -->
This subsection explains **11. Sensitivity Analysis 1: IPTW** and shows how it contributes to the overall Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin workflow.


Instead of matching, estimate the average treatment effect using inverse probability of treatment weighting.


<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r iptw}
cohort <- cohort %>%
  mutate(
    treat_num = as.integer(treatment == "Insulin"),
    weight_iptw = ifelse(
      treat_num == 1,
      1 / ps,
      1 / (1 - ps)
    )
  )

# Truncate extreme weights at the 1st and 99th percentiles
limits <- quantile(
  cohort$weight_iptw,
  probs = c(0.01, 0.99)
)

cohort <- cohort %>%
  mutate(
    weight_trim = pmin(
      pmax(weight_iptw, limits[1]),
      limits[2]
    )
  )

summary(cohort$weight_trim)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r iptw-outcome}
cox_iptw <- coxph(
  Surv(followup_time, event) ~ treatment,
  data = cohort,
  weights = weight_trim,
  robust = TRUE
)

summary(cox_iptw)

exp(
  cbind(
    HR = coef(cox_iptw),
    confint(cox_iptw)
  )
)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 12. Sensitivity Analysis 2: Conventional Multivariable Adjustment

<!-- generated-learning-note -->
This section prepares the information needed for **12. Sensitivity Analysis 2: Conventional Multivariable Adjustment**. Check variable definitions, units, coding, and missing values before moving to analysis.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r regression-adjustment}
cox_adjusted <- coxph(
  Surv(followup_time, event) ~
    treatment +
    age +
    male +
    bmi +
    hba1c +
    creatinine +
    systolic_bp +
    smoking +
    prior_cvd +
    hypertension +
    ckd,
  data = cohort
)

summary(cox_adjusted)

exp(
  cbind(
    HR = coef(cox_adjusted),
    confint(cox_adjusted)
  )
)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### 13. Compare Results Across Methods

<!-- generated-learning-note -->
This section develops **13. Compare Results Across Methods**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r compare-results}
extract_hr <- function(model, label) {
  est <- coef(model)["treatmentInsulin"]
  ci <- confint(model)["treatmentInsulin", ]

  data.frame(
    Method = label,
    HR = exp(est),
    Lower95 = exp(ci[1]),
    Upper95 = exp(ci[2])
  )
}

results <- bind_rows(
  extract_hr(cox_fit, "Propensity-score matching"),
  extract_hr(cox_iptw, "IPTW"),
  extract_hr(cox_adjusted, "Multivariable Cox")
)

results
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


If the estimates are reasonably similar across multiple prespecified approaches, confidence in robustness is increased. However, agreement does not eliminate the possibility of unmeasured confounding or other biases.

#### 14. Interpretation and Limitations

<!-- generated-learning-note -->
This section focuses on **14. Interpretation and Limitations**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.


The real-world evidence workflow is:

1. Define the causal question.
2. Identify the real-world data source.
3. Define eligibility criteria and the index date.
4. Define treatment strategies.
5. Define outcomes and follow-up.
6. Prespecify baseline confounders using clinical knowledge and causal reasoning.
7. Assess data quality.
8. Evaluate baseline imbalance.
9. Adjust measured confounding using matching, weighting, or regression.
10. Check covariate balance and positivity.
11. Estimate the treatment effect.
12. Perform sensitivity analyses.
13. Interpret the findings while acknowledging residual and unmeasured confounding.

## Interpreting the Results

Evaluate data provenance, eligibility, treatment assignment, follow-up, outcome ascertainment, confounding, and missingness. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
