---
title: "Real-World Evidence Study: Insulin vs Sulfonylurea After Metformin"
author: "Daniel He"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# 1. Background and Research Question

This educational example mirrors the design of a published real-world retrospective cohort study among patients with type 2 diabetes who were initially treated with metformin and subsequently added either insulin or a sulfonylurea.

**Research question:**  
Among eligible patients with type 2 diabetes receiving metformin, is adding insulin associated with a different risk of myocardial infarction, stroke, or all-cause death compared with adding a sulfonylurea?

Because treatment is not randomized, baseline confounding must be addressed.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(MatchIt)
library(cobalt)
library(survival)
library(tableone)

set.seed(2026)
```

# 2. Create / Import Real-World Data

In a real study, this step would import EHR, claims, registry, pharmacy, laboratory, or linked administrative data.

Here we simulate a dataset with the same basic structure.

```{r simulate-data}
n <- 5000

rwd <- data.frame(
  id = 1:n,
  age = round(rnorm(n, 62, 9)),
  male = rbinom(n, 1, 0.94),
  bmi = round(rnorm(n, 31, 5), 1),
  hba1c = round(rnorm(n, 8.1, 1.3), 1),
  creatinine = round(pmax(rnorm(n, 1.1, 0.35), 0.4), 2),
  systolic_bp = round(rnorm(n, 132, 15)),
  smoking = rbinom(n, 1, 0.25),
  prior_cvd = rbinom(n, 1, 0.25),
  hypertension = rbinom(n, 1, 0.65),
  ckd = rbinom(n, 1, 0.18)
)

# Treatment selection depends on baseline patient characteristics.
# This intentionally creates confounding by indication.
lp_treat <-
  -3.0 +
  0.02 * (rwd$age - 60) +
  0.55 * (rwd$hba1c - 8) +
  0.35 * rwd$prior_cvd +
  0.45 * rwd$ckd +
  0.02 * (rwd$bmi - 30)

p_insulin <- plogis(lp_treat)

rwd$treatment <- ifelse(
  rbinom(n, 1, p_insulin) == 1,
  "Insulin",
  "Sulfonylurea"
)

rwd$treatment <- factor(
  rwd$treatment,
  levels = c("Sulfonylurea", "Insulin")
)

table(rwd$treatment)
```

# 3. Define the Study Cohort

A real-world study should prespecify eligibility criteria, index date, baseline period, treatment strategy, follow-up, and outcome.

For demonstration, we restrict the cohort to adults with plausible baseline values.

```{r cohort-definition}
cohort <- rwd %>%
  filter(
    age >= 40,
    age <= 85,
    hba1c >= 5,
    hba1c <= 14,
    creatinine > 0
  )

nrow(cohort)
table(cohort$treatment)
```

# 4. Define Exposure, Outcome, and Baseline Covariates

The exposure is treatment intensification:

- Insulin
- Sulfonylurea

Potential baseline confounders include age, sex, BMI, HbA1c, creatinine, blood pressure, smoking, prior cardiovascular disease, hypertension, and chronic kidney disease.

We now simulate a time-to-event outcome. In a real study, outcome events would be identified from diagnosis/procedure codes, death records, or adjudication.

```{r outcome-generation}
lp_event <-
  -4.3 +
  0.22 * (cohort$treatment == "Insulin") +
  0.025 * (cohort$age - 60) +
  0.30 * cohort$prior_cvd +
  0.20 * cohort$ckd +
  0.15 * cohort$smoking +
  0.08 * (cohort$hba1c - 8)

event_rate <- exp(lp_event)

true_event_time <- rexp(
  nrow(cohort),
  rate = event_rate
)

censor_time <- runif(
  nrow(cohort),
  min = 1,
  max = 5
)

cohort$followup_time <- pmin(true_event_time, censor_time)

cohort$event <- as.integer(
  true_event_time <= censor_time
)

table(cohort$event)
```

# 5. Data Quality Checks

Before analysis, check missingness, implausible values, duplicates, coding consistency, and follow-up.

```{r data-quality}
# Duplicate patient IDs
sum(duplicated(cohort$id))

# Missingness
colSums(is.na(cohort))

# Basic distributions
summary(
  cohort[, c(
    "age",
    "bmi",
    "hba1c",
    "creatinine",
    "systolic_bp",
    "followup_time"
  )]
)
```

# 6. Examine Baseline Imbalance Before Adjustment

Because treatment was not randomized, the treatment groups may differ substantially at baseline.

```{r baseline-before}
vars <- c(
  "age",
  "male",
  "bmi",
  "hba1c",
  "creatinine",
  "systolic_bp",
  "smoking",
  "prior_cvd",
  "hypertension",
  "ckd"
)

table1_before <- CreateTableOne(
  vars = vars,
  strata = "treatment",
  data = cohort,
  test = FALSE
)

print(
  table1_before,
  smd = TRUE
)
```

# 7. Estimate the Propensity Score

The propensity score estimates each patient's probability of receiving insulin conditional on measured baseline covariates.

```{r propensity-score}
ps_model <- glm(
  I(treatment == "Insulin") ~
    age +
    male +
    bmi +
    hba1c +
    creatinine +
    systolic_bp +
    smoking +
    prior_cvd +
    hypertension +
    ckd,
  family = binomial(),
  data = cohort
)

cohort$ps <- predict(
  ps_model,
  type = "response"
)

summary(cohort$ps)
```

# 8. Propensity Score Matching

Perform 1:1 nearest-neighbor propensity-score matching.

```{r matching}
m.out <- matchit(
  treatment ~
    age +
    male +
    bmi +
    hba1c +
    creatinine +
    systolic_bp +
    smoking +
    prior_cvd +
    hypertension +
    ckd,
  data = cohort,
  method = "nearest",
  distance = "glm",
  ratio = 1,
  caliper = 0.2
)

summary(m.out)

matched <- match.data(m.out)

table(matched$treatment)
```

# 9. Check Covariate Balance

A common rule of thumb is an absolute standardized mean difference (SMD) below 0.10.

```{r balance}
bal.tab(
  m.out,
  un = TRUE,
  thresholds = c(m = 0.10)
)

love.plot(
  m.out,
  threshold = 0.10,
  abs = TRUE
)
```

Good balance among measured covariates does **not** prove that unmeasured confounding has been eliminated.

# 10. Primary Outcome Analysis

Use a Cox proportional hazards model to compare time to the composite event between treatment groups.

```{r primary-analysis}
cox_fit <- coxph(
  Surv(followup_time, event) ~ treatment,
  data = matched,
  weights = weights,
  robust = TRUE
)

summary(cox_fit)

exp(
  cbind(
    HR = coef(cox_fit),
    confint(cox_fit)
  )
)
```

Interpretation:

- HR > 1: higher event hazard in the insulin group.
- HR < 1: lower event hazard in the insulin group.
- A 95% confidence interval containing 1 indicates that the difference is not statistically significant at the conventional 0.05 level.

Because this is an observational study, the estimate should be interpreted as an adjusted association / causal estimate under the required causal assumptions rather than proof of causality.

# 11. Sensitivity Analysis 1: IPTW

Instead of matching, estimate the average treatment effect using inverse probability of treatment weighting.

```{r iptw}
cohort <- cohort %>%
  mutate(
    treat_num = as.integer(treatment == "Insulin"),
    weight_iptw = ifelse(
      treat_num == 1,
      1 / ps,
      1 / (1 - ps)
    )
  )

# Truncate extreme weights at the 1st and 99th percentiles
limits <- quantile(
  cohort$weight_iptw,
  probs = c(0.01, 0.99)
)

cohort <- cohort %>%
  mutate(
    weight_trim = pmin(
      pmax(weight_iptw, limits[1]),
      limits[2]
    )
  )

summary(cohort$weight_trim)
```

```{r iptw-outcome}
cox_iptw <- coxph(
  Surv(followup_time, event) ~ treatment,
  data = cohort,
  weights = weight_trim,
  robust = TRUE
)

summary(cox_iptw)

exp(
  cbind(
    HR = coef(cox_iptw),
    confint(cox_iptw)
  )
)
```

# 12. Sensitivity Analysis 2: Conventional Multivariable Adjustment

```{r regression-adjustment}
cox_adjusted <- coxph(
  Surv(followup_time, event) ~
    treatment +
    age +
    male +
    bmi +
    hba1c +
    creatinine +
    systolic_bp +
    smoking +
    prior_cvd +
    hypertension +
    ckd,
  data = cohort
)

summary(cox_adjusted)

exp(
  cbind(
    HR = coef(cox_adjusted),
    confint(cox_adjusted)
  )
)
```

# 13. Compare Results Across Methods

```{r compare-results}
extract_hr <- function(model, label) {
  est <- coef(model)["treatmentInsulin"]
  ci <- confint(model)["treatmentInsulin", ]

  data.frame(
    Method = label,
    HR = exp(est),
    Lower95 = exp(ci[1]),
    Upper95 = exp(ci[2])
  )
}

results <- bind_rows(
  extract_hr(cox_fit, "Propensity-score matching"),
  extract_hr(cox_iptw, "IPTW"),
  extract_hr(cox_adjusted, "Multivariable Cox")
)

results
```

If the estimates are reasonably similar across multiple prespecified approaches, confidence in robustness is increased. However, agreement does not eliminate the possibility of unmeasured confounding or other biases.

# 14. Interpretation and Limitations

The real-world evidence workflow is:

1. Define the causal question.
2. Identify the real-world data source.
3. Define eligibility criteria and the index date.
4. Define treatment strategies.
5. Define outcomes and follow-up.
6. Prespecify baseline confounders using clinical knowledge and causal reasoning.
7. Assess data quality.
8. Evaluate baseline imbalance.
9. Adjust measured confounding using matching, weighting, or regression.
10. Check covariate balance and positivity.
11. Estimate the treatment effect.
12. Perform sensitivity analyses.
13. Interpret the findings while acknowledging residual and unmeasured confounding.


