---
title: "R shiny practice"
description: "A practical R learning guide to R shiny practice, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Data Management & Programming"
category: "SAS & R Programming"
topic: "R shiny practice"
type: article
tags: ["SAS & R Programming", "Data Management & Programming", "R", "shiny", "practice"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Clear statistical programs make an analysis reproducible, reviewable, and easier to validate.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **R shiny practice**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. R shiny practice

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **R shiny practice** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### R shiny practice

<!-- generated-learning-note -->
This subsection explains **R shiny practice** and shows how it contributes to the overall R shiny practice workflow.


- [Interactive Correlation Explorer + Heatmap](https://pl8ev5-dian-he.shinyapps.io/shiny/)

- [Cumulative Incidence by uploading your data](https://pl8ev5-dian-he.shinyapps.io/surv_curv/)

E.g.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# Create the dataset
data <- data.frame(
  time = c(5, 8, 12, 3, 7, 2, 10, 6, 4, 9,
           6, 11, 7, 13, 2, 5, 9, 3, 8, 12),
  status = c(1, 0, 1, 1, 0, 1, 0, 1, 0, 1,
             0, 1, 0, 1, 1, 0, 1, 0, 1, 0),
  group = c(rep("A", 10), rep("B", 10))
)

# View the dataset
print(data)

```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

## Interpreting the Results

Read each example as a sequence of inputs, transformations, and outputs; then change one element at a time. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
