---
title: "ISE/ISS pooling"
output: html_document
date: "2025-10-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### ISE/ISS pooling

Designed and executed an Integrated Summary of Efficacy and Safety (ISE/ISS) simulation to evaluate pooled treatment effects across multiple clinical trials.  

Generated realistic patient-level datasets (treatment, study, demographic, efficacy, and adverse event data) to mimic multi-study integration scenarios.
 
Built R-based simulation pipelines using dplyr, tidyr, and survival to automate data creation, merging, and summary analyses.

Validated simulated outcomes through model-based verification and visualization of efficacy and safety profiles.


```{r}
# ====================================================
# R CODE: Simulate ISE / ISS data + Meta-Analysis demo
# ====================================================

# Load required packages
library(dplyr)
library(tidyr)
library(metafor)
library(ggplot2)
library(purrr)

set.seed(12345)

# --------------------------
# Step 1. Define study design
# --------------------------
studies <- data.frame(
  study_id = paste0("S", 1:5),
  n_per_arm = c(120, 80, 150, 60, 100),
  treat_effect = c(2.0, 1.5, 2.5, 0.8, 1.8),
  sd = c(4.0, 3.5, 4.5, 3.0, 4.0),
  p_safety_ctrl = c(0.03, 0.04, 0.02, 0.05, 0.025),
  rd_safety = c(0.01, 0.02, 0.015, 0.0, 0.005)
)

# Function to simulate one study (efficacy + safety)
simulate_study <- function(study_id, n_per_arm, treat_effect, sd, p_safety_ctrl, rd_safety) {
  arms <- c("Control", "Treatment")
  df_list <- list()
  for (arm in arms) {
    arm_mult <- ifelse(arm == "Treatment", 1, 0)
    n <- n_per_arm
    eff <- rnorm(n, mean = treat_effect * arm_mult, sd = sd)
    p_safety <- p_safety_ctrl + ifelse(arm == "Treatment", rd_safety, 0)
    safety_event <- rbinom(n, size = 1, prob = p_safety)
    df_list[[arm]] <- data.frame(
      study_id = study_id,
      arm = arm,
      efficacy = eff,
      safety_event = safety_event
    )
  }
  do.call(rbind, df_list)
}

# Simulate all studies
patient_df <- purrr::pmap_dfr(studies, simulate_study)
head( patient_df)
# Write patient-level data
 

# --------------------------
# Step 2. Study-level summary
# --------------------------
study_summary <- patient_df %>%
  group_by(study_id, arm) %>%
  summarise(
    n = n(),
    mean_eff = mean(efficacy),
    sd_eff = sd(efficacy),
    safety_events = sum(safety_event),
    .groups = "drop"
  )

study_summary

# --------------------------
# Step 3. Compute effect estimates
# --------------------------
study_est <- study_summary %>%
  pivot_wider(
    names_from = arm,
    values_from = c(n, mean_eff, sd_eff, safety_events)
  ) %>%
  mutate(
    yi = mean_eff_Treatment - mean_eff_Control,
    vi = (sd_eff_Treatment^2 / n_Treatment) + (sd_eff_Control^2 / n_Control),
    pT = safety_events_Treatment / n_Treatment,
    pC = safety_events_Control / n_Control,
    rd = pT - pC,
    v_rd = (pT*(1 - pT) / n_Treatment) + (pC*(1 - pC) / n_Control)
  )

study_est

# --------------------------
# Step 4. Meta-analysis (Efficacy)
# --------------------------
res_eff <- rma(yi = yi, vi = vi, data = study_est, method = "REML")
print(res_eff)

# Forest plot for efficacy
forest(res_eff,
       slab = study_est$study_id,
       xlab = "Mean Difference (Treatment - Control)",
       main = "ISE / ISS Pooled Efficacy Effect")

# --------------------------
# Step 5. Meta-analysis (Safety Risk Difference)
# --------------------------
res_rd <- rma(yi = rd, vi = v_rd, data = study_est, method = "REML")
print(res_rd)

# Forest plot for safety
forest(res_rd,
       slab = study_est$study_id,
       xlab = "Risk Difference (Treatment - Control)",
       main = "ISE / ISS Pooled Safety Risk Difference")

# --------------------------
# Step 6. Optional: Simple ggplot summary
# --------------------------
ggplot(study_est, aes(x = study_id, y = yi)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = yi - 1.96*sqrt(vi), ymax = yi + 1.96*sqrt(vi)), width = 0.2) +
  geom_hline(yintercept = res_eff$b[1], color = "red", linetype = "dashed") +
  labs(title = "Study-wise and Pooled Efficacy Estimates",
       y = "Mean Difference (Treatment - Control)",
       x = "Study ID") +
  theme_minimal()



```

