---
title: "BE_Comparision_Report"
output: html_document
date: "2025-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

This document demonstrates a simulation-based approach to evaluate bioequivalence (BE) between a **Test** and **Reference** product using pharmacokinetic (PK) parameters such as Cmax and AUC.  

Monte Carlo simulations are used to incorporate between-subject variability and assess the robustness of BE conclusions.
 
We simulate individual PK metrics assuming log-normal distributions for both treatments, reflecting realistic between-subject variability.  

The bioequivalence criterion follows the standard **90% confidence interval (CI)** of the Test/Reference ratio within **80%–125%** on the geometric scale.


### BE_Comparision_Report

```{r}
# ============================================================
# Example: Simulated Bioequivalence Comparison (Parallel Design)
# ============================================================

set.seed(2025)

# --- Step 1. Define time points and reference mean profile ---
time_hr <- c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 12, 24)
mean_conc_ref <- c(0, 12.5, 28.3, 42.8, 36.0, 20.1, 12.6, 7.8, 2.1, 0.2)

# --- Step 2. Simulation parameters ---
n_subj <- 24            # per group
CV_between <- 0.25      # between-subject CV
CV_residual <- 0.20     # within-subject (residual) CV
true_GMR <- 0.95        # true T/R ratio for exposure, true_GMR = true geometric mean ratio (test/reference) for PK parameter (e.g., AUC or Cmax).

ln_sd_between <- sqrt(log(1 + CV_between^2))     #standard deviations in the log scale are calculated
ln_sd_resid   <- sqrt(log(1 + CV_residual^2))

# --- Step 3. Helper function for AUC ---
trap_auc <- function(t, conc){
  sum(0.5 * (conc[-1] + conc[-length(conc)]) * diff(t))
}
# Computes the area under the curve (AUC) using the trapezoidal rule.
# conc[-1] = all points except first, conc[-length(conc)] = all points except last.
# diff(t) = difference between consecutive time points.

# --- Step 4. Simulate individual concentration-time data ---
simulate_subject <- function(mean_conc, ln_sd_between, ln_sd_resid, gmr = 1){
  eta <- exp(rnorm(1, 0, ln_sd_between))
  eps <- exp(rnorm(length(mean_conc), 0, ln_sd_resid))
  mean_conc * eta * eps * gmr
}

# eta = subject-specific random effect (between-subject variability).
# eps = residual error at each time point (within-subject variability).
# The final simulated concentrations = mean_conc * eta * eps * gmr.
# gmr allows simulating a test product with a true geometric mean ratio different from 1.

ref_conc <- replicate(n_subj, simulate_subject(mean_conc_ref, ln_sd_between, ln_sd_resid))
test_conc <- replicate(n_subj, simulate_subject(mean_conc_ref, ln_sd_between, ln_sd_resid, gmr = true_GMR))
# replicate generates PK profiles for all subjects in reference and test groups.

# --- Step 5. Compute AUC and Cmax for each subject ---
ref_auc  <- apply(ref_conc, 2, function(x) trap_auc(time_hr, x))
test_auc <- apply(test_conc, 2, function(x) trap_auc(time_hr, x))
ref_cmax  <- apply(ref_conc, 2, max)
test_cmax <- apply(test_conc, 2, max)

# Compute AUC for each subject using trap_auc.
# Compute Cmax (maximum concentration) for each subject using max.
# apply(..., 2, ...) = apply function across columns (each column = one subject).

# --- Step 6. BE comparison using log-transformed data ---
be_stats <- function(yT, yR, alpha = 0.10){
  logT <- log(yT)
  logR <- log(yR)
  mean_diff <- mean(logT) - mean(logR)
  se_diff <- sqrt(var(logT)/length(yT) + var(logR)/length(yR))
  df <- length(yT) + length(yR) - 2
  tcrit <- qt(1 - alpha/2, df)
  lower <- mean_diff - tcrit * se_diff
  upper <- mean_diff + tcrit * se_diff
  data.frame(
    Ratio = exp(mean_diff),    #Exponentiate to return to original scale
    Lower90 = exp(lower),
    Upper90 = exp(upper)
  )
}

auc_BE  <- be_stats(test_auc, ref_auc)
cmax_BE <- be_stats(test_cmax, ref_cmax)

# --- Step 7. Output summary ---
cat("=== Bioequivalence Results ===\n")
print(auc_BE)
print(cmax_BE)

# --- Step 8. Plot example PK profiles ---
library(ggplot2)
library(dplyr)
plot_df <- data.frame(
  Time = rep(time_hr, 2 * n_subj),
  Conc = c(as.vector(ref_conc), as.vector(test_conc)),
  Subject = rep(1:(2*n_subj), each = length(time_hr)),
  Group = rep(c(rep("Reference", n_subj), rep("Test", n_subj)), each = length(time_hr))
)
# Includes Time, Conc, Subject, and Group columns.

ggplot(plot_df, aes(x = Time, y = Conc, color = Group, group = Subject)) +
  geom_line(alpha = 0.5) +
  scale_y_continuous(trans = "log10") +
  labs(title = "Simulated Concentration–Time Profiles",
       y = "Concentration (ng/mL, log scale)", x = "Time (hr)") +
  theme_minimal(base_size = 13)

```

