---
title: "BE_Comparision_Report"
description: "A practical R learning guide to BE_Comparision_Report, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Non-Inferiority & Equivalence"
topic: "BE_Comparision_Report"
type: article
tags: ["Non-Inferiority & Equivalence", "Biostatistics", "BE", "Comparision", "Report"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Non-inferiority and equivalence analyses compare effects with clinically justified margins rather than testing only for any difference.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **BE_Comparision_Report**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Introduction
2. BEComparisionReport

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Non-Inferiority & Equivalence",
  topic = "BE_Comparision_Report"
)
learning_overview
```

## Learning Material

The following material develops **BE_Comparision_Report** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Introduction

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **Introduction** in the context of BE_Comparision_Report.


This document demonstrates a simulation-based approach to evaluate bioequivalence (BE) between a **Test** and **Reference** product using pharmacokinetic (PK) parameters such as Cmax and AUC.  

Monte Carlo simulations are used to incorporate between-subject variability and assess the robustness of BE conclusions.
 
We simulate individual PK metrics assuming log-normal distributions for both treatments, reflecting realistic between-subject variability.  

The bioequivalence criterion follows the standard **90% confidence interval (CI)** of the Test/Reference ratio within **80%–125%** on the geometric scale.


#### BEComparisionReport

<!-- generated-learning-note -->
This subsection explains **BEComparisionReport** and shows how it contributes to the overall BE_Comparision_Report workflow.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# ============================================================
# Example: Simulated Bioequivalence Comparison (Parallel Design)
# ============================================================

set.seed(2025)

# --- Step 1. Define time points and reference mean profile ---
time_hr <- c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 12, 24)
mean_conc_ref <- c(0, 12.5, 28.3, 42.8, 36.0, 20.1, 12.6, 7.8, 2.1, 0.2)

# --- Step 2. Simulation parameters ---
n_subj <- 24            # per group
CV_between <- 0.25      # between-subject CV
CV_residual <- 0.20     # within-subject (residual) CV
true_GMR <- 0.95        # true T/R ratio for exposure, true_GMR = true geometric mean ratio (test/reference) for PK parameter (e.g., AUC or Cmax).

ln_sd_between <- sqrt(log(1 + CV_between^2))     #standard deviations in the log scale are calculated
ln_sd_resid   <- sqrt(log(1 + CV_residual^2))

# --- Step 3. Helper function for AUC ---
trap_auc <- function(t, conc){
  sum(0.5 * (conc[-1] + conc[-length(conc)]) * diff(t))
}
# Computes the area under the curve (AUC) using the trapezoidal rule.
# conc[-1] = all points except first, conc[-length(conc)] = all points except last.
# diff(t) = difference between consecutive time points.

# --- Step 4. Simulate individual concentration-time data ---
simulate_subject <- function(mean_conc, ln_sd_between, ln_sd_resid, gmr = 1){
  eta <- exp(rnorm(1, 0, ln_sd_between))
  eps <- exp(rnorm(length(mean_conc), 0, ln_sd_resid))
  mean_conc * eta * eps * gmr
}

# eta = subject-specific random effect (between-subject variability).
# eps = residual error at each time point (within-subject variability).
# The final simulated concentrations = mean_conc * eta * eps * gmr.
# gmr allows simulating a test product with a true geometric mean ratio different from 1.

ref_conc <- replicate(n_subj, simulate_subject(mean_conc_ref, ln_sd_between, ln_sd_resid))
test_conc <- replicate(n_subj, simulate_subject(mean_conc_ref, ln_sd_between, ln_sd_resid, gmr = true_GMR))
# replicate generates PK profiles for all subjects in reference and test groups.

# --- Step 5. Compute AUC and Cmax for each subject ---
ref_auc  <- apply(ref_conc, 2, function(x) trap_auc(time_hr, x))
test_auc <- apply(test_conc, 2, function(x) trap_auc(time_hr, x))
ref_cmax  <- apply(ref_conc, 2, max)
test_cmax <- apply(test_conc, 2, max)

# Compute AUC for each subject using trap_auc.
# Compute Cmax (maximum concentration) for each subject using max.
# apply(..., 2, ...) = apply function across columns (each column = one subject).

# --- Step 6. BE comparison using log-transformed data ---
be_stats <- function(yT, yR, alpha = 0.10){
  logT <- log(yT)
  logR <- log(yR)
  mean_diff <- mean(logT) - mean(logR)
  se_diff <- sqrt(var(logT)/length(yT) + var(logR)/length(yR))
  df <- length(yT) + length(yR) - 2
  tcrit <- qt(1 - alpha/2, df)
  lower <- mean_diff - tcrit * se_diff
  upper <- mean_diff + tcrit * se_diff
  data.frame(
    Ratio = exp(mean_diff),    #Exponentiate to return to original scale
    Lower90 = exp(lower),
    Upper90 = exp(upper)
  )
}

auc_BE  <- be_stats(test_auc, ref_auc)
cmax_BE <- be_stats(test_cmax, ref_cmax)

# --- Step 7. Output summary ---
cat("=== Bioequivalence Results ===\n")
print(auc_BE)
print(cmax_BE)

# --- Step 8. Plot example PK profiles ---
library(ggplot2)
library(dplyr)
plot_df <- data.frame(
  Time = rep(time_hr, 2 * n_subj),
  Conc = c(as.vector(ref_conc), as.vector(test_conc)),
  Subject = rep(1:(2*n_subj), each = length(time_hr)),
  Group = rep(c(rep("Reference", n_subj), rep("Test", n_subj)), each = length(time_hr))
)
# Includes Time, Conc, Subject, and Group columns.

ggplot(plot_df, aes(x = Time, y = Conc, color = Group, group = Subject)) +
  geom_line(alpha = 0.5) +
  scale_y_continuous(trans = "log10") +
  labs(title = "Simulated Concentration–Time Profiles",
       y = "Concentration (ng/mL, log scale)", x = "Time (hr)") +
  theme_minimal(base_size = 13)

```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

## Interpreting the Results

Interpret confidence intervals relative to the margin and confirm that the analysis scale matches the scientific claim. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
