---
title: "Least-squares means with interaction"
description: "A practical R learning guide to Least-squares means with interaction, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Subgroup & Interaction Analysis"
topic: "Least-squares means with interaction"
type: article
tags: ["Subgroup & Interaction Analysis", "Biostatistics", "Least-squares", "means", "interaction"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Interaction analyses assess whether an effect differs across levels of another variable; subgroup estimates alone do not establish heterogeneity.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Least-squares means with interaction**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Least-squares means with interaction

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Least-squares means with interaction** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Reference](https://www.quantargo.com/help/r/latest/packages/merTools/0.5.2/predictInterval)

#### Least-squares means with interaction

<!-- generated-learning-note -->
This subsection explains **Least-squares means with interaction** and shows how it contributes to the overall Least-squares means with interaction workflow.


Without interaction

at a special value

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# Compute least-squares means for specified factors or factor combinations in a linear model

require(lsmeans)

# Use dataset 'fiber'

### model 1 without interaction, Covariance example 
fiber.lm <- lm(strength ~ diameter + machine, data = fiber)

# adjusted means and comparison
lsmeans(fiber.lm,   list (pairwise ~ machine ) ) 

# adjusted means and contrast, treating machine C as control
lsmeans (fiber.lm, "machine", contr = "trt.vs.ctrlk")

# adjusted means and comparison, after setting "diameter" as a specific value
lsmeans(fiber.lm,   list (pairwise ~ machine ), at = list(diameter =20) ) 
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


With interaction

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
### model 2 with interaction
fiber.lm2 <- lm(strength ~ diameter*machine, data=fiber)

# means when diameter is its mean
(lsmeans(fiber.lm2, ~ machine | diameter))

# means when setting diameter as its mean
lsmeans(fiber.lm2,   list (pairwise ~ machine ), at = list(diameter =24.1) ) 
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
### model 3 for category covariates
warp.lm <- lm(breaks ~ wool * tension, data = warpbreaks)

# means after adjusting the interaction effect
(lsmeans(warp.lm, ~ wool | tension))

# set tension ="M"
lsmeans(warp.lm,   list (pairwise ~ wool ), at = list(tension ="M")) 
lsmeans(warp.lm,   list (pairwise ~ wool ), at = list(tension ="H")) 
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


Compare slopes and test

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
### Trends, obtain slopes  
fiber.lst=( lstrends(fiber.lm, "machine", var="diameter") )
fiber.lst
# pairwise comparisons thereof slopes 
pairs(fiber.lst)

fiber.lst2=( lstrends(fiber.lm2, "machine", var="diameter") )
fiber.lst2
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


Least-squares means with interaction for LMM

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(lme4)
##############
m1 <- lmer(Reaction ~ Days + (1 | Subject), sleepstudy)
lsmeans(m1,    ~ Days  , at = list(Days =20)  ) 

lsmeans(m1,    ~ Days    ) 
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

## Interpreting the Results

Prioritize interaction contrasts and uncertainty over separate within-group significance tests. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
