---
title: "Linear regression MCMC"
description: "A practical R learning guide to Linear regression MCMC, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Adaptive & Bayesian Methods"
topic: "Linear regression MCMC"
type: article
tags: ["Adaptive & Bayesian Methods", "Biostatistics", "Linear", "regression", "MCMC"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Adaptive and Bayesian methods update evidence as data accumulate and express uncertainty through probability distributions or decision rules.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Linear regression MCMC**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Simple linear regression
2. Metropolis-Hastings algorithm
3. Plot parameters
4. Linear regression MCMC in JAGS
5. Plot(fitlm1)
6. Prediction

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Linear regression MCMC** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Simple linear regression

<!-- generated-learning-note -->
This section develops **Simple linear regression**. Read the model specification together with its assumptions and the scale of each reported effect.


`makes the posterior to be near prior = MCMC stability`

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
# Creating data
trueA = 5
trueB = 0
trueSd = 10
sampleSize = 31
# create independent x-values
x =(-(sampleSize-1)/2):((sampleSize-1)/2)
# create dependent values according to ax + b + N(0,sd)
y = trueA * x + trueB + rnorm(n=sampleSize,mean=0,sd=trueSd)
plot(x,y, main="Test Data")

```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.



<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r}
# For estimating parameters in a Bayesian analysis, we need to derive the likelihood function for the model 
likelihood = function(param){
  a = param[1]
  b = param[2]
  sd = param[3]
  
  pred = a*x + b
  singlelikelihoods = dnorm(y, mean = pred, sd = sd, log = T)
  sumll = sum(singlelikelihoods)
  return(sumll)
}

# Defining the prior
# Prior distribution
prior = function(param){
  a = param[1]
  b = param[2]
  sd = param[3]
  aprior = dunif(a, min=0, max=10, log = T)
  bprior = dnorm(b, sd = 5, log = T)
  sdprior = dunif(sd, min=0, max=30, log = T)
  return(aprior+bprior+sdprior)
}

# The posterior
posterior = function(param){
  return (likelihood(param) + prior(param))
}
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.


#### Metropolis-Hastings algorithm

<!-- generated-learning-note -->
This subsection explains **Metropolis-Hastings algorithm** and shows how it contributes to the overall Linear regression MCMC workflow.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
# proposal function
proposalfunction = function(param){
  return(rnorm(3,mean = param, sd= c(0.1,0.5,0.3)))
}

# MH algorithm 
run_metropolis_MCMC = function(startvalue, iterations){
  chain = array(dim = c(iterations+1,3))
  chain[1,] = startvalue
  for (i in 1:iterations){
    proposal = proposalfunction(chain[i,])
    
    probab = exp(posterior(proposal) - posterior(chain[i,]))     #p1/p2 = exp[log(p1)-log(p2)].
    if (runif(1) < probab){   #get the uniform distribution points which meet the requirement / accept rate 
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}

# run MH
startvalue = c(4,0,10)
chain = run_metropolis_MCMC(startvalue, 10000)

# acceptance rate
burnIn = 5000
acceptance = 1-mean(duplicated(chain[-(1:burnIn),]))  #The acceptance rate can be influenced by the proposal function
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.




<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
acceptance
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Plot parameters

<!-- generated-learning-note -->
This section uses **Plot parameters** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}

### Summary: #######################

par(mfrow = c(2,3))
hist(chain[-(1:burnIn),1],nclass=30, , main="Posterior of a", xlab="True value = red line" )
abline(v = mean(chain[-(1:burnIn),1]))
abline(v = trueA, col="red" )
hist(chain[-(1:burnIn),2],nclass=30, main="Posterior of b", xlab="True value = red line")
abline(v = mean(chain[-(1:burnIn),2]))
abline(v = trueB, col="red" )
hist(chain[-(1:burnIn),3],nclass=30, main="Posterior of sd", xlab="True value = red line")
abline(v = mean(chain[-(1:burnIn),3]) )
abline(v = trueSd, col="red" )
plot(chain[-(1:burnIn),1], type = "l", xlab="True value = red line" , main = "Chain values of a", )
abline(h = trueA, col="red" )
plot(chain[-(1:burnIn),2], type = "l", xlab="True value = red line" , main = "Chain values of b", )
abline(h = trueB, col="red" )
plot(chain[-(1:burnIn),3], type = "l", xlab="True value = red line" , main = "Chain values of sd", )
abline(h = trueSd, col="red" )
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
# slope
mean(chain[-(1:burnIn),1])
# intercept
mean(chain[-(1:burnIn),2])

```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.



#### Linear regression MCMC in JAGS

<!-- generated-learning-note -->
This section develops **Linear regression MCMC in JAGS**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# create data
snakes1 <- data.frame(b_length = x, b_mass = y)
head(snakes1)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(R2jags)
jagsdata_s1 <- with(snakes1, list(b_mass = b_mass, b_length = b_length, N = length(b_mass)))

lm1_jags <- function(){
  # Likelihood:
  for (i in 1:N){
    b_mass[i] ~ dnorm(mu[i], tau) # tau is precision (1 / variance)
    mu[i] <- alpha + beta * b_length[i]
  }
  # Priors:
  alpha ~ dnorm(0, 0.01) # intercept
  beta ~ dnorm(0, 0.01) # slope
  sigma ~ dunif(0, 100) # standard deviation
  tau <- 1 / (sigma * sigma) # sigma^2 doesn't work in JAGS
}

init_values <- function(){
  list(alpha = rnorm(1), beta = rnorm(1), sigma = runif(1))
}

params <- c("alpha", "beta", "sigma")

fit_lm1 <- jags(data = jagsdata_s1, inits = init_values, parameters.to.save = params, model.file = lm1_jags,
                n.chains = 3, n.iter = 12000, n.burnin = 5000, n.thin = 10, DIC = F)

fit_lm1
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Plot(fitlm1)

<!-- generated-learning-note -->
This section develops **Plot(fitlm1)**. Read the model specification together with its assumptions and the scale of each reported effect.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
# plot
# plot(fit_lm1)
traceplot(fit_lm1, mfrow = c(2, 2), ask = F)

lm1_mcmc <- as.mcmc(fit_lm1)
plot(lm1_mcmc)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


`If the Rhat values are substantially larger than one, the chains haven’t mixed properly and the posterior estimates cannot be trusted. `

#### Prediction

<!-- generated-learning-note -->
This subsection explains **Prediction** and shows how it contributes to the overall Linear regression MCMC workflow.


<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
# new dataset
nvalues <- 100
b_length_new <- seq(min(snakes1$b_length), max(snakes1$b_length), length.out = nvalues)

# we combine the three chains into one
lm1_mcmc_combi <- as.mcmc(rbind(lm1_mcmc[[1]], lm1_mcmc[[2]], lm1_mcmc[[3]]))

# calculate the expected value of y for each of the new x
pred_mean_mean <- mean(lm1_mcmc_combi[, "alpha"]) + b_length_new * mean(lm1_mcmc_combi[, "beta"])
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.




<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
head(pred_mean_mean)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### A kind of CI

<!-- generated-learning-note -->
This subsection explains **A kind of CI** and shows how it contributes to the overall Linear regression MCMC workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# the uncertainty around this mean. 
# ci
pred_mean_dist <- matrix(NA, nrow = nrow(lm1_mcmc_combi), ncol = nvalues)
for (i in 1:nrow(pred_mean_dist)){
  pred_mean_dist[i,] <- lm1_mcmc_combi[i,"alpha"] + b_length_new * lm1_mcmc_combi[i,"beta"]
}
credible_lower <- apply(pred_mean_dist, MARGIN = 2, quantile, prob = 0.025)
credible_upper <- apply(pred_mean_dist, MARGIN = 2, quantile, prob = 0.975)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Percentiles

<!-- generated-learning-note -->
This subsection explains **Percentiles** and shows how it contributes to the overall Linear regression MCMC workflow.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
# percentile
lm1_mcmc_combi_rep <- do.call(rbind, rep(list(lm1_mcmc_combi), 50)) # replication

# Draw random values for all parameter combinations (rows) and body length values (columns):
pred_data_dist <- matrix(NA, nrow = nrow(lm1_mcmc_combi_rep), ncol = nvalues)
for (i in 1:nrow(pred_data_dist)){
  pred_data_dist[i,] <- lm1_mcmc_combi_rep[i,"alpha"] + b_length_new * lm1_mcmc_combi_rep[i,"beta"] +
    rnorm(nvalues, mean = 0, sd = lm1_mcmc_combi_rep[i, "sigma"])
}

# Calculate quantiles:
uncertain_lower <- apply(pred_data_dist, MARGIN = 2, quantile, prob = 0.025)
uncertain_upper <- apply(pred_data_dist, MARGIN = 2, quantile, prob = 0.975)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.



#### Plot CI and percentiles

<!-- generated-learning-note -->
This section uses **Plot CI and percentiles** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
# plot
par(mfrow = c(1, 1))
plot(b_mass ~ b_length, data = snakes1)
lines(b_length_new, pred_mean_mean)
lines(b_length_new, credible_lower, lty = 2)
lines(b_length_new, credible_upper, lty = 2)
lines(b_length_new, uncertain_lower, lty = 2, col = "red")
lines(b_length_new, uncertain_upper, lty = 2, col = "red")
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


Comparison with SLR:

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
summary(lm(snakes1$b_mass~snakes1$b_length))

```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.




**The above three methonds have similar results. **

#### Multiple regression in JAGS

<!-- generated-learning-note -->
This section develops **Multiple regression in JAGS**. Read the model specification together with its assumptions and the scale of each reported effect.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# multiple regression 
# create dataset
snakes2 <- data.frame(b_length = x, b_mass = y, sex = rep(c(0,1),times=16)[1:31] )
head(snakes2)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
summary(lm( snakes2$b_mass~snakes2$b_length +snakes2$sex,data=snakes2))
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
plot(b_mass ~ b_length, col = (sex + 1), data = snakes2)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
# format the data
jagsdata_s2 <- with(snakes2, list(b_mass = b_mass, b_length = b_length, sex = sex, N = length(b_mass)))

# an R function containing the JAGS code:
  lm2_jags <- function(){
    # Likelihood:
    for (i in 1:N){
      b_mass[i] ~ dnorm(mu[i], tau) # tau is precision (1 / variance)
      mu[i] <- alpha[1]   + beta[1] * sex[i] + beta[2] * b_length[i]
    }
    # Priors:
    alpha[1] ~ dnorm(0, 0.01)
    for (i in 1:2){
      
      beta[i] ~ dnorm(0, 0.01)
    }
    sigma ~ dunif(0, 100)
    tau <- 1 / (sigma * sigma)
  }

  
  # Markov chains and initial values
  init_values <- function(){
    list(alpha = rnorm(1), beta = rnorm(2), sigma = runif(1))
  }
  
  params <- c("alpha", "beta", "sigma")
  
  fit_lm2 <- jags(data = jagsdata_s2, inits = init_values, parameters.to.save = params, model.file = lm2_jags,
                  n.chains = 3, n.iter = 12000, n.burnin = 2000, n.thin = 10, DIC = F)

  fit_lm2
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

**Results are similar to the multiple regression**

References

[1](https://biometry.github.io/APES//LectureNotes/StatsCafe/Linear_models_jags.html)

[2](https://theoreticalecology.wordpress.com/2010/09/17/metropolis-hastings-mcmc-in-r/)

[3](https://pubmed.ncbi.nlm.nih.gov/31777998/)

[4]()

## Interpreting the Results

Inspect priors, likelihoods, posterior summaries, convergence, simulations, and operating characteristics. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
