---
title: "Linear regression MCMC"
output: html_document
date: '2024-09-05'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simple linear regression

`makes the posterior to be near prior = MCMC stability` 


```{r}
# Creating data
trueA = 5
trueB = 0
trueSd = 10
sampleSize = 31
# create independent x-values
x =(-(sampleSize-1)/2):((sampleSize-1)/2)
# create dependent values according to ax + b + N(0,sd)
y = trueA * x + trueB + rnorm(n=sampleSize,mean=0,sd=trueSd)
plot(x,y, main="Test Data")

```

```{r}
# For estimating parameters in a Bayesian analysis, we need to derive the likelihood function for the model 
likelihood = function(param){
  a = param[1]
  b = param[2]
  sd = param[3]
  
  pred = a*x + b
  singlelikelihoods = dnorm(y, mean = pred, sd = sd, log = T)
  sumll = sum(singlelikelihoods)
  return(sumll)
}

# Defining the prior
# Prior distribution
prior = function(param){
  a = param[1]
  b = param[2]
  sd = param[3]
  aprior = dunif(a, min=0, max=10, log = T)
  bprior = dnorm(b, sd = 5, log = T)
  sdprior = dunif(sd, min=0, max=30, log = T)
  return(aprior+bprior+sdprior)
}

# The posterior
posterior = function(param){
  return (likelihood(param) + prior(param))
}
```

### Metropolis-Hastings algorithm
```{r}
# proposal function
proposalfunction = function(param){
  return(rnorm(3,mean = param, sd= c(0.1,0.5,0.3)))
}

# MH algorithm 
run_metropolis_MCMC = function(startvalue, iterations){
  chain = array(dim = c(iterations+1,3))
  chain[1,] = startvalue
  for (i in 1:iterations){
    proposal = proposalfunction(chain[i,])
    
    probab = exp(posterior(proposal) - posterior(chain[i,]))     #p1/p2 = exp[log(p1)-log(p2)].
    if (runif(1) < probab){   #get the uniform distribution points which meet the requirement / accept rate 
      chain[i+1,] = proposal
    }else{
      chain[i+1,] = chain[i,]
    }
  }
  return(chain)
}

# run MH
startvalue = c(4,0,10)
chain = run_metropolis_MCMC(startvalue, 10000)

# acceptance rate
burnIn = 5000
acceptance = 1-mean(duplicated(chain[-(1:burnIn),]))  #The acceptance rate can be influenced by the proposal function
```


```{r}
acceptance
```
### Plot parameters 
```{r}

### Summary: #######################

par(mfrow = c(2,3))
hist(chain[-(1:burnIn),1],nclass=30, , main="Posterior of a", xlab="True value = red line" )
abline(v = mean(chain[-(1:burnIn),1]))
abline(v = trueA, col="red" )
hist(chain[-(1:burnIn),2],nclass=30, main="Posterior of b", xlab="True value = red line")
abline(v = mean(chain[-(1:burnIn),2]))
abline(v = trueB, col="red" )
hist(chain[-(1:burnIn),3],nclass=30, main="Posterior of sd", xlab="True value = red line")
abline(v = mean(chain[-(1:burnIn),3]) )
abline(v = trueSd, col="red" )
plot(chain[-(1:burnIn),1], type = "l", xlab="True value = red line" , main = "Chain values of a", )
abline(h = trueA, col="red" )
plot(chain[-(1:burnIn),2], type = "l", xlab="True value = red line" , main = "Chain values of b", )
abline(h = trueB, col="red" )
plot(chain[-(1:burnIn),3], type = "l", xlab="True value = red line" , main = "Chain values of sd", )
abline(h = trueSd, col="red" )
```
```{r}
# slope
mean(chain[-(1:burnIn),1])
# intercept
mean(chain[-(1:burnIn),2])

```


## Linear regression MCMC in JAGS

```{r}
# create data
snakes1 <- data.frame(b_length = x, b_mass = y)
head(snakes1)
```

```{r}
library(R2jags)
jagsdata_s1 <- with(snakes1, list(b_mass = b_mass, b_length = b_length, N = length(b_mass)))

lm1_jags <- function(){
  # Likelihood:
  for (i in 1:N){
    b_mass[i] ~ dnorm(mu[i], tau) # tau is precision (1 / variance)
    mu[i] <- alpha + beta * b_length[i]
  }
  # Priors:
  alpha ~ dnorm(0, 0.01) # intercept
  beta ~ dnorm(0, 0.01) # slope
  sigma ~ dunif(0, 100) # standard deviation
  tau <- 1 / (sigma * sigma) # sigma^2 doesn't work in JAGS
}

init_values <- function(){
  list(alpha = rnorm(1), beta = rnorm(1), sigma = runif(1))
}

params <- c("alpha", "beta", "sigma")

fit_lm1 <- jags(data = jagsdata_s1, inits = init_values, parameters.to.save = params, model.file = lm1_jags,
                n.chains = 3, n.iter = 12000, n.burnin = 5000, n.thin = 10, DIC = F)

fit_lm1
```

### Plot(fit_lm1)
```{r}
# plot
# plot(fit_lm1)
traceplot(fit_lm1, mfrow = c(2, 2), ask = F)

lm1_mcmc <- as.mcmc(fit_lm1)
plot(lm1_mcmc)
```

`If the Rhat values are substantially larger than one, the chains haven’t mixed properly and the posterior estimates cannot be trusted. `

### Prediction
```{r}
# new dataset
nvalues <- 100
b_length_new <- seq(min(snakes1$b_length), max(snakes1$b_length), length.out = nvalues)

# we combine the three chains into one
lm1_mcmc_combi <- as.mcmc(rbind(lm1_mcmc[[1]], lm1_mcmc[[2]], lm1_mcmc[[3]]))

# calculate the expected value of y for each of the new x
pred_mean_mean <- mean(lm1_mcmc_combi[, "alpha"]) + b_length_new * mean(lm1_mcmc_combi[, "beta"])
```


```{r}
head(pred_mean_mean)
```

### A kind of CI
```{r}
# the uncertainty around this mean. 
# ci
pred_mean_dist <- matrix(NA, nrow = nrow(lm1_mcmc_combi), ncol = nvalues)
for (i in 1:nrow(pred_mean_dist)){
  pred_mean_dist[i,] <- lm1_mcmc_combi[i,"alpha"] + b_length_new * lm1_mcmc_combi[i,"beta"]
}
credible_lower <- apply(pred_mean_dist, MARGIN = 2, quantile, prob = 0.025)
credible_upper <- apply(pred_mean_dist, MARGIN = 2, quantile, prob = 0.975)
```

### Percentiles
```{r}
# percentile
lm1_mcmc_combi_rep <- do.call(rbind, rep(list(lm1_mcmc_combi), 50)) # replication

# Draw random values for all parameter combinations (rows) and body length values (columns):
pred_data_dist <- matrix(NA, nrow = nrow(lm1_mcmc_combi_rep), ncol = nvalues)
for (i in 1:nrow(pred_data_dist)){
  pred_data_dist[i,] <- lm1_mcmc_combi_rep[i,"alpha"] + b_length_new * lm1_mcmc_combi_rep[i,"beta"] +
    rnorm(nvalues, mean = 0, sd = lm1_mcmc_combi_rep[i, "sigma"])
}

# Calculate quantiles:
uncertain_lower <- apply(pred_data_dist, MARGIN = 2, quantile, prob = 0.025)
uncertain_upper <- apply(pred_data_dist, MARGIN = 2, quantile, prob = 0.975)
```


### Plot CI and percentiles
```{r}
# plot
par(mfrow = c(1, 1))
plot(b_mass ~ b_length, data = snakes1)
lines(b_length_new, pred_mean_mean)
lines(b_length_new, credible_lower, lty = 2)
lines(b_length_new, credible_upper, lty = 2)
lines(b_length_new, uncertain_lower, lty = 2, col = "red")
lines(b_length_new, uncertain_upper, lty = 2, col = "red")
```

Comparison with SLR:
```{r}
summary(lm(snakes1$b_mass~snakes1$b_length))

```



**The above three methonds have similar results. **

### Multiple regression in JAGS
```{r}
# multiple regression 
# create dataset
snakes2 <- data.frame(b_length = x, b_mass = y, sex = rep(c(0,1),times=16)[1:31] )
head(snakes2)
```
```{r}
summary(lm( snakes2$b_mass~snakes2$b_length +snakes2$sex,data=snakes2))
```

```{r}
plot(b_mass ~ b_length, col = (sex + 1), data = snakes2)
```

```{r}
# format the data
jagsdata_s2 <- with(snakes2, list(b_mass = b_mass, b_length = b_length, sex = sex, N = length(b_mass)))

# an R function containing the JAGS code:
  lm2_jags <- function(){
    # Likelihood:
    for (i in 1:N){
      b_mass[i] ~ dnorm(mu[i], tau) # tau is precision (1 / variance)
      mu[i] <- alpha[1]   + beta[1] * sex[i] + beta[2] * b_length[i]
    }
    # Priors:
    alpha[1] ~ dnorm(0, 0.01)
    for (i in 1:2){
      
      beta[i] ~ dnorm(0, 0.01)
    }
    sigma ~ dunif(0, 100)
    tau <- 1 / (sigma * sigma)
  }

  
  # Markov chains and initial values
  init_values <- function(){
    list(alpha = rnorm(1), beta = rnorm(2), sigma = runif(1))
  }
  
  params <- c("alpha", "beta", "sigma")
  
  fit_lm2 <- jags(data = jagsdata_s2, inits = init_values, parameters.to.save = params, model.file = lm2_jags,
                  n.chains = 3, n.iter = 12000, n.burnin = 2000, n.thin = 10, DIC = F)

  fit_lm2
```
**Results are similar to the multiple regression**

References

[1](https://biometry.github.io/APES//LectureNotes/StatsCafe/Linear_models_jags.html)

[2](https://theoreticalecology.wordpress.com/2010/09/17/metropolis-hastings-mcmc-in-r/)

[3](https://pubmed.ncbi.nlm.nih.gov/31777998/)

[4]()





