---
title: "Causal inference and associated regression"
output: html_document
date: '2024-01-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Causal inference and association regression

- Still need to consider the chronological order in design;
- Well designed observation study is suitable for causal inference;
- Generally associated analysis estimate is larger or smaller than the causal inference; 
- Associated analysis is used for strata and it is hard to deal with interaction (with exposure) and noncollapsibility so CI should be recommended;
- Data imputation is also needed;
- Double robust estimate is more robust;
- Bootstrapping sampling should be more than 80% with more than 1000 samples and the statistic is not necessarily normal distribution;
- Categorizing should be based on the special knowledge not quartile/tertile/ ROC/ ;
- Ltmle longitudinal study should be less than 30 times measurements; 
- Nonparameter method has fewer assumptions but is computation resource consuming (sometimes less precise); 
- The distribution will be changed after normalization;

#### Why RCT is gold standard

In a randomized control trial (RCT) we assume [that](https://stats.stackexchange.com/questions/308397/why-is-average-treatment-effect-different-from-average-treatment-effect-on-the-t). 

$$
\mathrm{E}[Y^0|D=1] = \mathrm{E}[Y^0|D=0]
$$
and
$$
\mathrm{E}[Y^1|D=1] = \mathrm{E}[Y^1|D=0]
$$


Since
$$
\begin{align}
\mathrm{E}[\delta] =& \{ \pi \mathrm{E}[Y^1|D=1] + (1-\pi) \mathrm{E}[Y^1|D=0] \} \\ -& \{ \pi \mathrm{E}[Y^0|D=1] + (1-\pi) \mathrm{E}[Y^0|D=0] \}=ATE
\end{align}
$$

then

$$
\mathrm{E}[Y^1|D=1] - \mathrm{E}[Y^0|D=0]=ATE
$$


```{r, message=F, warning=F}
# load packages
pacman::p_load(tidyverse, here, sandwich, lmtest, boot, ranger,
               ggplot2, broom, SuperLearner, tmle, AIPW,
               ranger, xgboost, e1071, nnet, glmnet, remotes)


library(tlverse)
library(tmle3)
library(sl3)
library(data.table)
```

Data preparation 
```{r}
nhefs <- read_csv("C:\\Users\\hed2\\Downloads\\mlci_shortcourse-main\\mlci_shortcourse-main\\data\\nhefs.csv") %>% 
  mutate(wt_delta = as.numeric(wt82_71>median(wt82_71)))
```

#### Fit a regression model
```{r}
formulaVars <- "qsmk + sex + age + income + sbp + dbp + price71 + tax71 + race"
modelForm0 <- as.formula(paste0("wt_delta ~", formulaVars))
modelForm0

#' This model can be used to quantify a conditionally adjusted 
#' odds ratio with correct standard error
modelOR <- glm(modelForm0,data=nhefs,family = binomial("logit"))

# summary(modelOR)

exp(tidy(modelOR)[2,2])
```

#### G compute and confidence interval
```{r}
# modeling account for any "potential" exposure-covariates interactions
formulaVars <- "sex + age + income + sbp + dbp + price71 + tax71 + race"
modelForm <- as.formula(paste0("wt_delta ~", formulaVars))
modelForm

#' Regress the outcome against the confounders among the unexposed and exposed sub datasets 
model0 <- glm(modelForm,data=subset(nhefs,qsmk==0),family=binomial("logit"))
model1 <- glm(modelForm,data=subset(nhefs,qsmk==1),family=binomial("logit"))

##' Generate predictions for everyone in the sample using the models
mu1 <- predict(model1,newdata=nhefs,type="response")
mu0 <- predict(model0,newdata=nhefs,type="response")

#' Marginally adjusted odds ratio
marg_stand_OR <- (mean(mu1)/mean(1-mu1))/(mean(mu0)/mean(1-mu0))
#' Marginally adjusted risk ratio
marg_stand_RR <- mean(mu1)/mean(mu0)
#' Marginally adjusted risk difference
marg_stand_RD <- mean(mu1)-mean(mu0)

#' Using the bootstrap to obtain confidence intervals 
bootfunc <- function(data,index){
  boot_dat <- data[index,]
  model0 <- glm(modelForm,data=subset(boot_dat,qsmk==0),family=binomial("logit"))
  model1 <- glm(modelForm,data=subset(boot_dat,qsmk==1),family=binomial("logit"))
  mu1 <- predict(model1,newdata=boot_dat,type="response")
  mu0 <- predict(model0,newdata=boot_dat,type="response")
  
  marg_stand_OR_ <- (mean(mu1)/mean(1-mu1))/(mean(mu0)/mean(1-mu0))
  marg_stand_RR_ <- mean(mu1)/mean(mu0)
  marg_stand_RD_ <- mean(mu1)-mean(mu0)
  res <- c(marg_stand_RD_,marg_stand_RR_,marg_stand_OR_)
  return(res)
}

#' Run the boot function. Set a seed to obtain reproducibility
set.seed(123)
boot_res <- boot(nhefs,bootfunc,R=2000)

boot_RD <- boot.ci(boot_res,index=1)
boot_RR <- boot.ci(boot_res,index=2)
boot_OR <- boot.ci(boot_res,index=3)

marg_stand_OR
# marg_stand_RD
```

#### Inverse probability weight

```{r}
# create the propensity score in the dataset
nhefs$propensity_score <- glm(qsmk ~ sex + age + income + sbp + dbp + price71 + tax71 + race, data = nhefs, family = binomial("logit"))$fitted.values

# stabilized inverse probability weights
nhefs$sw <- (mean(nhefs$qsmk)/nhefs$propensity_score)*nhefs$qsmk + 
  ((1-mean(nhefs$qsmk))/(1-nhefs$propensity_score))*(1-nhefs$qsmk)

# summary(nhefs$sw)

# F----------------------------------------------------------------------------

model_RD_weighted <- glm(wt_delta ~ qsmk, data = nhefs, weights=sw, family = quasibinomial("identity"))

# summary(model_RD_weighted)$coefficients

exp(summary(model_RD_weighted)$coefficients[2])

# t test----------------------------------------------------------------------------
library(lmtest)
library(sandwich)
# coeftest(model_RD_weighted, vcov. = vcovHC)
```

#### Double robust estimation- 1
```{r}
# using the sl3 function create the super learner, that contains a simple
# average estimator and a glm estimator.
lrnr_glm <- make_learner(Lrnr_glm)
lrnr_mean <- make_learner(Lrnr_mean)
sl <- Lrnr_sl$new(learners = list(lrnr_glm,lrnr_mean))

learner_list <- list(Y = sl, A = sl)

# PREPARE THE THINGS WE WANT TO FEED IN TO TMLE3
# ate_spec defines the exposure levels we want 
# to compare
ate_spec <- tmle_ATE(treatment_level = 1, control_level = 0)

# nodes define the variables we want to adjsut for, and the exposure and the outcome
nodes <- list(W = c("sex", "age", "income", "sbp", "dbp", "price71", "tax71", "race"), 
              A = "qsmk", 
              Y = "wt_delta")

# Run the tmle3 function using the pieces defined above 
tmle_fit <- tmle3(ate_spec, nhefs, nodes, learner_list)

exp(tmle_fit$summary$psi_transformed)

# tmle_fit$summary$lower_transformed
# 
# tmle_fit$summary$upper_transformed
```

#### Double robust estimation- 2
```{r}
# using the SuperLearner function with simple average estimator and a glm estimator.
sl <- c("SL.mean","SL.glm")

outcome <- nhefs$wt_delta
exposure <- nhefs$qsmk
covariates <- nhefs[,c("sex", "age", "income", "sbp", "dbp", "price71", "tax71", "race")]

AIPW_SL <- AIPW$new(Y = outcome,
                    A = exposure,
                    W = covariates,
                    Q.SL.library = sl,
                    g.SL.library = sl,
                    k_split = 3,
                    verbose=FALSE)$
  fit()$
  summary(g.bound = 0.025)$ 
  plot.p_score()$
  plot.ip_weights()

exp(AIPW_SL$result[3,1])
```

#### Non parameter- association 
```{r}
# random Forrest
#' Marginal Standardization with Random Forest
model0 <- ranger(modelForm0, num.trees=500, mtry=3, min.node.size = 50, data=(nhefs))

nhefs$mu0 <- predict(model0, data=nhefs, type="response")$pred

marg_stand_OR_reg <- mean(nhefs$mu0[nhefs$qsmk==1])/mean(1-nhefs$mu0[nhefs$qsmk==1])/( mean(nhefs$mu0[nhefs$qsmk==0])/mean(1-nhefs$mu0[nhefs$qsmk==0]))

marg_stand_OR_reg
```

`Non parameter- potential outcome`
```{r}
# random Forrest
#' Marginal Standardization with Random Forest
# model0 <- ranger(modelForm, num.trees=500, mtry=3, min.node.size = 50, data=subset(nhefs, qsmk==0))
# model1 <- ranger(modelForm, num.trees=500, mtry=3, min.node.size = 50, data=subset(nhefs, qsmk==1))
# 
# mu1 <- predict(model1, data=nhefs, type="response")$pred
# mu0 <- predict(model0, data=nhefs, type="response")$pred
# 
# marg_stand_RDrf <- mean(mu1) - mean(mu0)

# bootfunc <- function(data,index){
#   boot_dat <- data[index,]
#   model0 <- ranger(modelForm, num.trees=500, mtry=5, data=subset(boot_dat, qsmk==0))
#   model1 <- ranger(modelForm, num.trees=500, mtry=5, data=subset(boot_dat, qsmk==1))
#   mu1 <- predict(model1,data=boot_dat,type="response")$pred
#   mu0 <- predict(model0,data=boot_dat,type="response")$pred
#   
#   marg_stand_RD_ <- mean(mu1)-mean(mu0)
#   return(marg_stand_RD_)
# }
# 
# #' Run the boot function.  
# set.seed(123)
# boot_res <- boot(nhefs,bootfunc,R=200)
# boot_RDrf <- boot.ci(boot_res)

# boot_RDrf$bca
# marg_stand_RDrf 

```




