---
title: "Causal inference and associated regression"
description: "A practical R learning guide to Causal inference and associated regression, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Epidemiology & Real-World Evidence"
category: "Causal Inference"
topic: "Causal inference and associated regression"
type: article
tags: ["Causal Inference", "Epidemiology & Real-World Evidence", "Causal", "inference", "associated"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Causal inference defines the contrast of interest and the assumptions needed to learn it from observed data.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Causal inference and associated regression**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Causal inference and association regression
2. Why RCT is gold standard
3. Fit a regression model
4. G compute and confidence interval
5. Inverse probability weight
6. Double robust estimation- 1

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Causal inference and associated regression** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Causal inference and association regression

<!-- generated-learning-note -->
This section develops **Causal inference and association regression**. Read the model specification together with its assumptions and the scale of each reported effect.


- Still need to consider the chronological order in design;
- Well designed observation study is suitable for causal inference;
- Generally associated analysis estimate is larger or smaller than the causal inference; 
- Associated analysis is used for strata and it is hard to deal with interaction (with exposure) and noncollapsibility so CI should be recommended;
- Data imputation is also needed;
- Double robust estimate is more robust;
- Bootstrapping sampling should be more than 80% with more than 1000 samples and the statistic is not necessarily normal distribution;
- Categorizing should be based on the special knowledge not quartile/tertile/ ROC/ ;
- Ltmle longitudinal study should be less than 30 times measurements; 
- Nonparameter method has fewer assumptions but is computation resource consuming (sometimes less precise); 
- The distribution will be changed after normalization;

#### Why RCT is gold standard

<!-- generated-learning-note -->
This subsection explains **Why RCT is gold standard** and shows how it contributes to the overall Causal inference and associated regression workflow.


In a randomized control trial (RCT) we assume [that](https://stats.stackexchange.com/questions/308397/why-is-average-treatment-effect-different-from-average-treatment-effect-on-the-t). 

$$
\mathrm{E}[Y^0|D=1] = \mathrm{E}[Y^0|D=0]
$$
and
$$
\mathrm{E}[Y^1|D=1] = \mathrm{E}[Y^1|D=0]
$$


Since
$$
\begin{align}
\mathrm{E}[\delta] =& \{ \pi \mathrm{E}[Y^1|D=1] + (1-\pi) \mathrm{E}[Y^1|D=0] \} \\ -& \{ \pi \mathrm{E}[Y^0|D=1] + (1-\pi) \mathrm{E}[Y^0|D=0] \}=ATE
\end{align}
$$

then

$$
\mathrm{E}[Y^1|D=1] - \mathrm{E}[Y^0|D=0]=ATE
$$

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r, message=F, warning=F}
# load packages
pacman::p_load(tidyverse, here, sandwich, lmtest, boot, ranger,
               ggplot2, broom, SuperLearner, tmle, AIPW,
               ranger, xgboost, e1071, nnet, glmnet, remotes)


library(tlverse)
library(tmle3)
library(sl3)
library(data.table)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


Data preparation 

<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r}
#| eval: false
#| code-fold: true
nhefs <- read_csv("C:\\Users\\hed2\\Downloads\\mlci_shortcourse-main\\mlci_shortcourse-main\\data\\nhefs.csv") %>% 
  mutate(wt_delta = as.numeric(wt82_71>median(wt82_71)))
```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.


#### Fit a regression model

<!-- generated-learning-note -->
This section develops **Fit a regression model**. Read the model specification together with its assumptions and the scale of each reported effect.


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
formulaVars <- "qsmk + sex + age + income + sbp + dbp + price71 + tax71 + race"
modelForm0 <- as.formula(paste0("wt_delta ~", formulaVars))
modelForm0

#' This model can be used to quantify a conditionally adjusted 
#' odds ratio with correct standard error
modelOR <- glm(modelForm0,data=nhefs,family = binomial("logit"))

# summary(modelOR)

exp(tidy(modelOR)[2,2])
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


#### G compute and confidence interval

<!-- generated-learning-note -->
This subsection explains **G compute and confidence interval** and shows how it contributes to the overall Causal inference and associated regression workflow.


<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r}
# modeling account for any "potential" exposure-covariates interactions
formulaVars <- "sex + age + income + sbp + dbp + price71 + tax71 + race"
modelForm <- as.formula(paste0("wt_delta ~", formulaVars))
modelForm

#' Regress the outcome against the confounders among the unexposed and exposed sub datasets 
model0 <- glm(modelForm,data=subset(nhefs,qsmk==0),family=binomial("logit"))
model1 <- glm(modelForm,data=subset(nhefs,qsmk==1),family=binomial("logit"))

##' Generate predictions for everyone in the sample using the models
mu1 <- predict(model1,newdata=nhefs,type="response")
mu0 <- predict(model0,newdata=nhefs,type="response")

#' Marginally adjusted odds ratio
marg_stand_OR <- (mean(mu1)/mean(1-mu1))/(mean(mu0)/mean(1-mu0))
#' Marginally adjusted risk ratio
marg_stand_RR <- mean(mu1)/mean(mu0)
#' Marginally adjusted risk difference
marg_stand_RD <- mean(mu1)-mean(mu0)

#' Using the bootstrap to obtain confidence intervals 
bootfunc <- function(data,index){
  boot_dat <- data[index,]
  model0 <- glm(modelForm,data=subset(boot_dat,qsmk==0),family=binomial("logit"))
  model1 <- glm(modelForm,data=subset(boot_dat,qsmk==1),family=binomial("logit"))
  mu1 <- predict(model1,newdata=boot_dat,type="response")
  mu0 <- predict(model0,newdata=boot_dat,type="response")
  
  marg_stand_OR_ <- (mean(mu1)/mean(1-mu1))/(mean(mu0)/mean(1-mu0))
  marg_stand_RR_ <- mean(mu1)/mean(mu0)
  marg_stand_RD_ <- mean(mu1)-mean(mu0)
  res <- c(marg_stand_RD_,marg_stand_RR_,marg_stand_OR_)
  return(res)
}

#' Run the boot function. Set a seed to obtain reproducibility
set.seed(123)
boot_res <- boot(nhefs,bootfunc,R=2000)

boot_RD <- boot.ci(boot_res,index=1)
boot_RR <- boot.ci(boot_res,index=2)
boot_OR <- boot.ci(boot_res,index=3)

marg_stand_OR
# marg_stand_RD
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.


#### Inverse probability weight

<!-- generated-learning-note -->
This subsection explains **Inverse probability weight** and shows how it contributes to the overall Causal inference and associated regression workflow.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# create the propensity score in the dataset
nhefs$propensity_score <- glm(qsmk ~ sex + age + income + sbp + dbp + price71 + tax71 + race, data = nhefs, family = binomial("logit"))$fitted.values

# stabilized inverse probability weights
nhefs$sw <- (mean(nhefs$qsmk)/nhefs$propensity_score)*nhefs$qsmk + 
  ((1-mean(nhefs$qsmk))/(1-nhefs$propensity_score))*(1-nhefs$qsmk)

# summary(nhefs$sw)

# F----------------------------------------------------------------------------

model_RD_weighted <- glm(wt_delta ~ qsmk, data = nhefs, weights=sw, family = quasibinomial("identity"))

# summary(model_RD_weighted)$coefficients

exp(summary(model_RD_weighted)$coefficients[2])

# t test----------------------------------------------------------------------------
library(lmtest)
library(sandwich)
# coeftest(model_RD_weighted, vcov. = vcovHC)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Double robust estimation- 1

<!-- generated-learning-note -->
This subsection explains **Double robust estimation- 1** and shows how it contributes to the overall Causal inference and associated regression workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# using the sl3 function create the super learner, that contains a simple
# average estimator and a glm estimator.
lrnr_glm <- make_learner(Lrnr_glm)
lrnr_mean <- make_learner(Lrnr_mean)
sl <- Lrnr_sl$new(learners = list(lrnr_glm,lrnr_mean))

learner_list <- list(Y = sl, A = sl)

# PREPARE THE THINGS WE WANT TO FEED IN TO TMLE3
# ate_spec defines the exposure levels we want 
# to compare
ate_spec <- tmle_ATE(treatment_level = 1, control_level = 0)

# nodes define the variables we want to adjsut for, and the exposure and the outcome
nodes <- list(W = c("sex", "age", "income", "sbp", "dbp", "price71", "tax71", "race"), 
              A = "qsmk", 
              Y = "wt_delta")

# Run the tmle3 function using the pieces defined above 
tmle_fit <- tmle3(ate_spec, nhefs, nodes, learner_list)

exp(tmle_fit$summary$psi_transformed)

# tmle_fit$summary$lower_transformed
# 
# tmle_fit$summary$upper_transformed
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Double robust estimation- 2

<!-- generated-learning-note -->
This subsection explains **Double robust estimation- 2** and shows how it contributes to the overall Causal inference and associated regression workflow.


<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# using the SuperLearner function with simple average estimator and a glm estimator.
sl <- c("SL.mean","SL.glm")

outcome <- nhefs$wt_delta
exposure <- nhefs$qsmk
covariates <- nhefs[,c("sex", "age", "income", "sbp", "dbp", "price71", "tax71", "race")]

AIPW_SL <- AIPW$new(Y = outcome,
                    A = exposure,
                    W = covariates,
                    Q.SL.library = sl,
                    g.SL.library = sl,
                    k_split = 3,
                    verbose=FALSE)$
  fit()$
  summary(g.bound = 0.025)$ 
  plot.p_score()$
  plot.ip_weights()

exp(AIPW_SL$result[3,1])
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


#### Non parameter- association

<!-- generated-learning-note -->
This subsection explains **Non parameter- association** and shows how it contributes to the overall Causal inference and associated regression workflow.


<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# random Forrest
#' Marginal Standardization with Random Forest
model0 <- ranger(modelForm0, num.trees=500, mtry=3, min.node.size = 50, data=(nhefs))

nhefs$mu0 <- predict(model0, data=nhefs, type="response")$pred

marg_stand_OR_reg <- mean(nhefs$mu0[nhefs$qsmk==1])/mean(1-nhefs$mu0[nhefs$qsmk==1])/( mean(nhefs$mu0[nhefs$qsmk==0])/mean(1-nhefs$mu0[nhefs$qsmk==0]))

marg_stand_OR_reg
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


`Non parameter- potential outcome`

<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r}
# random Forrest
#' Marginal Standardization with Random Forest
# model0 <- ranger(modelForm, num.trees=500, mtry=3, min.node.size = 50, data=subset(nhefs, qsmk==0))
# model1 <- ranger(modelForm, num.trees=500, mtry=3, min.node.size = 50, data=subset(nhefs, qsmk==1))
# 
# mu1 <- predict(model1, data=nhefs, type="response")$pred
# mu0 <- predict(model0, data=nhefs, type="response")$pred
# 
# marg_stand_RDrf <- mean(mu1) - mean(mu0)

# bootfunc <- function(data,index){
#   boot_dat <- data[index,]
#   model0 <- ranger(modelForm, num.trees=500, mtry=5, data=subset(boot_dat, qsmk==0))
#   model1 <- ranger(modelForm, num.trees=500, mtry=5, data=subset(boot_dat, qsmk==1))
#   mu1 <- predict(model1,data=boot_dat,type="response")$pred
#   mu0 <- predict(model0,data=boot_dat,type="response")$pred
#   
#   marg_stand_RD_ <- mean(mu1)-mean(mu0)
#   return(marg_stand_RD_)
# }
# 
# #' Run the boot function.  
# set.seed(123)
# boot_res <- boot(nhefs,bootfunc,R=200)
# boot_RDrf <- boot.ci(boot_res)

# boot_RDrf$bca
# marg_stand_RDrf 

```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.

## Interpreting the Results

State the target estimand, causal structure, exchangeability assumptions, positivity, and consistency before fitting a model. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
