---
title: "Principle components analysis"
description: "A practical R learning guide to Principle components analysis, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Statistical Models"
topic: "Principle components analysis"
type: article
tags: ["Statistical Models", "Biostatistics", "Principle", "components"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

A statistical model is a structured approximation that links observed data to an estimand or prediction target.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Principle components analysis**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Principle components analysis
2. Load the data
3. Calculate principal components
4. Visualize the Results
5. Find Variance Explained by Each Principal Component
6. Applications:

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Statistical Models",
  topic = "Principle components analysis"
)
learning_overview
```

## Learning Material

The following material develops **Principle components analysis** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Principle components analysis

<!-- generated-learning-note -->
This subsection explains **Principle components analysis** and shows how it contributes to the overall Principle components analysis workflow.


#### Load the data

<!-- generated-learning-note -->
This section prepares the information needed for **Load the data**. Check variable definitions, units, coding, and missing values before moving to analysis.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(tidyverse)
#load data
data("iris")
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Calculate principal components

<!-- generated-learning-note -->
This subsection explains **Calculate principal components** and shows how it contributes to the overall Principle components analysis workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
#calculate principal components
results <- prcomp(iris[,colnames(iris)[1:4]], scale = TRUE)

#reverse the signs
results$rotation <- -1*results$rotation

#display principal components
results$rotation
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
#reverse the signs of the scores
results$x <- -1*results$x

#display the first six scores
head(results$x)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.




#### Visualize the Results

<!-- generated-learning-note -->
This section focuses on **Visualize the Results**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
biplot(results, scale = 0)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
data_pot=data.frame(iris,results$x)

library(ggplot2)
# Scatter plot by group, upPCA
ggplot(data_pot, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
# Scatter plot by group, PCA
ggplot(data_pot, aes(x = PC1, y = PC2, color = Species)) +
  geom_point()
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
data_spedal=data.frame(results$rotation)
# Scatter plot 
ggplot(data_spedal, aes(x = PC1, y = PC2)) +
  geom_point()
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



#### Find Variance Explained by Each Principal Component

<!-- generated-learning-note -->
This subsection explains **Find Variance Explained by Each Principal Component** and shows how it contributes to the overall Principle components analysis workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
#calculate total variance explained by each principal component
results$sdev^2 / sum(results$sdev^2)

#calculate total variance explained by each principal component
var_explained = results$sdev^2 / sum(results$sdev^2)

#create scree plot
qplot(c(1:4), var_explained) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Applications

<!-- generated-learning-note -->
This subsection explains **Applications** and shows how it contributes to the overall Principle components analysis workflow.

- reduce dimensions and biplot, then understand which observations are similar to each other.  
- principal components regression to prevent multicolinearity

## Interpreting the Results

Interpret coefficients with their coding and scale, examine uncertainty, and use diagnostics to evaluate assumptions. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
