---
title: "Expectation maximum"
output: html_document
date: '2023-05-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The parameters that maximize the likelihood function, as an algebraic function of the data - difficult or impossible. A common example is missing data; other examples include 'hidden' or 'latent' variables.



[See reference](http://rstudio-pubs-static.s3.amazonaws.com/1001_3177e85f5e4840be840c84452780db52.html)

Steps: 

- 1 Expectation Step: Given the current estimates of the parameters θ^
   , we take the expected value of the labels given the data.

- 2 Maximization Step: We plug back these expected values into the complete-data likelihood and maximize over the parameters θ^ .

- The EM algorithm is an iterative method for approximating the maximum of a likelihood function.

`A typical example is that of mixture distributions, whereby we have a set of distributions, each observation comes from one of these distributions with some probability, but we only observe the value of each observation, and not what distribution it came from.`

Generate a simulation data

```{r}
# whether we arce drawing from distribution 1 or 2 - ie, Bernoulli trials.
set.seed(123)
tau_1_true <- 0.25
x <- y <- rep(0,1000)
for( i in 1:1000 ) {
  if( runif(1) < tau_1_true ) {
    x[i] <- rnorm(1, mean=1)
    y[i] <- "heads"
  } else {
    x[i] <- rnorm(1, mean=7)
    y[i] <- "tails"
  }
}

library(lattice)
densityplot( ~x, par.settings = list( plot.symbol=list( col=as.factor(y) ) ) )


```

EM by hand

```{r}
############## iterations
# Now, we simply repeat the following process, until the difference between our successive estimates of the mean are very small.
# We'll run 10 iterations.
 
## set the initial guesses for the distribution parameters (arbitrarily, p1,2 and means1,2) 
mu_1 <- 0
mu_2 <- 1

## as well as the latent variable parameters
tau_1 <- 0.5
tau_2 <- 0.5

for( i in 1:10 ) {

  ## Given the observed data, as well as the distribution parameters,
  ## what are the latent variables?

  T_1 <- tau_1 * dnorm( x, mu_1 )
  T_2 <- tau_2 * dnorm( x, mu_2 )
  
  # updated probability
  P_1 <- T_1 / (T_1 + T_2)
  P_2 <- T_2 / (T_1 + T_2)  

  tau_1 <- mean(P_1)
  tau_2 <- mean(P_2)

  ## Given the observed data, as well as the latent variables,
  ## what are the population parameters?

  mu_1 <- sum( P_1 * x ) / sum(P_1)
  mu_2 <- sum( P_2 * x ) / sum(P_2)

  ## print the current estimates

  print( c(mu_1, mu_2, mean(P_1), mean(P_2)) )

  # converge quickly, after just one iteration
}
```

EM by package
```{r}
########### using the R package mixtools to apply the EM algorithm
library("mixtools")
myEM <- normalmixEM( x, mu = c(0,1), sigma=c(1,1), sd.constr=c(1,1) )

myEM$mu

myEM$lambda

# What if the guesses initials aren't so similar?
# The algorithm converges much more slowly, but still gives good estimates.

########### using the R package mixtools to apply the EM algorithm
library("mixtools")
myEM <- normalmixEM( x, mu = c(0,10), sigma=c(1,1), sd.constr=c(1,1) )

myEM$mu

myEM$lambda
```

