---
title: "Expectation maximum"
description: "A practical R learning guide to Expectation maximum, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Statistical Principles & Estimands"
topic: "Expectation maximum"
type: article
tags: ["Statistical Principles & Estimands", "Biostatistics", "Expectation", "maximum"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Statistical principles define what is being estimated, under which assumptions, and how uncertainty should be represented.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Expectation maximum**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Introduction

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Expectation maximum** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Introduction

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **Introduction** in the context of Expectation maximum.


The parameters that maximize the likelihood function, as an algebraic function of the data - difficult or impossible. A common example is missing data; other examples include 'hidden' or 'latent' variables.



[See reference](http://rstudio-pubs-static.s3.amazonaws.com/1001_3177e85f5e4840be840c84452780db52.html)

Steps: 

- 1 Expectation Step: Given the current estimates of the parameters θ^
   , we take the expected value of the labels given the data.

- 2 Maximization Step: We plug back these expected values into the complete-data likelihood and maximize over the parameters θ^ .

- The EM algorithm is an iterative method for approximating the maximum of a likelihood function.

`A typical example is that of mixture distributions, whereby we have a set of distributions, each observation comes from one of these distributions with some probability, but we only observe the value of each observation, and not what distribution it came from.`

Generate a simulation data

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# whether we arce drawing from distribution 1 or 2 - ie, Bernoulli trials.
set.seed(123)
tau_1_true <- 0.25
x <- y <- rep(0,1000)
for( i in 1:1000 ) {
  if( runif(1) < tau_1_true ) {
    x[i] <- rnorm(1, mean=1)
    y[i] <- "heads"
  } else {
    x[i] <- rnorm(1, mean=7)
    y[i] <- "tails"
  }
}

library(lattice)
densityplot( ~x, par.settings = list( plot.symbol=list( col=as.factor(y) ) ) )


```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


EM by hand


<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
############## iterations
# Now, we simply repeat the following process, until the difference between our successive estimates of the mean are very small.
# We'll run 10 iterations.
 
## set the initial guesses for the distribution parameters (arbitrarily, p1,2 and means1,2) 
mu_1 <- 0
mu_2 <- 1

## as well as the latent variable parameters
tau_1 <- 0.5
tau_2 <- 0.5

for( i in 1:10 ) {

  ## Given the observed data, as well as the distribution parameters,
  ## what are the latent variables?

  T_1 <- tau_1 * dnorm( x, mu_1 )
  T_2 <- tau_2 * dnorm( x, mu_2 )
  
  # updated probability
  P_1 <- T_1 / (T_1 + T_2)
  P_2 <- T_2 / (T_1 + T_2)  

  tau_1 <- mean(P_1)
  tau_2 <- mean(P_2)

  ## Given the observed data, as well as the latent variables,
  ## what are the population parameters?

  mu_1 <- sum( P_1 * x ) / sum(P_1)
  mu_2 <- sum( P_2 * x ) / sum(P_2)

  ## print the current estimates

  print( c(mu_1, mu_2, mean(P_1), mean(P_2)) )

  # converge quickly, after just one iteration
}
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.


EM by package

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
########### using the R package mixtools to apply the EM algorithm
library("mixtools")
myEM <- normalmixEM( x, mu = c(0,1), sigma=c(1,1), sd.constr=c(1,1) )

myEM$mu

myEM$lambda

# What if the guesses initials aren't so similar?
# The algorithm converges much more slowly, but still gives good estimates.

########### using the R package mixtools to apply the EM algorithm
library("mixtools")
myEM <- normalmixEM( x, mu = c(0,10), sigma=c(1,1), sd.constr=c(1,1) )

myEM$mu

myEM$lambda
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

## Interpreting the Results

Link every numerical result to its target quantity, assumptions, and sampling interpretation. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
