---
title: "Standard Operating Procedure for Epidemiologic Survival Analysis"
author: "[Author / Study Team]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Purpose

This SOP provides a standard and reusable framework for survival analysis in observational epidemiologic studies, including prospective and retrospective cohort studies, registry studies, EHR studies, claims-based studies, and other real-world data studies.

The objective is to estimate an exposure-outcome association or effect while appropriately addressing time zero, follow-up, censoring, confounding, missing data, model assumptions, and sensitivity to alternative analytic assumptions.

# 2. Research Question and Target Effect

The primary research question should be stated before analysis.

**General form:**

> Among [target population], is [exposure] associated with or causally related to the risk of [time-to-event outcome] during [follow-up period]?

Prespecify:

- target population,
- exposure and reference condition,
- outcome,
- time zero,
- follow-up period,
- censoring rules,
- primary effect measure,
- and covariate-adjustment strategy.

The primary relative effect measure will usually be the **hazard ratio (HR)** with a **95% confidence interval (CI)**.

If causal interpretation is intended, specify the target effect, such as a total effect, conditional effect, or marginal population-level effect.

# 3. Study Design and Data Source

Describe:

- study design,
- source population,
- data source,
- study period,
- data availability,
- and relevant linkage procedures.

Common sources include population cohorts, epidemiologic surveys, EHR databases, claims databases, disease registries, and linked administrative data.

# 4. Study Population and Analytic Cohort

Define inclusion and exclusion criteria before outcome modeling.

Typical criteria may include:

- eligible age range,
- availability of exposure information,
- availability of baseline covariates,
- absence of the study outcome at baseline when incident disease is studied,
- and eligibility for follow-up.

Document cohort derivation using a flow diagram when appropriate.

# 5. Exposure, Outcome, Time Zero, Follow-up, and Censoring

## 5.1 Exposure

Specify:

- exposure definition,
- data source,
- ascertainment window,
- reference category,
- coding,
- and whether exposure is fixed or time-varying.

## 5.2 Outcome

Specify:

- event definition,
- event date,
- source of outcome information,
- validation/adjudication method if applicable,
- and whether the analysis is time-to-first-event or recurrent-event.

## 5.3 Time Zero

Time zero is the date at which follow-up begins.

Exposure status, eligibility criteria, and baseline confounders should be aligned relative to time zero.

Incorrect alignment may introduce immortal-time bias, prevalent-user bias, reverse temporality, or other selection bias.

## 5.4 Follow-up and Censoring

Follow-up will end at the earliest of:

1. occurrence of the outcome,
2. death when not part of the outcome,
3. loss to follow-up,
4. disenrollment or loss of data availability,
5. administrative study end,
6. or another prespecified censoring event.

Censoring rules should be defined before the primary analysis.

# 6. Confounder Selection and Covariate Adjustment

Potential confounders should be identified using:

- subject-matter knowledge,
- prior epidemiologic evidence,
- temporal ordering,
- and causal reasoning or DAGs.

A variable should not be included solely because it is statistically significant, differs between exposure groups, or improves model fit.

The primary covariate set should preferably be prespecified.

## 6.1 Baseline Confounders

Typical baseline confounders may include age, sex, smoking, BMI, blood pressure, diabetes, hypertension, cholesterol, physical activity, socioeconomic factors, healthcare utilization, and relevant comorbidities.

## 6.2 Post-Exposure Variables

Variables occurring after exposure should not automatically be included in the primary adjustment model.

If the objective is the **total effect** of the exposure, mediators or downstream variables should generally not be adjusted for.

For example:

\[
Diabetes \rightarrow Diabetes\ Treatment \rightarrow CVD
\]

If the study objective is the total effect of diabetes on CVD, post-diabetes treatment would generally not be included as a baseline confounder.

If the scientific objective concerns an effect under a fixed or hypothetical treatment condition, that target effect should be defined separately and may require methods beyond ordinary regression adjustment.

# 7. Data Quality Assessment

Before modeling, assess:

- duplicate participants,
- missing data,
- implausible values,
- inconsistent dates,
- temporal inconsistencies,
- exposure misclassification,
- outcome misclassification,
- completeness of follow-up,
- and number of observed events.

```{r data-quality-example, eval=FALSE}
sum(duplicated(analysis_data$id))
colSums(is.na(analysis_data))
summary(analysis_data)
```

# 8. Descriptive Analysis

Baseline characteristics will be summarized by exposure group.

Continuous variables:

- mean and standard deviation, or
- median and interquartile range.

Categorical variables:

- number and percentage.

Standardized mean differences (SMDs) may be reported to describe baseline imbalance.

An absolute SMD below approximately 0.10 may indicate acceptable measured balance, but SMD should not determine whether a variable is a true confounder.

Also summarize:

- number of participants,
- number of events,
- person-time,
- median follow-up,
- and crude incidence rate.

# 9. Kaplan-Meier Analysis

Kaplan-Meier curves will be used to describe event-free survival by exposure group when standard survival methods are appropriate.

Report, when useful:

- survival probabilities at prespecified time points,
- median survival if estimable,
- number at risk,
- and 95% CIs.

The log-rank test may be reported descriptively, but the adjusted survival model will generally serve as the primary inferential analysis.

```{r km-example, eval=FALSE}
library(survival)

km_fit <- survfit(
  Surv(followup_time, event) ~ exposure,
  data = analysis_data
)

plot(km_fit)
```

# 10. Primary Cox Proportional Hazards Analysis

## 10.1 Crude Model

First estimate the unadjusted exposure-outcome association.

```{r crude-cox, eval=FALSE}
fit_crude <- coxph(
  Surv(followup_time, event) ~ exposure,
  data = analysis_data
)

summary(fit_crude)
```

## 10.2 Adjusted Model

The primary model will adjust for the prespecified confounder set.

```{r adjusted-cox, eval=FALSE}
fit_primary <- coxph(
  Surv(followup_time, event) ~
    exposure +
    age +
    sex +
    bmi +
    smoking +
    hypertension,
  data = analysis_data
)

summary(fit_primary)
```

Report the exposure HR, 95% CI, and P-value when appropriate.

```{r hr-ci, eval=FALSE}
exp(
  cbind(
    HR = coef(fit_primary),
    confint(fit_primary)
  )
)
```

## 10.3 Conditional vs Marginal Effect

A Cox regression coefficient is generally a **conditional hazard ratio**, conditional on the covariates included in the model.

If the scientific objective is instead a **marginal population-level effect**, methods such as standardization, g-computation, or propensity-score weighting may be more appropriate.

# 11. Model Assumptions and Diagnostics

## 11.1 Proportional Hazards Assumption

Evaluate the proportional hazards assumption using:

- Schoenfeld residuals,
- `cox.zph()`,
- graphical assessment,
- and scientific judgment.

```{r ph-check, eval=FALSE}
ph_test <- cox.zph(fit_primary)
ph_test
plot(ph_test)
```

If substantial non-proportional hazards are present, consider:

- time-varying coefficients,
- exposure-by-time interaction,
- stratified Cox models,
- time-specific HRs,
- or restricted mean survival time (RMST).

## 11.2 Functional Form of Continuous Covariates

Evaluate the functional form of continuous variables when appropriate.

Possible approaches include restricted cubic splines, fractional polynomials, or clinically justified transformations.

Arbitrary categorization of continuous variables should generally be avoided.

## 11.3 Model Stability

Assess when relevant:

- sparse categories,
- influential observations,
- multicollinearity,
- convergence,
- and adequacy of the number of events relative to model complexity.

# 12. Missing Data

Summarize missingness overall, by exposure group, and for key covariates.

The primary missing-data strategy should be prespecified.

Common approaches include:

- complete-case analysis when missingness is minimal and defensible,
- multiple imputation,
- or another justified method.

If multiple imputation is used, the imputation model should include variables related to exposure, outcome, missingness, and relevant auxiliary predictors.

Sensitivity analyses should be considered when missing-data assumptions may materially influence conclusions.

# 13. Propensity-Score Analysis

Propensity-score methods may be used as a secondary or sensitivity analysis.

The propensity score is the probability of exposure conditional on measured baseline covariates.

Possible methods include:

- IPTW,
- stabilized IPTW,
- matching,
- stratification,
- and overlap weighting.

Assess:

- propensity-score overlap,
- positivity,
- extreme weights,
- post-weighting covariate balance,
- and effective sample size.

```{r ps-example, eval=FALSE}
ps_model <- glm(
  exposure ~ age + sex + bmi + smoking + hypertension,
  family = binomial(),
  data = analysis_data
)

analysis_data$ps <- predict(ps_model, type = "response")
```

For weighted Cox models, robust variance estimation should generally be used.

```{r weighted-cox-example, eval=FALSE}
fit_iptw <- coxph(
  Surv(followup_time, event) ~ exposure,
  data = analysis_data,
  weights = iptw,
  robust = TRUE
)
```

Good measured balance does not rule out residual or unmeasured confounding.

# 14. Competing Risks

If a competing event prevents occurrence of the primary outcome, competing-risk methods should be considered.

Depending on the scientific question:

- **cause-specific Cox regression** may be used for etiologic questions,
- **Fine-Gray subdistribution hazards models** may be used for cumulative-incidence or prognostic questions.

When competing risks are important, cumulative incidence functions should generally be reported.

# 15. Effect Modification and Subgroup Analysis

Potential effect modifiers should be prespecified based on scientific rationale.

Examples include age, sex, disease severity, major comorbidities, or clinically important risk strata.

Effect modification should preferably be evaluated using an interaction term.

```{r interaction-example, eval=FALSE}
fit_interaction <- coxph(
  Surv(followup_time, event) ~
    exposure * sex +
    age +
    bmi +
    smoking,
  data = analysis_data
)
```

Report subgroup-specific HRs, 95% CIs, and interaction P-value when relevant.

Exploratory subgroup findings should be interpreted cautiously.

# 16. Sensitivity Analyses

Sensitivity analyses should evaluate the robustness of the primary estimate to reasonable alternative assumptions.

Common analyses include:

1. alternative covariate adjustment sets;
2. alternative exposure definitions;
3. alternative outcome definitions;
4. lagged exposure analyses;
5. restriction of the analytic cohort;
6. multiple imputation;
7. propensity-score weighting or matching;
8. weight trimming or truncation;
9. inverse probability of censoring weighting;
10. alternative censoring rules;
11. competing-risk analyses;
12. alternative handling of post-exposure variables;
13. assessment of unmeasured confounding when appropriate.

Primary and sensitivity estimates should be compared with respect to direction, magnitude, precision, and clinical interpretation.

# 17. Statistical Inference

Unless otherwise specified:

- tests will be two-sided,
- alpha = 0.05,
- and 95% confidence intervals will be reported.

The primary analysis should be clearly distinguished from secondary and exploratory analyses.

Formal multiplicity adjustment is not always required in observational epidemiologic studies, but extensive testing across many outcomes, exposures, or subgroups should be acknowledged.

# 18. Tables and Figures

Recommended outputs include:

## Table 1
Baseline characteristics by exposure group.

## Table 2
Number of participants, events, person-time, follow-up, and incidence rates.

## Table 3
Crude and adjusted survival-model estimates.

| Analysis | HR | 95% CI | P-value |
|---|---:|---:|---:|
| Crude Cox | | | |
| Primary adjusted Cox | | | |
| IPTW analysis | | | |
| Sensitivity analysis | | | |

## Figures

Recommended figures include:

1. cohort flow diagram;
2. Kaplan-Meier or cumulative-incidence curves;
3. forest plot of primary and sensitivity analyses;
4. covariate-balance plot when propensity-score methods are used.

# 19. Interpretation and Limitations

Results should be interpreted with emphasis on:

- magnitude of the estimate,
- confidence interval,
- epidemiologic and clinical relevance,
- consistency across sensitivity analyses,
- and underlying assumptions.

An adjusted observational estimate should not automatically be interpreted as causal.

Potential limitations include:

- unmeasured confounding,
- residual confounding,
- exposure misclassification,
- outcome misclassification,
- measurement error,
- uncertain temporal ordering,
- reverse causation,
- selection bias,
- informative censoring,
- missing data,
- positivity or overlap problems,
- model misspecification,
- and limited generalizability.

# 20. Quality Control and Reproducibility

Analysis code should be reproducible and independently reviewed when feasible.

Key variables requiring validation include:

- participant ID,
- exposure,
- index date,
- event indicator,
- event date,
- follow-up time,
- censoring reason,
- and major confounders.

Material deviations from the prespecified analysis should be documented with the original plan, revised method, rationale, timing of the change, and potential impact on interpretation.

# Appendix A. Minimal Analysis Workflow

```{r workflow, eval=FALSE}
library(dplyr)
library(survival)

# Define analytic cohort
analysis_data <- raw_data %>%
  filter(eligible == 1, followup_time >= 0)

# Event summary
analysis_data %>%
  group_by(exposure) %>%
  summarise(
    n = n(),
    events = sum(event == 1, na.rm = TRUE),
    person_time = sum(followup_time, na.rm = TRUE),
    incidence_rate = events / person_time
  )

# Kaplan-Meier
km_fit <- survfit(
  Surv(followup_time, event) ~ exposure,
  data = analysis_data
)

# Crude Cox
fit_crude <- coxph(
  Surv(followup_time, event) ~ exposure,
  data = analysis_data
)

# Adjusted Cox
fit_primary <- coxph(
  Surv(followup_time, event) ~
    exposure + age + sex + bmi + smoking + hypertension,
  data = analysis_data
)

# HR and 95% CI
exp(cbind(HR = coef(fit_primary), confint(fit_primary)))

# PH assumption
ph_test <- cox.zph(fit_primary)
print(ph_test)
plot(ph_test)
```

# Appendix B. Pre-Analysis Checklist

Before final analysis, confirm:

- [ ] Research question is finalized
- [ ] Target population is defined
- [ ] Exposure definition is finalized
- [ ] Reference group is defined
- [ ] Outcome definition is finalized
- [ ] Time zero is defined
- [ ] Follow-up rules are defined
- [ ] Censoring rules are defined
- [ ] Primary effect measure is specified
- [ ] Primary confounder set is prespecified
- [ ] Post-exposure variables are classified
- [ ] Missing-data strategy is specified
- [ ] PH-assumption strategy is specified
- [ ] Competing-risk strategy is specified if relevant
- [ ] Sensitivity analyses are prespecified
- [ ] Subgroup/effect-modification analyses are prespecified
- [ ] Data cutoff date is finalized
- [ ] Final table and figure shells are prepared
