---
title: "Bayesian safety monitoring"
description: "A practical R learning guide to Bayesian safety monitoring, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Clinical Trial Process & Design"
category: "Safety Monitoring"
topic: "Bayesian safety monitoring"
type: article
tags: ["Safety Monitoring", "Clinical Trial Process & Design", "Bayesian", "safety", "monitoring"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Safety monitoring turns accumulating adverse-event data into timely, pre-specified decisions while protecting trial integrity.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Bayesian safety monitoring**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Bayesian safety monitoring
2. Blinded pooled AE data
3. Unblinded pooled AE data by treatment groups

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Bayesian safety monitoring** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Bayesian safety monitoring

<!-- generated-learning-note -->
This subsection explains **Bayesian safety monitoring** and shows how it contributes to the overall Bayesian safety monitoring workflow.


#### Blinded pooled AE data

<!-- generated-learning-note -->
This section prepares the information needed for **Blinded pooled AE data**. Check variable definitions, units, coding, and missing values before moving to analysis.


Input blinded pooled AE data.
Define Bayesian beta prior distribution.
Separate the data by AE preferred term.
Process interim looks in chronological order.
Update the posterior distribution using pooled AE counts and total subjects.
Treat the previous posterior as the prior for the next interim look.
Generate posterior simulations for the pooled AE rate.
Estimate the posterior mean AE rate and 95% credible interval.
Calculate the posterior probability that the pooled AE rate exceeds the safety threshold (rate_threshold = 0.10).
Apply predefined decision rules.
Classify each AE term as:
  NO SIGNAL
WATCH
ALERT
Generate a blinded pooled Bayesian safety monitoring table.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r cars}
# =========================================================
# Blinded / Pooled Bayesian Safety Monitoring
# Aggregated AE Monitoring without Treatment Group
# Beta-Binomial Sequential Updating
# =========================================================

bayes_pooled_safety_monitor <- function(
    pooled_data,
    alpha_prior = 0.5,
    beta_prior  = 0.5,
    rate_threshold = 0.10,
    prob_threshold_watch = 0.80,
    prob_threshold_alert = 0.90,
    nsim = 100000,
    seed = 123
) {
  
  set.seed(seed)
  
  ae_terms <- unique(pooled_data$ae_term)
  looks <- sort(unique(pooled_data$look))
  
  results <- data.frame()
  
  for (ae in ae_terms) {
    
    dat_ae <- pooled_data[pooled_data$ae_term == ae, ]
    
    alpha_post <- alpha_prior
    beta_post  <- beta_prior
    
    for (lk in looks) {
      
      dat_lk <- dat_ae[dat_ae$look == lk, ]
      
      if (nrow(dat_lk) == 0) next
      
      ae_count <- dat_lk$ae_count
      n_total  <- dat_lk$n_total
      
      # Sequential Bayesian updating
      # Previous posterior becomes current prior
      alpha_post <- alpha_post + ae_count
      beta_post  <- beta_post  + n_total - ae_count
      
      # Posterior simulation for pooled AE rate
      p_pooled <- rbeta(nsim, alpha_post, beta_post)
      
      post_mean_rate <- mean(p_pooled)
      ci_rate <- quantile(p_pooled, c(0.025, 0.975))
      
      prob_exceed <- mean(p_pooled > rate_threshold)
      
      signal_level <- ifelse(
        prob_exceed >= prob_threshold_alert,
        "ALERT",
        ifelse(
          prob_exceed >= prob_threshold_watch,
          "WATCH",
          "NO SIGNAL"
        )
      )
      
      ae_cum <- alpha_post - alpha_prior
      n_cum  <- alpha_post + beta_post - alpha_prior - beta_prior
      
      results <- rbind(
        results,
        data.frame(
          look = lk,
          ae_term = ae,
          ae_cum = ae_cum,
          n_cum = n_cum,
          observed_pooled_rate = ae_cum / n_cum,
          posterior_alpha = alpha_post,
          posterior_beta = beta_post,
          posterior_mean_rate = post_mean_rate,
          rate_ci_lower = ci_rate[1],
          rate_ci_upper = ci_rate[2],
          rate_threshold = rate_threshold,
          prob_rate_gt_threshold = prob_exceed,
          signal_level = signal_level
        )
      )
    }
  }
  
  return(results)
}


# =========================================================
# Example blinded pooled safety data
# Each row = one AE term at one interim look
# =========================================================

pooled_data <- data.frame(
  look = c(
    1,1,1,
    2,2,2,
    3,3,3,
    4,4,4
  ),
  
  ae_term = c(
    "Nausea", "Headache", "Liver enzyme increased",
    "Nausea", "Headache", "Liver enzyme increased",
    "Nausea", "Headache", "Liver enzyme increased",
    "Nausea", "Headache", "Liver enzyme increased"
  ),
  
  ae_count = c(
    37, 32, 9,
    28, 27, 12,
    31, 29, 15,
    35, 31, 17
  ),
  
  n_total = c(
    400, 400, 400,
    300, 300, 300,
    300, 300, 300,
    400, 400, 400
  )
)



pooled_results <- bayes_pooled_safety_monitor(
  pooled_data = pooled_data,
  alpha_prior = 0.5,
  beta_prior = 0.5,
  rate_threshold = 0.10,
  prob_threshold_watch = 0.80,
  prob_threshold_alert = 0.90,
  nsim = 100000,
  seed = 123
)

print(pooled_results)
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.




  
 
#### Unblinded pooled AE data by treatment groups

<!-- generated-learning-note -->
This section prepares the information needed for **Unblinded pooled AE data by treatment groups**. Check variable definitions, units, coding, and missing values before moving to analysis.


Input interim safety data for multiple AE preferred terms.
Specify Bayesian beta prior distributions.
Separate data by AE term.
Perform sequential Bayesian updating at each interim look.
Update posterior distributions using AE counts and non-AE counts.
Generate posterior simulations using beta distributions.
Estimate treatment and control AE rates.
Calculate posterior Risk Difference (RD) and Risk Ratio (RR).
Compute posterior probabilities:
  Pr(RD > threshold)
Pr(RR > threshold) (rd_threshold = 0.03, rr_threshold = 1.5), 
Apply predefined Bayesian safety decision rules.
Classify signals as:
  NO SIGNAL
WATCH
ALERT
Generate interim Bayesian safety monitoring tables for DMC/safety review.



<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r   }
# =========================================================
# Industrial-style Bayesian Safety Signal Monitoring
# Multiple AE Preferred Terms
# Beta-Binomial Sequential Updating
# =========================================================

bayes_safety_monitor_industry <- function(
    safety_data,
    alpha_prior = 0.5,
    beta_prior  = 0.5,
    rd_threshold = 0.03,
    rr_threshold = 1.5,
    prob_threshold_watch = 0.80,
    prob_threshold_alert = 0.90,
    nsim = 100000,
    seed = 123
) {
  
  set.seed(seed)
  
  ae_terms <- unique(safety_data$ae_term)
  looks <- sort(unique(safety_data$look))
  
  results <- data.frame()
  
  for (ae in ae_terms) {
    
    dat_ae <- safety_data[safety_data$ae_term == ae, ]
    
    alpha_trt <- alpha_prior
    beta_trt  <- beta_prior
    
    alpha_ctl <- alpha_prior
    beta_ctl  <- beta_prior
    
    for (lk in looks) {
      
      dat_lk <- dat_ae[dat_ae$look == lk, ]
      
      if (nrow(dat_lk) == 0) next
      
      ae_trt <- dat_lk$ae_trt
      n_trt  <- dat_lk$n_trt
      
      ae_ctl <- dat_lk$ae_ctl
      n_ctl  <- dat_lk$n_ctl
      
      # Sequential Bayesian updating
      alpha_trt <- alpha_trt + ae_trt
      beta_trt  <- beta_trt  + n_trt - ae_trt
      
      alpha_ctl <- alpha_ctl + ae_ctl
      beta_ctl  <- beta_ctl  + n_ctl - ae_ctl
      
      # Posterior simulation
      p_trt <- rbeta(nsim, alpha_trt, beta_trt)
      p_ctl <- rbeta(nsim, alpha_ctl, beta_ctl)
      
      rd <- p_trt - p_ctl
      rr <- p_trt / p_ctl
      
      prob_rd_signal <- mean(rd > rd_threshold)
      prob_rr_signal <- mean(rr > rr_threshold)
      
      # Industrial-style traffic light decision
      signal_level <- ifelse(
        prob_rd_signal >= prob_threshold_alert | prob_rr_signal >= prob_threshold_alert,
        "ALERT",
        ifelse(
          prob_rd_signal >= prob_threshold_watch | prob_rr_signal >= prob_threshold_watch,
          "WATCH",
          "NO SIGNAL"
        )
      )
      
      results <- rbind(
        results,
        data.frame(
          look = lk,
          ae_term = ae,
          
          trt_ae_cum = alpha_trt - alpha_prior,
          trt_n_cum  = alpha_trt + beta_trt - alpha_prior - beta_prior,
          trt_rate   = (alpha_trt - alpha_prior) /
            (alpha_trt + beta_trt - alpha_prior - beta_prior),
          
          ctl_ae_cum = alpha_ctl - alpha_prior,
          ctl_n_cum  = alpha_ctl + beta_ctl - alpha_prior - beta_prior,
          ctl_rate   = (alpha_ctl - alpha_prior) /
            (alpha_ctl + beta_ctl - alpha_prior - beta_prior),
          
          post_mean_rd = mean(rd),
          rd_ci_lower  = quantile(rd, 0.025),
          rd_ci_upper  = quantile(rd, 0.975),
          
          post_mean_rr = mean(rr),
          rr_ci_lower  = quantile(rr, 0.025),
          rr_ci_upper  = quantile(rr, 0.975),
          
          prob_rd_gt_threshold = prob_rd_signal,
          prob_rr_gt_threshold = prob_rr_signal,
          
          signal_level = signal_level
        )
      )
    }
  }
  
  return(results)
}


# =========================================================
# Example safety data
# Each row = one AE term at one interim look
# =========================================================

safety_data <- data.frame(
  look = c(
    1,1,1,
    2,2,2,
    3,3,3,
    4,4,4
  ),
  
  ae_term = c(
    "Nausea", "Headache", "Liver enzyme increased",
    "Nausea", "Headache", "Liver enzyme increased",
    "Nausea", "Headache", "Liver enzyme increased",
    "Nausea", "Headache", "Liver enzyme increased"
  ),
  
  ae_trt = c(
    25, 18, 6,
    18, 15, 8,
    20, 16, 10,
    22, 17, 12
  ),
  
  n_trt = c(
    200, 200, 200,
    150, 150, 150,
    150, 150, 150,
    200, 200, 200
  ),
  
  ae_ctl = c(
    12, 14, 3,
    10, 12, 4,
    11, 13, 5,
    13, 14, 5
  ),
  
  n_ctl = c(
    200, 200, 200,
    150, 150, 150,
    150, 150, 150,
    200, 200, 200
  )
)


results <- bayes_safety_monitor_industry(
  safety_data = safety_data,
  alpha_prior = 0.5,
  beta_prior = 0.5,
  rd_threshold = 0.03,
  rr_threshold = 1.5,
  prob_threshold_watch = 0.80,
  prob_threshold_alert = 0.90,
  nsim = 100000,
  seed = 123
)

print(results)
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.


Based on a comprehensive assessment of these factors—including AE severity, SAEs, AESIs, time-to-onset, exposure-adjusted incidence, medical review, multiplicity, and benefit-risk—a judgment is rendered.

## Interpreting the Results

Interpret posterior probabilities or monitoring boundaries together with event counts, exposure, and clinical relevance. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
