---
title: "Bayesian Linear Regression Using Metropolis–Hastings MCMC"
author: "Daniel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# 1. Objective

This example first simulates a simple linear regression dataset and then uses a manually implemented Metropolis–Hastings Markov chain Monte Carlo (MCMC) algorithm to estimate three unknown parameters:

- the slope $a$;
- the intercept $b$;
- the residual standard deviation $\sigma$.

The assumed linear regression model is

$$
y_i = a x_i + b + \varepsilon_i,
$$

where

$$
\varepsilon_i \sim N(0,\sigma^2).
$$

Equivalently,

$$
y_i \mid a,b,\sigma \sim N(a x_i+b,\sigma^2).
$$

The true parameter values used to generate the simulated data are

$$
a=5,\qquad b=0,\qquad \sigma=10.
$$

In a real analysis, these parameters are unknown. They are specified here only to generate simulated data and to evaluate whether the Bayesian estimates recover values close to the truth.

# 2. Set the Random Seed

```{r, eval=FALSE}
set.seed(123)
```

`set.seed(123)` fixes the random-number seed.

This program uses random-number generators such as:

- `rnorm()` for normal random variables;
- `runif()` for uniform random variables.

Without a fixed seed, the simulated data and MCMC results will change every time the code is run.

Using the same seed makes the analysis reproducible and is useful for:

- reproducing results;
- debugging code;
- comparing different methods;
- teaching demonstrations;
- documenting simulation studies.

# 3. Generate Simulated Data

## 3.1 Define the True Parameters

```{r, eval=FALSE}
trueA  <- 5
trueB  <- 0
trueSd <- 10
```

These values define the true data-generating parameters:

$$
a=5,
$$

$$
b=0,
$$

and

$$
\sigma=10.
$$

Therefore, the true data-generating model is

$$
y_i = 5x_i + \varepsilon_i,
$$

where

$$
\varepsilon_i \sim N(0,10^2).
$$

The objects `trueA`, `trueB`, and `trueSd` are used only to simulate data. The Bayesian analysis below estimates the parameters from the simulated observations.

## 3.2 Define the Sample Size

```{r, eval=FALSE}
sampleSize <- 31
```

The sample size is

$$
n=31.
$$

Therefore, the dataset contains 31 values of $x$ and 31 corresponding values of $y$.

## 3.3 Generate the Predictor Values

```{r, eval=FALSE}
x <- (-(sampleSize - 1) / 2):
      ((sampleSize - 1) / 2)

x
```

Because

$$
\frac{31-1}{2}=15,
$$

the predictor values are

$$
-15,-14,-13,\ldots,-1,0,1,\ldots,13,14,15.
$$

These values are symmetric around zero, so

$$
\bar{x}=0.
$$

Centering the predictor around zero has two common advantages:

1. The intercept $b$ represents the expected response when $x=0$.
2. It often reduces posterior correlation between the slope and intercept.

## 3.4 Generate the Outcome Values

```{r, eval=FALSE}
y <- trueA * x +
     trueB +
     rnorm(
       n = sampleSize,
       mean = 0,
       sd = trueSd
     )
```

This code generates the outcome according to

$$
y_i = a x_i+b+\varepsilon_i.
$$

Substituting the true parameter values gives

$$
y_i = 5x_i+0+\varepsilon_i,
$$

where

$$
\varepsilon_i\sim N(0,10^2).
$$

The following code generates 31 independent normal errors:

```{r, eval=FALSE}
rnorm(
  n = sampleSize,
  mean = 0,
  sd = trueSd
)
```

Each outcome therefore consists of a systematic component and a random component:

$$
y_i=\text{linear trend}+\text{random error}.
$$

In this example,

$$
y_i=5x_i+\varepsilon_i.
$$

## 3.5 Plot the Simulated Data

```{r, eval=FALSE}
plot(
  x,
  y,
  main = "Simulated Data",
  xlab = "x",
  ylab = "y",
  pch = 19
)
```

Because the true slope is positive,

$$
a=5,
$$

the scatterplot should show a clear increasing trend.

However, because the residual standard deviation is 10, the observations do not fall exactly on a straight line.

# 4. Define the Log-Likelihood Function

```{r, eval=FALSE}
log_likelihood <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (!is.finite(sigma) || sigma <= 0) {
    return(-Inf)
  }

  predicted_mean <- a * x + b

  log_density <- dnorm(
    y,
    mean = predicted_mean,
    sd = sigma,
    log = TRUE
  )

  return(sum(log_density))
}
```

## 4.1 Input Parameters

The input `param` is a vector containing three parameters:

$$
\text{param}=(a,b,\sigma).
$$

The following lines extract the parameters:

```{r, eval=FALSE}
a     <- param[1]
b     <- param[2]
sigma <- param[3]
```

Thus:

- `param[1]` is the slope $a$;
- `param[2]` is the intercept $b$;
- `param[3]` is the residual standard deviation $\sigma$.

For example,

```{r, eval=FALSE}
param <- c(5, 0, 10)
```

represents

$$
a=5,\qquad b=0,\qquad \sigma=10.
$$

## 4.2 Check Whether the Standard Deviation Is Valid

```{r, eval=FALSE}
if (!is.finite(sigma) || sigma <= 0) {
  return(-Inf)
}
```

A normal standard deviation must satisfy

$$
\sigma>0.
$$

The parameter is invalid if $\sigma$ is:

- less than or equal to zero;
- `NA`;
- `NaN`;
- `Inf`;
- `-Inf`.

In these cases, the function returns

$$
-\infty.
$$

Because

$$
\log(0)=-\infty,
$$

returning `-Inf` tells the MCMC algorithm that this parameter combination has zero probability and should be rejected.

This check prevents `dnorm()` from receiving an invalid standard deviation.

## 4.3 Calculate the Predicted Mean

```{r, eval=FALSE}
predicted_mean <- a * x + b
```

For each observation, the conditional mean is

$$
\mu_i=ax_i+b.
$$

Therefore,

$$
E(y_i\mid a,b,\sigma)=ax_i+b.
$$

`predicted_mean` is a vector of length 31. Its $i$th element is

$$
a x_i+b.
$$

## 4.4 Calculate the Log-Density for Each Observation

```{r, eval=FALSE}
log_density <- dnorm(
  y,
  mean = predicted_mean,
  sd = sigma,
  log = TRUE
)
```

The model assumes

$$
y_i\mid a,b,\sigma\sim N(ax_i+b,\sigma^2).
$$

The normal density for a single observation is

$$
p(y_i\mid a,b,\sigma)
=
\frac{1}{\sqrt{2\pi}\sigma}
\exp\left[
-\frac{(y_i-ax_i-b)^2}{2\sigma^2}
\right].
$$

Taking logarithms gives

$$
\log p(y_i\mid a,b,\sigma)
=
-\frac{1}{2}\log(2\pi)
-\log(\sigma)
-\frac{(y_i-ax_i-b)^2}{2\sigma^2}.
$$

The argument

```{r, eval=FALSE}
log = TRUE
```

instructs R to calculate the log-density directly.

This is numerically more stable than multiplying many small probability densities.

## 4.5 Calculate the Joint Log-Likelihood

```{r, eval=FALSE}
return(sum(log_density))
```

Assuming conditional independence of observations, the joint likelihood is

$$
L(a,b,\sigma)
=
\prod_{i=1}^{n}
p(y_i\mid a,b,\sigma).
$$

Taking logarithms converts the product into a sum:

$$
\log L(a,b,\sigma)
=
\sum_{i=1}^{n}
\log p(y_i\mid a,b,\sigma).
$$

Therefore, `log_likelihood()` returns the joint log-likelihood for the complete dataset.

# 5. Define the Log-Prior Function

```{r, eval=FALSE}
log_prior <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (
    a < 0 ||
    a > 10 ||
    sigma <= 0 ||
    sigma > 30
  ) {
    return(-Inf)
  }

  log_prior_a <- dunif(
    a,
    min = 0,
    max = 10,
    log = TRUE
  )

  log_prior_b <- dnorm(
    b,
    mean = 0,
    sd = 5,
    log = TRUE
  )

  log_prior_sigma <- dunif(
    sigma,
    min = 0,
    max = 30,
    log = TRUE
  )

  return(
    log_prior_a +
    log_prior_b +
    log_prior_sigma
  )
}
```

## 5.1 Check the Prior Support

```{r, eval=FALSE}
if (
  a < 0 ||
  a > 10 ||
  sigma <= 0 ||
  sigma > 30
) {
  return(-Inf)
}
```

Under the current prior specification, the slope must satisfy

$$
0\leq a\leq10,
$$

and the residual standard deviation must satisfy

$$
0<\sigma\leq30.
$$

A proposed parameter value outside these ranges has prior density zero and log-prior

$$
-\infty.
$$

For example, the following values are automatically rejected:

$$
a=-1,
$$

$$
a=12,
$$

$$
\sigma=-2,
$$

and

$$
\sigma=35.
$$

## 5.2 Prior Distribution for the Slope

```{r, eval=FALSE}
log_prior_a <- dunif(
  a,
  min = 0,
  max = 10,
  log = TRUE
)
```

The slope prior is

$$
a\sim U(0,10).
$$

Its density is

$$
p(a)=
\begin{cases}
\frac{1}{10}, & 0\leq a\leq10,\\
0, & \text{otherwise}.
\end{cases}
$$

This prior means that:

- the slope is restricted to values between 0 and 10;
- every value in that interval has the same prior density;
- negative slopes are excluded.

Within the support,

$$
\log p(a)=\log\left(\frac{1}{10}\right).
$$

## 5.3 Prior Distribution for the Intercept

```{r, eval=FALSE}
log_prior_b <- dnorm(
  b,
  mean = 0,
  sd = 5,
  log = TRUE
)
```

The intercept prior is

$$
b\sim N(0,5^2).
$$

This means that:

- values of $b$ close to zero are considered more likely;
- both positive and negative intercepts are allowed;
- values far from zero receive lower prior density.

Approximately 95% of the prior probability lies within

$$
0\pm1.96\times5,
$$

or approximately

$$
(-9.8,9.8).
$$

## 5.4 Prior Distribution for the Standard Deviation

```{r, eval=FALSE}
log_prior_sigma <- dunif(
  sigma,
  min = 0,
  max = 30,
  log = TRUE
)
```

The prior for the residual standard deviation is

$$
\sigma\sim U(0,30).
$$

This assigns equal prior density to values between 0 and 30.

Because a standard deviation must be strictly positive, the code separately excludes

```{r, eval=FALSE}
sigma <= 0
```

Uniform priors are convenient for teaching. In formal Bayesian analyses, alternatives may include:

- half-normal priors;
- half-$t$ priors;
- half-Cauchy priors;
- inverse-gamma priors.

## 5.5 Joint Prior Distribution

```{r, eval=FALSE}
return(
  log_prior_a +
  log_prior_b +
  log_prior_sigma
)
```

The code assumes prior independence:

$$
p(a,b,\sigma)
=
p(a)p(b)p(\sigma).
$$

Taking logarithms gives

$$
\log p(a,b,\sigma)
=
\log p(a)
+
\log p(b)
+
\log p(\sigma).
$$

Therefore, the three log-prior contributions can be added.

# 6. Define the Log-Posterior Function

```{r, eval=FALSE}
log_posterior <- function(param) {

  lp <- log_prior(param)

  if (!is.finite(lp)) {
    return(-Inf)
  }

  return(
    log_likelihood(param) + lp
  )
}
```

## 6.1 Bayes' Theorem

Bayes' theorem is

$$
p(a,b,\sigma\mid y)
=
\frac{
p(y\mid a,b,\sigma)p(a,b,\sigma)
}{
p(y)
}.
$$

Because the denominator $p(y)$ does not depend on the parameters,

$$
p(a,b,\sigma\mid y)
\propto
p(y\mid a,b,\sigma)p(a,b,\sigma).
$$

Taking logarithms gives

$$
\log p(a,b,\sigma\mid y)
=
\log p(y\mid a,b,\sigma)
+
\log p(a,b,\sigma)
+
C,
$$

where $C$ does not depend on the parameters.

Therefore, the MCMC algorithm only needs

$$
\log p(y\mid a,b,\sigma)
+
\log p(a,b,\sigma).
$$

## 6.2 Check the Prior First

```{r, eval=FALSE}
lp <- log_prior(param)
```

The function first calculates the log-prior.

If the parameter vector lies outside the prior support, `log_prior()` returns `-Inf`.

```{r, eval=FALSE}
if (!is.finite(lp)) {
  return(-Inf)
}
```

In that situation, there is no need to calculate the likelihood.

This:

- avoids unnecessary computation;
- prevents invalid values from entering the likelihood;
- ensures invalid proposals are rejected.

## 6.3 Calculate the Unnormalized Log-Posterior

```{r, eval=FALSE}
return(
  log_likelihood(param) + lp
)
```

The function returns

$$
\text{log-likelihood}+\text{log-prior}.
$$

That is,

$$
\log \widetilde{p}(a,b,\sigma\mid y),
$$

the unnormalized log-posterior density.

Metropolis–Hastings only requires posterior-density ratios, so the normalizing constant does not need to be calculated.

# 7. Define the Proposal Distribution

```{r, eval=FALSE}
proposal_function <- function(param) {

  rnorm(
    n = 3,
    mean = param,
    sd = c(0.1, 0.5, 0.3)
  )
}
```

At each MCMC iteration, a new candidate parameter vector is generated near the current parameter vector.

Let the current state be

$$
\theta^{(t)}
=
\left(
a^{(t)},b^{(t)},\sigma^{(t)}
\right).
$$

The proposal distribution is

$$
a^*\sim N(a^{(t)},0.1^2),
$$

$$
b^*\sim N(b^{(t)},0.5^2),
$$

and

$$
\sigma^*\sim N(\sigma^{(t)},0.3^2).
$$

Equivalently,

$$
a^*=a^{(t)}+\varepsilon_a,
$$

$$
b^*=b^{(t)}+\varepsilon_b,
$$

and

$$
\sigma^*=\sigma^{(t)}+\varepsilon_\sigma,
$$

where

$$
\varepsilon_a\sim N(0,0.1^2),
$$

$$
\varepsilon_b\sim N(0,0.5^2),
$$

and

$$
\varepsilon_\sigma\sim N(0,0.3^2).
$$

This is called a random-walk proposal distribution.

## 7.1 Proposal Step Sizes

The code

```{r, eval=FALSE}
sd = c(0.1, 0.5, 0.3)
```

specifies:

- a proposal standard deviation of 0.1 for the slope;
- a proposal standard deviation of 0.5 for the intercept;
- a proposal standard deviation of 0.3 for the residual standard deviation.

If the proposal step sizes are too small:

- the acceptance rate may be high;
- each move will be short;
- autocorrelation may be high;
- exploration of the posterior may be slow.

If the proposal step sizes are too large:

- proposals may frequently fall in low-posterior-density regions;
- the acceptance rate may be low;
- the chain may remain at the same value for many iterations.

Proposal scales should therefore be assessed using acceptance rates, trace plots, autocorrelation, and effective sample sizes.

# 8. Define the Metropolis MCMC Function

```{r, eval=FALSE}
run_metropolis_MCMC <- function(
  start_value,
  iterations
) {

  chain <- matrix(
    NA_real_,
    nrow = iterations + 1,
    ncol = 3
  )

  colnames(chain) <- c(
    "slope",
    "intercept",
    "sigma"
  )

  chain[1, ] <- start_value

  accepted <- 0

  for (i in seq_len(iterations)) {

    current <- chain[i, ]

    proposal <- proposal_function(current)

    log_acceptance_ratio <-
      log_posterior(proposal) -
      log_posterior(current)

    if (
      is.finite(log_acceptance_ratio) &&
      log(runif(1)) < min(0, log_acceptance_ratio)
    ) {

      chain[i + 1, ] <- proposal
      accepted <- accepted + 1

    } else {

      chain[i + 1, ] <- current
    }
  }

  return(
    list(
      chain = chain,
      acceptance_rate = accepted / iterations
    )
  )
}
```

## 8.1 Function Inputs

The function has two inputs:

```{r, eval=FALSE}
start_value
```

which specifies the initial parameter vector, and

```{r, eval=FALSE}
iterations
```

which specifies the number of Metropolis iterations.

## 8.2 Create the MCMC Storage Matrix

```{r, eval=FALSE}
chain <- matrix(
  NA_real_,
  nrow = iterations + 1,
  ncol = 3
)
```

The matrix `chain` stores the parameter values at every iteration.

The three columns correspond to

$$
a,\quad b,\quad \sigma.
$$

The number of rows is

$$
\text{iterations}+1,
$$

because the first row stores the initial value.

For example, if

```{r, eval=FALSE}
iterations <- 10000
```

then the matrix has dimension

$$
10001\times3.
$$

The first row stores the initial parameter vector, and the final row stores the result after 10,000 updates.

`NA_real_` initializes the matrix as a numeric matrix.

## 8.3 Assign Column Names

```{r, eval=FALSE}
colnames(chain) <- c(
  "slope",
  "intercept",
  "sigma"
)
```

The three columns are named:

- `slope`;
- `intercept`;
- `sigma`.

The slope can then be extracted using

```{r, eval=FALSE}
chain[, "slope"]
```

rather than only

```{r, eval=FALSE}
chain[, 1]
```

## 8.4 Store the Initial Value

```{r, eval=FALSE}
chain[1, ] <- start_value
```

The initial parameter vector is stored in the first row.

## 8.5 Initialize the Acceptance Counter

```{r, eval=FALSE}
accepted <- 0
```

`accepted` records the number of accepted proposals.

Each accepted proposal increases the counter:

```{r, eval=FALSE}
accepted <- accepted + 1
```

The acceptance rate is

$$
\text{Acceptance rate}
=
\frac{\text{number of accepted proposals}}
{\text{total number of iterations}}.
$$

## 8.6 Begin the Iterations

```{r, eval=FALSE}
for (i in seq_len(iterations)) {
```

`seq_len(iterations)` generates

$$
1,2,\ldots,\text{iterations}.
$$

Therefore, the loop runs the specified number of times.

## 8.7 Extract the Current State

```{r, eval=FALSE}
current <- chain[i, ]
```

At iteration $i$, the current state is

$$
\theta^{(t)}
=
\left(
a^{(t)},b^{(t)},\sigma^{(t)}
\right).
$$

## 8.8 Generate a Proposal

```{r, eval=FALSE}
proposal <- proposal_function(current)
```

A candidate parameter vector is generated:

$$
\theta^*
=
(a^*,b^*,\sigma^*).
$$

## 8.9 Calculate the Log Acceptance Ratio

```{r, eval=FALSE}
log_acceptance_ratio <-
  log_posterior(proposal) -
  log_posterior(current)
```

The general Metropolis–Hastings acceptance probability is

$$
\alpha
=
\min\left[
1,
\frac{
p(\theta^*\mid y)
q(\theta^{(t)}\mid\theta^*)
}{
p(\theta^{(t)}\mid y)
q(\theta^*\mid\theta^{(t)})
}
\right].
$$

Because the proposal distribution is a symmetric normal random walk,

$$
q(\theta^*\mid\theta^{(t)})
=
q(\theta^{(t)}\mid\theta^*).
$$

The proposal-density ratio cancels, leaving

$$
\alpha
=
\min\left[
1,
\frac{
p(\theta^*\mid y)
}{
p(\theta^{(t)}\mid y)
}
\right].
$$

Define

$$
r=
\frac{
p(\theta^*\mid y)
}{
p(\theta^{(t)}\mid y)
}.
$$

Then

$$
\log r
=
\log p(\theta^*\mid y)
-
\log p(\theta^{(t)}\mid y).
$$

This is exactly what `log_acceptance_ratio` calculates.

## 8.10 Why Use the Log Scale?

A less stable implementation might calculate

```{r, eval=FALSE}
exp(
  log_posterior(proposal) -
  log_posterior(current)
)
```

However, large differences can produce numerical overflow or underflow:

- `exp(1000)` becomes `Inf`;
- `exp(-1000)` becomes 0.

The revised implementation performs the comparison directly on the log scale and avoids this problem.

## 8.11 Accept or Reject the Proposal

```{r, eval=FALSE}
if (
  is.finite(log_acceptance_ratio) &&
  log(runif(1)) < min(0, log_acceptance_ratio)
)
```

The standard Metropolis rule generates

$$
u\sim U(0,1)
$$

and accepts the proposal if

$$
u<\alpha.
$$

Because

$$
\alpha=\min(1,r),
$$

we have

$$
\log\alpha=\min(0,\log r).
$$

Therefore, the acceptance rule can be written as

$$
\log u<\min(0,\log r).
$$

This corresponds to

```{r, eval=FALSE}
log(runif(1)) < min(0, log_acceptance_ratio)
```

### When the Proposal Has Higher Posterior Density

If

$$
\log r>0,
$$

then

$$
r>1.
$$

The proposal has higher posterior density than the current state, so

$$
\alpha=1.
$$

The proposal is always accepted.

### When the Proposal Has Lower Posterior Density

If

$$
\log r<0,
$$

then

$$
0<r<1.
$$

The proposal is still accepted with probability $r$.

For example, if

$$
\log r=-1.609,
$$

then

$$
r=\exp(-1.609)\approx0.20.
$$

The proposal is accepted approximately 20% of the time.

Allowing occasional moves to lower-posterior-density states helps the chain explore the full posterior distribution rather than becoming trapped near one high-density location.

## 8.12 Accept the Proposal

```{r, eval=FALSE}
chain[i + 1, ] <- proposal
accepted <- accepted + 1
```

If the proposal is accepted,

$$
\theta^{(t+1)}=\theta^*.
$$

The acceptance counter is increased by one.

## 8.13 Reject the Proposal

```{r, eval=FALSE}
chain[i + 1, ] <- current
```

If the proposal is rejected,

$$
\theta^{(t+1)}=\theta^{(t)}.
$$

Therefore, consecutive rows of the MCMC chain may be identical. This is expected and indicates that a proposal was rejected.

## 8.14 Return the Results

```{r, eval=FALSE}
return(
  list(
    chain = chain,
    acceptance_rate = accepted / iterations
  )
)
```

The function returns a list containing:

1. `chain`: the complete MCMC parameter chain;
2. `acceptance_rate`: the proportion of accepted proposals.

# 9. Run the Metropolis MCMC Algorithm

## 9.1 Set the Initial Values

```{r, eval=FALSE}
start_value <- c(
  slope = 4,
  intercept = 0,
  sigma = 10
)
```

The initial values are

$$
a^{(0)}=4,
$$

$$
b^{(0)}=0,
$$

and

$$
\sigma^{(0)}=10.
$$

The initial values must lie within the prior support:

$$
0\leq a\leq10
$$

and

$$
0<\sigma\leq30.
$$

Otherwise, the initial log-posterior may be `-Inf`, and the algorithm may not start properly.

## 9.2 Run the Algorithm

```{r, eval=FALSE}
fit <- run_metropolis_MCMC(
  start_value = start_value,
  iterations = 10000
)
```

The algorithm performs 10,000 Metropolis updates.

The returned object is stored in `fit`.

## 9.3 Extract the Parameter Chain

```{r, eval=FALSE}
chain <- fit$chain

dim(chain)
head(chain)
```

`chain` is a $10001\times3$ matrix.

The columns represent:

- the slope;
- the intercept;
- the residual standard deviation.

The first row contains the initial values, and the following 10,000 rows contain the updated parameter values.

## 9.4 Examine the Acceptance Rate

```{r, eval=FALSE}
fit$acceptance_rate
```

The acceptance rate is

$$
\text{Acceptance rate}
=
\frac{
\text{number of accepted proposals}
}{
\text{total number of iterations}
}.
$$

For a low-dimensional random-walk Metropolis algorithm, an acceptance rate around 20% to 50% may sometimes be reasonable, but this is only a rough guideline.

Acceptance rate alone cannot determine whether an MCMC chain is efficient.

For example:

- a very high acceptance rate may indicate proposal steps that are too small;
- a very low acceptance rate may indicate proposal steps that are too large;
- a seemingly reasonable acceptance rate can still occur with highly autocorrelated samples.

The acceptance rate should be considered together with:

- trace plots;
- autocorrelation plots;
- effective sample size;
- multiple-chain convergence diagnostics.

# 10. Remove the Burn-In Period

```{r, eval=FALSE}
burn_in <- 5000

post_chain <- chain[
  (burn_in + 1):nrow(chain),
  ,
  drop = FALSE
]

dim(post_chain)
```

The early part of an MCMC chain may be influenced by the initial values.

Samples generated before the chain reaches a stable region are often called burn-in or warm-up samples.

This code removes the first 5,000 rows and retains rows

$$
5001,5002,\ldots,10001.
$$

The argument

```{r, eval=FALSE}
drop = FALSE
```

ensures that the result remains a matrix.

Because the first row stores the initial value, the current code removes the first 5,000 rows of the stored chain.

To remove both the initial value and the first 5,000 updates explicitly, one could instead use

```{r, eval=FALSE}
post_chain <- chain[
  (burn_in + 2):nrow(chain),
  ,
  drop = FALSE
]
```   

The two approaches differ by only one sample and usually produce nearly identical summaries.



# 11. Calculate Posterior Means

```{r, eval=FALSE}
posterior_mean <- colMeans(post_chain)

posterior_mean
```

`colMeans()` calculates the mean of the MCMC draws for each parameter.

The posterior mean of the slope is approximated by

$$
E(a\mid y)
\approx
\frac{1}{M}
\sum_{m=1}^{M}a^{(m)}.
$$

The posterior mean of the intercept is approximated by

$$
E(b\mid y)
\approx
\frac{1}{M}
\sum_{m=1}^{M}b^{(m)}.
$$

The posterior mean of the standard deviation is approximated by

$$
E(\sigma\mid y)
\approx
\frac{1}{M}
\sum_{m=1}^{M}\sigma^{(m)}.
$$

Because the true values are

$$
a=5,\qquad b=0,\qquad \sigma=10,
$$

the posterior means should generally be close to these values, although they will not be exactly equal.

Differences may arise because of:

- finite sample size;
- random sampling variation;
- prior information;
- finite Monte Carlo error.    



# 12. Calculate Posterior Credible Intervals 

```{r, eval=FALSE}
posterior_interval <- apply(
  post_chain,
  2,
  quantile,
  probs = c(0.025, 0.5, 0.975)
)

posterior_interval
```

For each parameter, the code calculates:

- the 2.5th percentile;
- the 50th percentile;
- the 97.5th percentile.

The 50th percentile is the posterior median.

The 2.5th and 97.5th percentiles form an equal-tailed 95% Bayesian credible interval:

$$
\left[
q_{0.025},q_{0.975}
\right].
$$

For example, if the 95% credible interval for the slope is

$$
(4.6,5.4),
$$

the Bayesian interpretation is:

> Given the model, prior distributions, and observed data, the posterior probability that the slope lies between 4.6 and 5.4 is 95%.

This differs from the interpretation of a frequentist confidence interval.



# 13. Create a Posterior Summary Table

```{r, eval=FALSE}
posterior_summary <- data.frame(
  Parameter = colnames(post_chain),
  Mean = colMeans(post_chain),
  Median = apply(post_chain, 2, median),
  SD = apply(post_chain, 2, sd),
  Lower_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.025
  ),
  Upper_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.975
  )
)

posterior_summary
```

The summary table includes:

- posterior mean;
- posterior median;
- posterior standard deviation;
- lower limit of the 95% credible interval;
- upper limit of the 95% credible interval.



# 14. Compare Posterior Estimates with the True Values

```{r, eval=FALSE}
comparison <- data.frame(
  Parameter = c(
    "slope",
    "intercept",
    "sigma"
  ),
  True_value = c(
    trueA,
    trueB,
    trueSd
  ),
  Posterior_mean = posterior_mean
)

comparison
```

Because this is a simulation study, the true parameter values are known.

The table compares

$$
\text{true parameter values}
$$

with

$$
\text{posterior means}.
$$

In real applications, the true parameter values are unknown, so this direct comparison is not available.
 


# 15. Plot the MCMC Trace Plots

```{r, eval=FALSE}
par(mfrow = c(3, 1))

plot(
  chain[, "slope"],
  type = "l",
  main = "Trace Plot: Slope",
  xlab = "Iteration",
  ylab = "Slope"
)

abline(
  h = trueA,
  lty = 2
)

plot(
  chain[, "intercept"],
  type = "l",
  main = "Trace Plot: Intercept",
  xlab = "Iteration",
  ylab = "Intercept"
)

abline(
  h = trueB,
  lty = 2
)

plot(
  chain[, "sigma"],
  type = "l",
  main = "Trace Plot: Sigma",
  xlab = "Iteration",
  ylab = "Sigma"
)

abline(
  h = trueSd,
  lty = 2
)

par(mfrow = c(1, 1))
```

A trace plot displays parameter values across MCMC iterations.

A well-behaved chain generally:

- fluctuates around a stable level;
- does not show a persistent upward or downward trend;
- does not show major structural changes;
- does not remain trapped in one narrow region for a long time;
- resembles stable random variation.

The horizontal dashed lines show the true parameter values.

In a real analysis, the true parameter values are unknown, so these reference lines would not be available.
``` 
```

# 16. Plot the Posterior Distributions

```{r, eval=FALSE}
par(mfrow = c(3, 1))

hist(
  post_chain[, "slope"],
  breaks = 30,
  main = "Posterior Distribution of Slope",
  xlab = "Slope",
  probability = TRUE
)

lines(
  density(post_chain[, "slope"]),
  lwd = 2
)

abline(
  v = trueA,
  lty = 2
)

hist(
  post_chain[, "intercept"],
  breaks = 30,
  main = "Posterior Distribution of Intercept",
  xlab = "Intercept",
  probability = TRUE
)

lines(
  density(post_chain[, "intercept"]),
  lwd = 2
)

abline(
  v = trueB,
  lty = 2
)

hist(
  post_chain[, "sigma"],
  breaks = 30,
  main = "Posterior Distribution of Sigma",
  xlab = "Sigma",
  probability = TRUE
)

lines(
  density(post_chain[, "sigma"]),
  lwd = 2
)

abline(
  v = trueSd,
  lty = 2
)

par(mfrow = c(1, 1))
```

The posterior histograms show the posterior distributions of the three parameters.

They can be used to examine:

- the center of each posterior distribution;
- posterior uncertainty;
- symmetry or skewness;
- whether the true value lies in a high-density region.
```  
``` 


# 17. Examine Autocorrelation

```{r, eval=FALSE}
par(mfrow = c(3, 1))

acf(
  post_chain[, "slope"],
  main = "Autocorrelation: Slope"
)

acf(
  post_chain[, "intercept"],
  main = "Autocorrelation: Intercept"
)

acf(
  post_chain[, "sigma"],
  main = "Autocorrelation: Sigma"
)

par(mfrow = c(1, 1))
```

MCMC samples are generally not independent.

Autocorrelation plots measure the dependence between a draw and earlier draws in the chain.

If autocorrelation decreases slowly, this may indicate that:

- neighboring samples are highly correlated;
- the chain moves slowly;
- the effective sample size is much smaller than the number of iterations;
- proposal standard deviations may need adjustment;
- more iterations may be required.
```
```
# 18. Complete Runnable Code

The following section contains the complete analysis without the detailed explanation.

```{r complete-analysis}
set.seed(123)

#-----------------------------
# 1. Generate simulated data
#-----------------------------

trueA  <- 5
trueB  <- 0
trueSd <- 10

sampleSize <- 31

x <- (-(sampleSize - 1) / 2):
      ((sampleSize - 1) / 2)

y <- trueA * x +
     trueB +
     rnorm(
       n = sampleSize,
       mean = 0,
       sd = trueSd
     )

plot(
  x,
  y,
  main = "Simulated Data",
  xlab = "x",
  ylab = "y",
  pch = 19
)

#-----------------------------
# 2. Log-likelihood
#-----------------------------

log_likelihood <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (!is.finite(sigma) || sigma <= 0) {
    return(-Inf)
  }

  predicted_mean <- a * x + b

  log_density <- dnorm(
    y,
    mean = predicted_mean,
    sd = sigma,
    log = TRUE
  )

  return(sum(log_density))
}

#-----------------------------
# 3. Log-prior
#-----------------------------

log_prior <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (
    a < 0 ||
    a > 10 ||
    sigma <= 0 ||
    sigma > 30
  ) {
    return(-Inf)
  }

  log_prior_a <- dunif(
    a,
    min = 0,
    max = 10,
    log = TRUE
  )

  log_prior_b <- dnorm(
    b,
    mean = 0,
    sd = 5,
    log = TRUE
  )

  log_prior_sigma <- dunif(
    sigma,
    min = 0,
    max = 30,
    log = TRUE
  )

  return(
    log_prior_a +
    log_prior_b +
    log_prior_sigma
  )
}

#-----------------------------
# 4. Log-posterior
#-----------------------------

log_posterior <- function(param) {

  lp <- log_prior(param)

  if (!is.finite(lp)) {
    return(-Inf)
  }

  return(
    log_likelihood(param) + lp
  )
}

#-----------------------------
# 5. Proposal distribution
#-----------------------------

proposal_function <- function(param) {

  rnorm(
    n = 3,
    mean = param,
    sd = c(0.1, 0.5, 0.3)
  )
}

#-----------------------------
# 6. Metropolis algorithm
#-----------------------------

run_metropolis_MCMC <- function(
  start_value,
  iterations
) {

  chain <- matrix(
    NA_real_,
    nrow = iterations + 1,
    ncol = 3
  )

  colnames(chain) <- c(
    "slope",
    "intercept",
    "sigma"
  )

  chain[1, ] <- start_value

  accepted <- 0

  for (i in seq_len(iterations)) {

    current <- chain[i, ]

    proposal <- proposal_function(current)

    log_acceptance_ratio <-
      log_posterior(proposal) -
      log_posterior(current)

    if (
      is.finite(log_acceptance_ratio) &&
      log(runif(1)) < min(0, log_acceptance_ratio)
    ) {

      chain[i + 1, ] <- proposal
      accepted <- accepted + 1

    } else {

      chain[i + 1, ] <- current
    }
  }

  return(
    list(
      chain = chain,
      acceptance_rate = accepted / iterations
    )
  )
}

#-----------------------------
# 7. Run MCMC
#-----------------------------

start_value <- c(
  slope = 4,
  intercept = 0,
  sigma = 10
)

fit <- run_metropolis_MCMC(
  start_value = start_value,
  iterations = 10000
)

chain <- fit$chain

fit$acceptance_rate

#-----------------------------
# 8. Remove burn-in
#-----------------------------

burn_in <- 5000

post_chain <- chain[
  (burn_in + 1):nrow(chain),
  ,
  drop = FALSE
]

#-----------------------------
# 9. Posterior summaries
#-----------------------------

posterior_mean <- colMeans(post_chain)

posterior_interval <- apply(
  post_chain,
  2,
  quantile,
  probs = c(0.025, 0.5, 0.975)
)

posterior_mean
posterior_interval

#-----------------------------
# 10. Summary table
#-----------------------------

posterior_summary <- data.frame(
  Parameter = colnames(post_chain),
  Mean = colMeans(post_chain),
  Median = apply(post_chain, 2, median),
  SD = apply(post_chain, 2, sd),
  Lower_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.025
  ),
  Upper_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.975
  )
)

posterior_summary

#-----------------------------
# 11. Trace plots
#-----------------------------

par(mfrow = c(3, 1))

plot(
  chain[, "slope"],
  type = "l",
  main = "Trace Plot: Slope",
  xlab = "Iteration",
  ylab = "Slope"
)

plot(
  chain[, "intercept"],
  type = "l",
  main = "Trace Plot: Intercept",
  xlab = "Iteration",
  ylab = "Intercept"
)

plot(
  chain[, "sigma"],
  type = "l",
  main = "Trace Plot: Sigma",
  xlab = "Iteration",
  ylab = "Sigma"
)

par(mfrow = c(1, 1))

#-----------------------------
# 12. Posterior distributions
#-----------------------------

par(mfrow = c(3, 1))

hist(
  post_chain[, "slope"],
  breaks = 30,
  probability = TRUE,
  main = "Posterior Distribution of Slope",
  xlab = "Slope"
)

lines(
  density(post_chain[, "slope"]),
  lwd = 2
)

hist(
  post_chain[, "intercept"],
  breaks = 30,
  probability = TRUE,
  main = "Posterior Distribution of Intercept",
  xlab = "Intercept"
)

lines(
  density(post_chain[, "intercept"]),
  lwd = 2
)

hist(
  post_chain[, "sigma"],
  breaks = 30,
  probability = TRUE,
  main = "Posterior Distribution of Sigma",
  xlab = "Sigma"
)

lines(
  density(post_chain[, "sigma"]),
  lwd = 2
)

par(mfrow = c(1, 1))
```



# 19. Summary

This program uses the Metropolis–Hastings algorithm to fit the Bayesian linear regression model

$$
y_i\mid a,b,\sigma
\sim
N(ax_i+b,\sigma^2).
$$

The main steps are:

1. Generate a candidate parameter vector near the current parameter vector.
2. Calculate the posterior density of the candidate and current states.
3. Use the posterior-density ratio to accept or reject the proposal.
4. Move to the proposal if it is accepted.
5. Remain at the current state if it is rejected.
6. Repeat the procedure for many iterations.
7. Remove the burn-in period.
8. Use the remaining samples to approximate the posterior distribution.

After sufficiently many iterations, the MCMC samples approximate

$$
p(a,b,\sigma\mid y).
$$

These samples can then be used to estimate:

- posterior means;
- posterior medians;
- posterior standard deviations;
- Bayesian credible intervals;
- posterior correlations;
- any posterior probability of interest.

For example, the posterior probability that the slope is greater than zero can be estimated by

```{r posterior-probability}
mean(post_chain[, "slope"] > 0)
```

This approximates

$$
P(a>0\mid y).
$$

In this example, the prior already restricts the slope to the interval from 0 to 10, so this probability will be equal or very close to 1.
