---
title: "Bayesian Linear Regression Using Metropolis–Hastings MCMC"
description: "A practical R learning guide to Bayesian Linear Regression Using Metropolis–Hastings MCMC, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Adaptive & Bayesian Methods"
topic: "Bayesian Linear Regression Using Metropolis–Hastings MCMC"
type: article
tags: ["Adaptive & Bayesian Methods", "Biostatistics", "Bayesian", "Linear", "Regression"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Adaptive and Bayesian methods update evidence as data accumulate and express uncertainty through probability distributions or decision rules.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Bayesian Linear Regression Using Metropolis–Hastings MCMC**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. 1. Objective
2. 2. Set the Random Seed
3. 3. Generate Simulated Data
4. 3.1 Define the True Parameters
5. 3.2 Define the Sample Size
6. 3.3 Generate the Predictor Values

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Adaptive & Bayesian Methods",
  topic = "Bayesian Linear Regression Using Metropolis\u2013Hastings MCMC"
)
learning_overview
```

## Learning Material

The following material develops **Bayesian Linear Regression Using Metropolis–Hastings MCMC** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

#### 1. Objective

<!-- generated-learning-note -->
This subsection explains **1. Objective** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


This example first simulates a simple linear regression dataset and then uses a manually implemented Metropolis–Hastings Markov chain Monte Carlo (MCMC) algorithm to estimate three unknown parameters:

- the slope $a$;
- the intercept $b$;
- the residual standard deviation $\sigma$.

The assumed linear regression model is

$$
y_i = a x_i + b + \varepsilon_i,
$$

where

$$
\varepsilon_i \sim N(0,\sigma^2).
$$

Equivalently,

$$
y_i \mid a,b,\sigma \sim N(a x_i+b,\sigma^2).
$$

The true parameter values used to generate the simulated data are

$$
a=5,\qquad b=0,\qquad \sigma=10.
$$

In a real analysis, these parameters are unknown. They are specified here only to generate simulated data and to evaluate whether the Bayesian estimates recover values close to the truth.

#### 2. Set the Random Seed

<!-- generated-learning-note -->
This subsection explains **2. Set the Random Seed** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
set.seed(123)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`set.seed(123)` fixes the random-number seed.

This program uses random-number generators such as:

- `rnorm()` for normal random variables;
- `runif()` for uniform random variables.

Without a fixed seed, the simulated data and MCMC results will change every time the code is run.

Using the same seed makes the analysis reproducible and is useful for:

- reproducing results;
- debugging code;
- comparing different methods;
- teaching demonstrations;
- documenting simulation studies.

#### 3. Generate Simulated Data

<!-- generated-learning-note -->
This section prepares the information needed for **3. Generate Simulated Data**. Check variable definitions, units, coding, and missing values before moving to analysis.


#### 3.1 Define the True Parameters

<!-- generated-learning-note -->
This subsection explains **3.1 Define the True Parameters** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
trueA  <- 5
trueB  <- 0
trueSd <- 10
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


These values define the true data-generating parameters:

$$
a=5,
$$

$$
b=0,
$$

and

$$
\sigma=10.
$$

Therefore, the true data-generating model is

$$
y_i = 5x_i + \varepsilon_i,
$$

where

$$
\varepsilon_i \sim N(0,10^2).
$$

The objects `trueA`, `trueB`, and `trueSd` are used only to simulate data. The Bayesian analysis below estimates the parameters from the simulated observations.

#### 3.2 Define the Sample Size

<!-- generated-learning-note -->
This subsection explains **3.2 Define the Sample Size** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
sampleSize <- 31
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The sample size is

$$
n=31.
$$

Therefore, the dataset contains 31 values of $x$ and 31 corresponding values of $y$.

#### 3.3 Generate the Predictor Values

<!-- generated-learning-note -->
This subsection explains **3.3 Generate the Predictor Values** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
x <- (-(sampleSize - 1) / 2):
      ((sampleSize - 1) / 2)

x
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Because

$$
\frac{31-1}{2}=15,
$$

the predictor values are

$$
-15,-14,-13,\ldots,-1,0,1,\ldots,13,14,15.
$$

These values are symmetric around zero, so

$$
\bar{x}=0.
$$

Centering the predictor around zero has two common advantages:

1. The intercept $b$ represents the expected response when $x=0$.
2. It often reduces posterior correlation between the slope and intercept.

#### 3.4 Generate the Outcome Values

<!-- generated-learning-note -->
This subsection explains **3.4 Generate the Outcome Values** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r, eval=FALSE}
y <- trueA * x +
     trueB +
     rnorm(
       n = sampleSize,
       mean = 0,
       sd = trueSd
     )
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


This code generates the outcome according to

$$
y_i = a x_i+b+\varepsilon_i.
$$

Substituting the true parameter values gives

$$
y_i = 5x_i+0+\varepsilon_i,
$$

where

$$
\varepsilon_i\sim N(0,10^2).
$$

The following code generates 31 independent normal errors:


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r, eval=FALSE}
rnorm(
  n = sampleSize,
  mean = 0,
  sd = trueSd
)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


Each outcome therefore consists of a systematic component and a random component:

$$
y_i=\text{linear trend}+\text{random error}.
$$

In this example,

$$
y_i=5x_i+\varepsilon_i.
$$

#### 3.5 Plot the Simulated Data

<!-- generated-learning-note -->
This section prepares the information needed for **3.5 Plot the Simulated Data**. Check variable definitions, units, coding, and missing values before moving to analysis.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r, eval=FALSE}
plot(
  x,
  y,
  main = "Simulated Data",
  xlab = "x",
  ylab = "y",
  pch = 19
)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


Because the true slope is positive,

$$
a=5,
$$

the scatterplot should show a clear increasing trend.

However, because the residual standard deviation is 10, the observations do not fall exactly on a straight line.

#### 4. Define the Log-Likelihood Function

<!-- generated-learning-note -->
This subsection explains **4. Define the Log-Likelihood Function** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r, eval=FALSE}
log_likelihood <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (!is.finite(sigma) || sigma <= 0) {
    return(-Inf)
  }

  predicted_mean <- a * x + b

  log_density <- dnorm(
    y,
    mean = predicted_mean,
    sd = sigma,
    log = TRUE
  )

  return(sum(log_density))
}
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.


#### 4.1 Input Parameters

<!-- generated-learning-note -->
This subsection explains **4.1 Input Parameters** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


The input `param` is a vector containing three parameters:

$$
\text{param}=(a,b,\sigma).
$$

The following lines extract the parameters:


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
a     <- param[1]
b     <- param[2]
sigma <- param[3]
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Thus:

- `param[1]` is the slope $a$;
- `param[2]` is the intercept $b$;
- `param[3]` is the residual standard deviation $\sigma$.

For example,


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
param <- c(5, 0, 10)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


represents

$$
a=5,\qquad b=0,\qquad \sigma=10.
$$

#### 4.2 Check Whether the Standard Deviation Is Valid

<!-- generated-learning-note -->
This subsection explains **4.2 Check Whether the Standard Deviation Is Valid** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
if (!is.finite(sigma) || sigma <= 0) {
  return(-Inf)
}
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


A normal standard deviation must satisfy

$$
\sigma>0.
$$

The parameter is invalid if $\sigma$ is:

- less than or equal to zero;
- `NA`;
- `NaN`;
- `Inf`;
- `-Inf`.

In these cases, the function returns

$$
-\infty.
$$

Because

$$
\log(0)=-\infty,
$$

returning `-Inf` tells the MCMC algorithm that this parameter combination has zero probability and should be rejected.

This check prevents `dnorm()` from receiving an invalid standard deviation.

#### 4.3 Calculate the Predicted Mean

<!-- generated-learning-note -->
This subsection explains **4.3 Calculate the Predicted Mean** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
predicted_mean <- a * x + b
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


For each observation, the conditional mean is

$$
\mu_i=ax_i+b.
$$

Therefore,

$$
E(y_i\mid a,b,\sigma)=ax_i+b.
$$

`predicted_mean` is a vector of length 31. Its $i$th element is

$$
a x_i+b.
$$

#### 4.4 Calculate the Log-Density for Each Observation

<!-- generated-learning-note -->
This subsection explains **4.4 Calculate the Log-Density for Each Observation** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
log_density <- dnorm(
  y,
  mean = predicted_mean,
  sd = sigma,
  log = TRUE
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The model assumes

$$
y_i\mid a,b,\sigma\sim N(ax_i+b,\sigma^2).
$$

The normal density for a single observation is

$$
p(y_i\mid a,b,\sigma)
=
\frac{1}{\sqrt{2\pi}\sigma}
\exp\left[
-\frac{(y_i-ax_i-b)^2}{2\sigma^2}
\right].
$$

Taking logarithms gives

$$
\log p(y_i\mid a,b,\sigma)
=
-\frac{1}{2}\log(2\pi)
-\log(\sigma)
-\frac{(y_i-ax_i-b)^2}{2\sigma^2}.
$$

The argument


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
log = TRUE
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


instructs R to calculate the log-density directly.

This is numerically more stable than multiplying many small probability densities.

#### 4.5 Calculate the Joint Log-Likelihood

<!-- generated-learning-note -->
This subsection explains **4.5 Calculate the Joint Log-Likelihood** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
return(sum(log_density))
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Assuming conditional independence of observations, the joint likelihood is

$$
L(a,b,\sigma)
=
\prod_{i=1}^{n}
p(y_i\mid a,b,\sigma).
$$

Taking logarithms converts the product into a sum:

$$
\log L(a,b,\sigma)
=
\sum_{i=1}^{n}
\log p(y_i\mid a,b,\sigma).
$$

Therefore, `log_likelihood()` returns the joint log-likelihood for the complete dataset.

#### 5. Define the Log-Prior Function

<!-- generated-learning-note -->
This subsection explains **5. Define the Log-Prior Function** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r, eval=FALSE}
log_prior <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (
    a < 0 ||
    a > 10 ||
    sigma <= 0 ||
    sigma > 30
  ) {
    return(-Inf)
  }

  log_prior_a <- dunif(
    a,
    min = 0,
    max = 10,
    log = TRUE
  )

  log_prior_b <- dnorm(
    b,
    mean = 0,
    sd = 5,
    log = TRUE
  )

  log_prior_sigma <- dunif(
    sigma,
    min = 0,
    max = 30,
    log = TRUE
  )

  return(
    log_prior_a +
    log_prior_b +
    log_prior_sigma
  )
}
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.


#### 5.1 Check the Prior Support

<!-- generated-learning-note -->
This subsection explains **5.1 Check the Prior Support** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
if (
  a < 0 ||
  a > 10 ||
  sigma <= 0 ||
  sigma > 30
) {
  return(-Inf)
}
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Under the current prior specification, the slope must satisfy

$$
0\leq a\leq10,
$$

and the residual standard deviation must satisfy

$$
0<\sigma\leq30.
$$

A proposed parameter value outside these ranges has prior density zero and log-prior

$$
-\infty.
$$

For example, the following values are automatically rejected:

$$
a=-1,
$$

$$
a=12,
$$

$$
\sigma=-2,
$$

and

$$
\sigma=35.
$$

#### 5.2 Prior Distribution for the Slope

<!-- generated-learning-note -->
This subsection explains **5.2 Prior Distribution for the Slope** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
log_prior_a <- dunif(
  a,
  min = 0,
  max = 10,
  log = TRUE
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The slope prior is

$$
a\sim U(0,10).
$$

Its density is

$$
p(a)=
\begin{cases}
\frac{1}{10}, & 0\leq a\leq10,\\
0, & \text{otherwise}.
\end{cases}
$$

This prior means that:

- the slope is restricted to values between 0 and 10;
- every value in that interval has the same prior density;
- negative slopes are excluded.

Within the support,

$$
\log p(a)=\log\left(\frac{1}{10}\right).
$$

#### 5.3 Prior Distribution for the Intercept

<!-- generated-learning-note -->
This subsection explains **5.3 Prior Distribution for the Intercept** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
log_prior_b <- dnorm(
  b,
  mean = 0,
  sd = 5,
  log = TRUE
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The intercept prior is

$$
b\sim N(0,5^2).
$$

This means that:

- values of $b$ close to zero are considered more likely;
- both positive and negative intercepts are allowed;
- values far from zero receive lower prior density.

Approximately 95% of the prior probability lies within

$$
0\pm1.96\times5,
$$

or approximately

$$
(-9.8,9.8).
$$

#### 5.4 Prior Distribution for the Standard Deviation

<!-- generated-learning-note -->
This subsection explains **5.4 Prior Distribution for the Standard Deviation** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
log_prior_sigma <- dunif(
  sigma,
  min = 0,
  max = 30,
  log = TRUE
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The prior for the residual standard deviation is

$$
\sigma\sim U(0,30).
$$

This assigns equal prior density to values between 0 and 30.

Because a standard deviation must be strictly positive, the code separately excludes


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
sigma <= 0
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Uniform priors are convenient for teaching. In formal Bayesian analyses, alternatives may include:

- half-normal priors;
- half-$t$ priors;
- half-Cauchy priors;
- inverse-gamma priors.

#### 5.5 Joint Prior Distribution

<!-- generated-learning-note -->
This subsection explains **5.5 Joint Prior Distribution** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
return(
  log_prior_a +
  log_prior_b +
  log_prior_sigma
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The code assumes prior independence:

$$
p(a,b,\sigma)
=
p(a)p(b)p(\sigma).
$$

Taking logarithms gives

$$
\log p(a,b,\sigma)
=
\log p(a)
+
\log p(b)
+
\log p(\sigma).
$$

Therefore, the three log-prior contributions can be added.

#### 6. Define the Log-Posterior Function

<!-- generated-learning-note -->
This subsection explains **6. Define the Log-Posterior Function** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r, eval=FALSE}
log_posterior <- function(param) {

  lp <- log_prior(param)

  if (!is.finite(lp)) {
    return(-Inf)
  }

  return(
    log_likelihood(param) + lp
  )
}
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.


#### 6.1 Bayes' Theorem

<!-- generated-learning-note -->
This subsection explains **6.1 Bayes' Theorem** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


Bayes' theorem is

$$
p(a,b,\sigma\mid y)
=
\frac{
p(y\mid a,b,\sigma)p(a,b,\sigma)
}{
p(y)
}.
$$

Because the denominator $p(y)$ does not depend on the parameters,

$$
p(a,b,\sigma\mid y)
\propto
p(y\mid a,b,\sigma)p(a,b,\sigma).
$$

Taking logarithms gives

$$
\log p(a,b,\sigma\mid y)
=
\log p(y\mid a,b,\sigma)
+
\log p(a,b,\sigma)
+
C,
$$

where $C$ does not depend on the parameters.

Therefore, the MCMC algorithm only needs

$$
\log p(y\mid a,b,\sigma)
+
\log p(a,b,\sigma).
$$

#### 6.2 Check the Prior First

<!-- generated-learning-note -->
This subsection explains **6.2 Check the Prior First** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
lp <- log_prior(param)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The function first calculates the log-prior.

If the parameter vector lies outside the prior support, `log_prior()` returns `-Inf`.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
if (!is.finite(lp)) {
  return(-Inf)
}
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


In that situation, there is no need to calculate the likelihood.

This:

- avoids unnecessary computation;
- prevents invalid values from entering the likelihood;
- ensures invalid proposals are rejected.

#### 6.3 Calculate the Unnormalized Log-Posterior

<!-- generated-learning-note -->
This subsection explains **6.3 Calculate the Unnormalized Log-Posterior** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
return(
  log_likelihood(param) + lp
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The function returns

$$
\text{log-likelihood}+\text{log-prior}.
$$

That is,

$$
\log \widetilde{p}(a,b,\sigma\mid y),
$$

the unnormalized log-posterior density.

Metropolis–Hastings only requires posterior-density ratios, so the normalizing constant does not need to be calculated.

#### 7. Define the Proposal Distribution

<!-- generated-learning-note -->
This subsection explains **7. Define the Proposal Distribution** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r, eval=FALSE}
proposal_function <- function(param) {

  rnorm(
    n = 3,
    mean = param,
    sd = c(0.1, 0.5, 0.3)
  )
}
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


At each MCMC iteration, a new candidate parameter vector is generated near the current parameter vector.

Let the current state be

$$
\theta^{(t)}
=
\left(
a^{(t)},b^{(t)},\sigma^{(t)}
\right).
$$

The proposal distribution is

$$
a^*\sim N(a^{(t)},0.1^2),
$$

$$
b^*\sim N(b^{(t)},0.5^2),
$$

and

$$
\sigma^*\sim N(\sigma^{(t)},0.3^2).
$$

Equivalently,

$$
a^*=a^{(t)}+\varepsilon_a,
$$

$$
b^*=b^{(t)}+\varepsilon_b,
$$

and

$$
\sigma^*=\sigma^{(t)}+\varepsilon_\sigma,
$$

where

$$
\varepsilon_a\sim N(0,0.1^2),
$$

$$
\varepsilon_b\sim N(0,0.5^2),
$$

and

$$
\varepsilon_\sigma\sim N(0,0.3^2).
$$

This is called a random-walk proposal distribution.

#### 7.1 Proposal Step Sizes

<!-- generated-learning-note -->
This subsection explains **7.1 Proposal Step Sizes** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


The code


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
sd = c(0.1, 0.5, 0.3)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


specifies:

- a proposal standard deviation of 0.1 for the slope;
- a proposal standard deviation of 0.5 for the intercept;
- a proposal standard deviation of 0.3 for the residual standard deviation.

If the proposal step sizes are too small:

- the acceptance rate may be high;
- each move will be short;
- autocorrelation may be high;
- exploration of the posterior may be slow.

If the proposal step sizes are too large:

- proposals may frequently fall in low-posterior-density regions;
- the acceptance rate may be low;
- the chain may remain at the same value for many iterations.

Proposal scales should therefore be assessed using acceptance rates, trace plots, autocorrelation, and effective sample sizes.

#### 8. Define the Metropolis MCMC Function

<!-- generated-learning-note -->
This subsection explains **8. Define the Metropolis MCMC Function** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r, eval=FALSE}
run_metropolis_MCMC <- function(
  start_value,
  iterations
) {

  chain <- matrix(
    NA_real_,
    nrow = iterations + 1,
    ncol = 3
  )

  colnames(chain) <- c(
    "slope",
    "intercept",
    "sigma"
  )

  chain[1, ] <- start_value

  accepted <- 0

  for (i in seq_len(iterations)) {

    current <- chain[i, ]

    proposal <- proposal_function(current)

    log_acceptance_ratio <-
      log_posterior(proposal) -
      log_posterior(current)

    if (
      is.finite(log_acceptance_ratio) &&
      log(runif(1)) < min(0, log_acceptance_ratio)
    ) {

      chain[i + 1, ] <- proposal
      accepted <- accepted + 1

    } else {

      chain[i + 1, ] <- current
    }
  }

  return(
    list(
      chain = chain,
      acceptance_rate = accepted / iterations
    )
  )
}
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### 8.1 Function Inputs

<!-- generated-learning-note -->
This subsection explains **8.1 Function Inputs** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


The function has two inputs:


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
start_value
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


which specifies the initial parameter vector, and


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
iterations
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


which specifies the number of Metropolis iterations.

#### 8.2 Create the MCMC Storage Matrix

<!-- generated-learning-note -->
This subsection explains **8.2 Create the MCMC Storage Matrix** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
chain <- matrix(
  NA_real_,
  nrow = iterations + 1,
  ncol = 3
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The matrix `chain` stores the parameter values at every iteration.

The three columns correspond to

$$
a,\quad b,\quad \sigma.
$$

The number of rows is

$$
\text{iterations}+1,
$$

because the first row stores the initial value.

For example, if


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
iterations <- 10000
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


then the matrix has dimension

$$
10001\times3.
$$

The first row stores the initial parameter vector, and the final row stores the result after 10,000 updates.

`NA_real_` initializes the matrix as a numeric matrix.

#### 8.3 Assign Column Names

<!-- generated-learning-note -->
This subsection explains **8.3 Assign Column Names** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
colnames(chain) <- c(
  "slope",
  "intercept",
  "sigma"
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The three columns are named:

- `slope`;
- `intercept`;
- `sigma`.

The slope can then be extracted using


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
chain[, "slope"]
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


rather than only


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
chain[, 1]
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### 8.4 Store the Initial Value

<!-- generated-learning-note -->
This subsection explains **8.4 Store the Initial Value** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
chain[1, ] <- start_value
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The initial parameter vector is stored in the first row.

#### 8.5 Initialize the Acceptance Counter

<!-- generated-learning-note -->
This subsection explains **8.5 Initialize the Acceptance Counter** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
accepted <- 0
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`accepted` records the number of accepted proposals.

Each accepted proposal increases the counter:


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
accepted <- accepted + 1
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The acceptance rate is

$$
\text{Acceptance rate}
=
\frac{\text{number of accepted proposals}}
{\text{total number of iterations}}.
$$

#### 8.6 Begin the Iterations

<!-- generated-learning-note -->
This subsection explains **8.6 Begin the Iterations** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
for (i in seq_len(iterations)) {
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`seq_len(iterations)` generates

$$
1,2,\ldots,\text{iterations}.
$$

Therefore, the loop runs the specified number of times.

#### 8.7 Extract the Current State

<!-- generated-learning-note -->
This subsection explains **8.7 Extract the Current State** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
current <- chain[i, ]
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


At iteration $i$, the current state is

$$
\theta^{(t)}
=
\left(
a^{(t)},b^{(t)},\sigma^{(t)}
\right).
$$

#### 8.8 Generate a Proposal

<!-- generated-learning-note -->
This subsection explains **8.8 Generate a Proposal** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
proposal <- proposal_function(current)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


A candidate parameter vector is generated:

$$
\theta^*
=
(a^*,b^*,\sigma^*).
$$

#### 8.9 Calculate the Log Acceptance Ratio

<!-- generated-learning-note -->
This subsection explains **8.9 Calculate the Log Acceptance Ratio** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
log_acceptance_ratio <-
  log_posterior(proposal) -
  log_posterior(current)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The general Metropolis–Hastings acceptance probability is

$$
\alpha
=
\min\left[
1,
\frac{
p(\theta^*\mid y)
q(\theta^{(t)}\mid\theta^*)
}{
p(\theta^{(t)}\mid y)
q(\theta^*\mid\theta^{(t)})
}
\right].
$$

Because the proposal distribution is a symmetric normal random walk,

$$
q(\theta^*\mid\theta^{(t)})
=
q(\theta^{(t)}\mid\theta^*).
$$

The proposal-density ratio cancels, leaving

$$
\alpha
=
\min\left[
1,
\frac{
p(\theta^*\mid y)
}{
p(\theta^{(t)}\mid y)
}
\right].
$$

Define

$$
r=
\frac{
p(\theta^*\mid y)
}{
p(\theta^{(t)}\mid y)
}.
$$

Then

$$
\log r
=
\log p(\theta^*\mid y)
-
\log p(\theta^{(t)}\mid y).
$$

This is exactly what `log_acceptance_ratio` calculates.

#### 8.10 Why Use the Log Scale?

<!-- generated-learning-note -->
This subsection explains **8.10 Why Use the Log Scale?** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


A less stable implementation might calculate


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
exp(
  log_posterior(proposal) -
  log_posterior(current)
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


However, large differences can produce numerical overflow or underflow:

- `exp(1000)` becomes `Inf`;
- `exp(-1000)` becomes 0.

The revised implementation performs the comparison directly on the log scale and avoids this problem.

#### 8.11 Accept or Reject the Proposal

<!-- generated-learning-note -->
This subsection explains **8.11 Accept or Reject the Proposal** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r, eval=FALSE}
if (
  is.finite(log_acceptance_ratio) &&
  log(runif(1)) < min(0, log_acceptance_ratio)
)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


The standard Metropolis rule generates

$$
u\sim U(0,1)
$$

and accepts the proposal if

$$
u<\alpha.
$$

Because

$$
\alpha=\min(1,r),
$$

we have

$$
\log\alpha=\min(0,\log r).
$$

Therefore, the acceptance rule can be written as

$$
\log u<\min(0,\log r).
$$

This corresponds to


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r, eval=FALSE}
log(runif(1)) < min(0, log_acceptance_ratio)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### When the Proposal Has Higher Posterior Density

<!-- generated-learning-note -->
This subsection explains **When the Proposal Has Higher Posterior Density** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


If

$$
\log r>0,
$$

then

$$
r>1.
$$

The proposal has higher posterior density than the current state, so

$$
\alpha=1.
$$

The proposal is always accepted.

#### When the Proposal Has Lower Posterior Density

<!-- generated-learning-note -->
This subsection explains **When the Proposal Has Lower Posterior Density** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


If

$$
\log r<0,
$$

then

$$
0<r<1.
$$

The proposal is still accepted with probability $r$.

For example, if

$$
\log r=-1.609,
$$

then

$$
r=\exp(-1.609)\approx0.20.
$$

The proposal is accepted approximately 20% of the time.

Allowing occasional moves to lower-posterior-density states helps the chain explore the full posterior distribution rather than becoming trapped near one high-density location.

#### 8.12 Accept the Proposal

<!-- generated-learning-note -->
This subsection explains **8.12 Accept the Proposal** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
chain[i + 1, ] <- proposal
accepted <- accepted + 1
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


If the proposal is accepted,

$$
\theta^{(t+1)}=\theta^*.
$$

The acceptance counter is increased by one.

#### 8.13 Reject the Proposal

<!-- generated-learning-note -->
This subsection explains **8.13 Reject the Proposal** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
chain[i + 1, ] <- current
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


If the proposal is rejected,

$$
\theta^{(t+1)}=\theta^{(t)}.
$$

Therefore, consecutive rows of the MCMC chain may be identical. This is expected and indicates that a proposal was rejected.

#### 8.14 Return the Results

<!-- generated-learning-note -->
This section focuses on **8.14 Return the Results**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
return(
  list(
    chain = chain,
    acceptance_rate = accepted / iterations
  )
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The function returns a list containing:

1. `chain`: the complete MCMC parameter chain;
2. `acceptance_rate`: the proportion of accepted proposals.

#### 9. Run the Metropolis MCMC Algorithm

<!-- generated-learning-note -->
This subsection explains **9. Run the Metropolis MCMC Algorithm** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


#### 9.1 Set the Initial Values

<!-- generated-learning-note -->
This subsection explains **9.1 Set the Initial Values** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
start_value <- c(
  slope = 4,
  intercept = 0,
  sigma = 10
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The initial values are

$$
a^{(0)}=4,
$$

$$
b^{(0)}=0,
$$

and

$$
\sigma^{(0)}=10.
$$

The initial values must lie within the prior support:

$$
0\leq a\leq10
$$

and

$$
0<\sigma\leq30.
$$

Otherwise, the initial log-posterior may be `-Inf`, and the algorithm may not start properly.

#### 9.2 Run the Algorithm

<!-- generated-learning-note -->
This subsection explains **9.2 Run the Algorithm** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
fit <- run_metropolis_MCMC(
  start_value = start_value,
  iterations = 10000
)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The algorithm performs 10,000 Metropolis updates.

The returned object is stored in `fit`.

#### 9.3 Extract the Parameter Chain

<!-- generated-learning-note -->
This subsection explains **9.3 Extract the Parameter Chain** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
chain <- fit$chain

dim(chain)
head(chain)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`chain` is a $10001\times3$ matrix.

The columns represent:

- the slope;
- the intercept;
- the residual standard deviation.

The first row contains the initial values, and the following 10,000 rows contain the updated parameter values.

#### 9.4 Examine the Acceptance Rate

<!-- generated-learning-note -->
This subsection explains **9.4 Examine the Acceptance Rate** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
fit$acceptance_rate
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The acceptance rate is

$$
\text{Acceptance rate}
=
\frac{
\text{number of accepted proposals}
}{
\text{total number of iterations}
}.
$$

For a low-dimensional random-walk Metropolis algorithm, an acceptance rate around 20% to 50% may sometimes be reasonable, but this is only a rough guideline.

Acceptance rate alone cannot determine whether an MCMC chain is efficient.

For example:

- a very high acceptance rate may indicate proposal steps that are too small;
- a very low acceptance rate may indicate proposal steps that are too large;
- a seemingly reasonable acceptance rate can still occur with highly autocorrelated samples.

The acceptance rate should be considered together with:

- trace plots;
- autocorrelation plots;
- effective sample size;
- multiple-chain convergence diagnostics.

#### 10. Remove the Burn-In Period

<!-- generated-learning-note -->
This subsection explains **10. Remove the Burn-In Period** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
burn_in <- 5000

post_chain <- chain[
  (burn_in + 1):nrow(chain),
  ,
  drop = FALSE
]

dim(post_chain)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The early part of an MCMC chain may be influenced by the initial values.

Samples generated before the chain reaches a stable region are often called burn-in or warm-up samples.

This code removes the first 5,000 rows and retains rows

$$
5001,5002,\ldots,10001.
$$

The argument


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
drop = FALSE
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


ensures that the result remains a matrix.

Because the first row stores the initial value, the current code removes the first 5,000 rows of the stored chain.

To remove both the initial value and the first 5,000 updates explicitly, one could instead use


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
post_chain <- chain[
  (burn_in + 2):nrow(chain),
  ,
  drop = FALSE
]
```   

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The two approaches differ by only one sample and usually produce nearly identical summaries.



#### 11. Calculate Posterior Means

<!-- generated-learning-note -->
This subsection explains **11. Calculate Posterior Means** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
posterior_mean <- colMeans(post_chain)

posterior_mean
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


`colMeans()` calculates the mean of the MCMC draws for each parameter.

The posterior mean of the slope is approximated by

$$
E(a\mid y)
\approx
\frac{1}{M}
\sum_{m=1}^{M}a^{(m)}.
$$

The posterior mean of the intercept is approximated by

$$
E(b\mid y)
\approx
\frac{1}{M}
\sum_{m=1}^{M}b^{(m)}.
$$

The posterior mean of the standard deviation is approximated by

$$
E(\sigma\mid y)
\approx
\frac{1}{M}
\sum_{m=1}^{M}\sigma^{(m)}.
$$

Because the true values are

$$
a=5,\qquad b=0,\qquad \sigma=10,
$$

the posterior means should generally be close to these values, although they will not be exactly equal.

Differences may arise because of:

- finite sample size;
- random sampling variation;
- prior information;
- finite Monte Carlo error.    



#### 12. Calculate Posterior Credible Intervals

<!-- generated-learning-note -->
This subsection explains **12. Calculate Posterior Credible Intervals** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
posterior_interval <- apply(
  post_chain,
  2,
  quantile,
  probs = c(0.025, 0.5, 0.975)
)

posterior_interval
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


For each parameter, the code calculates:

- the 2.5th percentile;
- the 50th percentile;
- the 97.5th percentile.

The 50th percentile is the posterior median.

The 2.5th and 97.5th percentiles form an equal-tailed 95% Bayesian credible interval:

$$
\left[
q_{0.025},q_{0.975}
\right].
$$

For example, if the 95% credible interval for the slope is

$$
(4.6,5.4),
$$

the Bayesian interpretation is:

> Given the model, prior distributions, and observed data, the posterior probability that the slope lies between 4.6 and 5.4 is 95%.

This differs from the interpretation of a frequentist confidence interval.



#### 13. Create a Posterior Summary Table

<!-- generated-learning-note -->
This section focuses on **13. Create a Posterior Summary Table**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
posterior_summary <- data.frame(
  Parameter = colnames(post_chain),
  Mean = colMeans(post_chain),
  Median = apply(post_chain, 2, median),
  SD = apply(post_chain, 2, sd),
  Lower_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.025
  ),
  Upper_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.975
  )
)

posterior_summary
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


The summary table includes:

- posterior mean;
- posterior median;
- posterior standard deviation;
- lower limit of the 95% credible interval;
- upper limit of the 95% credible interval.



#### 14. Compare Posterior Estimates with the True Values

<!-- generated-learning-note -->
This section develops **14. Compare Posterior Estimates with the True Values**. Read the model specification together with its assumptions and the scale of each reported effect.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
comparison <- data.frame(
  Parameter = c(
    "slope",
    "intercept",
    "sigma"
  ),
  True_value = c(
    trueA,
    trueB,
    trueSd
  ),
  Posterior_mean = posterior_mean
)

comparison
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


Because this is a simulation study, the true parameter values are known.

The table compares

$$
\text{true parameter values}
$$

with

$$
\text{posterior means}.
$$

In real applications, the true parameter values are unknown, so this direct comparison is not available.
 


#### 15. Plot the MCMC Trace Plots

<!-- generated-learning-note -->
This section uses **15. Plot the MCMC Trace Plots** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r, eval=FALSE}
par(mfrow = c(3, 1))

plot(
  chain[, "slope"],
  type = "l",
  main = "Trace Plot: Slope",
  xlab = "Iteration",
  ylab = "Slope"
)

abline(
  h = trueA,
  lty = 2
)

plot(
  chain[, "intercept"],
  type = "l",
  main = "Trace Plot: Intercept",
  xlab = "Iteration",
  ylab = "Intercept"
)

abline(
  h = trueB,
  lty = 2
)

plot(
  chain[, "sigma"],
  type = "l",
  main = "Trace Plot: Sigma",
  xlab = "Iteration",
  ylab = "Sigma"
)

abline(
  h = trueSd,
  lty = 2
)

par(mfrow = c(1, 1))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


A trace plot displays parameter values across MCMC iterations.

A well-behaved chain generally:

- fluctuates around a stable level;
- does not show a persistent upward or downward trend;
- does not show major structural changes;
- does not remain trapped in one narrow region for a long time;
- resembles stable random variation.

The horizontal dashed lines show the true parameter values.

In a real analysis, the true parameter values are unknown, so these reference lines would not be available.
``` 
```

#### 16. Plot the Posterior Distributions

<!-- generated-learning-note -->
This section uses **16. Plot the Posterior Distributions** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r, eval=FALSE}
par(mfrow = c(3, 1))

hist(
  post_chain[, "slope"],
  breaks = 30,
  main = "Posterior Distribution of Slope",
  xlab = "Slope",
  probability = TRUE
)

lines(
  density(post_chain[, "slope"]),
  lwd = 2
)

abline(
  v = trueA,
  lty = 2
)

hist(
  post_chain[, "intercept"],
  breaks = 30,
  main = "Posterior Distribution of Intercept",
  xlab = "Intercept",
  probability = TRUE
)

lines(
  density(post_chain[, "intercept"]),
  lwd = 2
)

abline(
  v = trueB,
  lty = 2
)

hist(
  post_chain[, "sigma"],
  breaks = 30,
  main = "Posterior Distribution of Sigma",
  xlab = "Sigma",
  probability = TRUE
)

lines(
  density(post_chain[, "sigma"]),
  lwd = 2
)

abline(
  v = trueSd,
  lty = 2
)

par(mfrow = c(1, 1))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


The posterior histograms show the posterior distributions of the three parameters.

They can be used to examine:

- the center of each posterior distribution;
- posterior uncertainty;
- symmetry or skewness;
- whether the true value lies in a high-density region.
```  
``` 


#### 17. Examine Autocorrelation

<!-- generated-learning-note -->
This subsection explains **17. Examine Autocorrelation** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.



<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r, eval=FALSE}
par(mfrow = c(3, 1))

acf(
  post_chain[, "slope"],
  main = "Autocorrelation: Slope"
)

acf(
  post_chain[, "intercept"],
  main = "Autocorrelation: Intercept"
)

acf(
  post_chain[, "sigma"],
  main = "Autocorrelation: Sigma"
)

par(mfrow = c(1, 1))
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


MCMC samples are generally not independent.

Autocorrelation plots measure the dependence between a draw and earlier draws in the chain.

If autocorrelation decreases slowly, this may indicate that:

- neighboring samples are highly correlated;
- the chain moves slowly;
- the effective sample size is much smaller than the number of iterations;
- proposal standard deviations may need adjustment;
- more iterations may be required.
```
```
#### 18. Complete Runnable Code

<!-- generated-learning-note -->
This subsection explains **18. Complete Runnable Code** and shows how it contributes to the overall Bayesian Linear Regression Using Metropolis–Hastings MCMC workflow.


The following section contains the complete analysis without the detailed explanation.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r complete-analysis}
set.seed(123)

#-----------------------------
# 1. Generate simulated data
#-----------------------------

trueA  <- 5
trueB  <- 0
trueSd <- 10

sampleSize <- 31

x <- (-(sampleSize - 1) / 2):
      ((sampleSize - 1) / 2)

y <- trueA * x +
     trueB +
     rnorm(
       n = sampleSize,
       mean = 0,
       sd = trueSd
     )

plot(
  x,
  y,
  main = "Simulated Data",
  xlab = "x",
  ylab = "y",
  pch = 19
)

#-----------------------------
# 2. Log-likelihood
#-----------------------------

log_likelihood <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (!is.finite(sigma) || sigma <= 0) {
    return(-Inf)
  }

  predicted_mean <- a * x + b

  log_density <- dnorm(
    y,
    mean = predicted_mean,
    sd = sigma,
    log = TRUE
  )

  return(sum(log_density))
}

#-----------------------------
# 3. Log-prior
#-----------------------------

log_prior <- function(param) {

  a     <- param[1]
  b     <- param[2]
  sigma <- param[3]

  if (
    a < 0 ||
    a > 10 ||
    sigma <= 0 ||
    sigma > 30
  ) {
    return(-Inf)
  }

  log_prior_a <- dunif(
    a,
    min = 0,
    max = 10,
    log = TRUE
  )

  log_prior_b <- dnorm(
    b,
    mean = 0,
    sd = 5,
    log = TRUE
  )

  log_prior_sigma <- dunif(
    sigma,
    min = 0,
    max = 30,
    log = TRUE
  )

  return(
    log_prior_a +
    log_prior_b +
    log_prior_sigma
  )
}

#-----------------------------
# 4. Log-posterior
#-----------------------------

log_posterior <- function(param) {

  lp <- log_prior(param)

  if (!is.finite(lp)) {
    return(-Inf)
  }

  return(
    log_likelihood(param) + lp
  )
}

#-----------------------------
# 5. Proposal distribution
#-----------------------------

proposal_function <- function(param) {

  rnorm(
    n = 3,
    mean = param,
    sd = c(0.1, 0.5, 0.3)
  )
}

#-----------------------------
# 6. Metropolis algorithm
#-----------------------------

run_metropolis_MCMC <- function(
  start_value,
  iterations
) {

  chain <- matrix(
    NA_real_,
    nrow = iterations + 1,
    ncol = 3
  )

  colnames(chain) <- c(
    "slope",
    "intercept",
    "sigma"
  )

  chain[1, ] <- start_value

  accepted <- 0

  for (i in seq_len(iterations)) {

    current <- chain[i, ]

    proposal <- proposal_function(current)

    log_acceptance_ratio <-
      log_posterior(proposal) -
      log_posterior(current)

    if (
      is.finite(log_acceptance_ratio) &&
      log(runif(1)) < min(0, log_acceptance_ratio)
    ) {

      chain[i + 1, ] <- proposal
      accepted <- accepted + 1

    } else {

      chain[i + 1, ] <- current
    }
  }

  return(
    list(
      chain = chain,
      acceptance_rate = accepted / iterations
    )
  )
}

#-----------------------------
# 7. Run MCMC
#-----------------------------

start_value <- c(
  slope = 4,
  intercept = 0,
  sigma = 10
)

fit <- run_metropolis_MCMC(
  start_value = start_value,
  iterations = 10000
)

chain <- fit$chain

fit$acceptance_rate

#-----------------------------
# 8. Remove burn-in
#-----------------------------

burn_in <- 5000

post_chain <- chain[
  (burn_in + 1):nrow(chain),
  ,
  drop = FALSE
]

#-----------------------------
# 9. Posterior summaries
#-----------------------------

posterior_mean <- colMeans(post_chain)

posterior_interval <- apply(
  post_chain,
  2,
  quantile,
  probs = c(0.025, 0.5, 0.975)
)

posterior_mean
posterior_interval

#-----------------------------
# 10. Summary table
#-----------------------------

posterior_summary <- data.frame(
  Parameter = colnames(post_chain),
  Mean = colMeans(post_chain),
  Median = apply(post_chain, 2, median),
  SD = apply(post_chain, 2, sd),
  Lower_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.025
  ),
  Upper_95 = apply(
    post_chain,
    2,
    quantile,
    probs = 0.975
  )
)

posterior_summary

#-----------------------------
# 11. Trace plots
#-----------------------------

par(mfrow = c(3, 1))

plot(
  chain[, "slope"],
  type = "l",
  main = "Trace Plot: Slope",
  xlab = "Iteration",
  ylab = "Slope"
)

plot(
  chain[, "intercept"],
  type = "l",
  main = "Trace Plot: Intercept",
  xlab = "Iteration",
  ylab = "Intercept"
)

plot(
  chain[, "sigma"],
  type = "l",
  main = "Trace Plot: Sigma",
  xlab = "Iteration",
  ylab = "Sigma"
)

par(mfrow = c(1, 1))

#-----------------------------
# 12. Posterior distributions
#-----------------------------

par(mfrow = c(3, 1))

hist(
  post_chain[, "slope"],
  breaks = 30,
  probability = TRUE,
  main = "Posterior Distribution of Slope",
  xlab = "Slope"
)

lines(
  density(post_chain[, "slope"]),
  lwd = 2
)

hist(
  post_chain[, "intercept"],
  breaks = 30,
  probability = TRUE,
  main = "Posterior Distribution of Intercept",
  xlab = "Intercept"
)

lines(
  density(post_chain[, "intercept"]),
  lwd = 2
)

hist(
  post_chain[, "sigma"],
  breaks = 30,
  probability = TRUE,
  main = "Posterior Distribution of Sigma",
  xlab = "Sigma"
)

lines(
  density(post_chain[, "sigma"]),
  lwd = 2
)

par(mfrow = c(1, 1))
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.




#### 19. Summary

<!-- generated-learning-note -->
This section focuses on **19. Summary**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.


This program uses the Metropolis–Hastings algorithm to fit the Bayesian linear regression model

$$
y_i\mid a,b,\sigma
\sim
N(ax_i+b,\sigma^2).
$$

The main steps are:

1. Generate a candidate parameter vector near the current parameter vector.
2. Calculate the posterior density of the candidate and current states.
3. Use the posterior-density ratio to accept or reject the proposal.
4. Move to the proposal if it is accepted.
5. Remain at the current state if it is rejected.
6. Repeat the procedure for many iterations.
7. Remove the burn-in period.
8. Use the remaining samples to approximate the posterior distribution.

After sufficiently many iterations, the MCMC samples approximate

$$
p(a,b,\sigma\mid y).
$$

These samples can then be used to estimate:

- posterior means;
- posterior medians;
- posterior standard deviations;
- Bayesian credible intervals;
- posterior correlations;
- any posterior probability of interest.

For example, the posterior probability that the slope is greater than zero can be estimated by


<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r posterior-probability}
mean(post_chain[, "slope"] > 0)
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.


This approximates

$$
P(a>0\mid y).
$$

In this example, the prior already restricts the slope to the interval from 0 to 10, so this probability will be equal or very close to 1.

## Interpreting the Results

Inspect priors, likelihoods, posterior summaries, convergence, simulations, and operating characteristics. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
