---
title: "MCMC sampling"
output: html_document
date: '2023-05-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

If the new solution costs less than the old solution then it is guaranteed to be accepted and a poorer solution is unlikely to be accepted. it is to calculate the integral, proportion, and sampling.
 
Data simulation
```{r}
# Plot the exponential distribution:
x <- seq(from = 0, to = 10, length=20)
y <- exp(-x)  #y is only density function
plot(x,y)
lines(x,y)

mean(x )
```

Set up a target function
```{r}
# set up a target function, we want to sample from in this case, is exponent function 
target <- function(x){
  if (x < 0) {
    return(0)
  } else {
    return(exp(-x))   #target function
  }
}

target(3)
target(-1)
```
Sampling from a proposal distribution proportional =normal to the target =exponent
```{r}
# program a Metropolis–Hastings scheme to sample 
x <- rep(0, 1000)
x[1] <- 30     # this is just a starting value, which I’ve set arbitrarily to 3

for (i in 2:1000) {
  # i=2
  currentx <- x[i - 1]
  proposedx <-   rnorm(1, mean = currentx, sd = 1)
  A <- target(proposedx) / target(currentx)
  if (runif(1) < A) {
    x[i] <- proposedx
  } else {
    x[i] <- currentx
  }
}
```

```{r}
hist(x )
plot(x)
```

Wrap this up in a function
```{r}
easyMCMC <- function(niter, startval, proposalsd) {
  x <- rep(0, niter)
  x[1] <- startval
  for (i in 2:niter) {
    currentx <- x[i - 1]  #status 1
    proposedx <-  rnorm(1, mean = currentx, sd = proposalsd)  #status 2
    #transform matrix, that is the matter, to sample target sampling
    
    A <- target(proposedx) / target(currentx)  
    if (runif(1) < A) {
      x[i] <- proposedx
    } else {
      x[i] <- currentx
    }
  }
  return(x)
}
```

Run the MCMC scheme 3 times with different start values

It will be stable even if the start value is large. 
```{r}
# 
par(mfrow = c(1, 1))
z1 <- easyMCMC(1000,50,1)
z2 <- easyMCMC(1000,5,1)
z3 <- easyMCMC(1000,0.5,1)
plot(z1, type = "l",col = "red")
lines(z2, col = "black")
lines(z3, col = "green")

# posterior mean
mean(z1)
```

How to calculate pie
```{r}
# generate random sample based on normal distribution or uniform
# x=rnorm(10000, 0, 1.7) 
x=runif(10000, -1, 1) 

# target function
y= sqrt (1-x^2)
plot(x,y)

# create uniform random numbers based on the value of density function
s=numeric()
for (i in 1:10000){
  s[i]=runif(1, min = 0, max = 1)
}
s <- unlist(s)  
plot(x,s)

# calculate the dots including in the target function
diff=y-s
data=data.frame(x,s,y,diff)

# select the qualified sample
data2=data[data$diff>=0,]
plot(data2$x,data2$s)
hist(data2$x,100)

# calculate the approximate integral, p* common function integral 
p=nrow(data2)/10000

# area of half circle
p*2
# vs
pi/2
```

Monte Carlo Simulations Using the MonteCarlo Package 

```{r MonteCarlo-package}
library(pacman)
pacman::p_load(data.table, fixest, stargazer, dplyr, magrittr, MonteCarlo)
 
## Define the function for our OLS estimator

ols_estimator <- function(n, beta_0, beta_1){
  
  # Generate data
  U_i = rnorm(n, mean = 0, sd = 2) # Error
  X_i = rnorm(n, mean = 5, sd = 5) # Independent variable
  Y_i = beta_0 + beta_1*X_i + U_i  # Dependent variable
  
  # Formulate data.table
  data_i = data.table(Y = Y_i, X = X_i)
  
  # Run regressions
  ols_i <- feols(data = data_i, Y ~ X)
  
  # Extract slope coefficient and save
```


```{r MonteCarlo-package}
slope
```


```{r MonteCarlo-package}
<- ols_i$coefficients[2]
  
  # Return Results in a list
  return(list("beta_1" = unname(slope)))
  
  
}

# set start values at only one interation 
ols_estimator(n = 100, beta_0 = 1.5, beta_1 = 2.0)
# $beta_1
# [1] 1.947254
```

MonteCarlo interation
```{r}
## Define parameter grid for MonteCarlo interation
n_grid <- c(100, 500, 1000)   # Sample sizes
beta0_grid <- c(2.0)          # Intercept
beta1_grid <- c(1.5)          # Slope coefficient

## Collect parameter grids in list
param_list = list("n" = n_grid, "beta_0" = beta0_grid, "beta_1" = beta1_grid)

## Run simulation
MC_result_OLS <- MonteCarlo(func = ols_estimator,    #  iteration estimator
                            nrep = 500,              
                            ######## set times of iterates, 500 results
                            
                            param_list = param_list, # Parameter grid (Sample size and betas)
                            ncpus = 1,               # Number of computers
                            time_n_test = TRUE)      
```


```{r review-results, warning = FALSE, message = FALSE}
## View results
res <- MC_result_OLS$results
res <- res$beta_1

n_100 <- res[1,,,]   # Sample size of 100
n_500 <- res[2,,,]   # Sample size of 500
n_1000 <- res[3,,,]  # Sample size of 1000 

hist(n_100, xlim = c(1.20,1.70))
hist(n_500, xlim = c(1.20,1.70))
hist(n_1000, xlim = c(1.20,1.70))
```


```{r}
# Summary statistics
estimates_DT <- data.table(n_100 = n_100, n_500 = n_500, n_1000=n_1000)

stargazer(estimates_DT[, c("n_100", "n_500","n_1000")], type = "text")
```

