---
title: "Adaptive randomization probabilities and Interim Analysis"
description: "A practical R learning guide to Adaptive randomization probabilities and Interim Analysis, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Interim Analysis & Stopping Rules"
topic: "Adaptive randomization probabilities and Interim Analysis"
type: article
tags: ["Interim Analysis & Stopping Rules", "Biostatistics", "Adaptive", "randomization", "probabilities"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Interim rules allow pre-planned adaptation while preserving operating characteristics and trial credibility.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Adaptive randomization probabilities and Interim Analysis**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Adaptive randomization probabilities and Interim Analysis

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Interim Analysis & Stopping Rules",
  topic = "Adaptive randomization probabilities and Interim Analysis"
)
learning_overview
```

## Learning Material

The following material develops **Adaptive randomization probabilities and Interim Analysis** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Adaptive randomization probabilities and Interim Analysis

<!-- generated-learning-note -->
This subsection explains **Adaptive randomization probabilities and Interim Analysis** and shows how it contributes to the overall Adaptive randomization probabilities and Interim Analysis workflow.


Adaptive randomization (favoring the better-performing arm).

One interim analysis for sample size re-estimation. 

Stopping rules based on per-arm maximum. 

Final estimation of treatment effect and trial power.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# ==========================================================
# FDA-style Adaptive Clinical Trial Simulation
# Adaptive Randomization + Sample Size Re-estimation
# ==========================================================

# Adaptive randomization slightly favors the arm showing better outcomes (here, treatment arm).

library(dplyr)
library(ggplot2)

set.seed(12345)

# --------------------------
# Step 1: Define trial parameters
# --------------------------
n_initial <- 50           # initial per arm
n_max <- 200              # max total sample size per arm
effect_true <- 1.5        # true treatment effect
sd <- 4.0                 # standard deviation
alpha <- 0.05             # significance level
interim_frac <- 0.5       # interim fraction
adapt_start <- 20         # start adaptive randomization after this many patients, allocation changes based on observed outcomes
n_sim <- 500              # number of simulations

# adapt_start = 20 specifies that adaptive randomization only begins after 20 patients in total have been enrolled (treatment + control).
# 
# Before 20 patients, randomization is fixed at 50% per arm (assign_prob = 0.5).
# 
# After 20 patients, the allocation probability can change based on observed outcomes:
#   
#   If treatment mean > control mean → probability of assigning new patients to treatment increases.
# 
# If treatment mean < control mean → probability of assigning new patients to treatment decreases.
# 
# Bounded between 0.3 and 0.7 to avoid extreme imbalance.

# --------------------------
# Step 2: Function to simulate one adaptive trial
# --------------------------
simulate_adaptive_trial <- function(n_initial, n_max, effect_true, sd, alpha, interim_frac, adapt_start) {
  
  control <- c()
  treatment <- c()
  
  # initial randomization probability
  assign_prob <- 0.5
  final_n_per_arm <- n_initial
  
  # simulate patient-by-patient
  for (i in 1:n_max) {
    # adaptive randomization
    arm <- sample(c("Treatment","Control"), 1, prob=c(assign_prob, 1-assign_prob))
    
    # generate outcome
    outcome <- if(arm=="Treatment") rnorm(1, mean=effect_true, sd=sd) else rnorm(1, mean=0, sd=sd)
    
    # store outcome
    if(arm=="Treatment") treatment <- c(treatment, outcome)
    if(arm=="Control") control <- c(control, outcome)
    
    # update randomization probability after adapt_start patients
    if(length(treatment)+length(control) >= adapt_start) {
      mean_treat <- mean(treatment)
      mean_ctrl <- mean(control)
      assign_prob <- 0.5 + 0.2 * sign(mean_treat - mean_ctrl)
      assign_prob <- min(max(assign_prob, 0.3), 0.7)
    }
    
    # interim analysis for sample size re-estimation
    total_n <- length(treatment) + length(control)
    if(total_n >= floor(n_max*interim_frac)) {
      t_test <- t.test(treatment, control)
      # if p-value > 0.1 and we can still increase sample size, double the final per-arm n
      if(t_test$p.value > 0.1 & final_n_per_arm < n_max) {
        final_n_per_arm <- min(n_max, final_n_per_arm*2)
      }
    }
#     p-value > 0.1 → stop for futility
# p-value ≤ 0.1 → continue trial / possibly increase sample size

    # stop if reached current adaptive max per-arm sample size
    if(length(treatment) >= final_n_per_arm & length(control) >= final_n_per_arm) break
  }
  
  # final t-test
  t_test <- t.test(treatment, control)
  
  return(c(
    n_treat = length(treatment),
    n_control = length(control),
    p_value = t_test$p.value,
    est_diff = mean(treatment) - mean(control),
    significant = as.numeric(t_test$p.value < alpha)
  ))
}

# --------------------------
# Step 3: Run multiple simulations
# --------------------------
results <- replicate(n_sim, simulate_adaptive_trial(n_initial, n_max, effect_true, sd, alpha, interim_frac, adapt_start))
results_df <- as.data.frame(t(results))
colnames(results_df) <- c("n_treat","n_control","p_value","est_diff","significant")
head(results_df,20)

# --------------------------
# Step 4: Summarize results
# --------------------------
summary_df <- results_df %>%
  summarise(
    mean_n_treat = mean(n_treat),
    mean_n_control = mean(n_control),
    power = mean(significant),
    mean_est_diff = mean(est_diff),
    sd_est_diff = sd(est_diff)
  )
print(summary_df)
# power
# Proportion of trials that were statistically significant (p < 0.05).

# mean_n_treat =
# Some trials triggered sample size re-estimation at interim (p-value > 0.1), which doubled the per-arm N.
# Some trials reached stopping criteria after treatment arm hit final_n_per_arm, leaving fewer patients in control.

# --------------------------
# Step 5: Visualizations
# --------------------------
# Estimated treatment effect distribution
ggplot(results_df, aes(x=est_diff)) +
  geom_histogram(binwidth=0.2, fill="steelblue", color="white") +
  geom_vline(xintercept=effect_true, color="red", linetype="dashed") +
  labs(title="Estimated Treatment Effect Across Simulations",
       x="Estimated Treatment Effect", y="Frequency") +
  theme_minimal()

# Treatment allocation distribution
ggplot(results_df, aes(x=n_treat)) +
  geom_bar(fill="coral", color="white") +
  labs(title="Treatment Arm Sample Size Distribution Across Simulations",
       x="Number of Patients in Treatment Arm", y="Count") +
  theme_minimal()

# Control allocation distribution
ggplot(results_df, aes(x=n_control)) +
  geom_bar(fill="green", color="white") +
  labs(title="Control Arm Sample Size Distribution Across Simulations",
       x="Number of Patients in Control Arm", y="Count") +
  theme_minimal()

```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

## Interpreting the Results

Examine stopping probabilities, error rates, expected sample size, and behavior under several plausible treatment effects. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
