---
title: "Probability basics"
description: "A practical R learning guide to Probability basics, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Resources"
category: "Reference Resources"
topic: "Probability basics"
type: article
tags: ["Reference Resources", "Resources", "Probability", "basics"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Mathematical and statistical foundations make model behavior and computational results easier to understand.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Probability basics**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Probability basics
2. Events
3. Probability formulas
4. Calculation of probability (operations)
5. Bayes's theorem
6. Random variables and distribution functions

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Resources",
  category = "Reference Resources",
  topic = "Probability basics"
)
learning_overview
```

## Learning Material

The following material develops **Probability basics** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Probability basics

<!-- generated-learning-note -->
This subsection explains **Probability basics** and shows how it contributes to the overall Probability basics workflow.


#### Events

<!-- generated-learning-note -->
This subsection explains **Events** and shows how it contributes to the overall Probability basics workflow.

- Probability is the likelihood of an outcome. 
e.g. {any combination of two dice}

- Event is a collection of outcomes. 
e.g. {both dice show the same face}

- The outcome space is all the possible outcomes. 
e.g. {all the possible outcomes die show}

<!-- mutual independence of events,mutual independence of random variables, mutual independence of observations, `confusion`-->  

#### Probability formulas

<!-- generated-learning-note -->
This subsection explains **Probability formulas** and shows how it contributes to the overall Probability basics workflow.


- `binomial variable`

$$
P(E) = \frac{number\ of\ outcomes\ in\ E}{number\ of\ possible\ outcomes}
$$

- `continuous varable` 

$$
P\begin{pmatrix}a\leq X \leq b \end{pmatrix} = \int_a^b f(x) dx
$$
*Probabilities of continuous random variables (X) are the area under the curve. The probability of any value is always zero. when X = k,*

$$
P\begin{pmatrix}X = k \end{pmatrix} = 0
$$

#### Calculation of probability (operations)

<!-- generated-learning-note -->
This subsection explains **Calculation of probability (operations)** and shows how it contributes to the overall Probability basics workflow.


`union probability, addition rule +`
$$
\begin{aligned}
&P(A \cup B)=P(A)+P(B)-P(A \cap B) \\
&P(A \cup B)=P(A)+P(B) \quad \text { if } \mathrm{A} \text { and } \mathrm{B} \text { are mutually exclusive }
\end{aligned}
$$
`joint probability, multiple rule x` 
$$
\begin{aligned}
&P(A \cap B)=P(A \mid B) P(B)=P(B \mid A) P(A) \\
&P(A \cap B)=P(A) P(B) \quad \text { if } \mathrm{A} \text { and } \mathrm{B} \text { are independent }
\end{aligned}
$$
`Marginal Probability` is without reference to any other event or events
$$
 P(A)  or  P(B)
$$

`conditional Probability`
$$
P(A\mid B)=\dfrac{P(A \: \cap\: B)}{P(B)}
$$
`p-values are conditional probabilities.`

#### Bayes's theorem

<!-- generated-learning-note -->
This subsection explains **Bayes's theorem** and shows how it contributes to the overall Probability basics workflow.

- multiple law
$$
P(A \cap B) = P(A\ |\ B) P(B) = P(B\ |\ A) P(A)
$$

- bayes's formula  

$$
P(B_j\ |\ A) = \frac{P(A\ |\ B_j) P(B_j)}{P(A)}\
$$

- law of total probability $P(A)$

$$
P(A) = P(A\ |\ B_1) P(B_1) + \cdots + P(A\ |\ B_n) P(B_n).\notag
$$

#### Random variables and distribution functions

<!-- generated-learning-note -->
This section prepares the information needed for **Random variables and distribution functions**. Check variable definitions, units, coding, and missing values before moving to analysis.


Random variable takes on different values determined by chance. we can use random variables' mathematical (distribution) function to find their probability. 

- probability mass function (PMF, discrete), e.g. Binomial Distribution. upon some conditions are satisfied, the sampling distribution of the sample proportion is approximately normal. 
- Probability Density Function (PDF, continuous), e.g. normal distribution, t, chi-squre, f...
- Cumulative Distribution Function (CDF). 


#### Probability distribution

<!-- generated-learning-note -->
This subsection explains **Probability distribution** and shows how it contributes to the overall Probability basics workflow.

`joint distribution, discrete variables` 
$$
P\left(X=x_{i}, Y=y_{j}\right)=p_{i j}, i, j=1,2, \ldots
$$
`Marginal distribution, discrete variables` 
$$
P\left(X=x_{i}\right)=\sum_{j=1}^{\infty} p_{i j}=p_{i}, i=1,2, \ldots
$$
`conditional probability, discrete variables`
$$
P\left(Y=y_{j} \mid X=x_{i}\right)=\frac{p_{i j}}{p_{i}}, j=1,2, \ldots
$$

#### Conditional expectation

<!-- generated-learning-note -->
This subsection explains **Conditional expectation** and shows how it contributes to the overall Probability basics workflow.

Conditional expectation is the mathematical expectation of a conditional distribution. 

`The discrete variable`

$$
E(Y|X_i)=\sum_{i=1}^{N}{(Y_i|X_i)}\cdot p(Y_i|X_i)
$$

`The continuous variable`

$$
E(Y|X)=\int{(Y|X)}\cdot g(Y|X)dY
$$

`expectation formula, discrete variable`
$$
\mu=E(X)=\sum x_if(x_i)
$$

#### Conditional variance

<!-- generated-learning-note -->
This subsection explains **Conditional variance** and shows how it contributes to the overall Probability basics workflow.


`variance formula, discrete variable`
$$
\sigma^2=\text{Var}(X)=\sum (x_i-\mu)^2f(x_i)
$$

*Conditional expectation and conditional variance exist and can be estimated by regression models.*


#### Sampling

<!-- generated-learning-note -->
This subsection explains **Sampling** and shows how it contributes to the overall Probability basics workflow.


We make inferences about the population based on the sample (inference) after summarizing data (description). 

The error is resulting from using a sample characteristic (statistic) to estimate a population characteristic (parameter).

`Standard Error`
$$
SD(\bar{X})=SE(\bar{X})=\dfrac{\sigma}{\sqrt{n}}
$$

**Central limit theorem and law of large numbers**

- For a large sample size,  x mean is approximately normally distributed, regardless of the distribution of the population one samples from. so, the population parameter can be estimated using the sample. 

- With large samples, The mean of the sampling distribution is very close to the population mean. 


#### Confidence interval

<!-- generated-learning-note -->
This subsection explains **Confidence interval** and shows how it contributes to the overall Probability basics workflow.


The higher the confidence level, the wider the width of the interval and thus the poorer the precision.

$$
\text{point estimate }\pm M\times \hat{SE}(\text{estimate})
$$
`the margin of error`
$$
E=z_{\alpha/2}\sqrt{\dfrac{\hat{p}(1-\hat{p})}{n}}
$$

#### Introduction to Hypothesis Testing

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **Introduction to Hypothesis Testing** in the context of Probability basics.


-Set up the hypotheses and decide on the significance level.

-Construct and calculate the test statistic.

-Calculate probability value (p-value).

-Make a decision and state an overall conclusion.  

#### Probability R practice

<!-- generated-learning-note -->
This subsection explains **Probability R practice** and shows how it contributes to the overall Probability basics workflow.

  - pros: don't need complicated probability theory, easy (simulation)
  - cons: hard to get the exact solution 

#### Integrate

<!-- generated-learning-note -->
This subsection explains **Integrate** and shows how it contributes to the overall Probability basics workflow.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
integrate(function(x){x^2},0,2)$value
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Derivation

<!-- generated-learning-note -->
This subsection explains **Derivation** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
fxy = expression(2*x^2+y+3*x*y^2)
dxy = deriv(fxy, c("x", "y"), func = TRUE)
dxy(1,2)  
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Create random variables with specific distributions

<!-- generated-learning-note -->
This section prepares the information needed for **Create random variables with specific distributions**. Check variable definitions, units, coding, and missing values before moving to analysis.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
 dnorm(0)# density at a number 
 pnorm(1.28)# cumulative possibility 
 qnorm(0.95)#  quantile
 rnorm(10)# random numbers
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

 
`using covariance matrix to generate Gaussian multiple variables `

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
 library(MASS)
 Sigma <- matrix(c(10,3,3,2),2,2)
 mvrnorm(n=20, rep(0, 2), Sigma)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


 
#### Prob function

<!-- generated-learning-note -->
This subsection explains **Prob function** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
pnorm(1.96,    0,1)
qnorm(0.025,    0,1)
pchisq(3.84,1,lower.tail=F)
mean(rchisq(10000,1)>3.84) #simulation 

```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.



#### Vector and operations

<!-- generated-learning-note -->
This subsection explains **Vector and operations** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
seq(1,10,  2)
x=rep(1:3,6)
x
y=rep(1:3, each = 6)
y

x+y
x-y
x*y
x/y
x%*%y
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Select and substitute elements of vector

<!-- generated-learning-note -->
This subsection explains **Select and substitute elements of vector** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
x[c(2,3)]
x[-1]
x[x>2]
x[x==2]

# substitute 
x[x == 2] <- 0.5
x
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Matrix and operations

<!-- generated-learning-note -->
This subsection explains **Matrix and operations** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
matrix(1:10,2,5)
matrix(1:10,5,2)
a <- matrix(12:20,3,3)
a[2,]
a[,2]
a[-2,]
a[,-2]
a[2,1]=21
a 
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Compute inverse, determinant and eigen values of matrix

<!-- generated-learning-note -->
This subsection explains **Compute inverse, determinant and eigen values of matrix** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
a<-matrix(c(11,21,31,21,32,43,12,32,54),3,3)
solve(a)
det(a)
solve(a)*det(a)

t(a)
eigen(a)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Dataframe

<!-- generated-learning-note -->
This section prepares the information needed for **Dataframe**. Check variable definitions, units, coding, and missing values before moving to analysis.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
name<-c('A','B','C')
chinese<-c(92,96,95)
math<-c(86, 85, 92)
score<-data.frame(name, chinese, math)
score
score[2]
score$math
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.





#### Solve problems using simulation

<!-- generated-learning-note -->
This subsection explains **Solve problems using simulation** and shows how it contributes to the overall Probability basics workflow.

`for loop`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
sim<-10000
p<-numeric(sim)
# numeric=NULL
for (i in 1:sim){
  p[i]<-  abs(mean(rnorm(10,20,sqrt(3)))-mean(rnorm(15,20,sqrt(3))))<0.1
}

mean(p)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

`using replicate`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
mean(replicate(10000,abs(mean(rnorm(10,20,sqrt(3)))-mean(rnorm(15,20,sqrt(3))))<0.1))

```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

`using apply function`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
A<-matrix(rnorm(250000,  20,sqrt(3)),10000,25)
head(A)

f<-function(x)   {abs(mean(x[1:10])-mean(x[11:25]))}
# solve the mean by apply

mean(apply(A,1,f)>0.1)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


`using probability method`

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
 pnorm(0.1,0,sqrt(0.5))-pnorm(-0.1,0,sqrt(0.5))
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Permutations and combinations

<!-- generated-learning-note -->
This subsection explains **Permutations and combinations** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
choose(10,2)
# combn(10,2)
factorial(10)
prod(1:10)
 
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.

#### Search value position in vector

<!-- generated-learning-note -->
This subsection explains **Search value position in vector** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
a<-c(1,2,3,5,0,9)
which(a==min(a))

sum(a)
unique(a)
length(a)
min(a)
max(a)
all(c(3,4) %in% a)

```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Solve directly and optimize

<!-- generated-learning-note -->
This subsection explains **Solve directly and optimize** and shows how it contributes to the overall Probability basics workflow.

`plot and find the range of solve`

<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r}
f<-function(x){x^2-exp(x)} 
uniroot(f,c(-0.8,-0.6))  $root

f2<-function(x){abs(x^2-exp(x))} 
optimize   (f2,lower=-0.8,upper=-0.6)$minimum
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.

#### Calculate probability using simulation method

<!-- generated-learning-note -->
This section develops **Calculate probability using simulation method**. Read the model specification together with its assumptions and the scale of each reported effect.

`questions: randomly select 3 numbers out of 1:10, the sum is 9. `

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
badge<-1:10 
sim<-10000 
p<-numeric(sim) 

for (i in 1:sim){ a<-sample(badge,3,replace=F) 
p[i]<-sum(a)==9 }

mean(p)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

`questions: eat three flavors tangyuan. `

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
Tangyuan<-c(rep('A',8),rep('B',8),rep('C',8))
sim<-10000
p<-numeric(sim)
# how to do it according to the condition
for (i in 1:sim){
  a<-sample(Tangyuan,24,replace=F)
  p[i]<-(length(unique(a[1:6]))==3)&(length(unique(a[7:12]))==3)&(length(unique(a[13:18]))==3)&(length(unique(a[19:24]))==3)
}
mean(p)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

`question: select 2 balls when they are the same color. `

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
box1<-c(rep('white',5), rep("black",11), rep('red',8))
box2<-c(rep('white',10), rep("black",8), rep('red',6))
sim<-10000
p<-numeric(sim)

for (i in 1:sim){
  a<-sample(box1, 1)
  b<-sample(box2, 1)
  p[i]<- a==b
}
mean(p)
 
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

`select after putting them back`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
box<-c(rep("white",4),rep("red",2))
sim<-10000
t<-numeric(sim)

for (i in 1:sim){
  a<-sample(box, 2 ,replace=T)
  #  there are two white balls
  t[i]<-length(a[a=="white"])==2
}
mean(t)

```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

`question: two students have the same birthday out of 30 students`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
n<-30
sim<-10000
t<-numeric(sim)

for (i in 1:sim){
  a<-sample(1:365, n, replace=T)
  t[i]<-n-length(unique(a))
}
1-mean(t==0)
 

#  probability
1-prod(365:(365-30+1))/365^30

```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


`An event is a set of outcomes. You can describe certain events using random variables (x, distribution). the distribution of random variable function. `

#### Discrete random variable

<!-- generated-learning-note -->
This section prepares the information needed for **Discrete random variable**. Check variable definitions, units, coding, and missing values before moving to analysis.

`question: choose correct one out of four answers`

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
x<-0:5
y<-dbinom(x,5,1/4)
plot(x,y,col=2,type='h')
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


- using plot
`the probability of shooting is 0.02, what is the most likelihood of hit with 400 shootings`

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
k<-0:400
p<-dbinom(k,400,0.02)
plot(k,p,type='h',col=2)
plot(k,p,type='h',col=2,xlim=c(0,20))

dbinom(7,400,0.02)
dbinom(8,400,0.02)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



#### Exponent distribution

<!-- generated-learning-note -->
This subsection explains **Exponent distribution** and shows how it contributes to the overall Probability basics workflow.

`question: lifetime of a light (lamda=1/2000)`

<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r}
integrate(dexp,rate=1/2000,1000,Inf)$value
 
f<-function(x){dexp(x,rate=1/2000)}
integrate(f,1000,Inf) $value
 
1-pexp(1000,rate=1/2000)
 
mean(rexp(10000,rate=1/2000)>1000)
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.

#### Normal distribution plot

<!-- generated-learning-note -->
This section uses **Normal distribution plot** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.

`continue variable`

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
x<-seq(-3,3,0.01)
plot(x, dnorm(x,mean=0, sd=2),type="l",xlab="x",ylab = "f(x)", col=1,lwd=2,ylim=c(0,1))   #density function

lines(x, dnorm(x,mean=0, sd=1),lty=2, col=2,lwd=2)

lines(x, dnorm(x,mean=0, sd=0.5), lty=3,col=3,lwd=2)

exbeta<-c(expression(paste(mu,"=0,", sigma,"=2")), expression(paste(mu,"=0,",sigma,"=1")), expression(paste(mu,"=0,", sigma,"=0.5")))
legend("topright", exbeta, lty = c(1, 2,3),col=c(1,2,3),lwd=2)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
x<-seq(-3,3,0.01)
plot(x, dnorm(x,mean=-1, sd=1),type="l",xlab="x",ylab = "f(x)", col=1,lwd=2,ylim=c(0,0.6))
lines(x, dnorm(x,mean=0, sd=1),lty=2, col=2,lwd=2)
lines(x, dnorm(x,mean=1, sd=1), lty=3,col=3,lwd=2)
exbeta<-c(expression(paste(mu,"=-1,", sigma,"=1")), expression(paste(mu,"=0,",sigma,"=1")), expression(paste(mu,"=1,", sigma,"=1")))
legend("topright", exbeta, lty = c(1, 2,3),col=c(1,2,3),lwd=2)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

`question: solve sigma using nomoral distribution  `

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
sigma<-1
repeat{
sigma<-sigma+0.01
if (pnorm(200,160,sigma)-pnorm(120,160,sigma)<0.80) break
}
sigma
# alternative 
sigma<-1
while( pnorm(200,160,sigma)-pnorm(120,160,sigma)>=0.80){sigma<-sigma+0.01}
sigma
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


#### Distribution of random variable function

<!-- generated-learning-note -->
This section prepares the information needed for **Distribution of random variable function**. Check variable definitions, units, coding, and missing values before moving to analysis.

`qestion: x^2 and 2x distributions `

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
x<-c(-1,0,1,2,2.5)
weight<-c(0.2,0.1,0.1,0.3,0.3)
toss<-sample(x,10000,replace=T,weight)
table(toss^2)/length(toss^2)

table(2*toss)/length(2*toss)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

`quetsion: continous vairable density   `

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
x <- seq(0,5,0.01)
truth<-rep(0,length(x))
truth[0<=x&x<1]<-2/3
truth[1<=x&x<2]<-1/3

plot(density(abs(runif(1000000,-1,2))),main=NA, ylim=c(0,1),lwd=3,lty=3)

lines(x,truth,col="red",lwd=2)
legend("topright",c("True Density","Estimated Density"), col=c("red","black"),lwd=3,lty=c(1,3))
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### Join and margin probability

<!-- generated-learning-note -->
This subsection explains **Join and margin probability** and shows how it contributes to the overall Probability basics workflow.

`question: x is randomly selected from 1:4, y randomly select from x`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
p<-function(x,y) {
sim<-10000
t<-numeric(sim)
for (i in 1:sim) {
a<-sample(1:4,1)
b<-sample(1:a,1)
t[i]<-(a==x)&(b==y) }
mean(t)
}
PF<-matrix(0,4,4)
for (i in 1:4) {
for (j in 1:4) {
PF[i,j]<-p(i, j) } }
PF
apply(PF,1,sum)
apply(PF,2,sum)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

#### Multiple random variables plots

<!-- generated-learning-note -->
This section prepares the information needed for **Multiple random variables plots**. Check variable definitions, units, coding, and missing values before moving to analysis.

`2 discrete variables distribution `

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
x<-sample(1:4, 10000, replace=T, prob=c(1/4, 1/4, 1/4, 1/4))
y<-numeric(10000)
for(i in 1:10000) {
if(x[i]==1) {y[i]<-sample(1:4,1,replace=T, prob=c(1,0,0,0))}
if(x[i]==2) {y[i]<-sample(1:4,1,replace=T, prob=c(1/2,1/2,0,0))}
if(x[i]==3) {y[i]<-sample(1:4,1,replace=T, prob=c(1/3,1/3,1/3,0))}
if(x[i]==4) {y[i]<-sample(1:4,1,replace=T, prob=c(1/4,1/4,1/4,1/4))}
}
z1<-x+y
table(z1)/length(z1)
z2<-x*y
table(z2)/length(z2)
z3<-pmax(x,y)
table(z3)/length(z3)

z4<-x/y
table(z4)/length(z4)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


`two normal distributions  `
`X～N(0,1),Y～N(0,1), Z=X+Y, therefre Z~N(0,2)`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
Z<-function(n){
x<-seq(-4,4,0.01)
truth<-dnorm(x,0,sqrt(2))

plot(density(rnorm(n)+rnorm(n)),main="Density Estimate of the Normal Addition Model",ylim=c(0,0.4),lwd=2,lty=2)

lines(x,truth,col="red",lwd=2)
legend("topright",c("True","Estimated"),col=c("red","black"),lwd=2,lty=c(1,2))
}

Z(10000)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### Generate a circle using simulated random dots

<!-- generated-learning-note -->
This section prepares the information needed for **Generate a circle using simulated random dots**. Check variable definitions, units, coding, and missing values before moving to analysis.

`D={(x,y)|x^2 + y^2 <= 1}`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
x<-runif(10000,-1,1)
y<-runif(10000,-1,1)

a<-x[x^2+y^2<=1]
b<-y[x^2+y^2<=1]
plot(a,b,col=4)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


`oval`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
a<-3
b<-1
x<-runif(10000,-a,a)
y<-runif(10000,-b,b)
x1<-x[x^2/a^2+y^2/b^2<=1]
y1<-y[x^2/a^2+y^2/b^2<=1]
plot(x1,y1,col=3)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### Expectation

<!-- generated-learning-note -->
This subsection explains **Expectation** and shows how it contributes to the overall Probability basics workflow.

`discrete variable`
`question: the benefit of products is different.` 

<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
sim<-10000
t<-numeric(sim)

for (i in 1:sim) {
Y<-1500
X<-rexp(1,rate=1/10)

Y[1<X&X<=2]<-2000
Y[2<X&X<=3]<-2500
Y[3<X]<-3000
t[i]<-Y
}

mean(t)
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.

`continue variable`
 
#### Central Limit Theorem

<!-- generated-learning-note -->
This subsection explains **Central Limit Theorem** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
###Central Limit Theorem for Expotential distribution
layout(matrix(c(1,3,2,4 ),ncol=2))
r<-1000
lambda<-1/100

for (n in c(1,5,10,30)){
mu<-1/lambda
xbar<-numeric(r)
sxbar<-1/(sqrt(n)*lambda)

for(i in 1:r){
xbar[i]<-mean(rexp(n,rate=lambda))
}

hist(xbar,prob=T,main=paste('SampDist.Xbar,n=',n),col=gray(.8))
Npdf<-dnorm(seq(mu-3*sxbar,mu+3*sxbar,0.01),mu,sxbar)
lines(seq(mu-3*sxbar,mu+3*sxbar,0.01),Npdf,lty=2,col=2)
box()
}
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

 

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
#####The central limit theorem for uniform distribution
layout(matrix(c(1,3,2,4),ncol=2))
r<-10000
mu<-5
sigma<-10/sqrt(12)
for (n in c(1,5,10,30)){
xbar<-numeric(r)
sxbar<-sigma/sqrt(n)
for (i in 1:r){
xbar[i]<-mean(runif(n,0,10))}
hist(xbar,prob=T,main=paste('SampDist.Xbar,n=',n),col=gray(0.8),ylim=c(0,1/(sqrt(1*pi)*sxbar)))
XX<-seq(mu-3*sxbar,mu+3*sxbar,0.01)
Npdf<-dnorm(XX,mu,sxbar)
lines(XX,Npdf,lty=2,col=2)
box()}
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### Law of large numbers

<!-- generated-learning-note -->
This subsection explains **Law of large numbers** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
N <- 5000
set.seed(123) 

x <- sample(1:10, N,    replace = T)
s <- cumsum(x)    
r.avg <- s/(1:N)

options(scipen = 10)  

plot(r.avg, ylim=c(1, 10), type = "l", xlab = "Observations"
     ,ylab = "Probability", lwd = 2)
lines(c(0,N), c(5.5,5.5),col="red", lwd = 2)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### Empirical distribution

<!-- generated-learning-note -->
This subsection explains **Empirical distribution** and shows how it contributes to the overall Probability basics workflow.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
x<-c(-2,-1.2,1.5,2.3,3.5)
plot(ecdf(x),col=2)
abline(v=0,col=3)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


`question: three numbers are from N(2,9), and what is the prob of their mean >3?`

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
A<-matrix(rnorm(30000,2,3),10000,3)
mean(apply(A,1,mean)>3)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


#### Maximum likelihood estimate

<!-- generated-learning-note -->
This section develops **Maximum likelihood estimate**. Read the model specification together with its assumptions and the scale of each reported effect.

`question: eatimate mean and variance` 

<!-- generated-learning-note -->
**Purpose.** This block defines a reusable function. Separate its inputs, calculations, and returned value before applying it to new data.

```{r}
sample<-c(1.38, 3.96, -0.16, 8.12, 6.30, 2.61, -1.35, 0.03, 3.94, 1.11)
n<-length(sample)
muhat<-mean(sample)
sigsqhat<-sum((sample-muhat)^2)/n
muhat
sigsqhat

loglike<-function(theta){
a<--n/2*log(2*pi)-n/2*log(theta[2])-sum((sample-theta[1])^2)/(2*theta[2])
return(-a)
}
optim(c(2,2),loglike,method="BFGS")$par
```

<!-- generated-learning-note -->
**How to read the output.** Test the function with simple inputs whose expected answer is known before relying on it in a larger workflow.

#### t distribution, F distribution plots, and common distributions

<!-- generated-learning-note -->
This section uses **t distribution, F distribution plots, and common distributions** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
n<-30
x<-seq(-6,6,0.01)
y<-seq(-6,6,1)

Truth<-df(x,1,n)

plot(density(rt(10000,n)^2),main="PDF",ylim=c(0,1),lty=2,xlim=c(-6,6)) #simulation 

lines(x, dt(x,n), col=3) #t dist
 
lines(x, dchisq(x,2), col=4) #chisq dist
 
lines(x,Truth,col=2) #f dist
abline (v=0 ,col=7)  

 
points(y,dbinom(y, size=12, prob=0.2),col=1) #binomial dist
points(y,dpois(y, 6),col=2) #poisson dist

lines(y,dunif(-6:6,min=-6,max=6 ),col=5) #uniform

 
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

## Interpreting the Results

Work through the notation first, then use the R examples to connect definitions with numerical behavior. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
