---
title: "Missing Data Analysis: MAR and MNAR Sensitivity"
author: "Dian He"
date: "`r Sys.Date()`"
output: html_document
---

# 1. Objective

This example demonstrates:

1. Primary analysis under **MAR** using MMRM / MI.
2. Sensitivity analysis under **MNAR** using a pattern-mixture model with delta adjustment.

Clinical example:

- Drug vs Placebo
- HbA1c measured at Baseline, Week 4, Week 8, and Week 12
- Some patients discontinue because of adverse events
- Week 12 HbA1c may therefore be missing

# 2. Packages

```{r}
library(dplyr)
library(tidyr)
library(nlme)
library(mice)
library(emmeans)

set.seed(123)
```

# 3. Simulate Clinical Trial Data

```{r}
n <- 300

dat <- data.frame(
  ID = 1:n,
  TRT = sample(c("Placebo", "Drug"), n, replace = TRUE),
  Baseline = rnorm(n, 8.5, 0.8)
)

dat <- dat %>%
  mutate(
    Y4 = Baseline +
      ifelse(TRT == "Drug", -0.4, -0.1) +
      rnorm(n, 0, 0.35),

    Y8 = Baseline +
      ifelse(TRT == "Drug", -0.8, -0.2) +
      rnorm(n, 0, 0.35),

    Y12_true = Baseline +
      ifelse(TRT == "Drug", -1.1, -0.3) +
      rnorm(n, 0, 0.35),

    AE = rbinom(
      n,
      1,
      ifelse(TRT == "Drug", 0.30, 0.12)
    )
  )
```

# 4. Generate MAR Dropout

Missingness depends on observed information:

- Treatment
- Week 8 HbA1c
- Adverse event status

Therefore, this can still satisfy MAR.

```{r}
# Target approximately 10% Week 12 missingness.
# Calibrate the intercept while preserving the MAR mechanism.

target_missing <- 0.10

lp_without_intercept <-
  1.2 * dat$AE +
  0.35 * (dat$Y8 - 8) +
  0.3 * (dat$TRT == "Drug")

calibrate_fun <- function(intercept) {
  mean(plogis(intercept + lp_without_intercept)) - target_missing
}

dropout_intercept <- uniroot(
  calibrate_fun,
  interval = c(-10, 2)
)$root

drop_prob <- plogis(
  dropout_intercept + lp_without_intercept
)

dat$Dropout <- rbinom(n, 1, drop_prob)

dat$Y12 <- dat$Y12_true
dat$Y12[dat$Dropout == 1] <- NA

# Overall and treatment-specific missing rates
overall_missing_rate <- mean(is.na(dat$Y12))

missing_by_trt <- dat %>%
  group_by(TRT) %>%
  summarise(
    MissingRate = mean(is.na(Y12)),
    .groups = "drop"
  )

overall_missing_rate
missing_by_trt
```

# 5. Primary Analysis: MMRM Under MAR

```{r}
longdat <- dat %>%
  select(ID, TRT, Baseline, Y4, Y8, Y12) %>%
  pivot_longer(
    cols = c(Y4, Y8, Y12),
    names_to = "Visit",
    values_to = "HbA1c"
  )

longdat$Visit <- factor(
  longdat$Visit,
  levels = c("Y4", "Y8", "Y12")
)

mmrm_fit <- lme(
  HbA1c ~ TRT * Visit + Baseline,
  random = ~1 | ID,
  data = longdat,
  na.action = na.exclude,
  method = "REML"
)

summary(mmrm_fit)

# Week 12 Drug - Placebo contrast
mmrm_emm <- emmeans(
  mmrm_fit,
  ~ TRT | Visit
)

mmrm_contrasts <- contrast(
  mmrm_emm,
  method = "revpairwise"
)

# Request confidence intervals explicitly.
# Depending on the model, emmeans may name CI columns
# lower.CL/upper.CL or asymp.LCL/asymp.UCL.
mmrm_contrast_summary <- as.data.frame(
  summary(
    mmrm_contrasts,
    infer = c(TRUE, TRUE)
  )
)

mmrm_week12 <- mmrm_contrast_summary %>%
  filter(Visit == "Y12")

mmrm_week12
```

The coefficient `TRTDrug` in an MMRM with a `TRT * Visit` interaction is the treatment difference at the **reference visit**, not necessarily the Week 12 effect.

Therefore, the clinically relevant Week 12 effect should be obtained from a visit-specific contrast.

# 6. Primary Analysis: Multiple Imputation Under MAR

```{r}
mi_dat <- dat %>%
  select(TRT, Baseline, Y4, Y8, AE, Y12)

mi_dat$TRT <- factor(
  mi_dat$TRT,
  levels = c("Placebo", "Drug")
)

imp <- mice(
  mi_dat,
  m = 20,
  method = "pmm",
  seed = 123,
  printFlag = FALSE
)

fit_mi <- with(
  imp,
  lm(Y12 ~ TRT + Baseline)
)

mi_pooled <- summary(
  pool(fit_mi),
  conf.int = TRUE
)

mi_pooled

mi_week12 <- mi_pooled %>%
  filter(term == "TRTDrug") %>%
  transmute(
    Method = "MI under MAR",
    Estimate = estimate,
    SE = std.error,
    Lower95 = `2.5 %`,
    Upper95 = `97.5 %`
  )

mi_week12
```

The MI analysis is a Week 12 ANCOVA. Because Placebo is the reference group, `TRTDrug` directly estimates the adjusted **Drug - Placebo difference at Week 12**.

# 7. Compare MMRM and MI Week 12 Treatment Effects

The two methods should be compared using the **same Week 12 contrast**, rather than comparing the raw `TRTDrug` coefficient from the MMRM output.

```{r}
# Identify the confidence-interval column names returned by emmeans
lower_col <- intersect(
  c("lower.CL", "asymp.LCL"),
  names(mmrm_week12)
)[1]

upper_col <- intersect(
  c("upper.CL", "asymp.UCL"),
  names(mmrm_week12)
)[1]

if (is.na(lower_col) || is.na(upper_col)) {
  stop("Confidence interval columns were not found in the emmeans output.")
}

mmrm_compare <- data.frame(
  Method = "MMRM under MAR",
  Estimate = mmrm_week12$estimate,
  SE = mmrm_week12$SE,
  Lower95 = mmrm_week12[[lower_col]],
  Upper95 = mmrm_week12[[upper_col]]
)

comparison_table <- bind_rows(
  mmrm_compare,
  mi_week12
)

comparison_table
```

Interpretation:

- **MMRM** uses all available repeated outcomes and estimates the Week 12 Drug - Placebo contrast from the longitudinal model.
- **MI** imputes missing Week 12 outcomes using observed information and then fits a Week 12 ANCOVA.
- Both are MAR-based approaches, but the estimates do not have to be exactly identical because the model structures differ.
- Similar estimates support consistency of the primary analysis.

# 8. MNAR Sensitivity Analysis: Delta Adjustment

Under MNAR, assume that patients with missing Week 12 HbA1c have outcomes worse than predicted under MAR.

For example:

\[
Y_{MNAR} = Y_{MAR} + \delta
\]

where a positive delta means higher, and therefore worse, HbA1c.

```{r}
run_delta <- function(delta, imp_object, original_data) {

  missing_y12 <- is.na(original_data$Y12)

  estimates <- numeric(imp_object$m)
  variances <- numeric(imp_object$m)

  for (i in 1:imp_object$m) {

    temp <- complete(imp_object, i)

    # Keep Placebo as the reference group
    temp$TRT <- factor(
      temp$TRT,
      levels = c("Placebo", "Drug")
    )

    # MNAR delta adjustment:
    # higher HbA1c is worse, so positive delta is unfavorable
    # to patients with originally missing Week 12 outcomes
    temp$Y12[missing_y12] <-
      temp$Y12[missing_y12] + delta

    fit <- lm(
      Y12 ~ TRT + Baseline,
      data = temp
    )

    estimates[i] <- unname(coef(fit)["TRTDrug"])

    vc <- vcov(fit)

    if (!"TRTDrug" %in% rownames(vc)) {
      stop(
        "TRTDrug coefficient was not found. ",
        "Check treatment factor levels."
      )
    }

    variances[i] <- vc["TRTDrug", "TRTDrug"]
  }

  # Rubin's rules
  Qbar <- mean(estimates)
  Ubar <- mean(variances)
  B <- var(estimates)

  total_var <-
    Ubar + (1 + 1 / imp_object$m) * B

  SE <- sqrt(total_var)

  data.frame(
    Delta = delta,
    Estimate = Qbar,
    SE = SE,
    Lower95 = Qbar - 1.96 * SE,
    Upper95 = Qbar + 1.96 * SE
  )
}
```

# 9. Test Several MNAR Assumptions

```{r}
delta_values <- c(
  0,
  0.2,
  0.4,
  0.6,
  0.8,
  1.0
)

results <- bind_rows(
  lapply(
    delta_values,
    run_delta,
    imp_object = imp,
    original_data = mi_dat
  )
)

results
```

Interpretation:

The treatment coefficient is **Drug - Placebo** because Placebo is explicitly set as the reference group. For HbA1c, a negative estimate favors Drug.

- `Delta = 0`: MAR assumption
- `Delta = 0.2`: missing HbA1c values are assumed to be 0.2% worse than MAR predictions
- `Delta = 0.5`: 0.5% worse
- Larger delta values represent increasingly unfavorable MNAR assumptions

# 10. Interpretation

The overall strategy is:

\[
\text{Primary Analysis}
\rightarrow
\text{MMRM or MI under MAR}
\]

followed by:

\[
\text{Sensitivity Analysis}
\rightarrow
\text{Pattern-Mixture Model + Delta Adjustment}
\]

If the treatment conclusion remains similar across clinically plausible delta values, the result is considered robust to departures from MAR.

If the conclusion changes only after a sufficiently large delta, that value can be interpreted as a **tipping point**.

# 11. Key Takeaway

**MAR allows missingness to depend on observed covariates and prior outcomes.**

For the primary analysis, MMRM or MI can therefore be used under a reasonable MAR assumption.

Because MAR cannot be verified from the observed data alone, MNAR sensitivity analyses such as **pattern-mixture models with delta adjustment** are used to assess robustness.
