---
title: "Matching and Weighting analyses"
output: html_document
date: '2023-05-15'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1) Standardized `weighting` and propensity score analysis; 2) Stratification analysis and `matching` are used to calculate the causal effect..

### Matching w/o stratification

Using `matchthem` function for imputed data. 

It also can apply to 1 imputed dataset if letting m=1.


```{r}
set.seed(123)
# load dataset
library(mice)
library(MatchThem)
library(tidyverse)
load("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\osteoarthritis.RData")

# check NA,
# a quick view data summary
summary(osteoarthritis)
md.pattern(osteoarthritis)
```

- Imputing the missing data, produces the 5 imputed datasets.

  
```{r}
imputed.datasets <- mice(osteoarthritis, m = 1)  #you can impute a no missing variable

# Get completed data sets
CDF = complete(imputed.datasets, 'long')
md.pattern(CDF)
```

```{r}
imputed.datasets <- mice(CDF, m = 2)
```


- Matching the imputed datasets
```{r}
# Matching the imputed datasets on the propensity score
matched.datasets <- matchthem(OSP ~ AGE + SEX + BMI + RAC + SMK,
                              datasets = imputed.datasets,
                              approach = 'within',
                              method = 'nearest',     #k nearest method
                              caliper = 0.05,
                              ratio = 1)
```
- Assessing balance on the matched datasets
```{r}
library(cobalt)
bal.tab(matched.datasets, stats = c('m', 'ks'),
        imp.fun = 'max')
```

Differneces are close to zero. 

- Analyzing the matched datasets
```{r}
library(survey)
matched.models <- with(matched.datasets,
                       svyglm(KOA ~ OSP, family = quasibinomial()),
                       cluster = TRUE)
```

no svydesign() or weights arguments need to be specified as these are automatically supplied by with(). 

- Pooling the causal effect estimates
```{r}
# we must pool the estimated models using pool(), containing the models we fit above
matched.results <- pool(matched.models)

summary(matched.results, conf.int = TRUE)
```

#### Stratification on the propensity score 

```{r}
library("readxl")
nhefs <- read.csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\nhefs.csv")

fit3 <- glm(qsmk ~ sex + race + age + I(age*age) + as.factor(education)
            + smokeintensity + I(smokeintensity*smokeintensity) + smokeyrs
            + I(smokeyrs*smokeyrs) + as.factor(exercise) + as.factor(active)
            + wt71 + I(wt71*wt71), data=nhefs, family=binomial())
summary(fit3)
```

```{r}
nhefs$ps <- predict(fit3, nhefs, type="response")
```


```{r}
# calculation of deciles
nhefs$ps.dec <- cut(nhefs$ps, 
                    breaks=c(quantile(nhefs$ps, probs=seq(0,1,0.1))),
                    labels=seq(1:10),
                    include.lowest=TRUE)

```


```{r}
# regression on PS deciles, not allowing for effect modification
fit.psdec <- glm(wt82_71 ~ qsmk + as.factor(ps.dec), data = nhefs)
summary(fit.psdec)
```

```{r}
confint.lm(fit.psdec)
```


- Standardized the outcome

```{r}
# regression on the propensity score (linear term) p.qsmk instead by ps
model6 <- glm(wt82_71 ~ qsmk + ps, data = nhefs)
summary(model6)
```


```{r}
# standardization on the propensity score
# (step 1) create two new datasets, one with all treated and one with all untreated
treated <- nhefs
  treated$qsmk <- 1

untreated <- nhefs
  untreated$qsmk <- 0

# (step 2) predict values for everyone in each new dataset based on above model
treated$pred.y <- predict(model6, treated)
untreated$pred.y <- predict(model6, untreated)

# (step 3) compare mean weight loss had all been treated vs. that had all been untreated
mean1 <- mean(treated$pred.y, na.rm = TRUE)
mean0 <- mean(untreated$pred.y, na.rm = TRUE)
mean1 - mean0
```



### Weighting using ps

- Imputing the missing data, produces the 5 imputed datasets.

- Weighting the imputed datasets.

```{r}
weighted.datasets <- weightthem(OSP ~ AGE + SEX + BMI + RAC + SMK,
                                datasets = imputed.datasets,
                                approach = 'across',
                                method = 'ps',         #propensity score
                                estimand = 'ATM')
# and the output of the calls applied to each imputed dataset.
```

- Assessing balance on the weighted datasets
```{r}
bal.tab(weighted.datasets, stats = c('m', 'ks'),
        imp.fun = 'max')
```
- Analyzing the weighted datasets
```{r}
library(survey)
weighted.models <- with(weighted.datasets,
                        svyglm(KOA ~ OSP, family = quasibinomial()))
#doesn't work for only 1 imputation
```

- Pooling the causal effect estimates
```{r}
weighted.results <- pool(weighted.models)
summary(weighted.results, conf.int = TRUE)
```



#### PSM using matchit function

Table 1 in RHC data 
```{r}
# load the dataset
ObsData <- read.csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\obsdata.csv", header = TRUE)
# add column for outcome Y: length of stay 
# Y = date of discharge - study admission date
# Y = date of death - study admission date if date of discharge not available
ObsData$Y <- ObsData$dschdte - ObsData$sadmdte
ObsData$Y[is.na(ObsData$Y)] <- ObsData$dthdte[is.na(ObsData$Y)] - 
  ObsData$sadmdte[is.na(ObsData$Y)]
# remove outcomes we are not examining in this example
ObsData <- dplyr::select(ObsData, 
                         !c(dthdte, lstctdte, dschdte, death, t3d30, dth30, surv2md1))
# remove unnecessary and problematic variables 
ObsData <- dplyr::select(ObsData, 
                         !c(sadmdte, ptid, X, adld3p, urin1, cat2))

# convert all categorical variables to factors 
factors <- c("cat1", "ca", "cardiohx", "chfhx", "dementhx", "psychhx", 
             "chrpulhx", "renalhx", "liverhx", "gibledhx", "malighx", 
             "immunhx", "transhx", "amihx", "sex", "dnr1", "ninsclas", 
             "resp", "card", "neuro", "gastr", "renal", "meta", "hema", 
             "seps", "trauma", "ortho", "race", "income")
ObsData[factors] <- lapply(ObsData[factors], as.factor)
# convert our treatment A (RHC vs. No RHC) to a binary variable
ObsData$A <- ifelse(ObsData$swang1 == "RHC", 1, 0)
ObsData <- dplyr::select(ObsData, !swang1)
# Categorize the variables to match with the original paper
ObsData$age <- cut(ObsData$age,breaks=c(-Inf, 50, 60, 70, 80, Inf),right=FALSE)
ObsData$race <- factor(ObsData$race, levels=c("white","black","other"))
ObsData$sex <- as.factor(ObsData$sex)
ObsData$sex <- relevel(ObsData$sex, ref = "Male")
ObsData$cat1 <- as.factor(ObsData$cat1)
levels(ObsData$cat1) <- c("ARF","CHF","Other","Other","Other",
                          "Other","Other","MOSF","MOSF")
ObsData$ca <- as.factor(ObsData$ca)
levels(ObsData$ca) <- c("Metastatic","None","Localized (Yes)")
ObsData$ca <- factor(ObsData$ca, levels=c("None",
                                          "Localized (Yes)","Metastatic"))
# Rename variables
names(ObsData) <- c("Disease.category", "Cancer", "Cardiovascular", 
                    "Congestive.HF", "Dementia", "Psychiatric", "Pulmonary", 
                    "Renal", "Hepatic", "GI.Bleed", "Tumor", 
                    "Immunosupperssion", "Transfer.hx", "MI", "age", "sex", 
                    "edu", "DASIndex", "APACHE.score", "Glasgow.Coma.Score", 
                    "blood.pressure", "WBC", "Heart.rate", "Respiratory.rate", 
                    "Temperature", "PaO2vs.FIO2", "Albumin", "Hematocrit", 
                    "Bilirubin", "Creatinine", "Sodium", "Potassium", "PaCo2", 
                    "PH", "Weight", "DNR.status", "Medical.insurance", 
                    "Respiratory.Diag", "Cardiovascular.Diag", 
                    "Neurological.Diag", "Gastrointestinal.Diag", "Renal.Diag",
                    "Metabolic.Diag", "Hematologic.Diag", "Sepsis.Diag", 
                    "Trauma.Diag", "Orthopedic.Diag", "race", "income", 
                    "Y", "A")
```

```{r}
baselinevars <- names(dplyr::select(ObsData, 
                                    !c(A,Y)))

require(tableone)
tab0 <- CreateTableOne(vars = c("age", "sex", "race", "Disease.category", "Cancer"),
                       data = ObsData, 
                       strata = "A", 
                       test = T)
print(tab0, showAllLevels = FALSE, )
```


Matching caliper setup
```{r}
set.seed(111)
require(MatchIt)
ps.formula <- as.formula(paste("A~", 
                               paste(baselinevars, collapse = "+")))
PS.fit <- glm(ps.formula,family="binomial", 
              data=ObsData)
ObsData$PS <- predict(PS.fit, 
                      newdata = ObsData, type="response") 

# propensity score
logitPS <-  -log(1/ObsData$PS - 1)  

match.obj <- matchit(ps.formula, data =ObsData,
                     distance = ObsData$PS,
                     method = "nearest", replace=FALSE,
                     ratio = 1,
                     caliper = .2*sd(logitPS))  #matching caliper 
```

PSM diagnostics
```{r}
require(cobalt)
bal.plot(match.obj,  
         var.name = "distance", 
         which = "both", 
         type = "histogram",  
         mirror = TRUE)
```
Plot 
```{r}
bal.tab(match.obj, un = TRUE, 
        thresholds = c(m = .1))

love.plot(match.obj, binary = "std", 
          thresholds = c(m = .1))  
```

Treatment effect calculation
```{r}
library(Publish)
matched.data <- match.data(match.obj)   

fit.matched <- glm(Y~A,
                   family=gaussian,  
                   data = matched.data)  
publish(fit.matched)

```


### Propensity scores analysis

`not inverse probability`
```{r}
library("readxl")
nhefs <- read.csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\nhefs.csv")

fit3 <- glm(qsmk ~ sex + race + age + I(age*age) + as.factor(education)
            + smokeintensity + I(smokeintensity*smokeintensity) + smokeyrs
            + I(smokeyrs*smokeyrs) + as.factor(exercise) + as.factor(active)
            + wt71 + I(wt71*wt71), data=nhefs, family=binomial())
summary(fit3)
```

```{r}
nhefs$ps <- predict(fit3, nhefs, type="response")

summary(nhefs$ps[nhefs$qsmk==0])
summary(nhefs$ps[nhefs$qsmk==1])
```


```{r}
# alternative plot with histograms
nhefs <- nhefs %>% mutate(qsmklabel = ifelse(qsmk == 1,
                       yes = 'Quit Smoking 1971-1982',
                       no = 'Did Not Quit Smoking 1971-1982'))
ggplot(nhefs, aes(x = ps, fill = as.factor(qsmk), color = as.factor(qsmk))) +
  geom_histogram(alpha = 0.3, position = 'identity', bins=15) +
  facet_grid(as.factor(qsmk) ~ .) +
  xlab('Probability of Quitting Smoking During Follow-up') +
  ggtitle('Propensity Score Distribution by Treatment Group') +
  scale_fill_discrete('') +
  scale_color_discrete('') +
  theme(legend.position = 'bottom', legend.direction = 'vertical')
```



### Differences between Matching and Regression

Compare to regression
```{r}
model=glm(KOA ~OSP+AGE + SEX + BMI + RAC + SMK,family = binomial,data=CDF)
summary(model)
```

Regression has more bias, while Matching has more variance. Matching may be more conservative than regression. Meanwhile, for Matching, there would be no need to make any functional form assumptions. The relationships between the covariates, the treatment and the outcome could be highly non-linear, as long as there is common support and the CIA holds. 

At last, `are the confounders directly available as variables you can condition on?` 










