---
title: "Sample size calculation"
description: "A practical R learning guide to Sample size calculation, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Sample Size & Power"
topic: "Sample size calculation"
type: article
tags: ["Sample Size & Power", "Biostatistics", "Sample", "size", "calculation"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Sample-size planning balances statistical sensitivity, scientific relevance, feasibility, and uncertainty in design assumptions.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Sample size calculation**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Sample size calculation

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Sample size calculation** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Sample size calculation

<!-- generated-learning-note -->
This subsection explains **Sample size calculation** and shows how it contributes to the overall Sample size calculation workflow.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
#given proportion 
px3=0.11
px4=0.07
px5=0.08
px6=0.06
px7=0.05

###################

  Outcome0 = NULL
  
  for (t in seq(350, 600, by=2)){  # change the possible sample size graudally from 400 to 700.  
    
    n=t
    count=0
    M=500   # times of simulation
    for (i in 1:M){
      
      # for a given sample size (400), for the first simulation,
      # we use rbinom function simulate the contingencey table as below with above given proportions,
      # and calculate the p value by using Chi square test for the simulation
      x3=rbinom(1,n,px3)
      x4=rbinom(1,n,px4)
      x5=rbinom(1,n,px5)
      x6=rbinom(1,n,px6)
      x7=rbinom(1,n,px7)
            
      data=matrix(c(x3,n-x3,x4,n-x4,x5,n-x5,x6,n-x6,x7,n-x7 ),ncol=2,byrow=T)   
      
      # if the p value of this simulation is less than 0.05, it means we get the significant result for the given this sample size this time. it means we detected the difference in this simulation experiment when the difference is true. 
      pv=prop.test(data)$p.value      
      
      count=as.numeric(pv<0.05)+count  #sum of pv<0.05
    }  #end loop of p value 
    power0=count/M 
    
    temp <- data.frame(size=t,power=power0)
    Outcome0 =  rbind(Outcome0, temp)
  }  #end loop of power from different sample size
  
  # generate a new variable named "power_loess" by loess method because the curve of sample size and power is not enough smooth
  power_loess <- round(predict(loess(Outcome0$power ~ Outcome0$size,data=Outcome0,span=0.6)),3)
  Outcome0 =  data.frame(Outcome0, power_loess)
  
  # plot a line chart of the power and sample size, and smooth the curve by loess method
  plot (Outcome0$size, Outcome0$power,type = "n", ylab = "Power",xlab="Sample size",
        main = expression(paste("Power analysis for GDM"))) 
  abline(h=0.9,col='red',lwd=2,lty=2)
  abline(h=0.85, col='red',lwd=2,lty=2)
  abline(h=0.8,col='red',lwd=2,lty=2)
  lines(Outcome0$size,power_loess,col="blue",lwd=2)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

## Interpreting the Results

Inspect how power or required sample size changes when the effect size, variability, alpha level, or dropout rate changes. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
