---
title: "Standard Operating Procedure for Real-World Evidence Studies"
author: "[Author / Study Team]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# 1. Purpose

This SOP provides a standard framework for the design, conduct, analysis, and reporting of observational real-world evidence (RWE) studies using routinely collected data.

It is intended for studies using:

- electronic health records (EHR),
- insurance claims,
- pharmacy data,
- registries,
- laboratory databases,
- administrative data,
- population-based cohorts,
- or linked real-world data sources.

The primary objective is to estimate treatment, exposure, or comparative-effectiveness effects as validly as possible while addressing major sources of bias inherent in nonrandomized data.

# 2. Research Question and Causal Objective

The research question should be specified before analysis.

**General form:**

> Among [target population], what is the effect of [treatment/exposure A] compared with [treatment/exposure B] on [outcome] during [follow-up period]?

The study should define:

- target population,
- treatment/exposure strategies,
- comparator,
- outcome,
- time zero / index date,
- follow-up period,
- censoring rules,
- effect measure,
- and target estimand.

When causal interpretation is intended, the analysis should emulate the key elements of a hypothetical target trial as closely as possible.

# 3. Target Trial Framework

Whenever applicable, prespecify the following target-trial components:

1. eligibility criteria;
2. treatment strategies;
3. treatment assignment procedure;
4. time zero;
5. follow-up;
6. outcome;
7. causal contrast / estimand;
8. analysis method.

The observational emulation should align eligibility, treatment assignment, and start of follow-up at the same time point to minimize time-related biases.

# 4. Data Source and Data Provenance

Describe:

- data source(s),
- calendar period,
- population coverage,
- coding systems,
- linkage procedures,
- laboratory availability,
- medication capture,
- mortality capture,
- and data refresh/cutoff date.

Important limitations of each source should be documented.

Examples include:

- incomplete capture of outside care,
- lack of over-the-counter medication use,
- missing laboratory measurements,
- changes in coding practices,
- incomplete death information,
- or limited medication adherence information.

# 5. Study Population and Cohort Definition

Define inclusion and exclusion criteria before outcome analysis.

Typical criteria may include:

- age range,
- diagnosis or disease status,
- minimum baseline observation period,
- continuous enrollment when using claims data,
- availability of treatment/exposure information,
- no prior occurrence of an incident outcome if required,
- and sufficient follow-up eligibility.

A cohort flow diagram should document the number of individuals excluded at each step.

# 6. Index Date / Time Zero

The index date is the date on which:

- eligibility is satisfied,
- treatment/exposure strategy is assigned,
- baseline covariates are defined,
- and follow-up begins.

These elements should be aligned as closely as possible.

Incorrect alignment can produce:

- immortal-time bias,
- prevalent-user bias,
- selection bias,
- and temporal ambiguity.

# 7. Exposure or Treatment Strategy Definition

Specify:

- exposure/treatment definition,
- comparator,
- coding algorithm,
- washout or baseline period,
- initiation date,
- treatment switching rules,
- discontinuation rules,
- grace periods,
- adherence definition if used,
- and whether exposure is fixed or time-varying.

For comparative-effectiveness studies, a **new-user active-comparator design** should be considered when feasible because it often improves comparability and temporal clarity.

# 8. Outcome Definition

Specify:

- primary outcome,
- secondary outcomes,
- coding algorithm,
- event date,
- validation status,
- adjudication procedure if available,
- and outcome ascertainment window.

Outcome definitions should be based on clinically meaningful and reproducible algorithms.

Where possible, use validated diagnosis/procedure algorithms.

# 9. Baseline Period and Covariates

The baseline period should be prespecified, for example:

- 180 days,
- 365 days,
- or another justified interval before index.

Baseline covariates may include:

- demographics,
- disease severity,
- prior diagnoses,
- prior procedures,
- prior medication use,
- laboratory values,
- vital signs,
- healthcare utilization,
- lifestyle factors,
- socioeconomic characteristics,
- and relevant comorbidities.

Covariates should be defined using information available before or at time zero.

# 10. Confounding Control

Because treatment or exposure is not randomized, confounding must be explicitly addressed.

## 10.1 Confounder Selection

Potential confounders should be selected using:

- clinical knowledge,
- prior literature,
- temporal ordering,
- causal diagrams / DAGs,
- and known predictors of treatment and outcome.

Variables should not be included solely because they are statistically significant or imbalanced.

## 10.2 Confounding by Indication

Special attention should be given to **confounding by indication**.

For example, patients receiving a more intensive therapy may be systematically sicker than patients receiving the comparator treatment.

Important markers of disease severity should therefore be identified and included when available.

## 10.3 Post-Exposure Variables

Post-exposure variables should not automatically be adjusted for.

If the target estimand is the **total treatment effect**, variables affected by treatment should generally not be included in the baseline confounder set.

If the objective instead concerns a direct or hypothetical effect, the estimand and analytic method should be explicitly defined.

# 11. Descriptive Analysis

Baseline characteristics will be summarized by treatment/exposure group.

Continuous variables:

- mean and SD,
- or median and IQR.

Categorical variables:

- frequency and percentage.

Standardized mean differences (SMDs) should be used to assess baseline imbalance.

An absolute SMD below approximately 0.10 is commonly used as a practical indicator of acceptable measured balance.

Baseline imbalance alone does not determine whether a variable is a confounder.

# 12. Propensity Score Estimation

When propensity-score methods are used, estimate:

\[
PS(X) = P(A=1 \mid X)
\]

where \(A\) is treatment/exposure and \(X\) is the set of measured baseline covariates.

Possible approaches include:

- logistic regression,
- generalized boosted models,
- machine-learning propensity models,
- high-dimensional propensity scores.

Example:

```{r ps-example, eval=FALSE}
ps_model <- glm(
  treatment ~ age + sex + bmi + hba1c +
    prior_cvd + ckd + smoking,
  family = binomial(),
  data = analysis_data
)

analysis_data$ps <- predict(
  ps_model,
  type = "response"
)
```

# 13. Propensity Score Adjustment Strategies

Common methods include:

## 13.1 Matching

Examples:

- 1:1 nearest-neighbor matching,
- variable-ratio matching,
- caliper matching.

Balance should be assessed after matching.

## 13.2 IPTW

For estimation of the average treatment effect:

\[
w_i =
\begin{cases}
1/PS_i, & A_i=1 \\
1/(1-PS_i), & A_i=0
\end{cases}
\]

Stabilized weights may be preferred.

Extreme weights should be evaluated and may be truncated or trimmed using prespecified rules.

## 13.3 Overlap Weighting

Overlap weighting may be considered when treatment groups have limited overlap or when the clinical target population is the region of common support.

# 14. Positivity and Overlap

Assess whether patients in all covariate strata have a reasonable probability of receiving each treatment.

Evaluate:

- propensity-score distributions,
- common support,
- extreme PS values,
- extreme weights,
- effective sample size,
- and overlap plots.

Serious positivity violations should be reported and may require:

- restriction,
- trimming,
- alternative weighting,
- or redefining the target population.

# 15. Covariate Balance Assessment

After matching or weighting, assess balance using:

- standardized mean differences,
- variance ratios when useful,
- propensity-score overlap,
- and love plots.

Example:

```{r balance-example, eval=FALSE}
library(cobalt)

bal.tab(
  treatment ~ age + sex + bmi + hba1c +
    prior_cvd + ckd + smoking,
  data = analysis_data,
  weights = analysis_data$weight,
  un = TRUE,
  thresholds = c(m = 0.10)
)
```

Good measured balance does not establish the absence of unmeasured confounding.

# 16. Primary Outcome Analysis

The primary analysis method depends on outcome type.

## 16.1 Time-to-Event Outcome

Use Cox proportional hazards regression or another prespecified survival model.

```{r cox-example, eval=FALSE}
library(survival)

fit <- coxph(
  Surv(followup_time, event) ~ treatment,
  data = analysis_data,
  weights = weight,
  robust = TRUE
)
```

Report:

- HR,
- 95% CI,
- and P-value when appropriate.

## 16.2 Binary Outcome

Possible methods:

- logistic regression,
- log-binomial regression,
- modified Poisson regression.

Report:

- OR,
- RR,
- or RD,
depending on the estimand.

## 16.3 Continuous Outcome

Possible methods:

- linear regression,
- weighted regression,
- mixed models for repeated measurements.

## 16.4 Count Outcome

Possible methods:

- Poisson regression,
- negative binomial regression.

# 17. Conditional vs Marginal Effect

The analysis should distinguish between:

- **conditional effects**, typically obtained from multivariable regression,
- and **marginal population-level effects**, often estimated using standardization or weighting.

For example:

- multivariable Cox regression usually estimates a conditional HR,
- IPTW more naturally targets a marginal population-level effect.

The estimand should be specified in advance.

# 18. Follow-up and Censoring

Follow-up should begin at the index date and end at the earliest of:

- outcome,
- death,
- treatment discontinuation when required by estimand,
- treatment switching,
- loss to follow-up,
- disenrollment,
- end of data availability,
- or administrative study end.

Censoring rules depend on the target estimand.

For example:

- **intention-to-treat-like / treatment-policy-like analysis** may ignore later switching,
- **as-treated / per-protocol-like analysis** may censor at switching or discontinuation.

Potential informative censoring should be evaluated.

Inverse probability of censoring weighting may be considered when appropriate.

# 19. Missing Data

Summarize:

- amount of missingness,
- variables affected,
- missingness by treatment group,
- and likely missing-data mechanisms.

Possible methods include:

- complete-case analysis,
- multiple imputation,
- missing-indicator methods only when justified,
- model-based approaches.

The primary method should be prespecified.

Sensitivity analyses should evaluate departures from missing-at-random assumptions when relevant.

# 20. Measurement Error and Misclassification

Evaluate the potential for:

- exposure misclassification,
- outcome misclassification,
- inaccurate diagnosis codes,
- incomplete medication capture,
- imperfect adherence information,
- laboratory measurement error,
- and date uncertainty.

When possible:

- use validated coding algorithms,
- require repeated codes,
- use confirmation windows,
- or perform quantitative bias analysis.

# 21. Model Diagnostics

Depending on the primary model, assess:

- proportional hazards,
- linearity,
- goodness-of-fit,
- influential observations,
- sparse cells,
- multicollinearity,
- and model convergence.

For Cox models:

```{r ph-example, eval=FALSE}
ph_test <- cox.zph(fit)
print(ph_test)
plot(ph_test)
```

# 22. Competing Risks

When competing events are clinically important, consider:

- cause-specific Cox regression,
- Fine-Gray subdistribution hazards models,
- and cumulative-incidence functions.

Choice of method should reflect the scientific question.

# 23. Effect Modification and Subgroup Analyses

Prespecified subgroups may include:

- age,
- sex,
- disease severity,
- kidney function,
- prior cardiovascular disease,
- or other clinically justified factors.

Effect modification should preferably be tested with interaction terms.

Exploratory subgroup analyses should be labeled as such.

# 24. Sensitivity Analyses

Common RWE sensitivity analyses include:

1. alternative exposure definitions;
2. alternative outcome definitions;
3. alternative baseline windows;
4. alternative confounder sets;
5. propensity-score matching;
6. IPTW;
7. overlap weighting;
8. multivariable regression;
9. weight trimming/truncation;
10. alternative censoring rules;
11. as-treated analysis;
12. treatment-policy-like analysis;
13. multiple imputation;
14. negative-control outcomes or exposures;
15. lagged exposure analyses;
16. competing-risk analyses;
17. inverse probability of censoring weighting;
18. assessment of unmeasured confounding;
19. quantitative bias analysis;
20. E-value analysis when appropriate.

Results should be compared across methods for:

- direction,
- magnitude,
- precision,
- and clinical interpretation.

# 25. Negative Controls and Falsification Analyses

When feasible, use:

- negative-control outcomes,
- negative-control exposures,
- falsification endpoints.

These analyses can help detect residual confounding, surveillance bias, or systematic error.

A null negative-control result does not prove that confounding is absent.

# 26. Unmeasured Confounding

Residual and unmeasured confounding should be explicitly acknowledged.

Possible approaches include:

- E-values,
- quantitative bias analysis,
- negative controls,
- instrumental-variable analysis when assumptions are defensible,
- external validation data.

No single method can prove the absence of unmeasured confounding.

# 27. Generalizability and Transportability

Describe the population to which results apply.

Assess:

- demographic representation,
- healthcare setting,
- insurance status,
- geography,
- disease severity,
- and calendar period.

The analytic population after matching or weighting may differ from the original source population.

# 28. Statistical Inference

Unless otherwise specified:

- tests will be two-sided,
- alpha = 0.05,
- and 95% confidence intervals will be reported.

The primary analysis should be distinguished from:

- secondary analyses,
- sensitivity analyses,
- exploratory analyses.

Multiplicity should be acknowledged when many outcomes or subgroup analyses are conducted.

# 29. Reporting of Results

Recommended outputs include:

## Table 1
Baseline characteristics before adjustment.

## Table 2
Baseline characteristics after matching or weighting.

## Table 3
Primary and sensitivity treatment-effect estimates.

| Method | Effect Estimate | 95% CI | P-value |
|---|---:|---:|---:|
| Primary analysis | | | |
| Propensity-score matching | | | |
| IPTW | | | |
| Multivariable regression | | | |
| Sensitivity analysis | | | |

## Figures

Recommended figures include:

1. cohort flow diagram;
2. propensity-score overlap plot;
3. love plot;
4. Kaplan-Meier or cumulative-incidence plot if relevant;
5. forest plot comparing primary and sensitivity estimates.

# 30. Interpretation

Interpretation should focus on:

- effect size,
- precision,
- clinical relevance,
- consistency across analyses,
- and assumptions required for causal interpretation.

Observational RWE findings should generally be described as adjusted associations unless sufficient design and causal assumptions support a causal interpretation.

# 31. Key Limitations

Common RWE limitations include:

- confounding by indication,
- unmeasured confounding,
- residual confounding,
- exposure misclassification,
- outcome misclassification,
- measurement error,
- missing data,
- incomplete medication adherence information,
- selection bias,
- immortal-time bias,
- informative censoring,
- positivity violations,
- model misspecification,
- data-source limitations,
- and limited generalizability.

# 32. Quality Control and Reproducibility

All key analysis steps should be reproducible.

Key variables requiring validation include:

- patient identifier,
- treatment/exposure,
- index date,
- baseline period,
- outcome,
- outcome date,
- follow-up,
- censoring,
- and major confounders.

Whenever feasible:

- analysis code should be version controlled,
- key outputs should be independently reviewed,
- cohort counts should be reconciled,
- and treatment/outcome algorithms should be validated.

# 33. Documentation of Deviations

Any material deviation from the prespecified protocol or SAP should be documented with:

- original plan,
- revised method,
- rationale,
- timing,
- and possible impact on interpretation.

# Appendix A. Minimal RWE Analysis Workflow

```{r minimal-workflow, eval=FALSE}
library(dplyr)
library(MatchIt)
library(cobalt)
library(survival)

# 1. Define analytic cohort
analysis_data <- raw_data %>%
  filter(
    eligible == 1
  )

# 2. Propensity score
ps_model <- glm(
  treatment ~ age + sex + bmi + hba1c +
    prior_cvd + ckd + smoking,
  family = binomial(),
  data = analysis_data
)

analysis_data$ps <- predict(
  ps_model,
  type = "response"
)

# 3. IPTW
analysis_data <- analysis_data %>%
  mutate(
    treat_num = as.integer(treatment == "TreatmentA"),
    iptw = ifelse(
      treat_num == 1,
      1 / ps,
      1 / (1 - ps)
    )
  )

# 4. Weighted outcome model
fit_iptw <- coxph(
  Surv(followup_time, event) ~ treatment,
  data = analysis_data,
  weights = iptw,
  robust = TRUE
)

summary(fit_iptw)
```

# Appendix B. Pre-Analysis Checklist

Before final analysis, confirm:

- [ ] Research question is finalized
- [ ] Target population is defined
- [ ] Target trial elements are specified
- [ ] Data source and study period are documented
- [ ] Eligibility criteria are finalized
- [ ] Index date is defined
- [ ] Treatment/exposure strategies are defined
- [ ] Comparator is defined
- [ ] Baseline period is defined
- [ ] Outcome algorithm is finalized
- [ ] Follow-up and censoring rules are defined
- [ ] Primary estimand is specified
- [ ] Primary confounder set is prespecified
- [ ] Post-exposure variables are classified
- [ ] Propensity-score method is specified if used
- [ ] Balance and positivity criteria are specified
- [ ] Missing-data method is specified
- [ ] Sensitivity analyses are prespecified
- [ ] Negative controls are considered
- [ ] Competing-risk strategy is specified if relevant
- [ ] Subgroup analyses are prespecified
- [ ] Data cutoff date is finalized
- [ ] Final tables and figures are defined
