---
title: "Manual simulating data for linear mix model"
description: "A practical R learning guide to Manual simulating data for linear mix model, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Longitudinal & Repeated Measures"
topic: "Manual simulating data for linear mix model"
type: article
tags: ["Longitudinal & Repeated Measures", "Biostatistics", "Manual", "simulating", "data"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Repeated observations from the same unit are correlated, so longitudinal models must represent both mean change and within-unit dependence.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Manual simulating data for linear mix model**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Introduction

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Manual simulating data for linear mix model** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Introduction

<!-- generated-learning-note -->
This section establishes the concepts and assumptions needed to understand **Introduction** in the context of Manual simulating data for linear mix model.


[Reference] (https://stats.stackexchange.com/questions/11233/how-to-simulate-data-based-on-a-linear-mixed-model-fit-object-in-r)

Manual simulating data for linear mix model with the random intercepts

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(nlme)
fm2 <- lme(distance ~ age + Sex, data = Orthodont, random = ~ 1) 

simfun <- function(n) {
  # n is the number of subjects, total rows will be 4*n
  # n=10
  # sig.b0 is the st. dev. of the random intercepts
  # it might be easier to just copy from the output
  sig.b0 <- exp(unlist(fm2$modelStruct$reStruct))*fm2$sigma
  
  b0 <- rnorm(n, 0, sig.b0)  #random intercepts
  
  sex <- rbinom(n, 1, 0.5)  # assign sex at random
  
  fe <- fixef(fm2)  #fixed coefficients
  
  my.df <- data.frame( Subject=as.factor(rep(1:n, each=4)), 
                       int = fe[1] + rep(b0, each=4), 
                       Sex=rep(sex,each=4), age=rep( c(8,10,12,14), n ) )       #create sex and age
  
  my.df$distance <- my.df$int + fe[2] * my.df$age + 
    fe[3]*my.df$Sex + rnorm(n*4, 0, fm2$sigma) #generate distance
  
  my.df$int <- NULL
  my.df$Sex <- factor( my.df$Sex, levels=0:1,
                       labels=c('Male','Female') ) #label sex
  my.df
}

Orthodont2 <- simfun(10)

```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


Compare simulated data and original data

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
summary(lme(distance ~ age +  (Sex), data = Orthodont, random = ~ 1|Subject))

summary(lme(distance ~ age +  (Sex), data = Orthodont2, random = ~ 1|Subject ))
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


Manual bootstrap sampling

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
###################

# set the number of participants in your big simulated study

nits=300 
# get the unique participants in the small study

sub=unique(Orthodont$Subject)
# sample the unique participants randomly with replacement

subs=sample(sub,nits,rep=T)
# make an empty data frame

df=Orthodont[-(1:dim(Orthodont)[1]),]  #empty data table
# loop through the sample size and bind it together.

for( i in 1:nits) {  
  df=rbind(df,Orthodont[which(Orthodont$Subject==subs[i]),])
}

#  nrow(df)
# [1] 1200
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.


Using simulate function to sample automatically (and aggregate the statistics )

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}
# simple simulation based on multi variates distribution 
# general linear
x <- 1:5
mod1 <- lm(c(1:3, 7, 6) ~ x)
S2 <- simulate(mod1, nsim = 200, seed = 123)
rowMeans(S2) 
fitted(mod1)

# generized linear
x2 <- sort(runif(100))
yb2 <- rbinom(100, prob = plogis(2*(x2-1)), size = 1)
yb2 <- factor(1 + yb2, labels = c("failure", "success"))
modb2 <- glm(yb2 ~ x2, family = binomial)
S4 <- simulate(modb2, nsim = 2)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.

## Interpreting the Results

Check fixed effects, random effects, covariance assumptions, and subject-specific versus population-level interpretation. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
