---
title: "BE Simulation Report"
description: "A practical R learning guide to BE Simulation Report, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Non-Inferiority & Equivalence"
topic: "BE Simulation Report"
type: article
tags: ["Non-Inferiority & Equivalence", "Biostatistics", "BE", "Simulation", "Report"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Non-inferiority and equivalence analyses compare effects with clinically justified margins rather than testing only for any difference.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **BE Simulation Report**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Summary of Simulation Workflow
2. Step 1 – Define Reference PK Profile
3. Step 2 – Set Simulation Parameters
4. Step 3 – Monte Carlo Trial Generation
5. Step 4 – Decision Probability Estimation
6. Step 5 – Decision Heatmap

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Non-Inferiority & Equivalence",
  topic = "BE Simulation Report"
)
learning_overview
```

## Learning Material

The following material develops **BE Simulation Report** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Summary of Simulation Workflow

<!-- generated-learning-note -->
This section focuses on **Summary of Simulation Workflow**. Emphasize effect size, uncertainty, diagnostics, and scientific relevance rather than a single threshold.


This R script implements a reproducible framework for virtual bioequivalence (BE) assessment using physiologically based pharmacokinetic (PBPK)–inspired concentration–time profiles and Monte Carlo methods.

#### Step 1 – Define Reference PK Profile

<!-- generated-learning-note -->
This subsection explains **Step 1 – Define Reference PK Profile** and shows how it contributes to the overall BE Simulation Report workflow.

A mean reference concentration–time profile is provided (e.g., `ref <- data.frame(time_hr, mean_conc_ng_per_mL)`).
This serves as the baseline for simulating both **Test** and **Reference** arms under user-specified variability assumptions.

#### Step 2 – Set Simulation Parameters

<!-- generated-learning-note -->
This subsection explains **Step 2 – Set Simulation Parameters** and shows how it contributes to the overall BE Simulation Report workflow.

Key parameters include:
- `n_per_arm` – sample size per treatment arm  
- `CV_between`, `CV_resid` – between-subject and residual variability (coefficient of variation)  
- `true_GMR` – assumed geometric mean ratio (Test/Reference)  
- BE acceptance limits (default 0.80 – 1.25) and type I error `alpha`

#### Step 3 – Monte Carlo Trial Generation

<!-- generated-learning-note -->
This subsection explains **Step 3 – Monte Carlo Trial Generation** and shows how it contributes to the overall BE Simulation Report workflow.

For each virtual trial:
1. Individual concentration profiles are generated with log-normal random effects (`CV_between`) and residual error (`CV_resid`).
2. Individual AUC and Cmax values are calculated using trapezoidal integration (`trap_auc()`).
3. Bioequivalence statistics (90 % CI of log-transformed ratios) are computed using a standard parallel-design t-test (`be_stats_parallel()`).
4. Pass/fail outcomes for AUC, Cmax, and joint BE are recorded.

#### Step 4 – Decision Probability Estimation

<!-- generated-learning-note -->
This subsection explains **Step 4 – Decision Probability Estimation** and shows how it contributes to the overall BE Simulation Report workflow.

Multiple (e.g., > 1000) virtual trials are simulated to estimate the empirical probability of meeting BE criteria for the specified design and parameters:
\[
P_\text{BE} = \Pr(90\%\,CI \subseteq [0.80,\,1.25]).
\]

#### Step 5 – Decision Heatmap

<!-- generated-learning-note -->
This subsection explains **Step 5 – Decision Heatmap** and shows how it contributes to the overall BE Simulation Report workflow.

A two-dimensional grid of **true GMR** values and **sample sizes** is explored.
For each grid cell, the Monte Carlo procedure is repeated, yielding a heatmap of \(P_\text{BE}\) for AUC and Cmax.
This allows visualization of power and design sensitivity across plausible conditions.

#### Step 6 – Parameter-Uncertainty Propagation

<!-- generated-learning-note -->
This subsection explains **Step 6 – Parameter-Uncertainty Propagation** and shows how it contributes to the overall BE Simulation Report workflow.

To evaluate robustness, uncertainty in variability parameters is propagated by repeatedly sampling plausible `CV_between` and `CV_resid` values (log-normal distribution around nominal).  
Each sampled pair triggers a short Monte Carlo run, generating a distribution of \(P_\text{BE}\) estimates.

#### Step 7 – Visualization and Reporting

<!-- generated-learning-note -->
This section uses **Step 7 – Visualization and Reporting** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.

Plots include:
- Histograms of simulated Cmax and AUC ratios with BE limits  
- Heatmaps of \(P_\text{BE}\) vs GMR and sample size  
- Sensitivity plots showing impact of variability on BE probability  

#### BE Simulation Report

<!-- generated-learning-note -->
This subsection explains **BE Simulation Report** and shows how it contributes to the overall BE Simulation Report workflow.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# virtual BE Monte Carlo with:
# ------------------------------------------------------------------------------

# ---------------------------
# PACKAGES
# ---------------------------
library(ggplot2)

# ---------------------------
# 0. EXAMPLE REFERENCE PROFILE
# ---------------------------
ref <- data.frame(
  time_hr = c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 12, 24),
  mean_conc_ng_per_mL = c(0, 12.5, 28.3, 42.8, 36.0, 20.1, 12.6, 7.8, 2.1, 0.2)
)

# ---------------------------
# 1. USER PARAMETERS (tweak for speed/precision)
# ---------------------------
set.seed(12345)
# Basic simulation parameters
N_trials_main <- 1200     # Monte Carlo trials for main simulation (per baseline scenario)
n_per_arm <- 24
CV_between_nom <- 0.25
CV_resid_nom <- 0.20
true_GMR_nom <- 0.95
alpha <- 0.10
BE_lower <- 0.80
BE_upper <- 1.25

# Extended analyses parameters
# 1) Heatmap grid for decision visualization
gmr_grid <- seq(0.88, 1.04, length.out = 9)      # range of true GMRs to explore
n_grid   <- seq(12, 60, by = 6)                  # sample sizes per arm to evaluate
reps_per_cell <- 250                             # Monte Carlo replicates per (GMR, n) cell

# 2) Parameter uncertainty propagation
param_sims <- 60                                 # number of different (CV_between, CV_resid) draws
trials_per_param_sim <- 600                      # MC trials within each param simulation

# Files
# results_dir <- "."
# results_csv_main <- file.path(results_dir, "results_summary_virtual_BE_main.csv")
# results_csv_heatmap <- file.path(results_dir, "results_heatmap_BE.csv")
# results_csv_param <- file.path(results_dir, "results_param_uncertainty.csv")
# report_rmd <- file.path(results_dir, "virtual_BE_report.Rmd")
# report_html <- file.path(results_dir, "virtual_BE_report.html")

# ---------------------------
# 2. HELPER FUNCTIONS
# ---------------------------
trap_auc <- function(t, conc){
  sum(0.5 * (conc[-1] + conc[-length(conc)]) * diff(t))
}
cv_to_lnsd <- function(cv) sqrt(log(1 + cv^2))
be_stats_parallel <- function(yT, yR, alpha = 0.10){
  # returns ratio, lower, upper (on original scale)
  lnT <- log(yT); lnR <- log(yR)
  nT <- length(lnT); nR <- length(lnR)
  meanT <- mean(lnT); meanR <- mean(lnR)
  s2T <- var(lnT); s2R <- var(lnR)
  df <- nT + nR - 2
  if (df <= 0) return(list(ratio = NA, lower = NA, upper = NA))
  s_pooled <- sqrt(((nT - 1) * s2T + (nR - 1) * s2R) / df)
  se_diff <- s_pooled * sqrt(1/nT + 1/nR)
  t_crit <- qt(1 - alpha/2, df)
  diff <- meanT - meanR
  lower_ln <- diff - t_crit * se_diff
  upper_ln <- diff + t_crit * se_diff
  list(ratio = exp(diff), lower = exp(lower_ln), upper = exp(upper_ln))
}

simulate_trial_parallel <- function(mean_profile, n_per_arm, CV_between, CV_resid, true_GMR, alpha){
  # Simulate one virtual two-arm (parallel) trial and return BE stats for AUC and Cmax
  times <- mean_profile$time_hr
  mean_conc <- mean_profile$mean_conc_ng_per_mL
  sd_between_ln <- cv_to_lnsd(CV_between)
  sd_resid_ln <- cv_to_lnsd(CV_resid)
  
  # subject-level scaling and residuals
  etaR <- exp(rnorm(n_per_arm, 0, sd_between_ln))
  etaT <- exp(rnorm(n_per_arm, 0, sd_between_ln))
  aucR <- aucT <- cmaxR <- cmaxT <- numeric(n_per_arm)
  
  for (i in seq_len(n_per_arm)) {
    residR <- exp(rnorm(length(mean_conc), 0, sd_resid_ln))
    residT <- exp(rnorm(length(mean_conc), 0, sd_resid_ln))
    concR <- mean_conc * etaR[i] * residR
    concT <- mean_conc * true_GMR * etaT[i] * residT
    aucR[i] <- trap_auc(times, concR); aucT[i] <- trap_auc(times, concT)
    cmaxR[i] <- max(concR); cmaxT[i] <- max(concT)
  }
  bs_auc <- be_stats_parallel(aucT, aucR, alpha)
  bs_cmax <- be_stats_parallel(cmaxT, cmaxR, alpha)
  pass_auc <- (bs_auc$lower >= BE_lower) & (bs_auc$upper <= BE_upper)
  pass_cmax <- (bs_cmax$lower >= BE_lower) & (bs_cmax$upper <= BE_upper)
  list(
    ratio_auc = bs_auc$ratio, lower_auc = bs_auc$lower, upper_auc = bs_auc$upper,
    ratio_cmax = bs_cmax$ratio, lower_cmax = bs_cmax$lower, upper_cmax = bs_cmax$upper,
    pass_auc = pass_auc, pass_cmax = pass_cmax, pass_both = (pass_auc & pass_cmax)
  )
}

# ---------------------------
# 3. MAIN / baseline simulation (single scenario) 
# ---------------------------
cat("Running main baseline Monte Carlo simulation...\n")
main_results <- data.frame(
  trial = integer(), ratio_auc = numeric(), lower_auc = numeric(), upper_auc = numeric(),
  ratio_cmax = numeric(), lower_cmax = numeric(), upper_cmax = numeric(),
  pass_auc = logical(), pass_cmax = logical(), pass_both = logical(),
  stringsAsFactors = FALSE
)

for (tr in seq_len(N_trials_main)){
  res <- simulate_trial_parallel(ref, n_per_arm, CV_between_nom, CV_resid_nom, true_GMR_nom, alpha)
  main_results[tr, ] <- c(tr, res$ratio_auc, res$lower_auc, res$upper_auc,
                          res$ratio_cmax, res$lower_cmax, res$upper_cmax,
                          res$pass_auc, res$pass_cmax, res$pass_both)
}
main_results$pass_auc <- as.logical(main_results$pass_auc)
main_results$pass_cmax <- as.logical(main_results$pass_cmax)
main_results$pass_both <- as.logical(main_results$pass_both)

print(head(main_results,100 ))


# Summary probabilities
prob_pass_auc_main  <- mean(main_results$pass_auc, na.rm = TRUE)
prob_pass_cmax_main <- mean(main_results$pass_cmax, na.rm = TRUE)
prob_pass_both_main <- mean(main_results$pass_both, na.rm = TRUE)
cat(sprintf("Baseline P(pass AUC) = %.3f | P(pass Cmax) = %.3f | P(pass both) = %.3f\n",
            prob_pass_auc_main, prob_pass_cmax_main, prob_pass_both_main))

# ---------------------------
# 4. Heatmap: P(BE) vs true_GMR and sample size (Cmax & AUC)
# ---------------------------
cat("Computing heatmap grid (this may take a while)...\n")
heatmap_rows <- expand.grid(true_GMR = gmr_grid, n = n_grid)
heatmap_rows$prob_cmax <- NA
heatmap_rows$prob_auc <- NA

# iterate cells
cell_idx <- 0
total_cells <- nrow(heatmap_rows)
for (r in seq_len(total_cells)){
  cell_idx <- cell_idx + 1
  gmr_val <- heatmap_rows$true_GMR[r]
  n_val <- heatmap_rows$n[r]
  pass_cmax_cnt <- 0
  pass_auc_cnt <- 0
  for (rep in seq_len(reps_per_cell)){
    res <- simulate_trial_parallel(ref, n_val, CV_between_nom, CV_resid_nom, gmr_val, alpha)
    pass_cmax_cnt <- pass_cmax_cnt + as.integer(res$pass_cmax)
    pass_auc_cnt  <- pass_auc_cnt  + as.integer(res$pass_auc)
  }
  heatmap_rows$prob_cmax[r] <- pass_cmax_cnt / reps_per_cell
  heatmap_rows$prob_auc[r]  <- pass_auc_cnt / reps_per_cell
  if (cell_idx %% 8 == 0) cat(sprintf("Completed %d / %d cells...\n", cell_idx, total_cells))
}
print(heatmap_rows )


# ---------------------------
# 5. Parameter uncertainty propagation
# ---------------------------
cat("Running parameter uncertainty propagation...\n")
# Assume uncertainty on CVs ~ lognormal around nominal with 20% CV on each CV
param_results <- data.frame(sim = integer(), CV_between = numeric(), CV_resid = numeric(),
                            prob_pass_auc = numeric(), prob_pass_cmax = numeric(), prob_pass_both = numeric(),
                            stringsAsFactors = FALSE)

for (s in seq_len(param_sims)){
  # sample CVs in a plausible range (lognormal multiplicative uncertainty)
  CV_between_s <- CV_between_nom * exp(rnorm(1, 0, 0.2))  # ~20% multiplicative uncertainty
  CV_resid_s   <- CV_resid_nom   * exp(rnorm(1, 0, 0.2))
  pass_auc_cnt <- pass_cmax_cnt <- pass_both_cnt <- 0
  for (tr in seq_len(trials_per_param_sim)){
    res <- simulate_trial_parallel(ref, n_per_arm, CV_between_s, CV_resid_s, true_GMR_nom, alpha)
    pass_auc_cnt <- pass_auc_cnt + as.integer(res$pass_auc)
    pass_cmax_cnt <- pass_cmax_cnt + as.integer(res$pass_cmax)
    pass_both_cnt <- pass_both_cnt + as.integer(res$pass_both)
  }
  param_results[s, ] <- c(s, CV_between_s, CV_resid_s,
                          pass_auc_cnt / trials_per_param_sim,
                          pass_cmax_cnt / trials_per_param_sim,
                          pass_both_cnt / trials_per_param_sim)
  if (s %% 10 == 0) cat(sprintf("Completed %d / %d parameter sims...\n", s, param_sims))
}
names(param_results) <- c("sim", "CV_between", "CV_resid", "prob_pass_auc", "prob_pass_cmax", "prob_pass_both")
print(param_results )


# ---------------------------
# 6. PLOTS (save to files & keep in environment)
# ---------------------------
cat("Creating plots... (files will be saved in working directory)\n")
# Main: distributions of ratios (AUC & Cmax)
p1 <- ggplot(main_results, aes(x = ratio_cmax)) +
  geom_histogram(binwidth = 0.02, fill = "skyblue", color = "black") +
  geom_vline(xintercept = c(BE_lower, BE_upper), linetype = "dashed", color = "red") +
  labs(title = "Cmax Ratio Distribution (Main Scenario)", x = "Cmax Ratio", y = "Count")

p2 <- ggplot(main_results, aes(x = ratio_auc)) +
  geom_histogram(binwidth = 0.02, fill = "lightcoral", color = "black") +
  geom_vline(xintercept = c(BE_lower, BE_upper), linetype = "dashed", color = "red") +
  labs(title = "AUC Ratio Distribution (Main Scenario)", x = "AUC Ratio", y = "Count")
p1
p2

# Heatmap (Cmax)
library(reshape2)
heat_cmax <- dcast(heatmap_rows, n ~ true_GMR, value.var = "prob_cmax")
# convert to plotting-friendly long form
heat_long_cmax <- heatmap_rows[, c("true_GMR", "n", "prob_cmax")]
p_heat_cmax <- ggplot(heat_long_cmax, aes(x = factor(true_GMR), y = factor(n), fill = prob_cmax)) +
  geom_tile() + scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "True GMR", y = "Sample Size per Arm", fill = "P(BE) (Cmax)",
       title = "Heatmap: P(BE) for Cmax vs True GMR & Sample Size") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_heat_cmax

# Heatmap (AUC)
heat_long_auc <- heatmap_rows[, c("true_GMR", "n", "prob_auc")]
p_heat_auc <- ggplot(heat_long_auc, aes(x = factor(true_GMR), y = factor(n), fill = prob_auc)) +
  geom_tile() + scale_fill_gradient(low = "white", high = "tomato") +
  labs(x = "True GMR", y = "Sample Size per Arm", fill = "P(BE) (AUC)",
       title = "Heatmap: P(BE) for AUC vs True GMR & Sample Size") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_heat_auc


# Parameter uncertainty plots
p_param_cv_between <- ggplot(param_results, aes(x = CV_between, y = prob_pass_both)) +
  geom_point() + geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Effect of CV_between on P(BE both)", x = "CV_between", y = "P(pass both)")

p_param_cv_resid <- ggplot(param_results, aes(x = CV_resid, y = prob_pass_both)) +
  geom_point() + geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Effect of CV_resid on P(BE both)", x = "CV_resid", y = "P(pass both)")
p_param_cv_between
p_param_cv_resid
# ggsave("param_cv_between_vs_prob.png", p_param_cv_between, width = 7, height = 4)
# ggsave("param_cv_resid_vs_prob.png", p_param_cv_resid, width = 7, height = 4)


```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

## Interpreting the Results

Interpret confidence intervals relative to the margin and confirm that the analysis scale matches the scientific claim. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
