---
title: "NHANES Diabetes Prediction: Complete Clinical Prediction Modeling Workflow in R"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

# Overview

This tutorial develops a complete clinical prediction-modeling workflow using the `NHANES` R package and diabetes as the binary outcome.

The workflow covers:

1. Research question and target population
2. Data import and data dictionary
3. Duplicate, type, coding, range, and logical checks
4. Missing-data assessment
5. Outlier and influential-observation assessment
6. Exploratory data analysis (EDA)
7. Continuous-variable handling and nonlinear effects
8. Multicollinearity assessment
9. Multiple imputation
10. Logistic regression model development
11. Interaction assessment
12. Model-selection approaches
13. Penalized regression with LASSO
14. Apparent model performance
15. Bootstrap internal validation
16. Discrimination
17. Calibration
18. Brier score
19. Decision-curve analysis
20. Threshold-based classification performance
21. Sensitivity analyses
22. Subgroup performance
23. Temporal validation using NHANES survey cycles
24. Recalibration
25. Individual prediction
26. Nomogram
27. Recommended reporting framework

> **Important:** `NHANES::NHANES` is a teaching dataset derived from NHANES. This example predicts **prevalent diabetes at the time of survey**, not future incident diabetes. For a publication-quality NHANES analysis, use the original CDC NHANES files and account for the complex survey design, weights, strata, and PSU variables.

# 1. Setup

## 1.1 Install required packages

The following chunk installs any missing packages. It only needs internet access the first time a package is installed.

```{r install-packages, message=FALSE, warning=FALSE}
required_packages <- c(
  "NHANES",
  "tidyverse",
  "gtsummary",
  "naniar",
  "mice",
  "car",
  "rms",
  "pROC",
  "glmnet",
  "broom"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}
```

## 1.2 Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(NHANES)
library(tidyverse)
library(gtsummary)
library(naniar)
library(mice)
library(car)
library(rms)
library(pROC)
library(glmnet)
library(broom)
```

## 1.3 Reproducibility settings

```{r reproducibility}
set.seed(2026)
options(stringsAsFactors = FALSE)
```

# 2. Research Question

We want to predict whether an adult participant has diabetes using demographic, anthropometric, blood-pressure, lipid, and lifestyle information.

The binary outcome is

\[
Y_i =
\begin{cases}
1, & \text{Diabetes = Yes} \\
0, & \text{Diabetes = No}
\end{cases}
\]

The candidate predictor set is

\[
\text{Age} + \text{BMI} + \text{BPSysAve} + \text{TotChol} +
\text{Gender} + \text{Race1} + \text{PhysActive} + \text{Smoke100}.
\]

We deliberately **do not use `DiabetesAge`**, because it is downstream of diabetes diagnosis and would create outcome leakage.

# 3. Load and Inspect the NHANES Data

```{r load-data}
data("NHANES")

cat("Rows:", nrow(NHANES), "\n")
cat("Columns:", ncol(NHANES), "\n")
```

```{r inspect-data}
glimpse(NHANES)
```

Check the variables that will be used:

```{r check-required-vars}
required_vars <- c(
  "ID", "SurveyYr", "Diabetes", "Age", "Gender", "Race1",
  "BMI", "BPSysAve", "BPDiaAve", "TotChol", "DirectChol",
  "PhysActive", "Smoke100", "Education", "Poverty", "SleepHrsNight"
)

missing_vars <- setdiff(required_vars, names(NHANES))

if (length(missing_vars) > 0) {
  stop(
    "The following expected NHANES variables are missing: ",
    paste(missing_vars, collapse = ", ")
  )
}
```

# 4. Create the Initial Analysis Dataset

We restrict the example to adults aged 20 years or older and participants with a defined diabetes outcome.

```{r create-analysis-data}
nhanes0 <- NHANES %>%
  select(all_of(required_vars))

nhanes1 <- nhanes0 %>%
  filter(
    Age >= 20,
    Diabetes %in% c("Yes", "No")
  ) %>%
  mutate(
    Diabetes_num = if_else(Diabetes == "Yes", 1L, 0L)
  )

cat("Adult analytical cohort N =", nrow(nhanes1), "\n")
```

Check outcome coding:

```{r outcome-check}
table(nhanes1$Diabetes, nhanes1$Diabetes_num, useNA = "ifany")
prop.table(table(nhanes1$Diabetes))
```

# 5. Basic Data-Quality Checks

## 5.1 Duplicate subject IDs

```{r duplicate-check}
duplicate_ids <- nhanes1 %>%
  count(ID) %>%
  filter(n > 1)

duplicate_ids
```

A duplicate ID should never be deleted automatically. It may represent a true duplicate, repeated measurement, merge error, or multiple records per participant.

## 5.2 Variable classes

```{r classes}
sapply(nhanes1, class)
```

## 5.3 Categorical coding

```{r categorical-levels}
list(
  Diabetes = levels(nhanes1$Diabetes),
  SurveyYr = levels(nhanes1$SurveyYr),
  Gender = levels(nhanes1$Gender),
  Race1 = levels(nhanes1$Race1),
  PhysActive = levels(nhanes1$PhysActive),
  Smoke100 = levels(nhanes1$Smoke100),
  Education = levels(nhanes1$Education)
)
```

## 5.4 Continuous-variable ranges

```{r range-check}
range_summary <- nhanes1 %>%
  summarise(
    across(
      c(Age, BMI, BPSysAve, BPDiaAve, TotChol, DirectChol, Poverty, SleepHrsNight),
      list(
        min = ~ min(.x, na.rm = TRUE),
        max = ~ max(.x, na.rm = TRUE)
      )
    )
  )

range_summary
```

## 5.5 Flag potentially impossible values

The following rules are **screening rules**, not automatic deletion rules.

```{r impossible-values}
flagged_values <- nhanes1 %>%
  filter(
    Age < 20 | Age > 100 |
      (!is.na(BMI) & BMI <= 0) |
      (!is.na(BPSysAve) & BPSysAve <= 0) |
      (!is.na(BPDiaAve) & BPDiaAve < 0) |
      (!is.na(TotChol) & TotChol <= 0)
  )

flagged_values %>%
  select(ID, Age, BMI, BPSysAve, BPDiaAve, TotChol) %>%
  head(20)
```

# 6. Missing-Data Assessment

## 6.1 Missing percentage by variable

```{r missing-summary}
missing_summary <- nhanes1 %>%
  summarise(
    across(
      everything(),
      ~ mean(is.na(.x)) * 100
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Percent"
  ) %>%
  arrange(desc(Missing_Percent))

missing_summary
```

## 6.2 Missingness visualization

```{r missing-plot, fig.height=5, fig.width=9}
vis_miss(nhanes1)
```

```{r missing-variable-plot, fig.height=5, fig.width=8}
gg_miss_var(nhanes1)
```

## 6.3 Missing-data pattern for modeling variables

```{r md-pattern}
model_vars_for_missing <- nhanes1 %>%
  select(
    Diabetes_num,
    Age,
    BMI,
    BPSysAve,
    TotChol,
    Gender,
    Race1,
    PhysActive,
    Smoke100
  )

md.pattern(model_vars_for_missing)
```

# 7. Descriptive Analysis and EDA

## 7.1 Table 1 by diabetes status

```{r table1}
nhanes1 %>%
  select(
    Diabetes,
    Age,
    Gender,
    Race1,
    BMI,
    BPSysAve,
    TotChol,
    PhysActive,
    Smoke100
  ) %>%
  tbl_summary(
    by = Diabetes,
    statistic = list(
      all_continuous() ~ "{mean} ({sd}); median {median} [{p25}, {p75}]",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany"
  )
```

For prediction modeling, a variable should not be excluded simply because its univariable P-value is greater than 0.05.

## 7.2 Distribution of age

```{r age-hist, fig.height=4, fig.width=7}
ggplot(nhanes1, aes(x = Age)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Age",
    x = "Age (years)",
    y = "Count"
  )
```

## 7.3 Distribution of BMI

```{r bmi-hist, fig.height=4, fig.width=7}
ggplot(nhanes1, aes(x = BMI)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of BMI",
    x = "BMI",
    y = "Count"
  )
```

## 7.4 Distribution of systolic blood pressure

```{r sbp-hist, fig.height=4, fig.width=7}
ggplot(nhanes1, aes(x = BPSysAve)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Average Systolic Blood Pressure",
    x = "Average systolic BP",
    y = "Count"
  )
```

## 7.5 Diabetes versus BMI

```{r diabetes-bmi-box, fig.height=4, fig.width=6}
ggplot(nhanes1, aes(x = Diabetes, y = BMI)) +
  geom_boxplot() +
  labs(
    title = "BMI by Diabetes Status",
    x = "Diabetes",
    y = "BMI"
  )
```

## 7.6 Diabetes versus age

```{r diabetes-age-box, fig.height=4, fig.width=6}
ggplot(nhanes1, aes(x = Diabetes, y = Age)) +
  geom_boxplot() +
  labs(
    title = "Age by Diabetes Status",
    x = "Diabetes",
    y = "Age"
  )
```

# 8. Outlier Assessment

## 8.1 IQR flagging for BMI

```{r bmi-iqr}
q1_bmi <- quantile(nhanes1$BMI, 0.25, na.rm = TRUE)
q3_bmi <- quantile(nhanes1$BMI, 0.75, na.rm = TRUE)
iqr_bmi <- q3_bmi - q1_bmi

bmi_lower <- q1_bmi - 1.5 * iqr_bmi
bmi_upper <- q3_bmi + 1.5 * iqr_bmi

bmi_outliers <- nhanes1 %>%
  filter(
    !is.na(BMI),
    BMI < bmi_lower | BMI > bmi_upper
  ) %>%
  select(ID, Age, Gender, BMI, Diabetes)

head(bmi_outliers, 20)
```

A statistical outlier is not automatically a data error. Clinically plausible extreme values should generally be retained unless there is evidence of measurement or data-entry error.

# 9. Correlation and Multicollinearity Screening

## 9.1 Correlation matrix among continuous variables

```{r correlation-matrix}
cor_data <- nhanes1 %>%
  select(
    Age,
    BMI,
    BPSysAve,
    BPDiaAve,
    TotChol,
    DirectChol
  )

round(
  cor(cor_data, use = "pairwise.complete.obs"),
  2
)
```

# 10. Construct the Modeling Dataset

```{r modeling-dataset}
model_data <- nhanes1 %>%
  select(
    ID,
    SurveyYr,
    Diabetes,
    Diabetes_num,
    Age,
    BMI,
    BPSysAve,
    TotChol,
    Gender,
    Race1,
    PhysActive,
    Smoke100
  )
```

# 11. Complete-Case Dataset for Sensitivity Analysis

```{r complete-case-data}
model_cc <- model_data %>%
  drop_na(
    Age,
    BMI,
    BPSysAve,
    TotChol,
    Gender,
    Race1,
    PhysActive,
    Smoke100
  )

cat("Original modeling N =", nrow(model_data), "\n")
cat("Complete-case N =", nrow(model_cc), "\n")
cat(
  "Complete-case retention =",
  round(100 * nrow(model_cc) / nrow(model_data), 1),
  "%\n"
)
```

## 11.1 Compare complete versus incomplete cases

```{r complete-vs-missing}
model_data <- model_data %>%
  mutate(
    CompleteCase = if_else(
      complete.cases(
        Age,
        BMI,
        BPSysAve,
        TotChol,
        Gender,
        Race1,
        PhysActive,
        Smoke100
      ),
      "Complete",
      "Missing"
    )
  )

model_data %>%
  group_by(CompleteCase) %>%
  summarise(
    N = n(),
    Mean_Age = mean(Age, na.rm = TRUE),
    Mean_BMI = mean(BMI, na.rm = TRUE),
    Diabetes_Rate = mean(Diabetes_num, na.rm = TRUE),
    .groups = "drop"
  )
```

# 12. Multiple Imputation with MICE

## 12.1 Prepare the imputation dataset

The outcome is included in the imputation model because it can help predict missing predictor values. The outcome itself is not imputed here.

```{r mi-data}
mi_data <- model_data %>%
  select(
    Diabetes_num,
    Age,
    BMI,
    BPSysAve,
    TotChol,
    Gender,
    Race1,
    PhysActive,
    Smoke100
  )
```

## 12.2 Inspect default MICE methods

```{r mice-initialization}
ini <- mice(
  mi_data,
  maxit = 0,
  printFlag = FALSE
)

method <- ini$method
predictor_matrix <- ini$predictorMatrix

method
```

Prevent imputation of the observed outcome:

```{r mice-method-adjustment}
method["Diabetes_num"] <- ""
predictor_matrix[, "Diabetes_num"] <- 1
predictor_matrix["Diabetes_num", ] <- 0

method
```

## 12.3 Run multiple imputation

```{r run-mice}
set.seed(2026)

imp <- mice(
  mi_data,
  m = 20,
  maxit = 20,
  method = method,
  predictorMatrix = predictor_matrix,
  seed = 2026,
  printFlag = FALSE
)

imp
```

## 12.4 Inspect convergence

```{r mice-convergence, fig.height=6, fig.width=9}
plot(imp)
```

## 12.5 Compare observed and imputed BMI distributions

```{r mice-density-bmi, fig.height=5, fig.width=8}
densityplot(imp, ~ BMI)
```

# 13. Fit the Prespecified Logistic Model Across Imputed Datasets

```{r fit-mi-model}
fit_mi <- with(
  imp,
  glm(
    Diabetes_num ~
      Age +
      BMI +
      BPSysAve +
      TotChol +
      Gender +
      Race1 +
      PhysActive +
      Smoke100,
    family = binomial()
  )
)

pooled_fit <- pool(fit_mi)
```

Pooled odds ratios:

```{r pooled-or}
summary(
  pooled_fit,
  conf.int = TRUE,
  exponentiate = TRUE
)
```

> **Important statistical distinction:** The pooled coefficients are useful for regression inference. A fully rigorous prediction-validation workflow with missing data should repeat imputation **inside each resampling iteration**. The remainder of this tutorial uses one completed dataset for transparent teaching of model diagnostics and validation. That simplification should not be mistaken for the strictest publication-level MI + bootstrap procedure.

# 14. Create a Completed Dataset for the Remaining Demonstration

```{r completed-data}
model_complete <- complete(imp, action = 1)

summary(model_complete)
```

# 15. Fit a Full Logistic Regression Model

```{r full-glm}
fit_full <- glm(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_complete,
  family = binomial()
)

summary(fit_full)
```

Odds ratios and confidence intervals:

```{r full-glm-or}
broom::tidy(
  fit_full,
  exponentiate = TRUE,
  conf.int = TRUE
)
```

For a prediction model, statistical significance is not the main criterion for retaining a prespecified predictor.

# 16. Multicollinearity Assessment

```{r vif}
vif_result <- car::vif(fit_full)
vif_result
```

If `vif()` returns generalized VIF values for multi-level factors, the commonly reviewed quantity is

\[
GVIF^{1/(2df)}.
\]

# 17. Assess Nonlinear Predictor Effects with Restricted Cubic Splines

## 17.1 Prepare the `rms` environment

```{r rms-setup}
dd <- datadist(model_complete)
options(datadist = "dd")
```

## 17.2 Linear logistic model using `lrm()`

```{r lrm-linear}
fit_linear <- lrm(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_complete,
  x = TRUE,
  y = TRUE
)

fit_linear
```

## 17.3 Restricted cubic spline model

```{r lrm-spline}
fit_spline <- lrm(
  Diabetes_num ~
    rcs(Age, 4) +
    rcs(BMI, 4) +
    rcs(BPSysAve, 4) +
    rcs(TotChol, 4) +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_complete,
  x = TRUE,
  y = TRUE
)

fit_spline
```

## 17.4 Test overall and nonlinear effects

```{r spline-anova}
anova(fit_spline)
```

The `Nonlinear` rows test whether a purely linear representation is inadequate for the corresponding continuous predictor.

## 17.5 Plot age effect

```{r spline-age, fig.height=5, fig.width=7}
plot(
  Predict(
    fit_spline,
    Age,
    fun = plogis
  ),
  xlab = "Age",
  ylab = "Predicted probability"
)
```

## 17.6 Plot BMI effect

```{r spline-bmi, fig.height=5, fig.width=7}
plot(
  Predict(
    fit_spline,
    BMI,
    fun = plogis
  ),
  xlab = "BMI",
  ylab = "Predicted probability"
)
```

## 17.7 Compare model AIC

```{r compare-aic}
AIC(fit_linear, fit_spline)
```

A lower AIC indicates a better tradeoff between likelihood fit and model complexity, but AIC should not be the only model-selection criterion.

# 18. Example Interaction Assessment

Suppose there is a clinically plausible question about whether the BMI association differs with age.

```{r interaction-model}
fit_interaction <- glm(
  Diabetes_num ~
    Age * BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_complete,
  family = binomial()
)
```

Likelihood-ratio comparison:

```{r interaction-lrt}
anova(
  fit_full,
  fit_interaction,
  test = "LRT"
)
```

Interactions should ideally be prespecified or clinically motivated rather than discovered by indiscriminately testing every possible pair.

# 19. Traditional Stepwise Selection: Demonstration Only

```{r stepwise}
fit_step <- step(
  fit_full,
  direction = "backward",
  trace = FALSE
)

formula(fit_step)
summary(fit_step)
```

Stepwise selection is shown because it is common in practice, but it is generally not preferred for modern clinical prediction modeling because it can produce unstable variable selection, biased coefficients, and optimistic performance.

# 20. Penalized Regression with LASSO

## 20.1 Create model matrix

```{r lasso-matrix}
x <- model.matrix(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_complete
)[, -1, drop = FALSE]

y <- model_complete$Diabetes_num
```

## 20.2 Cross-validated LASSO

```{r lasso-cv}
set.seed(2026)

cv_lasso <- cv.glmnet(
  x = x,
  y = y,
  family = "binomial",
  alpha = 1,
  nfolds = 10,
  type.measure = "deviance"
)

cv_lasso$lambda.min
cv_lasso$lambda.1se
```

```{r lasso-plot, fig.height=5, fig.width=7}
plot(cv_lasso)
```

## 20.3 LASSO coefficients

```{r lasso-coefs-min}
coef(cv_lasso, s = "lambda.min")
```

```{r lasso-coefs-1se}
coef(cv_lasso, s = "lambda.1se")
```

LASSO performs shrinkage and may set some coefficients exactly to zero.

# 21. Apparent Discrimination

For the remaining primary demonstration, we use the spline model.

```{r apparent-predictions}
model_complete$pred_prob <- as.numeric(
  predict(
    fit_spline,
    type = "fitted"
  )
)
```

## 21.1 ROC curve and AUC

```{r roc}
roc_obj <- roc(
  response = model_complete$Diabetes_num,
  predictor = model_complete$pred_prob,
  quiet = TRUE
)

roc_obj
auc(roc_obj)
ci.auc(roc_obj)
```

```{r roc-plot, fig.height=5, fig.width=6}
plot(
  roc_obj,
  print.auc = TRUE,
  main = "ROC Curve: Apparent Performance"
)
```

The AUC measures discrimination: how well the model ranks participants with diabetes above participants without diabetes.

# 22. Bootstrap Internal Validation

Bootstrap validation estimates model optimism by repeatedly rebuilding the model in bootstrap samples and testing performance in the original data.

```{r bootstrap-validation}
set.seed(2026)

val_boot <- validate(
  fit_spline,
  method = "boot",
  B = 500
)

val_boot
```

The most useful columns are typically:

- `index.orig`: apparent performance in the original data
- `training`: average performance in bootstrap samples
- `test`: performance when bootstrap-fitted models are applied back to the original data
- `optimism`: estimated over-optimism
- `index.corrected`: optimism-corrected performance

For a final analysis, increasing `B` to 1000 or more is common when computationally feasible.

# 23. Bootstrap Calibration

```{r bootstrap-calibration}
set.seed(2026)

cal_boot <- calibrate(
  fit_spline,
  method = "boot",
  B = 500
)

cal_boot
```

```{r calibration-plot, fig.height=6, fig.width=7}
plot(
  cal_boot,
  xlab = "Predicted Probability",
  ylab = "Observed Probability"
)
```

A perfectly calibrated model would lie on the 45-degree identity line.

# 24. Calibration Intercept and Calibration Slope

Calibration should preferably be evaluated on validation or resampled predictions. The following calculation on the development data is shown to explain the definitions.

```{r calibration-lp}
model_complete$lp <- as.numeric(
  predict(
    fit_spline,
    type = "lp"
  )
)
```

## 24.1 Calibration intercept with slope fixed at 1

```{r calibration-intercept}
cal_intercept_model <- glm(
  Diabetes_num ~ 1 + offset(lp),
  data = model_complete,
  family = binomial()
)

calibration_intercept <- unname(coef(cal_intercept_model)[1])
calibration_intercept
```

Ideal value:

\[
\text{Calibration intercept}=0.
\]

## 24.2 Calibration slope

```{r calibration-slope}
cal_slope_model <- glm(
  Diabetes_num ~ lp,
  data = model_complete,
  family = binomial()
)

calibration_slope <- unname(coef(cal_slope_model)[2])
calibration_slope
```

Ideal value:

\[
\text{Calibration slope}=1.
\]

A slope less than 1 often indicates overly extreme predictions and overfitting.

# 25. Brier Score

The Brier score is

\[
\text{Brier} = \frac{1}{n}\sum_{i=1}^{n}(\hat p_i-y_i)^2.
\]

```{r brier}
brier_model <- mean(
  (model_complete$pred_prob - model_complete$Diabetes_num)^2
)

prevalence <- mean(model_complete$Diabetes_num)

brier_null <- mean(
  (prevalence - model_complete$Diabetes_num)^2
)

brier_model
brier_null
```

A smaller Brier score indicates lower overall prediction error.

# 26. Decision-Curve Analysis

To avoid dependence on an additional package, this tutorial calculates net benefit directly.

For a threshold probability \(p_t\), net benefit is

\[
NB = \frac{TP}{N} - \frac{FP}{N}\frac{p_t}{1-p_t}.
\]

## 26.1 Function to calculate model net benefit

```{r dca-functions}
net_benefit_model <- function(y, p, threshold) {
  pred_positive <- p >= threshold

  tp <- sum(pred_positive & y == 1)
  fp <- sum(pred_positive & y == 0)
  n <- length(y)

  tp / n - fp / n * threshold / (1 - threshold)
}

net_benefit_all <- function(y, threshold) {
  prevalence <- mean(y)
  prevalence - (1 - prevalence) * threshold / (1 - threshold)
}
```

## 26.2 Calculate DCA across thresholds

```{r dca-calculate}
thresholds <- seq(0.01, 0.50, by = 0.01)

dca_results <- tibble(
  threshold = thresholds,
  model = map_dbl(
    thresholds,
    ~ net_benefit_model(
      y = model_complete$Diabetes_num,
      p = model_complete$pred_prob,
      threshold = .x
    )
  ),
  treat_all = map_dbl(
    thresholds,
    ~ net_benefit_all(
      y = model_complete$Diabetes_num,
      threshold = .x
    )
  ),
  treat_none = 0
) %>%
  pivot_longer(
    cols = c(model, treat_all, treat_none),
    names_to = "Strategy",
    values_to = "Net_Benefit"
  )
```

## 26.3 Plot decision curves

```{r dca-plot, fig.height=5, fig.width=8}
ggplot(
  dca_results,
  aes(
    x = threshold,
    y = Net_Benefit,
    linetype = Strategy
  )
) +
  geom_line(linewidth = 0.9) +
  labs(
    title = "Decision Curve Analysis",
    x = "Threshold Probability",
    y = "Net Benefit"
  ) +
  theme_minimal()
```

DCA is used to determine whether the model provides useful clinical net benefit over a clinically relevant threshold range. It is **not simply an automatic cutoff-selection procedure**.

# 27. Threshold-Based Classification Performance

Suppose, for illustration, a threshold of 20% is clinically relevant.

```{r cutoff-classification}
cutoff <- 0.20

model_complete <- model_complete %>%
  mutate(
    pred_class = if_else(pred_prob >= cutoff, 1L, 0L)
  )

confusion_matrix <- table(
  Predicted = model_complete$pred_class,
  Observed = model_complete$Diabetes_num
)

confusion_matrix
```

Calculate sensitivity, specificity, PPV, and NPV:

```{r classification-metrics}
tp <- sum(
  model_complete$pred_class == 1 &
    model_complete$Diabetes_num == 1
)

fp <- sum(
  model_complete$pred_class == 1 &
    model_complete$Diabetes_num == 0
)

tn <- sum(
  model_complete$pred_class == 0 &
    model_complete$Diabetes_num == 0
)

fn <- sum(
  model_complete$pred_class == 0 &
    model_complete$Diabetes_num == 1
)

classification_metrics <- tibble(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV"),
  Value = c(
    tp / (tp + fn),
    tn / (tn + fp),
    tp / (tp + fp),
    tn / (tn + fn)
  )
)

classification_metrics
```

## 27.1 Youden index for statistical comparison only

```{r youden-index}
coords(
  roc_obj,
  x = "best",
  best.method = "youden",
  ret = c("threshold", "sensitivity", "specificity")
)
```

A Youden-optimal cutoff maximizes sensitivity + specificity - 1, but it does not directly incorporate clinical consequences, treatment burden, resource use, or the relative cost of false positives and false negatives.

# 28. Influential Observations

For influence diagnostics, use the ordinary logistic model.

```{r cooks-distance}
cook <- cooks.distance(fit_full)

cook_threshold <- 4 / nrow(model_complete)

influential_indices <- which(cook > cook_threshold)

cat("Number above 4/n threshold:", length(influential_indices), "\n")
```

```{r cooks-plot, fig.height=5, fig.width=8}
plot(
  cook,
  type = "h",
  ylab = "Cook's Distance",
  main = "Influence Diagnostic"
)
abline(h = cook_threshold, lty = 2)
```

The `4/n` rule is only a screening heuristic. Large Cook's distance does not justify automatic deletion.

# 29. Sensitivity Analysis: Remove Highly Influential Observations

This is a sensitivity analysis, not the primary model.

```{r influence-sensitivity}
model_sens <- model_complete[
  cook <= cook_threshold,
  ,
  drop = FALSE
]

fit_sens <- glm(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_sens,
  family = binomial()
)

coef_comparison <- tibble(
  Term = names(coef(fit_full)),
  Primary = coef(fit_full),
  Sensitivity = coef(fit_sens)[names(coef(fit_full))]
)

coef_comparison
```

# 30. Sensitivity Analysis: Complete Case versus Multiple Imputation

Complete-case model:

```{r complete-case-model}
fit_cc <- glm(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_cc,
  family = binomial()
)

broom::tidy(
  fit_cc,
  exponentiate = TRUE,
  conf.int = TRUE
)
```

Compare this with the pooled MI estimates reported earlier.

# 31. Sensitivity Analysis: Linear versus Spline Functional Forms

```{r linear-vs-spline-validation}
set.seed(2026)

val_linear <- validate(
  fit_linear,
  method = "boot",
  B = 300
)

set.seed(2026)

val_spline <- validate(
  fit_spline,
  method = "boot",
  B = 300
)

val_linear
val_spline
```

If a more complex spline model provides little improvement after optimism correction, the simpler linear model may be preferable.

# 32. Subgroup Performance

Subgroup evaluation asks whether the **same model** performs similarly across clinically important groups. It is different from fitting a new model within each subgroup.

## 32.1 AUC by gender

```{r subgroup-gender-auc}
gender_auc <- model_complete %>%
  split(.$Gender) %>%
  imap_dfr(
    function(dat, group_name) {
      if (length(unique(dat$Diabetes_num)) < 2) {
        return(
          tibble(
            Gender = group_name,
            N = nrow(dat),
            AUC = NA_real_
          )
        )
      }

      r <- roc(
        dat$Diabetes_num,
        dat$pred_prob,
        quiet = TRUE
      )

      tibble(
        Gender = group_name,
        N = nrow(dat),
        AUC = as.numeric(auc(r))
      )
    }
  )

gender_auc
```

## 32.2 Mean predicted versus observed risk by race

```{r subgroup-race-calibration}
race_calibration <- model_complete %>%
  group_by(Race1) %>%
  summarise(
    N = n(),
    Observed = mean(Diabetes_num),
    Predicted = mean(pred_prob),
    Difference = Predicted - Observed,
    .groups = "drop"
  )

race_calibration
```

This is only a simple subgroup calibration summary. More complete subgroup validation would also examine calibration slopes and smooth calibration curves where sample size permits.

# 33. Temporal Validation Using Survey Cycle

The `NHANES` teaching dataset contains two survey cycles. We use the earlier cycle for development and the later cycle as a temporal validation sample.

Check the actual coding first:

```{r survey-year-levels}
levels(NHANES$SurveyYr)
table(NHANES$SurveyYr, useNA = "ifany")
```

## 33.1 Create temporal-validation data

For clarity, this section uses complete cases. In a publication-level analysis, the missing-data strategy should be carefully transported to the validation sample as well.

```{r temporal-data}
temporal_data <- NHANES %>%
  filter(
    Age >= 20,
    Diabetes %in% c("Yes", "No")
  ) %>%
  mutate(
    Diabetes_num = if_else(Diabetes == "Yes", 1L, 0L)
  ) %>%
  select(
    SurveyYr,
    Diabetes_num,
    Age,
    BMI,
    BPSysAve,
    TotChol,
    Gender,
    Race1,
    PhysActive,
    Smoke100
  ) %>%
  drop_na()

survey_levels <- levels(droplevels(temporal_data$SurveyYr))

if (length(survey_levels) != 2) {
  stop(
    "Expected exactly two SurveyYr levels, but found: ",
    paste(survey_levels, collapse = ", ")
  )
}

survey_levels
```

Use the first factor level as development and the second as validation:

```{r temporal-split}
development_level <- survey_levels[1]
validation_level <- survey_levels[2]

development <- temporal_data %>%
  filter(SurveyYr == development_level) %>%
  droplevels()

validation <- temporal_data %>%
  filter(SurveyYr == validation_level) %>%
  droplevels()

cat("Development cycle:", development_level, "N =", nrow(development), "\n")
cat("Validation cycle:", validation_level, "N =", nrow(validation), "\n")
```

# 34. Develop the Model Only in the Earlier Survey Cycle

```{r temporal-development-model}
fit_dev <- glm(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = development,
  family = binomial()
)

summary(fit_dev)
```

# 35. Apply the Frozen Model to the Validation Cycle

The validation model must **not be refitted** before performance is evaluated.

```{r temporal-predict}
validation$pred <- predict(
  fit_dev,
  newdata = validation,
  type = "response"
)

validation$lp <- predict(
  fit_dev,
  newdata = validation,
  type = "link"
)
```

# 36. Temporal Validation Discrimination

```{r temporal-auc}
roc_val <- roc(
  response = validation$Diabetes_num,
  predictor = validation$pred,
  quiet = TRUE
)

validation_auc <- as.numeric(auc(roc_val))
validation_auc_ci <- ci.auc(roc_val)

validation_auc
validation_auc_ci
```

```{r temporal-roc-plot, fig.height=5, fig.width=6}
plot(
  roc_val,
  print.auc = TRUE,
  main = "ROC Curve: Temporal Validation"
)
```

# 37. Temporal Validation Calibration

## 37.1 Calibration-in-the-large

```{r temporal-calibration-intercept}
cal_intercept_val_model <- glm(
  Diabetes_num ~ 1 + offset(lp),
  data = validation,
  family = binomial()
)

cal_intercept_val <- unname(
  coef(cal_intercept_val_model)[1]
)

cal_intercept_val
```

Ideal:

\[
\alpha = 0.
\]

## 37.2 Calibration slope

```{r temporal-calibration-slope}
cal_slope_val_model <- glm(
  Diabetes_num ~ lp,
  data = validation,
  family = binomial()
)

cal_slope_val <- unname(
  coef(cal_slope_val_model)[2]
)

cal_slope_val
```

Ideal:

\[
\gamma = 1.
\]

## 37.3 Decile calibration table

```{r temporal-calibration-table}
validation <- validation %>%
  mutate(
    risk_group = ntile(pred, 10)
  )

calibration_table <- validation %>%
  group_by(risk_group) %>%
  summarise(
    N = n(),
    Predicted = mean(pred),
    Observed = mean(Diabetes_num),
    .groups = "drop"
  )

calibration_table
```

```{r temporal-calibration-plot, fig.height=6, fig.width=7}
ggplot(
  calibration_table,
  aes(
    x = Predicted,
    y = Observed
  )
) +
  geom_point() +
  geom_line() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = 2
  ) +
  coord_equal() +
  labs(
    title = "Temporal Validation Calibration",
    x = "Mean Predicted Probability",
    y = "Observed Proportion"
  ) +
  theme_minimal()
```

# 38. Temporal Validation Brier Score

```{r temporal-brier}
brier_val <- mean(
  (validation$pred - validation$Diabetes_num)^2
)

brier_val
```

# 39. Recalibration

If discrimination remains acceptable but calibration has shifted, a recalibration model can be considered.

```{r recalibration-model}
recal_model <- glm(
  Diabetes_num ~ lp,
  data = validation,
  family = binomial()
)

coef(recal_model)
```

The updated linear predictor is

\[
LP_{new} = \alpha + \gamma LP_{old},
\]

where \(\alpha\) is the calibration intercept and \(\gamma\) is the calibration slope.

Predictions from the recalibrated model:

```{r recalibrated-predictions}
validation$pred_recalibrated <- predict(
  recal_model,
  type = "response"
)

head(
  validation %>%
    select(pred, pred_recalibrated)
)
```

# 40. Final Logistic Model for Individual Prediction

For an easy-to-deploy risk calculator, use the ordinary logistic model fitted to the completed modeling dataset.

```{r final-model}
final_fit <- glm(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_complete,
  family = binomial()
)

coef(final_fit)
```

The linear predictor is

\[
LP = \beta_0 + \sum_{j=1}^{p}\beta_j X_j,
\]

and the predicted probability is

\[
P(\text{Diabetes}) = \frac{e^{LP}}{1+e^{LP}}.
\]

# 41. Robust Construction of a New Patient

Factor levels can differ in capitalization across datasets. The following helper selects an available level using case-insensitive matching.

```{r factor-helper}
match_factor_level <- function(x, desired) {
  lvls <- levels(x)
  idx <- match(tolower(desired), tolower(lvls))

  if (is.na(idx)) {
    stop(
      "Requested level '", desired,
      "' not found. Available levels: ",
      paste(lvls, collapse = ", ")
    )
  }

  lvls[idx]
}
```

Inspect available levels:

```{r final-factor-levels}
levels(model_complete$Gender)
levels(model_complete$Race1)
levels(model_complete$PhysActive)
levels(model_complete$Smoke100)
```

Create an illustrative new patient:

```{r new-patient}
new_patient <- tibble(
  Age = 60,
  BMI = 32,
  BPSysAve = 145,
  TotChol = 5.2,
  Gender = factor(
    match_factor_level(model_complete$Gender, "male"),
    levels = levels(model_complete$Gender)
  ),
  Race1 = factor(
    match_factor_level(model_complete$Race1, "white"),
    levels = levels(model_complete$Race1)
  ),
  PhysActive = factor(
    match_factor_level(model_complete$PhysActive, "no"),
    levels = levels(model_complete$PhysActive)
  ),
  Smoke100 = factor(
    match_factor_level(model_complete$Smoke100, "yes"),
    levels = levels(model_complete$Smoke100)
  )
)

new_patient
```

# 42. Predict Diabetes Probability for the New Patient

```{r new-patient-prediction}
new_patient_probability <- predict(
  final_fit,
  newdata = new_patient,
  type = "response"
)

new_patient_probability
```

Because the `NHANES` example is cross-sectional, interpret this as a model-based probability of **prevalent diabetes status**, not future 5-year risk.

# 43. Nomogram

A nomogram can be created from the ordinary `lrm()` model.

```{r nomogram-model}
nom_fit <- lrm(
  Diabetes_num ~
    Age +
    BMI +
    BPSysAve +
    TotChol +
    Gender +
    Race1 +
    PhysActive +
    Smoke100,
  data = model_complete,
  x = TRUE,
  y = TRUE
)
```

```{r nomogram, fig.height=8, fig.width=11}
nom <- nomogram(
  nom_fit,
  fun = plogis,
  fun.at = c(0.05, 0.10, 0.20, 0.30, 0.50),
  funlabel = "Probability of Diabetes"
)

plot(nom)
```

# 44. Summary Performance Table

```{r performance-summary}
performance_summary <- tibble(
  Domain = c(
    "Development sample size",
    "Development event rate",
    "Apparent AUC",
    "Apparent Brier score",
    "Temporal validation sample size",
    "Temporal validation AUC",
    "Temporal validation calibration intercept",
    "Temporal validation calibration slope",
    "Temporal validation Brier score"
  ),
  Value = c(
    as.character(nrow(model_complete)),
    sprintf("%.3f", mean(model_complete$Diabetes_num)),
    sprintf("%.3f", as.numeric(auc(roc_obj))),
    sprintf("%.3f", brier_model),
    as.character(nrow(validation)),
    sprintf("%.3f", validation_auc),
    sprintf("%.3f", cal_intercept_val),
    sprintf("%.3f", cal_slope_val),
    sprintf("%.3f", brier_val)
  )
)

performance_summary
```

# 45. Recommended Interpretation Framework

A clinical prediction model should not be judged by AUC alone.

A useful summary is

\[
\boxed{
\text{Model Quality}
=
\text{Discrimination}
+
\text{Calibration}
+
\text{Prediction Error}
+
\text{Clinical Utility}
+
\text{Validation}
}
\]

## Discrimination

Typical measure:

\[
AUC/C\text{-statistic}.
\]

Question:

> Can the model rank participants with the outcome above those without the outcome?

## Calibration

Typical measures:

- Calibration plot
- Calibration intercept
- Calibration slope

Question:

> Are predicted probabilities numerically close to observed risks?

## Overall prediction error

Typical measure:

\[
\text{Brier Score}.
\]

## Clinical utility

Typical method:

\[
\text{Decision Curve Analysis}.
\]

Question:

> Does using the model provide greater clinical net benefit than reasonable default strategies over relevant threshold probabilities?

## Generalizability

Assess with:

- temporal validation
- geographic validation
- external institutional validation
- population validation

# 46. Full Workflow Recap

The complete process can be summarized as:

```text
Research question
        ↓
Define target population and outcome
        ↓
Choose clinically justified candidate predictors
        ↓
Check IDs and duplicates
        ↓
Check coding and variable types
        ↓
Check ranges and clinically impossible values
        ↓
Assess missingness
        ↓
Assess outliers
        ↓
EDA and descriptive statistics
        ↓
Check correlation and multicollinearity
        ↓
Assess continuous-variable functional forms
        ↓
Multiple imputation
        ↓
Fit prespecified/full model
        ↓
Consider nonlinear terms and interactions
        ↓
Consider penalization/shrinkage
        ↓
Estimate apparent performance
        ↓
Bootstrap internal validation
        ↓
Optimism-corrected discrimination
        ↓
Calibration assessment
        ↓
Brier score
        ↓
Decision-curve analysis
        ↓
Clinically justified threshold assessment
        ↓
Sensitivity analyses
        ↓
Subgroup performance
        ↓
Temporal/external validation
        ↓
Recalibration if needed
        ↓
Individual prediction
        ↓
Nomogram / calculator / deployment
```

# 47. Publication-Level Upgrades

For a real publication, consider the following upgrades beyond this teaching example:

1. Use the original CDC NHANES data rather than the teaching package dataset.
2. Account for NHANES sampling weights, strata, and PSU structure when the research question targets population inference.
3. Prespecify predictors based on clinical knowledge and prior literature.
4. Determine model sample size using prediction-model sample-size methods rather than relying only on the historical `10 events per variable` rule.
5. Avoid arbitrary categorization of continuous variables.
6. Use nonlinear modeling when clinically/statistically justified.
7. Perform multiple imputation within the resampling procedure for rigorous internal validation with missing data.
8. Preserve the complete model-building algorithm inside every bootstrap resample, including variable selection if variable selection is part of the algorithm.
9. Report optimism-corrected discrimination and calibration.
10. Validate the frozen model in independent data before clinical deployment.
11. Report according to current prediction-model reporting guidance applicable to the study.
12. Keep data-cleaning, derivation, modeling, validation, and reporting code reproducible and version controlled.

# 48. Session Information

Always save session information so the analysis can be reproduced later.

```{r session-info}
sessionInfo()
```
