---
title: "Clinical Statistical Modeling SOP"
author: "[Author / Study Team]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# 1. Purpose

This SOP provides a concise and reusable framework for clinical statistical modeling.

It is intended for common clinical and epidemiologic modeling tasks, including:

- binary outcomes,
- continuous outcomes,
- count outcomes,
- time-to-event outcomes,
- repeated-measures outcomes,
- and clinical prediction models.

The objective is to ensure that model development, assessment, validation, interpretation, and reporting are scientifically justified, reproducible, and clinically meaningful.

# 2. Define the Modeling Objective

Before analysis, clearly define:

- research question,
- target population,
- outcome,
- candidate predictors or covariates,
- time point or prediction horizon when relevant,
- intended model use,
- and primary estimand or performance target.

The modeling objective should be classified as one of the following:

1. **Explanation / association**
2. **Causal effect estimation**
3. **Prediction / risk estimation**

The model-building strategy and interpretation should follow the scientific objective.

# 3. Define the Analysis Population and Variables

Prespecify:

- inclusion criteria,
- exclusion criteria,
- outcome definition,
- predictor/covariate definitions,
- reference categories,
- baseline or measurement windows,
- and variable coding.

Variables that occur after the outcome or directly encode the outcome should not be used as predictors because they may introduce **outcome leakage**.

# 4. Data Quality Assessment

Before modeling, assess:

- duplicate subjects or records,
- variable types,
- categorical coding,
- clinically implausible values,
- inconsistent dates,
- logical inconsistencies,
- missing values,
- and unexpected distributions.

Potential errors should be investigated rather than automatically deleted.

A statistical outlier should not automatically be treated as a data error.

```{r data-quality-example, eval=FALSE}
# Duplicate IDs
sum(duplicated(data$id))

# Missing values
colSums(is.na(data))

# Basic distributions
summary(data)
```

# 5. Missing Data

Summarize missingness:

- overall,
- by variable,
- and, when useful, by outcome or treatment group.

The missing-data strategy should be prespecified.

Common approaches include:

- complete-case analysis when missingness is minimal and defensible,
- multiple imputation,
- or another justified model-based approach.

For multiple imputation, the imputation model should generally include:

- the outcome,
- candidate predictors/covariates,
- and variables related to missingness.

When prediction-model validation uses resampling, imputation should ideally be incorporated within the resampling procedure.

# 6. Descriptive Analysis and EDA

Before fitting the final model:

- summarize the outcome,
- summarize candidate predictors,
- inspect distributions,
- identify sparse categories,
- examine clinically important relationships,
- and evaluate potential outliers.

Continuous variables should generally be retained as continuous when possible rather than arbitrarily categorized.

# 7. Predictor / Covariate Selection

Candidate variables should preferably be selected using:

- clinical knowledge,
- prior literature,
- prespecified hypotheses,
- and scientific relevance.

Variables should not be removed solely because their univariable P-value exceeds 0.05.

Traditional stepwise selection may be used for exploratory purposes but is generally not preferred as the primary model-building strategy because it can produce unstable variable selection and optimistic performance.

When the number of candidate predictors is large relative to the available information, consider:

- LASSO,
- ridge regression,
- elastic net,
- or other shrinkage/penalization methods.

# 8. Model Selection by Outcome Type

The statistical model should match the outcome structure.

| Outcome | Common Model |
|---|---|
| Continuous | Linear regression |
| Binary | Logistic regression |
| Count | Poisson / Negative Binomial |
| Time-to-event | Cox proportional hazards model |
| Repeated continuous | Linear mixed model / MMRM |
| Repeated binary | GEE / GLMM |

The primary model should be prespecified whenever possible.

# 9. Functional Form of Continuous Variables

Do not automatically assume that continuous predictors have linear effects.

Assess nonlinear relationships when clinically or statistically plausible.

Common approaches include:

- restricted cubic splines,
- fractional polynomials,
- transformations,
- or clinically justified piecewise functions.

Example:

```{r spline-example, eval=FALSE}
library(rms)

fit_spline <- lrm(
  outcome ~ rcs(age, 4) + rcs(bmi, 4) + sex,
  data = model_data
)
```

Avoid arbitrary categorization unless there is a strong clinical reason.

# 10. Interaction Assessment

Interactions should be evaluated when there is a prespecified clinical or scientific rationale.

Example:

```{r interaction-example, eval=FALSE}
fit_interaction <- glm(
  outcome ~ treatment * sex + age + bmi,
  family = binomial(),
  data = model_data
)
```

When an interaction is important, interpretation should focus on the treatment or predictor effect within relevant levels of the interacting variable rather than interpreting the main effect alone.

Exploratory interaction testing should be interpreted cautiously.

# 11. Multicollinearity

Assess strong correlation among predictors when relevant.

Possible methods include:

- correlation matrices,
- variance inflation factors (VIF),
- condition indices,
- and clinical review of redundant variables.

Highly correlated variables may lead to unstable coefficients even when overall prediction remains acceptable.

# 12. Fit the Prespecified Model

Fit the prespecified or final candidate model using the selected modeling approach.

Example for a binary outcome:

```{r logistic-example, eval=FALSE}
fit <- glm(
  outcome ~ age + bmi + blood_pressure +
    cholesterol + sex + smoking,
  family = binomial(),
  data = model_data
)

summary(fit)
```

Model coefficients, standard errors, confidence intervals, and clinically relevant effect measures should be reported when inference is an objective.

# 13. Model Diagnostics

Model diagnostics should be appropriate to the model type.

Common checks include:

- residual diagnostics,
- influential observations,
- model convergence,
- sparse cells,
- proportional hazards for Cox models,
- distributional assumptions for linear models,
- overdispersion for count models,
- and random-effect assumptions for mixed models.

Influential observations should be investigated rather than automatically removed.

# 14. Model Performance

For prediction models, performance should generally be evaluated across several dimensions.

## 14.1 Discrimination

For binary models, commonly use:

- ROC curve,
- AUC / C-statistic.

For survival models, commonly use:

- Harrell's C-index,
- time-dependent AUC when appropriate.

Discrimination evaluates how well the model separates individuals with different outcomes.

## 14.2 Calibration

Assess agreement between predicted and observed outcomes.

Common measures include:

- calibration plot,
- calibration intercept,
- calibration slope,
- observed versus predicted risk.

Calibration is essential and should not be replaced by AUC alone.

## 14.3 Overall Prediction Error

For binary prediction models, the **Brier score** may be reported.

A lower Brier score indicates better overall prediction accuracy.

# 15. Internal Validation

Apparent performance calculated in the same data used to develop the model is usually optimistic.

Internal validation should be performed when the model is intended for prediction.

Preferred methods include:

- bootstrap validation,
- cross-validation.

Bootstrap validation can be used to estimate optimism-corrected:

- discrimination,
- calibration,
- and overall performance.

If variable selection, imputation, or tuning is part of the modeling algorithm, these steps should ideally be repeated within each resampling iteration.

# 16. External or Temporal Validation

When independent data are available, evaluate the frozen model in:

- a later time period,
- another institution,
- another geographic region,
- or another relevant population.

Assess at least:

- discrimination,
- calibration,
- and overall prediction error.

If discrimination remains acceptable but calibration has shifted, recalibration may be considered.

# 17. Clinical Utility

When the model is intended to support clinical decisions, evaluate whether predictions provide meaningful clinical benefit.

Decision-curve analysis (DCA) may be used to evaluate **net benefit** across clinically relevant threshold probabilities.

DCA should not be interpreted simply as an automatic method for choosing a cutoff.

Threshold-based measures such as:

- sensitivity,
- specificity,
- PPV,
- and NPV

should be reported only when the threshold is clinically justified.

# 18. Sensitivity Analyses

Perform sensitivity analyses that address important modeling assumptions.

Common examples include:

1. complete-case versus multiple-imputation analysis;
2. linear versus nonlinear functional forms;
3. alternative predictor sets;
4. exclusion versus inclusion of influential observations;
5. alternative outcome definitions;
6. penalized versus conventional regression;
7. alternative modeling assumptions.

Sensitivity analyses should evaluate whether conclusions are materially changed.

# 19. Subgroup Assessment

When clinically relevant, evaluate model performance or effects in prespecified subgroups.

Examples include:

- sex,
- age group,
- race/ethnicity when appropriate,
- disease severity,
- treatment group,
- or major comorbidity groups.

For prediction models, subgroup evaluation may include:

- AUC,
- calibration,
- observed versus predicted risk,
- and sample size/event counts.

Small subgroup results should be interpreted cautiously.

# 20. Reporting

The final report should clearly describe:

- modeling objective,
- study population,
- outcome definition,
- candidate variables,
- missing-data handling,
- model specification,
- nonlinear terms,
- interactions,
- model-selection strategy,
- diagnostics,
- validation method,
- performance measures,
- sensitivity analyses,
- and limitations.

For prediction models, recommended performance reporting includes:

| Domain | Common Measure |
|---|---|
| Discrimination | AUC / C-index |
| Calibration | Calibration plot, intercept, slope |
| Overall error | Brier score |
| Clinical utility | Decision-curve analysis |
| Validation | Bootstrap / external validation |

# 21. Reproducibility and Quality Control

The analysis should be reproducible.

Recommended practices include:

- set and document random seeds when needed,
- version-control analysis code,
- retain the data dictionary,
- document all derived variables,
- independently review key model code and outputs,
- retain model coefficients and model specification,
- and record software/package versions.

```{r session-info, eval=FALSE}
sessionInfo()
```

# Appendix A. Minimal Modeling Workflow

```text
Research question
        ↓
Define target population and outcome
        ↓
Define candidate predictors / covariates
        ↓
Data quality checks
        ↓
Missing-data assessment
        ↓
Descriptive analysis / EDA
        ↓
Check outliers and multicollinearity
        ↓
Assess continuous-variable functional forms
        ↓
Handle missing data
        ↓
Fit prespecified model
        ↓
Assess nonlinear terms / interactions
        ↓
Model diagnostics
        ↓
Evaluate performance
        ↓
Internal validation
        ↓
Sensitivity analyses
        ↓
External / temporal validation if available
        ↓
Clinical interpretation and reporting
```

# Appendix B. Pre-Modeling Checklist

Before final modeling, confirm:

- [ ] Research question is clearly defined
- [ ] Modeling objective is defined: association, causal effect, or prediction
- [ ] Target population is defined
- [ ] Outcome is clearly defined
- [ ] Candidate predictors/covariates are prespecified
- [ ] Outcome leakage variables are excluded
- [ ] Data quality checks are completed
- [ ] Missing-data strategy is specified
- [ ] Continuous-variable strategy is specified
- [ ] Interaction terms are prespecified when relevant
- [ ] Multicollinearity has been assessed
- [ ] Primary model is specified
- [ ] Model diagnostics are specified
- [ ] Validation strategy is specified
- [ ] Performance measures are specified
- [ ] Sensitivity analyses are specified
- [ ] Subgroup analyses are prespecified when relevant
- [ ] Reporting outputs are defined
- [ ] Analysis code is reproducible
