---
title: "Real-World Study"
description: "A practical R learning guide to Real-World Study, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Epidemiology & Real-World Evidence"
category: "Real-World Data & Evidence"
topic: "Real-World Study"
type: article
tags: ["Real-World Data & Evidence", "Epidemiology & Real-World Evidence", "Real-World", "Study"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Real-world evidence depends on whether routinely collected data can support a clearly specified causal or descriptive objective.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Real-World Study**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Real-World Study

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Real-World Study** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Real-World Study

<!-- generated-learning-note -->
This subsection explains **Real-World Study** and shows how it contributes to the overall Real-World Study workflow.

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r}
#| eval: false
#| code-fold: true
set.seed(20251006)
# setwd("C:\\Users\\hed2\\Downloads\\rws")

# Load packages
pkgs <- c("data.table", "dplyr", "tableone", "MatchIt", "cobalt",
          "survey", "survival", "survminer", "ggplot2", "broom", "sandwich", "lmtest")
# for (p in pkgs) if(!requireNamespace(p, quietly = TRUE)) install.packages(p)
lapply(pkgs, library, character.only = TRUE)

# ------------------------------------------------------
# 1. Simulate Real-World Dataset
# ------------------------------------------------------
n <- 1000
dat <- data.frame(
  patient_id = 1:n,
  exposure_group = sample(c(0, 1), n, replace = TRUE, prob = c(0.4, 0.6)),
  age_at_index = round(rnorm(n, 60, 10)),
  sex = sample(c("Male", "Female"), n, replace = TRUE),
  baseline_hba1c = round(rnorm(n, 7.5, 1.2), 1),
  comorbidity_score = pmax(0, round(rnorm(n, 2, 1))),
  region = sample(c("North", "South", "East", "West"), n, replace = TRUE)
)

# Simulate follow-up time and event
# treatment A Slightly reduce risk
dat <- dat %>%
  mutate(
    lambda = 0.03 + 0.002 * (age_at_index - 60) + 0.01 * comorbidity_score + ifelse(exposure_group=="A", -0.01, 0),
    follow_up_time = rexp(n, rate = lambda),
    event_flag = rbinom(n, 1, 1 - exp(-lambda * follow_up_time))
  )

# fwrite(dat, "analysis_dataset.csv")



# ------------------------------------------------------
# 2. Baseline Summary
# ------------------------------------------------------
vars <- c("age_at_index", "sex", "baseline_hba1c", "comorbidity_score", "region")
catVars <- c("sex", "region")
table1 <- CreateTableOne(vars = vars, strata = "exposure_group", data = dat, factorVars = catVars)
print(table1, smd = TRUE)



# ------------------------------------------------------
# 3. Propensity Score Modeling
# ------------------------------------------------------
ps_formula <- as.formula("exposure_group ~ age_at_index + sex + baseline_hba1c + comorbidity_score + region")
m.out <- matchit(ps_formula, data = dat, method = "nearest", ratio = 1, caliper = 0.2)
summary(m.out)

# Extract matched dataset
matched <- match.data(m.out)
cat("Matched sample size:", nrow(matched), "\n")

# Balance Visualization
love.plot(m.out, stats = "mean.diffs", threshold = 0.1, var.order = "unadjusted")



# ------------------------------------------------------
# 4. Cox Model on Matched Data
# ------------------------------------------------------
cox_mod <- coxph(Surv(follow_up_time, event_flag) ~ exposure_group + age_at_index + sex + comorbidity_score, data = matched)
summary(cox_mod)

# Kaplan-Meier Plot
fit_km <- survfit(Surv(follow_up_time, event_flag) ~ exposure_group, data = matched)
# summary(fit_km)
ggsurvplot(fit_km, data = matched, pval = T, risk.table = F, 
           title = "Kaplan-Meier Curve by Treatment Group")


# ------------------------------------------------------
# 5. Sensitivity Analysis: IPTW Inverse Probability Treatment Weighting
# ------------------------------------------------------
ps_model <- glm(exposure_group ~ age_at_index + sex + baseline_hba1c + comorbidity_score + region,
                data = dat, family = binomial)
dat$ps <- predict(ps_model, type = "response")
dat <- dat %>% mutate(iptw = ifelse(exposure_group == "A", 1/ps, 1/(1-ps)))   #Inverse Probability Treatment Weighting

# Cut off extreme weights
q <- quantile(dat$iptw, probs = c(0.01, 0.99))   #weight between q1 and q2
dat$iptw_trunc <- pmin(pmax(dat$iptw, q[1]), q[2])

# weight Cox
cox_ipw <- coxph(Surv(follow_up_time, event_flag) ~ exposure_group + age_at_index + sex + comorbidity_score,
                 data = dat,         weights = iptw_trunc)      
summary(cox_ipw)

```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.

## Interpreting the Results

Evaluate data provenance, eligibility, treatment assignment, follow-up, outcome ascertainment, confounding, and missingness. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
