---
title: "Longitudinal Targeted Maximum Likelihood Estimation"
output: html_document
date: '2023-05-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LMTLE: A Brief Introduction
This instruction (of causal inference) is for the complex longitudinal data.

In detail please see the reference by [Ashley I Naimi](https://github.com/ainaimi/mlci_shortcourse).  

### Inplementaion by hand
Constructing stabilized IP weights
```{r}
library(tidyverse)
# data management or manipulate to generate lag variables
a <- read_csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\2023_04_21-time-dependent.csv") %>% 
group_by(ID) %>%
  
# lag after group id
mutate(exposure_lag = lag(exposure, n = 1L,
default = 0), c1_lag = lag(c1, n = 1L,
default = 0), c2_lag = lag(c2, n = 1L,
default = 0)) %>%
ungroup()
a
```

Calculate stabilized IP weights by ID
```{r}
# numerator
num <- glm(exposure ~ factor(Int), data = a,
family = binomial("logit"))$fitted.values
# denominator
den <- glm(exposure ~ factor(Int) + exposure_lag +
c1 + c2 + c1_lag + c2_lag, data = a,
# using lag?
  
family = binomial("logit"))$fitted.values
```

Cumulative product wights by ID (each one has the different weights in different times)
```{r}
# ip
a <- a %>%
mutate(sw_ = (num/den) * exposure + ((1 -
num)/(1 - den)) * (1 - exposure)) %>%
  
# cumprod ip  
group_by(ID) %>%
mutate(sw = cumprod(sw_))  

head(cbind(a$ID, a$Int,a$sw,a$sw_))
  
a <- a %>% ungroup() %>%
select(-sw_)
```

Summary weights by time
```{r}
# summary weights
a %>%
group_by(Int) %>%
# group by time (close to 1) 
summarise(meanSW = mean(sw), maxSW = max(sw))

# a %>%
# group_by(ID) %>%
# # group by ID
# summarise(meanSW = mean(sw), maxSW = max(sw))  
```

Use these weights to fit an IP weighted MSM
```{r}
# aim to adjust the weight
library(lmtest)
library(sandwich)

modMSM <- lm(Y ~ factor(Int) + I(exposure/4),  #why divided by 4, otherwise, coefficient *4
data = a, weights = sw)

coeftest(modMSM, vcov. = vcovCL(modMSM, cluster = a$ID,
type = "HC3"))

```

### Longitudinal Targeted Maximum Likelihood Estimation by package

```{r}
library(ltmle)
```

```{r}
# read in data
a <- read_csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\2023_04_21-time-dependent.csv")
a %>%
print(n = 3)
```

Convert data from long to wide
```{r}
# data transform, without lag
# keep confounders
b <- a %>%
pivot_wider(names_from = Int, values_from = c(exposure,
c1, c2, Y)) %>%
mutate(Y_1 = if_else(is.na(Y_1), 1, Y_1),
Y_2 = if_else(is.na(Y_2), 1, Y_2),
Y_3 = if_else(is.na(Y_3), 1, Y_3),
Y_4 = if_else(is.na(Y_4), 1, Y_4)) %>%
select(exposure_1, c1_1, c2_1, Y_1, exposure_2,
c1_2, c2_2, Y_2, exposure_3, c1_3,
c2_3, Y_3, exposure_4, c1_4, c2_4,
Y_4)

head(b)
```

Setup parameters of model
```{r}
# ltmle
# super learner library, algorithm
sl.lib <- c("SL.mean",
"SL.glm",
"SL.ranger")

#ltmle
result <- ltmle(b,
Anodes=c(paste0("exposure_",1:4)),   #treatment
Lnodes=c("c1_1", "c2_1", "c1_2", "c2_2",  #
"c1_3", "c2_3", "c1_4", "c2_4"),
Ynodes=c("Y_1","Y_2","Y_3","Y_4"),

survivalOutcome = TRUE,
SL.library = list(Q = sl.lib, g = sl.lib), #g computation and ps?

abar=list(treament = c(1, 1, 1, 1),
control = c(0, 0, 0, 0)),
# estimate.time = T,
# need to comment out to run
stratify = T) 
```

```{r}
# summary(result)
summary(result)[[2]]$ATE
# result$fit$g
# result$fit$Q
```










