---
title: "Data wrangling"
description: "A practical R learning guide to Data wrangling, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Data Management & Programming"
category: "Data Cleaning & Quality Control"
topic: "Data wrangling"
type: article
tags: ["Data Cleaning & Quality Control", "Data Management & Programming", "Data", "wrangling"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Data cleaning and quality control make analytical results traceable and reduce errors caused by inconsistent or implausible values.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Data wrangling**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. How to do data wrangling
2. Load data and package
3. Select certain rows
4. Select certain columns
5. Rename variables
6. Sorting in ascending or descending order

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Data Management & Programming",
  category = "Data Cleaning & Quality Control",
  topic = "Data wrangling"
)
learning_overview
```

## Learning Material

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### How to do data wrangling  
We will use `tidyverse` package to work with data.  

##### Load data and package
```{r}
head (iris)
```
```{r}
library(tidyverse)
```
##### Select certain rows
```{r}
setosa <- filter(iris, Species == 'setosa')
setosa
```
##### Select certain columns
```{r}
select(iris, Sepal.Length, Species)
```

```{r}
select(iris, -Sepal.Length, -Species)
```
##### Rename variables
```{r}
rename(iris,  Sepal_Width= Sepal.Width,  Sepal_Length= Sepal.Length )
```
##### Sorting in ascending or descending order  
- put a minus in front of a variable for descending order
```{r}
arrange(iris, Petal.Length, -Petal.Width)
 
```
##### Transform variables
```{r}
mutate(iris,  newvar= Sepal.Width*10, Petal.Length=Petal.Length/100  )
```
##### Working with pipes %>% 
```{r}
iris %>% 
  filter(Species=="setosa") %>% 
  mutate (newvar=Sepal.Width*10) %>% 
  select (-Sepal.Width, -Petal.Width) %>% 
  arrange(-Sepal.Length, newvar)
```
##### Pivot wider (long to wide)  
- If no unique identifier row in each group **doesn't work**   

```{r}
iris %>%
pivot_wider(  names_from=Species, values_from= c(Sepal.Length))
```
- Create a unique identifier row for each name and then use pivot_wider 

```{r}
widedata <- iris %>%
  # create groups then assign unique identifier row number in each group
  group_by(Species) %>%
  mutate(row = row_number()) %>%
pivot_wider(names_from=Species, values_from= c(Petal.Length,Sepal.Length,Petal.Width,Sepal.Width))
widedata
```

```{r}
iris %>%
  group_by(Species) %>%
  mutate(row = row_number()) %>%
pivot_wider( names_from=Species, values_from= c(Petal.Length, Petal.Width))
```
##### Pivot longer (wide to long)   

```{r}
longdata = pivot_longer(widedata, -
                          c(  "row"   ,                  "Petal.Length_setosa" ,    "Petal.Length_versicolor",
  "Petal.Length_virginica",  "Sepal.Length_setosa"  ,   "Sepal.Length_versicolor",
  "Sepal.Length_virginica" , "Petal.Width_setosa"    ,  "Petal.Width_versicolor" ,
  "Petal.Width_virginica"    )        , names_to="Sepal.Width", values_to="Sepal.Width.value")
longdata
```
- Pivot wider again (long to wide)

```{r}
pivot_wider(longdata, names_from=Sepal.Width, values_from= c(Sepal.Width.value))

```
##### Separate columns       
```{r}
separate(iris, Species, into = c("integer","decimal","third"), sep="o")

```
##### Recode/relabel data    
```{r}
mutate(iris, Species2 = recode(Species, "setosa"="seto", "versicolor"="versi"))
```
##### Combine data sets  
- prepare data sets 
```{r}
data1 <- data.frame(ID = 1:4,                       
                    X1 = c("a1", "a2","a3", "a4"),
                    stringsAsFactors = FALSE)
data2 <- data.frame(ID = 2:5,                       
                    X2 = c("b1", "b2","b3", "b4"),
                    stringsAsFactors = FALSE)
```
- inner join   
```{r}
inner_join(data1, data2, by = "ID")    
```
- left join   
```{r}
left_join(data1, data2, by = "ID")   
```
- right join   
```{r}
right_join(data1, data2, by = "ID")  
```
- full join   
```{r}
full_join(data1, data2, by = "ID") 
```
- keep cases of left data table without in right data table 
```{r}
anti_join(data1, data2, by = "ID")  
```
- keep cases of left data table in right data table 
```{r}
semi_join(data1, data2, by = "ID")  
```
- multiple full join  
```{r}
full_join(data1, data2, by = "ID") %>%              
  full_join(., data2, by = "ID") 
```
- rbind **doesn't work** 
```{r}
# df1 <- data.frame(col1 = LETTERS[1:6],
#                   col2a = c(5:10),
#                   col3a = TRUE)
#   
# df2 <- data.frame(col1 = LETTERS[4:8],
#                   col2b= c(4:8),
#                   col3b = FALSE)
# rbind(df1,df2)
```
- append two data tables by using join and merge       
```{r}
data_frame1 <- data.frame(col1 = c(6:8),
                         col2 = letters[1:3],
                         col3 = c(1,4,NA))
 
data_frame2 <- data.frame(col1 = c(5:6),
                          col5 = letters[7:8])
 
data_frame_merge <- merge(data_frame1, data_frame2,
                          by = 'col1', all = TRUE)
 
print (data_frame_merge)
```
```{r}
full_join(data_frame1,data_frame2, by=c("col1"),)
```

## Interpreting the Results

Use summaries and visual checks to confirm ranges, distributions, duplicates, and missing-data patterns. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
