---
title: "Double Robust Estimators"
output: html_document
date: '2023-05-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using sl3 with AIPW

In detail please see the reference by [Ashley I Naimi](https://github.com/ainaimi/mlci_shortcourse).

```{r}
library(AIPW)
library(tmle3)
library(sl3)
library(tidyverse)
```

Scale the variables 
```{r}
# data manipulate,  standardization or centralization
scale_ <- function(x) {
(x - mean(x, na.rm = TRUE))/sd(x, na.rm = TRUE)
}

nhefs <- read.csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\nhefs_an.csv") %>%
mutate(wt_delta = as.numeric(wt82_71 >
median(wt82_71)), age = scale_(age),
sbp = scale_(sbp), dbp = scale_(dbp),
price71 = scale_(price71), tax71 = scale_(tax71)) %>%
select(-wt82_71)

head(nhefs)
```

Choose learners and configuration
```{r}
lrnr_mean <- make_learner(Lrnr_mean)
lrnr_glm <- make_learner(Lrnr_glm)

# ranger learner
grid_params <- list(num.trees = c(250, 500,
1000, 2000), mtry = c(2, 4, 6), min.node.size = c(50,
100))
grid <- expand.grid(grid_params, KEEP.OUT.ATTRS = FALSE)
lrnr_ranger <- vector("list", length = nrow(grid))
for (i in 1:nrow(grid)) {
lrnr_ranger[[i]] <- make_learner(Lrnr_ranger,
num.trees = grid[i, ]$num.trees,
mtry = grid[i, ]$mtry, min.node.size = grid[i,
]$min.node.size)
}
lrnr_ranger <- make_learner(Lrnr_ranger) ########### FLAG! ###########

# glmnet learner
grid_params <- seq(0, 1, by = 0.25)
lrnr_glmnet <- vector("list", length = length(grid_params))
for (i in 1:length(grid_params)) {
lrnr_glmnet[[i]] <- make_learner(Lrnr_glmnet,
                                 alpha = grid_params[i])
}
lrnr_glmnet <- make_learner(Lrnr_glmnet) ########### FLAG! ###########

# xgboost learner
grid_params <- list(max_depth = c(2, 4, 6,
8), eta = c(0.01, 0.1, 0.2), nrounds = c(50,
100, 500))
grid <- expand.grid(grid_params, KEEP.OUT.ATTRS = FALSE)
lrnr_xgboost <- vector("list", length = nrow(grid))
for (i in 1:nrow(grid)) {
lrnr_xgboost[[i]] <- make_learner(Lrnr_xgboost,
max_depth = grid[i, ]$max_depth,
eta = grid[i, ]$eta)
}
lrnr_xgboost <- make_learner(Lrnr_xgboost) ########## FLAG! ##########

# earth learner
grid_params <- c(2, 3, 4, 5, 6)
lrnr_earth <- vector("list", length = length(grid_params))
for (i in 1:length(grid_params)) {
lrnr_earth[[i]] <- make_learner(Lrnr_earth,
degree = grid_params[i])
}
lrnr_earth <- make_learner(Lrnr_earth) ############ FLAG! ############
```

Set a stack learners library
```{r}
stacklearner <- Stack$new(lrnr_mean, lrnr_glm,
lrnr_ranger, lrnr_glmnet, lrnr_xgboost,
lrnr_earth)

metalearner <- Lrnr_nnls$new(convex = T)

sl.lib <- Lrnr_sl$new(learners = stacklearner,
metalearner = metalearner)

```

Set parameters of the model
```{r}
outcome <- nhefs$wt_delta
exposure <- nhefs$qsmk
covariates <- nhefs[, c("age", "sbp", "dbp",
"price71", "tax71", "sex", "income",
"race")]

# set parameters
set.seed(123)
AIPW_SL <- AIPW$new(Y = outcome, A = exposure,
W = covariates, Q.SL.library = sl.lib,
g.SL.library = sl.lib, k_split = 10,
verbose = FALSE)$fit()$summary(g.bound = 0.025)$plot.p_score()
```

```{r}
# ouput ATE
print(AIPW_SL$result, digits = 2)
```


## Using sl3 with tmle3

Stack learners
```{r}
sl_ <- make_learner(Stack, unlist(list(lrnr_mean,
lrnr_glm, lrnr_ranger, lrnr_glmnet, lrnr_xgboost,lrnr_earth), recursive = TRUE))
```


```{r}
# DEFINE SL_Y AND SL_A We only need
# one, because they're the same
Q_learner <- Lrnr_sl$new(learners = sl_,  #stacking learners
metalearner = Lrnr_nnls$new(convex = T))

g_learner <- Lrnr_sl$new(learners = sl_,
metalearner = Lrnr_nnls$new(convex = T))

learner_list <- list(Y = Q_learner, A = g_learner)
```

Setup the TMLE3 model
```{r}
# PREPARE THE THINGS WE WANT TO FEED IN
# TO TMLE3
ate_spec <- tmle_ATE(treatment_level = 1,
control_level = 0)

nodes_ <- list(W = c("age", "sbp", "dbp",
"price71", "tax71", "sex", "income",
"race"), A = "qsmk", Y = "wt_delta")
 
# RUN TMLE3
set.seed(123)
tmle_fit_ <- tmle3(ate_spec, nhefs, nodes_,
learner_list)
print(tmle_fit_)
```

Model properties to diagnose (calculate PS and stabilized weights)

```{r}
tmle_task <- ate_spec$make_tmle_task(nhefs,
nodes_)

initial_likelihood <- ate_spec$make_initial_likelihood(tmle_task,
learner_list)

## save propensity score for diagnosis
# initial ps
propensity_score <- initial_likelihood$get_likelihoods(tmle_task)$A
# updated ps
propensity_score <- propensity_score * nhefs$qsmk +
(1 - propensity_score) * (1 - nhefs$qsmk)
```


```{r}
plap_ <- tibble(exposure = nhefs$qsmk, pscore = propensity_score)

# updated weight
plap_ <- plap_ %>%
mutate(sw = exposure * (mean(exposure)/propensity_score) +
(1 - exposure) * ((1 - mean(exposure))/(1 -
propensity_score)))

# distribution of 
summary(plap_$sw)
summary(propensity_score)
```

plot the final pscore  
```{r}
# ps overlap plot
ggplot(plap_) + geom_density(aes(pscore,
color = factor(exposure))) + scale_fill_manual(values = c("blue",
"orange")) + scale_x_continuous(expand = c(0,
0)) + scale_y_continuous(expand = c(0,
0))
```
Output p and q weights respectively by different learners
```{r}
# super learner coefficients for PS
# model
g_fit <- tmle_fit_$likelihood$factor_list[["A"]]$learner

g_fit$fit_object$full_fit$learner_fits$Lrnr_nnls_TRUE

# super learner coefficients for
# outcome model
Q_fit <- tmle_fit_$likelihood$factor_list[["Y"]]$learner
Q_fit$fit_object$full_fit$learner_fits$Lrnr_nnls_TRUE
```









