---
title: "IP Weighting and marginal structural models"
description: "A practical R learning guide to IP Weighting and marginal structural models, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Epidemiology & Real-World Evidence"
category: "Causal Methods"
topic: "IP Weighting and marginal structural models"
type: article
tags: ["Causal Methods", "Epidemiology & Real-World Evidence", "IP", "Weighting", "marginal"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Causal methods combine design assumptions and statistical adjustment to estimate well-defined treatment or exposure contrasts.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **IP Weighting and marginal structural models**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Inverse probability of treatment weight
2. Data preparation
3. Stochastic Weight Averaging
4. Treatment is a continuous variable
5. The outcome is a dichotomous variable
6. Effect Modification variable

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **IP Weighting and marginal structural models** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Reference book: Hernán, Miguel A, and James M Robins. 2020. Causal Inference: What If. Boca Raton: Chapman & Hall/CRC.](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)

#### Inverse probability of treatment weight

<!-- generated-learning-note -->
This subsection explains **Inverse probability of treatment weight** and shows how it contributes to the overall IP Weighting and marginal structural models workflow.


#### Data preparation

<!-- generated-learning-note -->
This section prepares the information needed for **Data preparation**. Check variable definitions, units, coding, and missing values before moving to analysis.


- Load data

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r pressure, echo=FALSE}
#| eval: false
#| code-fold: true
library("readxl")
library(tidyverse)
library(here)
nhefs <- read.csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\nhefs.csv")
```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.




- Data transform


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r cars}
nhefs$cens <- ifelse(is.na(nhefs$wt82), 1, 0)   

nhefs.nmv <- nhefs[which(!is.na(nhefs$wt82)),] #   pro visional ignore subjects with missing values for weight in 1982 

```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


- Data exploration

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}

lm(wt82_71 ~ qsmk, data=nhefs.nmv) # qsmk  
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


- Averages in different groups

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
predict(lm(wt82_71 ~ qsmk, data=nhefs.nmv), data.frame(qsmk=1)) # Smoking cessation
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
predict(lm(wt82_71 ~ qsmk, data=nhefs.nmv), data.frame(qsmk=0)) # No smoking cessation
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.

- Problem: age is different (unbalance between two groups)

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# Table
summary(nhefs.nmv[which(nhefs.nmv$qsmk==0),]$age)
summary(nhefs.nmv[which(nhefs.nmv$qsmk==1),]$age)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.



#### Inverse probability of treatment weight

<!-- generated-learning-note -->
This subsection explains **Inverse probability of treatment weight** and shows how it contributes to the overall IP Weighting and marginal structural models workflow.


(Weighting the exposure) 

- Using logistic regression to calculate the probability 


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# Estimation of ip weights via a logistic model
fit <- glm(qsmk ~ sex + race + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
           family = binomial(), data = nhefs.nmv)
summary(fit)
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


- Calculate the weight (1/probability)

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
p.qsmk.obs <- ifelse(nhefs.nmv$qsmk == 0, 1 - predict(fit, type = "response"),
                     predict(fit, type = "response"))

nhefs.nmv$w <- 1/p.qsmk.obs

summary(nhefs.nmv$w)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


- Modeling the weighted linear regression

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library("geepack")
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# what is id
msm.w <- geeglm(wt82_71 ~ qsmk, data=nhefs.nmv, weights=w,id = seqn,corstr="independence")
summary(msm.w)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.



- Balance check in this pseudo-population


<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# no association between sex and qsmk in pseudo-population after weighting
xtabs(nhefs.nmv$w ~ nhefs.nmv$sex + nhefs.nmv$qsmk)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
# "check" for positivity (White women)
table(nhefs.nmv$age[nhefs.nmv$race == 0 & nhefs.nmv$sex == 1], 
      nhefs.nmv$qsmk[nhefs.nmv$race == 0 & nhefs.nmv$sex == 1])
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.


`The inverse probability weight at this time can be: the treatment group is Pr[A=1]/f(A|L) (Or 1/f (A|L), Or 0.5/f (A|L)), and the non-treatment group is Pr[A=0]/f(A|L). WA =1/ f (A| L) is called non-stable weights, while SWA = f (A)/ f (A| L) is called stable weights.`

#### Stochastic Weight Averaging

<!-- generated-learning-note -->
This subsection explains **Stochastic Weight Averaging** and shows how it contributes to the overall IP Weighting and marginal structural models workflow.


- Estimation of denominator

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# estimation of denominator of ip weights (same as above, conditional probability)
denom.fit <- glm(qsmk ~ as.factor(sex) + as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                 family = binomial(), data = nhefs.nmv)
summary(denom.fit)

pd.qsmk <- predict(denom.fit, type = "response")
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


- Estimation of numerator

<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# Estimation of numerator of ip weights (probability (or mean) of treatment, marginal probability)
numer.fit <- glm(qsmk~1, family = binomial(), data = nhefs.nmv)
summary(numer.fit)

pn.qsmk <- predict(numer.fit, type = "response")
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


- Calculate the IPW

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
nhefs.nmv$sw <- ifelse(nhefs.nmv$qsmk == 0, ((1-pn.qsmk)/(1-pd.qsmk)),
                    (pn.qsmk/pd.qsmk))

# 0.257 is the probability of a treatment
summary(nhefs.nmv$sw)

summary(nhefs.nmv$w)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


- Calculate ATE

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
msm.sw <- geeglm(wt82_71 ~ qsmk, data=nhefs.nmv, weights=sw, id=seqn,
                corstr="independence")
summary(msm.sw)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


- Balance assessing

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
# no association between sex and qsmk in pseudo-population
xtabs(nhefs.nmv$sw ~ nhefs.nmv$sex + nhefs.nmv$qsmk)
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.



#### Treatment is a continuous variable

<!-- generated-learning-note -->
This section prepares the information needed for **Treatment is a continuous variable**. Check variable definitions, units, coding, and missing values before moving to analysis.

  
We assume that the probability density f (A|L) is a normal distribution, u=E(A|L), and the density f(A) in the molecule is also normally distributed. Because effect estimates are very sensitive to model choice for the conditional density f(A|L), we need to be very careful when using inverse probability weighting for continuous variables.


<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r}
# Analysis restricted to subjects reporting <=25 cig/day at baseline
nhefs.nmv.s <- subset(nhefs.nmv, smokeintensity <=25)

# estimation of denominator of ip weights
den.fit.obj <- lm(smkintensity82_71 ~ as.factor(sex) + 
  as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + I(smokeintensity^2) +
  smokeyrs + I(smokeyrs^2) + as.factor(exercise) + as.factor(active) + wt71  + 
  I(wt71^2), data = nhefs.nmv.s)

p.den <- predict(den.fit.obj, type = "response")

dens.den <- dnorm(nhefs.nmv.s$smkintensity82_71, p.den, summary(den.fit.obj)$sigma) # ipw denominator
# dnorm(x, mean, sd)

# estimation of numerator of ip weights
num.fit.obj <- lm(smkintensity82_71 ~ 1, data = nhefs.nmv.s) # ipw numerator
p.num <- predict(num.fit.obj, type = "response")
dens.num <- dnorm(nhefs.nmv.s$smkintensity82_71, p.num, summary(num.fit.obj)$sigma)

nhefs.nmv.s$sw.a = dens.num/dens.den
summary(nhefs.nmv.s$sw.a)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
msm.sw.cont <- geeglm(wt82_71 ~ smkintensity82_71 + I(smkintensity82_71*smkintensity82_71), 
                 data=nhefs.nmv.s, weights=sw.a, id=seqn, corstr="independence")
summary(msm.sw.cont)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.



#### The outcome is a dichotomous variable

<!-- generated-learning-note -->
This section prepares the information needed for **The outcome is a dichotomous variable**. Check variable definitions, units, coding, and missing values before moving to analysis.



<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
table(nhefs.nmv$qsmk, nhefs.nmv$death) # qsmk is treatment # death is outcome
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# First, estimation of stabilized weights sw.a  

# Second, fit logistic model below # nhefs.nmv.s/nhefs.nmv
msm.logistic <- geeglm(death ~ qsmk, data=nhefs.nmv.s, weights=sw.a, # sw /sw.a
                      id=seqn, family=binomial(), corstr="independence")
summary(msm.logistic)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.





#### Effect Modification variable

<!-- generated-learning-note -->
This section prepares the information needed for **Effect Modification variable**. Check variable definitions, units, coding, and missing values before moving to analysis.


However, when effect modification is present (exposure and a modifier, qsmk and sex), we can put other covariates in the marginal structure model, even if these covariates are not confounding variables. So our `numerator` of stable weights can be f(A) or f(A|V). 


<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
table(nhefs.nmv$sex)
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# estimation of denominator of ip weights
denom.fit <- glm(qsmk ~ as.factor(sex) + as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                 family = binomial(), data = nhefs.nmv) # denominator
summary(denom.fit)

pd.qsmk <- predict(denom.fit, type = "response")

```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# estimation of numerator of ip weights, sex is potient effect modification
numer.fit <- glm(qsmk~as.factor(sex), family = binomial(), data = nhefs.nmv) # numerator
summary(numer.fit)

pn.qsmk <- predict(numer.fit, type = "response")
```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
nhefs.nmv$sw.a <- ifelse(nhefs.nmv$qsmk == 0, ((1-pn.qsmk)/(1-pd.qsmk)),
                     (pn.qsmk/pd.qsmk))

summary(nhefs.nmv$sw.a)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# Estimating parameters of a marginal structural mean model
msm.emm <- geeglm(wt82_71~as.factor(qsmk) + as.factor(sex) 
                  + as.factor(qsmk):as.factor(sex), data=nhefs.nmv, 
                  weights=sw.a, id=seqn, corstr="independence")
summary(msm.emm)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


If we want to explore the interaction between two treatment variables A and B, we need to put both A and B into the marginal structure model. When calculating the inverse probability weight, the denominator will include these two joint distribution of variables.


#### Handling censoring and Missing Values

<!-- generated-learning-note -->
This subsection explains **Handling censoring and Missing Values** and shows how it contributes to the overall IP Weighting and marginal structural models workflow.


A quick view for censoring data


<!-- generated-learning-note -->
**Purpose.** This block calculates descriptive summaries. Check the denominator, grouping variables, units, and treatment of missing values.

```{r}
table(nhefs$qsmk, nhefs$cens)
```

<!-- generated-learning-note -->
**How to read the output.** Use the displayed values to understand center, spread, frequency, and missingness; investigate any implausible or unexpected result.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
summary(nhefs[which(nhefs$cens==0),]$wt71)
summary(nhefs[which(nhefs$cens==1),]$wt71)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


Calculate the weights for treatment and censor

- For completed data


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# for completed data

# estimation of denominator of ip weights for A
denom.fit <- glm(qsmk ~ as.factor(sex) + as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                 family = binomial(), data = nhefs)
summary(denom.fit)

pd.qsmk <- predict(denom.fit, type = "response")

```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.




<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# estimation of numerator of ip weights for A
numer.fit <- glm(qsmk~1, family = binomial(), data = nhefs)
summary(numer.fit)

pn.qsmk <- predict(numer.fit, type = "response")

```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


- For censoring data 


<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# for censoring data 
# as.factor(qsmk) likes a effect modification

# estimation of denominator of ip weights for C
denom.cens <- glm(cens ~ as.factor(qsmk) + as.factor(sex) + 
  as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                  family = binomial(), data = nhefs)
summary(denom.cens)

pd.cens <- 1-predict(denom.cens, type = "response")

```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.



<!-- generated-learning-note -->
**Purpose.** This block specifies and fits the statistical model. Identify the outcome, predictors, reference levels, link or distribution, and any correlation structure before interpreting coefficients.

```{r}
# estimation of numerator of ip weights for Censor data
numer.cens <- glm(cens~ as.factor(qsmk), family = binomial(), data = nhefs)
summary(numer.cens)

pn.cens <- 1-predict(numer.cens, type = "response")

```

<!-- generated-learning-note -->
**How to read the output.** Focus on coefficient direction and scale, uncertainty intervals, fit statistics, and diagnostics. Statistical significance alone does not establish practical importance or model adequacy.


- Time treatment_w and censor_w 

P(A|L)~P(A|L,C)*P(C|A,L) 

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
# times a_w and censor_w

nhefs$sw.a <- ifelse(nhefs$qsmk == 0, ((1-pn.qsmk)/(1-pd.qsmk)),
                     (pn.qsmk/pd.qsmk))

nhefs$sw.c <- pn.cens/pd.cens   

nhefs$sw <- nhefs$sw.c* nhefs$sw.a  #times

summary(nhefs$sw.a)

summary(nhefs$sw)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.



<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
msm.sw <- geeglm(wt82_71~qsmk, data=nhefs, 
                  weights=sw, id=seqn, corstr="independence")
summary(msm.sw)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.


- CI

<!-- generated-learning-note -->
**Purpose.** This block extracts inferential or predictive results from fitted objects. Match each estimate, interval, contrast, or diagnostic to the quantity it represents.

```{r}
beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[,2]
lcl <- beta-qnorm(0.975)*SE 
ucl <- beta+qnorm(0.975)*SE
cbind(beta, lcl, ucl)
```

<!-- generated-learning-note -->
**How to read the output.** Interpret estimates on the scale reported by the function, and use intervals or other uncertainty measures when drawing conclusions.

## Interpreting the Results

Check nuisance models, overlap, weights, balance, diagnostics, estimand alignment, and sensitivity to modeling choices. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
