---
title: "IP Weighting and marginal structural models"
output: html_document
date: '2023-05-03'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Reference book: Hernán, Miguel A, and James M Robins. 2020. Causal Inference: What If. Boca Raton: Chapman & Hall/CRC.](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)

### Inverse probability of treatment weight

#### Data preparation

- Load data

```{r pressure, echo=FALSE}
library("readxl")
library(tidyverse)
library(here)
nhefs <- read.csv("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2\\nhefs.csv")
```



- Data transform

```{r cars}
nhefs$cens <- ifelse(is.na(nhefs$wt82), 1, 0)   

nhefs.nmv <- nhefs[which(!is.na(nhefs$wt82)),] #   pro visional ignore subjects with missing values for weight in 1982 

```

- Data exploration
```{r}

lm(wt82_71 ~ qsmk, data=nhefs.nmv) # qsmk  
```

- Averages in different groups
```{r}
predict(lm(wt82_71 ~ qsmk, data=nhefs.nmv), data.frame(qsmk=1)) # Smoking cessation
```

```{r}
predict(lm(wt82_71 ~ qsmk, data=nhefs.nmv), data.frame(qsmk=0)) # No smoking cessation
```
- Problem: age is different (unbalance between two groups)
```{r}
# Table
summary(nhefs.nmv[which(nhefs.nmv$qsmk==0),]$age)
summary(nhefs.nmv[which(nhefs.nmv$qsmk==1),]$age)
```


### Inverse probability of treatment weight 

(Weighting the exposure) 

- Using logistic regression to calculate the probability 

```{r}
# Estimation of ip weights via a logistic model
fit <- glm(qsmk ~ sex + race + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
           family = binomial(), data = nhefs.nmv)
summary(fit)
```

- Calculate the weight (1/probability)
```{r}
p.qsmk.obs <- ifelse(nhefs.nmv$qsmk == 0, 1 - predict(fit, type = "response"),
                     predict(fit, type = "response"))

nhefs.nmv$w <- 1/p.qsmk.obs

summary(nhefs.nmv$w)
```

- Modeling the weighted linear regression
```{r}
library("geepack")
```

```{r}
# what is id
msm.w <- geeglm(wt82_71 ~ qsmk, data=nhefs.nmv, weights=w,id = seqn,corstr="independence")
summary(msm.w)
```


- Balance check in this pseudo-population

```{r}
# no association between sex and qsmk in pseudo-population after weighting
xtabs(nhefs.nmv$w ~ nhefs.nmv$sex + nhefs.nmv$qsmk)
```

```{r}
# "check" for positivity (White women)
table(nhefs.nmv$age[nhefs.nmv$race == 0 & nhefs.nmv$sex == 1], 
      nhefs.nmv$qsmk[nhefs.nmv$race == 0 & nhefs.nmv$sex == 1])
```

`The inverse probability weight at this time can be: the treatment group is Pr[A=1]/f(A|L) (Or 1/f (A|L), Or 0.5/f (A|L)), and the non-treatment group is Pr[A=0]/f(A|L). WA =1/ f (A| L) is called non-stable weights, while SWA = f (A)/ f (A| L) is called stable weights.`

#### Stochastic Weight Averaging

- Estimation of denominator
```{r}
# estimation of denominator of ip weights (same as above, conditional probability)
denom.fit <- glm(qsmk ~ as.factor(sex) + as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                 family = binomial(), data = nhefs.nmv)
summary(denom.fit)

pd.qsmk <- predict(denom.fit, type = "response")
```

- Estimation of numerator
```{r}
# Estimation of numerator of ip weights (probability (or mean) of treatment, marginal probability)
numer.fit <- glm(qsmk~1, family = binomial(), data = nhefs.nmv)
summary(numer.fit)

pn.qsmk <- predict(numer.fit, type = "response")
```

- Calculate the IPW
```{r}
nhefs.nmv$sw <- ifelse(nhefs.nmv$qsmk == 0, ((1-pn.qsmk)/(1-pd.qsmk)),
                    (pn.qsmk/pd.qsmk))

# 0.257 is the probability of a treatment
summary(nhefs.nmv$sw)

summary(nhefs.nmv$w)
```

- Calculate ATE
```{r}
msm.sw <- geeglm(wt82_71 ~ qsmk, data=nhefs.nmv, weights=sw, id=seqn,
                corstr="independence")
summary(msm.sw)
```

- Balance assessing
```{r}
# no association between sex and qsmk in pseudo-population
xtabs(nhefs.nmv$sw ~ nhefs.nmv$sex + nhefs.nmv$qsmk)
```


#### Treatment is a continuous variable
  
We assume that the probability density f (A|L) is a normal distribution, u=E(A|L), and the density f(A) in the molecule is also normally distributed. Because effect estimates are very sensitive to model choice for the conditional density f(A|L), we need to be very careful when using inverse probability weighting for continuous variables.

```{r}
# Analysis restricted to subjects reporting <=25 cig/day at baseline
nhefs.nmv.s <- subset(nhefs.nmv, smokeintensity <=25)

# estimation of denominator of ip weights
den.fit.obj <- lm(smkintensity82_71 ~ as.factor(sex) + 
  as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + I(smokeintensity^2) +
  smokeyrs + I(smokeyrs^2) + as.factor(exercise) + as.factor(active) + wt71  + 
  I(wt71^2), data = nhefs.nmv.s)

p.den <- predict(den.fit.obj, type = "response")

dens.den <- dnorm(nhefs.nmv.s$smkintensity82_71, p.den, summary(den.fit.obj)$sigma) # ipw denominator
# dnorm(x, mean, sd)

# estimation of numerator of ip weights
num.fit.obj <- lm(smkintensity82_71 ~ 1, data = nhefs.nmv.s) # ipw numerator
p.num <- predict(num.fit.obj, type = "response")
dens.num <- dnorm(nhefs.nmv.s$smkintensity82_71, p.num, summary(num.fit.obj)$sigma)

nhefs.nmv.s$sw.a = dens.num/dens.den
summary(nhefs.nmv.s$sw.a)
```

```{r}
msm.sw.cont <- geeglm(wt82_71 ~ smkintensity82_71 + I(smkintensity82_71*smkintensity82_71), 
                 data=nhefs.nmv.s, weights=sw.a, id=seqn, corstr="independence")
summary(msm.sw.cont)
```


#### The outcome is a dichotomous variable

```{r}
table(nhefs.nmv$qsmk, nhefs.nmv$death) # qsmk is treatment # death is outcome
```

```{r}
# First, estimation of stabilized weights sw.a  

# Second, fit logistic model below # nhefs.nmv.s/nhefs.nmv
msm.logistic <- geeglm(death ~ qsmk, data=nhefs.nmv.s, weights=sw.a, # sw /sw.a
                      id=seqn, family=binomial(), corstr="independence")
summary(msm.logistic)
```




#### Effect Modification variable

However, when effect modification is present (exposure and a modifier, qsmk and sex), we can put other covariates in the marginal structure model, even if these covariates are not confounding variables. So our `numerator` of stable weights can be f(A) or f(A|V). 

```{r}
table(nhefs.nmv$sex)
```

```{r}
# estimation of denominator of ip weights
denom.fit <- glm(qsmk ~ as.factor(sex) + as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                 family = binomial(), data = nhefs.nmv) # denominator
summary(denom.fit)

pd.qsmk <- predict(denom.fit, type = "response")

```

```{r}
# estimation of numerator of ip weights, sex is potient effect modification
numer.fit <- glm(qsmk~as.factor(sex), family = binomial(), data = nhefs.nmv) # numerator
summary(numer.fit)

pn.qsmk <- predict(numer.fit, type = "response")
```

```{r}
nhefs.nmv$sw.a <- ifelse(nhefs.nmv$qsmk == 0, ((1-pn.qsmk)/(1-pd.qsmk)),
                     (pn.qsmk/pd.qsmk))

summary(nhefs.nmv$sw.a)
```

```{r}
# Estimating parameters of a marginal structural mean model
msm.emm <- geeglm(wt82_71~as.factor(qsmk) + as.factor(sex) 
                  + as.factor(qsmk):as.factor(sex), data=nhefs.nmv, 
                  weights=sw.a, id=seqn, corstr="independence")
summary(msm.emm)
```

If we want to explore the interaction between two treatment variables A and B, we need to put both A and B into the marginal structure model. When calculating the inverse probability weight, the denominator will include these two joint distribution of variables.


#### Handling censoring and Missing Values

A quick view for censoring data

```{r}
table(nhefs$qsmk, nhefs$cens)
```

```{r}
summary(nhefs[which(nhefs$cens==0),]$wt71)
summary(nhefs[which(nhefs$cens==1),]$wt71)
```

Calculate the weights for treatment and censor

- For completed data

```{r}
# for completed data

# estimation of denominator of ip weights for A
denom.fit <- glm(qsmk ~ as.factor(sex) + as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                 family = binomial(), data = nhefs)
summary(denom.fit)

pd.qsmk <- predict(denom.fit, type = "response")

```


```{r}
# estimation of numerator of ip weights for A
numer.fit <- glm(qsmk~1, family = binomial(), data = nhefs)
summary(numer.fit)

pn.qsmk <- predict(numer.fit, type = "response")

```

- For censoring data 

```{r}
# for censoring data 
# as.factor(qsmk) likes a effect modification

# estimation of denominator of ip weights for C
denom.cens <- glm(cens ~ as.factor(qsmk) + as.factor(sex) + 
  as.factor(race) + age + I(age^2) + 
  as.factor(education) + smokeintensity + 
  I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + 
  as.factor(exercise) + as.factor(active) + wt71 + I(wt71^2), 
                  family = binomial(), data = nhefs)
summary(denom.cens)

pd.cens <- 1-predict(denom.cens, type = "response")

```

```{r}
# estimation of numerator of ip weights for Censor data
numer.cens <- glm(cens~ as.factor(qsmk), family = binomial(), data = nhefs)
summary(numer.cens)

pn.cens <- 1-predict(numer.cens, type = "response")

```

- Time treatment_w and censor_w 

P(A|L)~P(A|L,C)*P(C|A,L) 
```{r}
# times a_w and censor_w

nhefs$sw.a <- ifelse(nhefs$qsmk == 0, ((1-pn.qsmk)/(1-pd.qsmk)),
                     (pn.qsmk/pd.qsmk))

nhefs$sw.c <- pn.cens/pd.cens   

nhefs$sw <- nhefs$sw.c* nhefs$sw.a  #times

summary(nhefs$sw.a)

summary(nhefs$sw)
```

```{r}
msm.sw <- geeglm(wt82_71~qsmk, data=nhefs, 
                  weights=sw, id=seqn, corstr="independence")
summary(msm.sw)
```

- CI
```{r}
beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[,2]
lcl <- beta-qnorm(0.975)*SE 
ucl <- beta+qnorm(0.975)*SE
cbind(beta, lcl, ucl)
```



 

















