---
title: "How to calculate the prediction interval for LMM"
description: "A practical R learning guide to How to calculate the prediction interval for LMM, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Longitudinal & Repeated Measures"
topic: "How to calculate the prediction interval for LMM"
type: article
tags: ["Longitudinal & Repeated Measures", "Biostatistics", "calculate", "prediction", "interval"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  enabled: false
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Repeated observations from the same unit are correlated, so longitudinal models must represent both mean change and within-unit dependence.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **How to calculate the prediction interval for LMM**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. How to calculate the prediction interval for LMM

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Biostatistics",
  category = "Longitudinal & Repeated Measures",
  topic = "How to calculate the prediction interval for LMM"
)
learning_overview
```

## Learning Material

The following material develops **How to calculate the prediction interval for LMM** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### How to calculate the prediction interval for LMM

<!-- generated-learning-note -->
This subsection explains **How to calculate the prediction interval for LMM** and shows how it contributes to the overall How to calculate the prediction interval for LMM workflow.


Using ggpredict function

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(nlme)
library(ggeffects)
library(tidyverse)
 
m <- fm2 <- lme(distance ~ age + Sex, data = Orthodont, random = ~ 1|Subject) 

ggpredict(m, "age",interval = "confidence") %>% plot(rawdata = T, dot.alpha = 0.2)

ggpredict(m, "age", type = "re",interval = "prediction") %>% plot(rawdata = T, dot.alpha = 0.2)

outcome1=ggpredict(m, terms = "age [15:40]")
head(outcome1)

outcome2=ggpredict(m, "age", type = "re",interval = "prediction")
head(outcome2)

# read ggpredict docment
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

Using predictInterval function

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
# Lsmeans gives a Confidence Interval, and predictInterval gives a prediction interval. 

library(lsmeans)
library(lme4)
library(nlme)

# fit model
n1 <- lmer(distance ~ age + Sex+(1|Subject), data = Orthodont ) 
n2 <-  lme(distance ~ age + Sex, data = Orthodont, random = ~ 1|Subject) 

summary(n1)
summary(n2)

# lsmeans, confidence interval 
lsmeans(n1,   list (pairwise ~ Sex ), at = list(age=11))  
lsmeans(n2,   list (pairwise ~ Sex ), at = list(age=11))  
lsmeans(n1,   "age")  
 
# first method: predictInterval
library(merTools)
new.data <- data.frame(age=11, Subject='M01',Sex="Male")
pI <- predictInterval(n1,newdata=new.data, which='fixed',level=0.95)
pI  # no applicable method for 'getME' applied to an object of class "lme"

# 1 predictInterval, using the original data xs, then aggregate, the overall effect
pI <- predictInterval(n1,newdata=Orthodont ,level=0.95)
pred.data <- data.frame(Orthodont,pred=pI[1],pred_low=pI[3],pred_up=pI[2])
mean((pred.data$distance-pred.data$fit)^2)

# 2 fixed effect
pI <- predictInterval(n1,newdata=Orthodont, which='fixed',level=0.95)
pred.data <- data.frame(Orthodont,pred=pI[1],pred_low=pI[3],pred_up=pI[2])
mean((pred.data$distance-pred.data$fit)^2)

# 3 random effect 
pI <- predictInterval(n1,newdata=Orthodont, which='random',level=0.95)
pred.data <- data.frame(Orthodont,pred=pI[1],pred_low=pI[3],pred_up=pI[2])
mean((pred.data$distance-pred.data$fit)^2)

############## random and fixed effects
m1 <- lmer(Reaction ~ Days + (1 | Subject), sleepstudy)

intFit <- predictInterval(m1, newdata = sleepstudy ) # bounded values
head(intFit)

intFit2 <- predictInterval(m1, newdata = sleepstudy ,which = "random") # bounded values
head(intFit2)

intFit3 <- predictInterval(m1, newdata = sleepstudy ,which = "fixed") # bounded values
head(intFit3)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


Using Simulation

<!-- generated-learning-note -->
**Purpose.** This block generates controlled example data. Simulation makes the data-generating assumptions explicit and provides a known setting in which to study the method.

```{r}

# second method: Simulation. This has result similar to predictInterval
sim <- simulate(n1, nsim = 1, newdata=new.data) #based on distribution with random error, like MCMC

s <- numeric()

for (i in 1:500) {
  s[i] <- simulate(n1, nsim = 1, newdata=new.data)[1]
}
s <- unlist(s)
hist(s)
quantile(s,0.025)
quantile(s,0.975)

# third method:
#   calculate mean and std, then calculate percentile by using normal distribution

intFit4 <- predictInterval(n1, newdata = new.data,level=0.95 ) # bounded values
head(intFit4) #using the overall effect

```

<!-- generated-learning-note -->
**How to read the output.** Compare the generated sample with the parameters used to create it. Differences reflect random variation unless the code intentionally introduces bias or structure.



For general linear regression

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(ggplot2)
library(nlraa)
library(nlme)
library(mgcv)

data(Oats, package = "nlme")
## A subset for simplicity
Oats.I <- subset(Oats, subset = Block == "I")
plot(Oats.I)

# fitting a linear model
fm1 <- lm(yield ~ nitro, data = Oats.I)
fm1.prd <- predict(fm1, interval = "conf")
Oats.IA <- cbind(Oats.I, fm1.prd)

## Make a plot
ggplot(data = Oats.IA, aes(x = nitro, y = yield)) + 
  geom_point() + 
  geom_line(aes(y = fit)) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "purple", alpha = 0.4)

# bootstrapped confidence bands
fm1.boot <- boot_lm(fm1, fitted, R=200)  #randomly sampling data 
fm1.boot.prd <- summary_simulate(t(fm1.boot$t))
Oats.IAB <- cbind(Oats.I, fm1.boot.prd)

## Make a plot
ggplot(data = Oats.IAB, aes(x = nitro, y = yield)) + 
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "deepskyblue", alpha = 0.6) + 
  geom_line(aes(y = Estimate), color = "white", size = 1.5) +
  # geom_point() + 
  ggtitle("95% Bootstrapped Confidence Bands")

```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


For nonlinear

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(boot)
# Nonlinear Example: Loblolly, then using bootstrap or simulation  
Lob <- subset(Loblolly, Seed %in% c("301", "303", "305", "307", "309"))

fnlm1 <- nls(height ~ SSasymp(age, Asym, R0, lrc), data = Lob)

# bootstrapped confidence bands
fm1.boot <- boot(fnlm1, fitted, R=200)  #randomly sampling data 
fm1.boot.prd <- summary_simulate(t(fm1.boot$t))
Oats.IAB <- cbind(Lob, fm1.boot.prd)

## Plot of observed and fitted
## Make a plot
ggplot(data = Oats.IAB, aes(x = age, y = height)) + 
  geom_line(aes(y = Estimate), color = "white", size = 0.6) +
  # geom_point() + 
    geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "black", alpha = 1.5) +
  ggtitle("95% Bootstrapped Confidence Bands")

```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


[Bootstrapping](https://astrostatistics.psu.edu/su09/lecturenotes/simul.html) uses the original, initial sample as the population from which to resample, whereas Monte Carlo simulation is based on setting up a data generation process (parameters).

## Interpreting the Results

Check fixed effects, random effects, covariance assumptions, and subject-specific versus population-level interpretation. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
