---
title: "Data visualization course"
date: '2022-05-20'
author: Daniel 
output: 
  html_document:
    theme: cerulean
    toc: true 
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data visualization course

### Summarization 
```{r}
library(tidyverse)
mtcars %>% 
  mutate( kml = mpg * 0.42) %>% 
  group_by(cyl) %>% 
  summarize(avg_US = mean(mpg), avg_metric = mean(kml))
```

```{r}
mpg %>% 
  group_by(manufacturer, year) %>% 
  summarize_at(vars(cty, hwy), mean)
```
- change layout
```{r}
mpg %>% count(class, year)%>% 
  spread(class, n) 
```
- change all characters into factors
```{r}
mpg <- mpg %>% 
  mutate_if(is.character, as.factor) #if a column is a character, change to a factor
```
- wide to long data
```{r}
mpg1 <- mpg %>% 
  gather("key", "value", cty, hwy)
```

- convert wide data to long data using `pivot_longer `
```{r}
## Your code here. Naming choices for 1 and 2 are yours
dta <- mpg %>% 
  pivot_longer(cty:hwy, names_to = "var", values_to = "value") %>%  # Both of those are 
  # value label
   mutate(var = ifelse( var == 'cty', 'city','highway'))

ggplot(dta, aes(x = displ, y = value)) + 
  geom_point(aes(color = var))  +
  geom_smooth(aes(color = var), se = F)
```

- explore distribution  
```{r}
library(DataExplorer)
library(psych)
library(naniar)
plot_histogram(riskfactors)
```

- explore relationship/correlation
```{r}
library(psych)
pairs.panels(riskfactors[,1:10])
```

- create a individual theme
```{r}
my_theme <- function(){
  theme_bw() + 
 
  theme(axis.title = element_text(size=16),
  axis.text = element_text(size=14),
  text = element_text(size = 14))
}
```
### Explore missing values
```{r}
# install.packages("naniar")
library(naniar)
# head(riskfactors)
riskfactors <- riskfactors
```

```{r}
gg_miss_upset(riskfactors,nsets=10)
```

```{r}
# install.packages("DataExplorer")

plot_missing(riskfactors)
```

```{r}
# take a quick look at the data types of each column
visdat::vis_dat(riskfactors)
```
 



### Add statistical test 
```{r ,warning=F}
library(ggpubr)
plt <- ggplot(  data=mpg,
  mapping= aes(x =  as.factor(year),
               y = cty,
               color = as.factor(year) ) )+
  geom_boxplot() +
  geom_jitter(width=0.1)+
  labs(x = 'Year',
       y = "City mpg") +
  my_theme()+ facet_wrap( ~ manufacturer,nrow = 2)  
  
# add statistical test
  my_comparisons <- list(c('1999','2008'))
plt + stat_compare_means() +
stat_compare_means(comparisons = my_comparisons)
```


### Add texts to dots 
```{r}
USArrests <- USArrests %>% rownames_to_column('State')

ggplot(USArrests, aes(
x=UrbanPop,y=Murder))+
geom_point() +
  labs(x = "Percent of population that is urban", 
       y = "Murder arrests (per 100,000)",
       caption =  "McNeil (1997). Interactive Data Analysis")+
  geom_text(aes(label=State),size=3)
```

### Set the legend
```{r}
ggplot(iris, aes(x=  Sepal.Length , fill= as.factor( Species)) ) +  #whole plot's option
  
  geom_histogram(aes(y=..density..),alpha=0.5, position="identity" , bins = 50)+
  geom_density(aes(linetype=as.factor(Species)),alpha=.1 )+    #aesthetic's option
  
  scale_fill_manual( name = "Groups",values = c("grey", "black", "skyblue"),labels = c("setosa", "versicolor" , "virginica" ))+  
  scale_linetype_manual( name = "Groups" ,values = c(1,3,5),labels = c("setosa", "versicolor" , "virginica") )+       # common legend 
  
  labs(x = "Sepal.Length", 
       y = "Density", 
       title = "") 
```


### Create a panel of plots

- combine multiple plots into one
```{r}
p1=ggplot(data=riskfactors,aes(x=age))+
  geom_histogram(bins = 30 )

p2=ggplot(data=riskfactors,aes(x=sex))+
  geom_bar (aes(x=sex) )

p3=ggplot(riskfactors,aes(x = education, y = bmi))+
  geom_boxplot (  )

p4=ggplot(riskfactors, aes(x = marital )) + 
  geom_bar(aes(group = education, y = (..count..)/sum(..count..),fill = education)) + 
          scale_y_continuous(labels=scales::percent) 

# install.packages("ggpubr")
library(ggpubr)
ggarrange(p1, p2, p3, p4, ncol = 2, nrow=2)
```



### Plots in regression
- create linear regression model
```{r}
data("Boston", package = "MASS")
linear_reg <- glm(medv ~ ., data=Boston , family = gaussian())
summary(linear_reg)
```
- summary
```{r}
knitr::kable(broom::tidy(linear_reg))
```

- create logistical regression 
```{r}
# load the Pima Indians dataset from the mlbench dataset
library(mlbench)
data(PimaIndiansDiabetes) 
# rename dataset to have shorter name because lazy
diabetes <- PimaIndiansDiabetes

logistic_reg <- glm(diabetes ~ ., data=diabetes, family = binomial)

summary(logistic_reg)

```
- summary

```{r}
knitr::kable(broom::tidy(logistic_reg))
```


#### Create forest plots for `coefficients` or `OR`
```{r}
library(sjPlot)
plot_model(linear_reg, show.values = TRUE, value.offset = 0.5)
```

```{r}
plot_model(logistic_reg, show.values = TRUE, value.offset = .5, vline.color = "black")
```
**another way**
```{r}
library(finalfit) 

explanatory = c(  "crim"  ,  "zn"   ,   "indus"  ,   "nox"   ,  "rm"   ,   "age" ,    "dis"  ,   "rad"   ,  "tax"   ,"ptratio" ,"black"  , "lstat" )

dependent = "medv"


Boston %>%
 coefficient_plot(dependent, explanatory, table_text_size=3, 
                  title_text_size=12,
                   plot_opts=list(xlab("Beta, 95% CI"), 
                                  theme(axis.title = element_text(size=12))))
```

```{r}
 
library(finalfit) 

explanatory = c(  "pregnant", "glucose" , "pressure", "triceps"  ,"insulin" , "mass"   ,  "pedigree", "age"  )

dependent = "diabetes"

diabetes %>%
 or_plot(dependent, explanatory, table_text_size=3, 
                  title_text_size=12,
                   plot_opts=list(xlab("OR, 95% CI"), 
                                  theme(axis.title = element_text(size=12))))
 
```

- qq plot
```{r}
ggqqplot( (Boston$medv))
```

 


