---
title: "Data visualization course"
description: "A practical R learning guide to Data visualization course, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Data Management & Programming"
category: "Data Cleaning & Quality Control"
topic: "Data visualization course"
type: article
tags: ["Data Cleaning & Quality Control", "Data Management & Programming", "Data", "visualization", "course"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: false
---

## Why This Topic Matters

Data cleaning and quality control make analytical results traceable and reduce errors caused by inconsistent or implausible values.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Data visualization course**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Data visualization course
2. Summarization
3. Explore missing values
4. Add statistical test
5. Add texts to dots
6. Set the legend

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Reproducible Learning Check

This short R chunk confirms the scope of the lesson and always produces a visible output.

```{r}
#| label: learning-overview
learning_overview <- data.frame(
  section = "Data Management & Programming",
  category = "Data Cleaning & Quality Control",
  topic = "Data visualization course"
)
learning_overview
```

## Learning Material

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Data visualization course

##### Summarization 
```{r}
library(tidyverse)
mtcars %>% 
  mutate( kml = mpg * 0.42) %>% 
  group_by(cyl) %>% 
  summarize(avg_US = mean(mpg), avg_metric = mean(kml))
```

```{r}
mpg %>% 
  group_by(manufacturer, year) %>% 
  summarize_at(vars(cty, hwy), mean)
```
- change layout
```{r}
mpg %>% count(class, year)%>% 
  spread(class, n) 
```
- change all characters into factors
```{r}
mpg <- mpg %>% 
  mutate_if(is.character, as.factor) #if a column is a character, change to a factor
```
- wide to long data
```{r}
mpg1 <- mpg %>% 
  gather("key", "value", cty, hwy)
```

- convert wide data to long data using `pivot_longer `
```{r}
## Your code here. Naming choices for 1 and 2 are yours
dta <- mpg %>% 
  pivot_longer(cty:hwy, names_to = "var", values_to = "value") %>%  # Both of those are 
  # value label
   mutate(var = ifelse( var == 'cty', 'city','highway'))

ggplot(dta, aes(x = displ, y = value)) + 
  geom_point(aes(color = var))  +
  geom_smooth(aes(color = var), se = F)
```

- explore distribution  
```{r}
library(DataExplorer)
library(psych)
library(naniar)
plot_histogram(riskfactors)
```

- explore relationship/correlation
```{r}
library(psych)
pairs.panels(riskfactors[,1:10])
```

- create a individual theme
```{r}
my_theme <- function(){
  theme_bw() + 
 
  theme(axis.title = element_text(size=16),
  axis.text = element_text(size=14),
  text = element_text(size = 14))
}
```
##### Explore missing values
```{r}
#| eval: false
#| code-fold: true
# install.packages("naniar")
library(naniar)
# head(riskfactors)
riskfactors <- riskfactors
```

```{r}
gg_miss_upset(riskfactors,nsets=10)
```

```{r}
#| eval: false
#| code-fold: true
# install.packages("DataExplorer")

plot_missing(riskfactors)
```

```{r}
# take a quick look at the data types of each column
visdat::vis_dat(riskfactors)
```
 



##### Add statistical test 
```{r ,warning=F}
library(ggpubr)
plt <- ggplot(  data=mpg,
  mapping= aes(x =  as.factor(year),
               y = cty,
               color = as.factor(year) ) )+
  geom_boxplot() +
  geom_jitter(width=0.1)+
  labs(x = 'Year',
       y = "City mpg") +
  my_theme()+ facet_wrap( ~ manufacturer,nrow = 2)  
  
# add statistical test
  my_comparisons <- list(c('1999','2008'))
plt + stat_compare_means() +
stat_compare_means(comparisons = my_comparisons)
```


##### Add texts to dots 
```{r}
USArrests <- USArrests %>% rownames_to_column('State')

ggplot(USArrests, aes(
x=UrbanPop,y=Murder))+
geom_point() +
  labs(x = "Percent of population that is urban", 
       y = "Murder arrests (per 100,000)",
       caption =  "McNeil (1997). Interactive Data Analysis")+
  geom_text(aes(label=State),size=3)
```

##### Set the legend
```{r}
ggplot(iris, aes(x=  Sepal.Length , fill= as.factor( Species)) ) +  #whole plot's option
  
  geom_histogram(aes(y=..density..),alpha=0.5, position="identity" , bins = 50)+
  geom_density(aes(linetype=as.factor(Species)),alpha=.1 )+    #aesthetic's option
  
  scale_fill_manual( name = "Groups",values = c("grey", "black", "skyblue"),labels = c("setosa", "versicolor" , "virginica" ))+  
  scale_linetype_manual( name = "Groups" ,values = c(1,3,5),labels = c("setosa", "versicolor" , "virginica") )+       # common legend 
  
  labs(x = "Sepal.Length", 
       y = "Density", 
       title = "") 
```


##### Create a panel of plots

- combine multiple plots into one
```{r}
#| eval: false
#| code-fold: true
p1=ggplot(data=riskfactors,aes(x=age))+
  geom_histogram(bins = 30 )

p2=ggplot(data=riskfactors,aes(x=sex))+
  geom_bar (aes(x=sex) )

p3=ggplot(riskfactors,aes(x = education, y = bmi))+
  geom_boxplot (  )

p4=ggplot(riskfactors, aes(x = marital )) + 
  geom_bar(aes(group = education, y = (..count..)/sum(..count..),fill = education)) + 
          scale_y_continuous(labels=scales::percent) 

# install.packages("ggpubr")
library(ggpubr)
ggarrange(p1, p2, p3, p4, ncol = 2, nrow=2)
```



##### Plots in regression
- create linear regression model
```{r}
data("Boston", package = "MASS")
linear_reg <- glm(medv ~ ., data=Boston , family = gaussian())
summary(linear_reg)
```
- summary
```{r}
knitr::kable(broom::tidy(linear_reg))
```

- create logistical regression 
```{r}
# load the Pima Indians dataset from the mlbench dataset
library(mlbench)
data(PimaIndiansDiabetes) 
# rename dataset to have shorter name because lazy
diabetes <- PimaIndiansDiabetes

logistic_reg <- glm(diabetes ~ ., data=diabetes, family = binomial)

summary(logistic_reg)

```
- summary

```{r}
knitr::kable(broom::tidy(logistic_reg))
```


###### Create forest plots for `coefficients` or `OR`
```{r}
library(sjPlot)
plot_model(linear_reg, show.values = TRUE, value.offset = 0.5)
```

```{r}
plot_model(logistic_reg, show.values = TRUE, value.offset = .5, vline.color = "black")
```
**another way**
```{r}
library(finalfit) 

explanatory = c(  "crim"  ,  "zn"   ,   "indus"  ,   "nox"   ,  "rm"   ,   "age" ,    "dis"  ,   "rad"   ,  "tax"   ,"ptratio" ,"black"  , "lstat" )

dependent = "medv"


Boston %>%
 coefficient_plot(dependent, explanatory, table_text_size=3, 
                  title_text_size=12,
                   plot_opts=list(xlab("Beta, 95% CI"), 
                                  theme(axis.title = element_text(size=12))))
```

```{r}
 
library(finalfit) 

explanatory = c(  "pregnant", "glucose" , "pressure", "triceps"  ,"insulin" , "mass"   ,  "pedigree", "age"  )

dependent = "diabetes"

diabetes %>%
 or_plot(dependent, explanatory, table_text_size=3, 
                  title_text_size=12,
                   plot_opts=list(xlab("OR, 95% CI"), 
                                  theme(axis.title = element_text(size=12))))
 
```

- qq plot
```{r}
ggqqplot( (Boston$medv))
```

## Interpreting the Results

Use summaries and visual checks to confirm ranges, distributions, duplicates, and missing-data patterns. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
