---
title: "Imputing Missing Data with MICE"
output: html_document
date: '2023-12-11'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

MCAR and MAR observations are called ignorable, while MNAR values are called non-ignorable or informative.

- Work with the unbalanced data and use the ML or REML method for
parameter estimation. This approach works best if all missing values are
ignorable. However, for MNAR values, this strategy results in biased estimates.

- Delete all incomplete cases, which are the subjects with at least one missing
observation. It works best when the number of incomplete cases is not large, and there is no structure to the missing data (they are MCAR observations). 

- Impute the missing values and work with the resulting complete data set, including mean, last observation, and multiple imputation tech. The resulting parameter estimates are biased (unless that is MCAR).

### Imputing Missing Data with MICE

Substitute the means for other variables with missingness, run a regression to get the imputed values for the first variable; next, run a regression for the second variable and imputed using the first variable's updated values; repeat the above steps until all missing data was imputed by regression; that is the first iterate. repeat the above process by 50 iterates to get the first imputed dataset. repeat the above procedure 5 times to get 5 imputed datasets. 

- Data summary
```{r}
library(mice)
data <- airquality
summary(data)
```

- Take a look at the variable and its distribution
```{r}
md.pattern(data)
hist(data$Ozone)
```

- Imputed missing values using simple values
```{r}
imputed_zero = replace(data$Ozone, is.na(data$Ozone), 0) 
imputed_mean = replace(data$Ozone, is.na(data$Ozone), mean(data$Ozone, na.rm = TRUE))
imputed_median = replace(data$Ozone, is.na(data$Ozone), median(data$Ozone, na.rm = TRUE))
```

- Check the distribution again
```{r}
hist(imputed_zero)
hist(imputed_mean)
hist(imputed_median)
```

- Imputed missing values by different algorithm methods (MICE)
```{r}
imputed_pmm = complete(mice(data, method = "pmm",seed=500))$Ozone #default the first imputed dataset
imputed_cart = complete(mice(data, method = "cart",seed=500))$Ozone
imputed_lasso = complete(mice(data, method = "lasso.norm",seed=500))$Ozone
```

`lasso is different with other two`

- Check the distribution again
```{r}
hist(data$Ozone)
hist(imputed_pmm)
hist(imputed_cart)
hist(imputed_lasso)
```

- Imputing the missing data using MICE
```{r}
tempData <- mice(data,m=5,maxit=50,meth='pmm',seed=500) #m=5 refers to the number of imputed datasets.
summary(tempData)
tempData$imp$Ozone
```

- Get back a completed dataset
```{r}
completedData <- complete(tempData,1)
# completed <- complete(tempData) 
```

- Inspecting the distribution of original and imputed data
```{r}
xyplot(tempData,Ozone ~ Wind+Temp+Solar.R,pch=18,cex=1) #?
densityplot(tempData)
```

- Pooling the results
```{r}
modelFit1 <- with(tempData,lm(Temp~ Ozone+Solar.R+Wind))
summary(pool(modelFit1))
```



`Using logistic regression to indentify if it is missing completely at random;`

`How well is your imputation working? Remove all missing rows; Run statistics on your dataset from #1;Randomly remove values from your dataset; Impute values, calculate statistics, and compare to #2; Repeat steps 3 and 4 with many different removed values + imputation methods;`




