---
title: "Response-Adaptive Randomization: Stage 1 and Stage 2"
author: ""
date: "`r Sys.Date()`"
output:
html_document:
toc: true
toc_depth: 3
number_sections: true
---

# Overview

This example demonstrates a simple **response-adaptive randomization** design with two stages.

* **Stage 1:** The first 100 patients are randomized 1:1 to Treatment or Control.
* Observed response rates are calculated after Stage 1.
* The Stage 2 allocation probability is updated according to the observed response rates.
* **Stage 2:** The next 100 patients are randomized using the updated allocation probability.
* A randomization dataset is created for possible use in an EDC or IRT/RTSM workflow.

The adaptive allocation probability is defined as:

$$
P(\text{Treatment})
=
\frac{\hat p_{\text{Treatment}}}
{\hat p_{\text{Treatment}}+\hat p_{\text{Control}}}
$$

To avoid extreme allocation, the probability is restricted to:

$$
0.30 \leq P(\text{Treatment}) \leq 0.70
$$

# 1. Set Seed

```{r}
set.seed(123)
```

# 2. Stage 1: First 100 Patients

The first 100 patients are randomized equally between Treatment and Control.

Treatment coding is:

* `1 = Treatment`
* `0 = Control`

```{r}
n_stage1 <- 100

trt1 <- rbinom(
  n_stage1,
  1,
  0.50
)
```

# 3. Simulate Stage 1 Responses

For illustration, assume the true response rates are:

$$
P(\text{Response}|\text{Treatment}) = 0.70
$$

and

$$
P(\text{Response}|\text{Control}) = 0.50
$$

```{r}
response1 <- rbinom(
  n_stage1,
  1,
  prob = ifelse(
    trt1 == 1,
    0.70,
    0.50
  )
)
```

# 4. Calculate Observed Response Rates

```{r}
rate_trt <- mean(
  response1[trt1 == 1]
)

rate_ctl <- mean(
  response1[trt1 == 0]
)

cat(
  "Observed Treatment response rate =",
  round(rate_trt, 3),
  "\n"
)

cat(
  "Observed Control response rate =",
  round(rate_ctl, 3),
  "\n"
)
```

# 5. Calculate Adaptive Allocation Probability

The Stage 2 Treatment allocation probability is calculated as:

$$
P(\text{Treatment})
=
\frac{\hat p_T}
{\hat p_T+\hat p_C}
$$

where:

* (\hat p_T) is the observed Treatment response rate;
* (\hat p_C) is the observed Control response rate.

```{r}
prob_trt_next <- rate_trt /
  (rate_trt + rate_ctl)
```

To avoid extreme allocation, restrict the Treatment probability to the interval 0.30 to 0.70.

```{r}
prob_trt_next <- max(
  0.30,
  min(
    0.70,
    prob_trt_next
  )
)

prob_ctl_next <- 1 -
  prob_trt_next
```

Display the Stage 2 allocation probabilities.

```{r}
cat(
  "Stage 2 Treatment allocation probability =",
  round(prob_trt_next, 3),
  "\n"
)

cat(
  "Stage 2 Control allocation probability =",
  round(prob_ctl_next, 3),
  "\n"
)
```

# 6. Stage 2: Randomize the Next 100 Patients

```{r}
n_stage2 <- 100
```

# 7. Create Subject IDs

The next 100 patients are assigned Subject IDs from `SUBJ-101` through `SUBJ-200`.

```{r}
subject_id <- sprintf(
  "SUBJ-%03d",
  101:200
)
```

# 8. Create Randomization Numbers

```{r}
randomization_number <- sprintf(
  "R%04d",
  101:200
)
```

# 9. Generate Random Numbers

For each Stage 2 patient, generate one random number from a Uniform(0,1) distribution.

$$
U_i \sim Uniform(0,1)
$$

```{r}
u_random <- runif(
  n_stage2
)
```

# 10. Assign Treatment

The randomization rule is:

$$
\text{Treatment}_i =
\begin{cases}
\text{Treatment}, & U_i \leq P(\text{Treatment}) \
\text{Control}, & U_i > P(\text{Treatment})
\end{cases}
$$

```{r}
trt2 <- ifelse(
  u_random <= prob_trt_next,
  1,
  0
)
```

# 11. Create Treatment Labels

```{r}
treatment_assignment <- ifelse(
  trt2 == 1,
  "Treatment",
  "Control"
)
```

# 12. Create Blinded Treatment Codes

For illustration:

* `A = Treatment`
* `B = Control`

In a real double-blind trial, the actual mapping between blinded code and treatment would normally be restricted.

```{r}
treatment_code <- ifelse(
  trt2 == 1,
  "A",
  "B"
)
```

# 13. Create the Stage 2 Randomization Dataset

```{r}
stage2_randomization <- data.frame(
  
  Subject_ID =
    subject_id,
  
  Randomization_Number =
    randomization_number,
  
  Stage =
    2,
  
  Allocation_Probability_Treatment =
    rep(
      prob_trt_next,
      n_stage2
    ),
  
  Allocation_Probability_Control =
    rep(
      prob_ctl_next,
      n_stage2
    ),
  
  Random_Number =
    round(
      u_random,
      6
    ),
  
  Treatment_Code =
    treatment_code,
  
  Treatment_Assignment =
    treatment_assignment,
  
  stringsAsFactors = FALSE
)
```

# 14. Display the First 10 Randomized Patients

```{r}
head(
  stage2_randomization,
  10
)
```

# 15. Check Stage 2 Treatment Allocation

```{r}
table(
  stage2_randomization$
    Treatment_Assignment
)
```

# 16. Calculate Observed Stage 2 Allocation Proportions

```{r}
prop.table(
  table(
    stage2_randomization$
      Treatment_Assignment
  )
)
```

# Interpretation

The overall adaptive randomization process is:

$$
\text{Stage 1: 1:1 Randomization}
$$

$$
\downarrow
$$

$$
\text{Observe Stage 1 Responses}
$$

$$
\downarrow
$$

$$
\text{Calculate New Allocation Probability}
$$

$$
\downarrow
$$

$$
\text{Stage 2 Adaptive Randomization}
$$

For example, if the observed Stage 1 response rates are:

$$
\hat p_T = 0.70
$$

and

$$
\hat p_C = 0.50
$$

then:

$$
P(\text{Treatment})
=
 \frac{0.70}{0.70+0.50}
=
0.583
$$

Therefore, approximately 58.3% of Stage 2 patients would be randomized to Treatment and 41.7% to Control.

Importantly, the exact number of patients assigned to each group will not necessarily equal these proportions because each patient is still randomized individually.

# Practical Clinical Trial Note

This example uses one adaptation after the first 100 patients.

Therefore, all Stage 2 patients use the same updated allocation probability.

The design is:

$$
100\text{ patients}
\rightarrow
\text{Interim Analysis}
\rightarrow
\text{Update Allocation Probability}
\rightarrow
100\text{ additional patients}
$$

This is operationally simpler than updating the allocation probability after every individual patient.

In an actual clinical trial, the randomization algorithm, allocation restrictions, timing of adaptation, treatment-code mapping, and IRT/RTSM implementation should be prespecified and validated.
