---
title: "Nested mixed model"
output: html_document
date: '2024-04-09'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Nested mixed model in R

[reference](https://biostatmatt.com/archives/2718)

- Load package and data
```{r}
# load package
library(nlme)
# read in data
setwd("C:\\Users\\hed2\\OneDrive - National Institutes of Health\\Mixed model by SAS and R")
DF <- read.csv("Oxide.csv")
# specific the reference group
DF$Source=as.factor(DF$Source) 
DF <- within(DF, Source<- relevel(Source, ref = 2))
```

- Random intercept 

```{r}
# nested
# using lme package
model2 = lme(Thickness ~ Source  , 
           random = (~ 1   |Lot/Wafer), data = DF, method= "REML")  #wafer nested into lot
summary(model2)
# using lme4 package
library("lme4")
model3 = lmer(Thickness ~ Source+(1   | Lot/Wafer), data=DF) #wafer nested into lot
summary(model3)
```

- Random slope
```{r}
# nested
# using lme package
model2 = lme(Thickness ~ Source  , 
           random = (~ 1 +Source  |Lot/Wafer), data = DF, method= "REML")  #wafer nested into lot
summary(model2)

# using lme4 package
model3 = lmer(Thickness ~ Source+(1 +Source  | Lot/Wafer), data=DF) #wafer nested into lot
summary(model3)
```

### Nested mixed model in SAS

- Random intercept
```{r}
# /*nested*/
# PROC MIXED DATA=Oxide METHOD=REML;
# CLASS Lot    Wafer    Source  ;
# MODEL Thickness  = Source   /solution;
# RANDOM   int  /subject=Lot;
# RANDOM   int  / subject= Wafer(Lot)    type=un ;  /*wafer nested into lot*/
# RUN; QUIT; 
```

output

```{r}
# Solution for Fixed Effects 
# Effect Source Estimate Standard
# Error DF t Value Pr > |t| 
# Intercept   2005.19 5.7716 6 347.43 <.0001 
# Source 1 -10.0833 8.1622 48 -1.24 0.2227 
# Source 2 0 . . . .
```


- Random slope
```{r}
# /*nested*/
# PROC MIXED DATA=Oxide METHOD=REML;
# CLASS Lot    Wafer    Source  ;
# MODEL Thickness  = Source   /solution;
# RANDOM   int Source/subject=Lot;
# RANDOM   int Source/ subject= Wafer(Lot)    type=un ;  /*wafer nested into lot*/
# RUN; QUIT; 
```

output
```{r}
# Solution for Fixed Effects 
# Effect Source Estimate Standard
# Error DF t Value Pr > |t| 
# Intercept   2005.19 5.7596 0 348.15 . 
# Source 1 -10.0833 8.1143 0 -1.24 . 
# Source 2 0 . . . .
```


Summary:

- Generally, when you know nothing else making very strong assumptions about the correlation structure over time (e.g. type=AR(1) or type=CS or just having a random subject effect on the intercept) is not advisable. If you have sufficient data, it is usually recommended to use an unstructured covariance [matrix](https://stats.stackexchange.com/questions/233475/mixed-model-for-multiple-measurements-in-a-crossover-study-sas). 

- The values for any variable uniquely identify the levels of the physicians. Therefore, no nesting is [necessary](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://support.sas.com/resources/papers/proceedings15/SAS1919-2015.pdf).


- Nested designs arise when every level of one factor (B, say) is combined with (“nested inside”) only one level of another factor (A, say) in the design; this compares to the crossed design (or complete factorial designs) where all of the levels of factor A are combined with all of the levels of [factor B](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://www.pnw.edu/wp-content/uploads/2020/03/Lecture-Notes-10-6.pdf).














