---
title: "Data visualization"
date: '2022-05-20'
author: Daniel 
output: 
  html_document:
    theme: cerulean
    toc: true 
    toc_float: true
---


```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(echo = TRUE)

```

- Loading data set
```{r}
library(printr)
library(tidyverse)
head(iris)
```
## Scatter plot
### Create a empty canvas
- then create `aesthetic mapping`  
- tell the function which dataset and variables to use
```{r}
ggplot(data = iris,        # which data set? canvas? 
       aes(x=Sepal.Length , y=Petal.Length   ))  # which variables as aesthetics? x and y are mapped to columns of the data; different geoms can have different aesthetics (different variables). 
```

 
### Add a layer/geom of `points` to the canvas
```{r}
ggplot(data = iris,
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point()   # adding the geometrical representation 
```
 
```{r,eval=F}
# same plot as above
ggplot(data = iris) + 
  geom_point(  aes(x=Sepal.Length , y=Petal.Length  )) 
```
### Add another aesthetic
- add a curve/straight line to fit these points
- geom provides the aesthetic to ggplot
```{r}
# Loess curve
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point() +
  geom_smooth() 

```

```{r}
# Linear regression line
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

### Add other aesthetic
- set other aesthetics `colour, alpha (transparency), and size of points`

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width     ), 
             alpha = .5, 
             colour = "red")
```

 

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=Species), #white is a variable here 
             alpha=.9)
```

 
- categorize `Petal.Width` then map colour to this new variable
```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(data=iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=growth), 
             alpha=.9)
```

 
## Bar chart
```{r}
ggplot(data = iris) + 
  geom_bar(aes(x = growth))
```

- bar chart after `group_by` 
- then use `stat='identity'` 
```{r}
results  <- iris %>% 
  group_by(Species, growth) %>% 
  summarize(Sepal.Length.mean=mean (Sepal.Length ))

 gop <- results  %>% 
  filter(Species != "setosa_null"  )
 gop
```
- though meaningless below until line chart (just use the mean as the sum for demonstration)
```{r}
# We can also store parts of a plot in an object
plot1 <- ggplot(gop) + 
  geom_bar(aes(x=growth , y=Sepal.Length.mean), 
           
           stat='identity')
plot1
```
### Add some options for the whole ggplot rather than layers
- `switch` the x and y axes 
```{r}
plot1 + 
  coord_flip()
```

- `reorder` x categories (`-`means descending)
```{r}
ggplot( gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip()
```

- add x axis `label` and a `theme` 
```{r}
ggplot(gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip() + 
  xlab("Growth categories") + 
  guides(fill=F) +
  theme_minimal()
```

- set theme
```{r}
library(ggthemes)
ggplot(data = iris) + 
  geom_bar(aes(x = growth)) + 
  theme_economist()
```

### Grouped bar chart 
-bar chart with different panels
```{r}
ggplot(mpg, aes(x = class)) + 
  geom_bar() + 
  facet_wrap( ~ year)
```

- actual number (groups are stacked by default)
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species)
           , stat='identity'
           )
```

```{r}
ggplot(mpg, aes(x = class )) + 
  geom_bar(aes(group = year, fill = year), position = "stack")
```

- percentage
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='fill')
```
 
- groups are dodge with `actual number `
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='dodge')

```
- groups are dodge with `percentage`
```{r}
gop2 <- gop %>% 
  group_by(growth ) %>% 
  mutate(Sepal.Length.prop=Sepal.Length.mean/sum(Sepal.Length.mean))

ggplot(gop2) + 
  geom_bar(aes(x=growth, y=Sepal.Length.prop, fill=Species), 
           stat='identity', 
           position='dodge') + 
  ylab("Votes (%)")
```

 
## Line charts
 
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length))
```
 
### Grouped by `colour variable`
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length, colour = Species))
```

 
- grouped by state then set how many rows or columns
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length) ) + 
  
  facet_wrap(~Species, nrow = 1) +      #set how many rows
  coord_flip()  
```

### Multiple aesthetics
```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) + 
  geom_line(size=2,color="purple")+  
  # number format
  scale_x_log10(labels = scales::label_number())+
  geom_point( aes(size =  Sepal.Length,colour = as.factor(growth)),show.legend = F)+
  facet_wrap(~ Species) 

```



 
 

 
