---
title: "Data visualization"
description: "A practical R learning guide to Data visualization, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Data Management & Programming"
category: "Data Cleaning & Quality Control"
topic: "Data visualization"
type: article
tags: ["Data Cleaning & Quality Control", "Data Management & Programming", "Data", "visualization"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

Data cleaning and quality control make analytical results traceable and reduce errors caused by inconsistent or implausible values.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Data visualization**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Scatter plot
2. Create a empty canvas
3. Add a layer/geom of points to the canvas
4. Add another aesthetic
5. Add other aesthetic
6. Bar chart

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Data visualization** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
#| eval: false
#| code-fold: true
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(echo = TRUE)

```

- Loading data set

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(printr)
library(tidyverse)
head(iris)
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

#### Scatter plot

<!-- generated-learning-note -->
This section uses **Scatter plot** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.

#### Create a empty canvas

<!-- generated-learning-note -->
This subsection explains **Create a empty canvas** and shows how it contributes to the overall Data visualization workflow.

- then create `aesthetic mapping`  
- tell the function which dataset and variables to use

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(data = iris,        # which data set? canvas? 
       aes(x=Sepal.Length , y=Petal.Length   ))  # which variables as aesthetics? x and y are mapped to columns of the data; different geoms can have different aesthetics (different variables). 
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


 
#### Add a layer/geom of points to the canvas

<!-- generated-learning-note -->
This subsection explains **Add a layer/geom of points to the canvas** and shows how it contributes to the overall Data visualization workflow.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(data = iris,
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point()   # adding the geometrical representation 
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

 

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r,eval=F}
# same plot as above
ggplot(data = iris) + 
  geom_point(  aes(x=Sepal.Length , y=Petal.Length  )) 
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

#### Add another aesthetic

<!-- generated-learning-note -->
This subsection explains **Add another aesthetic** and shows how it contributes to the overall Data visualization workflow.

- add a curve/straight line to fit these points
- geom provides the aesthetic to ggplot

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
# Loess curve
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point() +
  geom_smooth() 

```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
# Linear regression line
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


#### Add other aesthetic

<!-- generated-learning-note -->
This subsection explains **Add other aesthetic** and shows how it contributes to the overall Data visualization workflow.

- set other aesthetics `colour, alpha (transparency), and size of points`


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width     ), 
             alpha = .5, 
             colour = "red")
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


 


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=Species), #white is a variable here 
             alpha=.9)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


 
- categorize `Petal.Width` then map colour to this new variable

<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(data=iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=growth), 
             alpha=.9)
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.


 
#### Bar chart

<!-- generated-learning-note -->
This subsection explains **Bar chart** and shows how it contributes to the overall Data visualization workflow.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(data = iris) + 
  geom_bar(aes(x = growth))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


- bar chart after `group_by` 
- then use `stat='identity'` 

<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r}
results  <- iris %>% 
  group_by(Species, growth) %>% 
  summarize(Sepal.Length.mean=mean (Sepal.Length ))

 gop <- results  %>% 
  filter(Species != "setosa_null"  )
 gop
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.

- though meaningless below until line chart (just use the mean as the sum for demonstration)

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
# We can also store parts of a plot in an object
plot1 <- ggplot(gop) + 
  geom_bar(aes(x=growth , y=Sepal.Length.mean), 
           
           stat='identity')
plot1
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

#### Add some options for the whole ggplot rather than layers

<!-- generated-learning-note -->
This section uses **Add some options for the whole ggplot rather than layers** to make patterns visible. Confirm what each axis, group, scale, and annotation represents.

- `switch` the x and y axes 

<!-- generated-learning-note -->
**Purpose.** This block carries out the next computational step in the workflow. Track the objects it creates and how they are used by later steps.

```{r}
plot1 + 
  coord_flip()
```

<!-- generated-learning-note -->
**How to read the output.** Read the printed object, table, or scalar in the context of the preceding definitions. Confirm its units and expected range before continuing.


- `reorder` x categories (`-`means descending)

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot( gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip()
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


- add x axis `label` and a `theme` 

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip() + 
  xlab("Growth categories") + 
  guides(fill=F) +
  theme_minimal()
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


- set theme

<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(ggthemes)
ggplot(data = iris) + 
  geom_bar(aes(x = growth)) + 
  theme_economist()
```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


#### Grouped bar chart

<!-- generated-learning-note -->
This subsection explains **Grouped bar chart** and shows how it contributes to the overall Data visualization workflow.

-bar chart with different panels

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(mpg, aes(x = class)) + 
  geom_bar() + 
  facet_wrap( ~ year)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


- actual number (groups are stacked by default)

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species)
           , stat='identity'
           )
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.



<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(mpg, aes(x = class )) + 
  geom_bar(aes(group = year, fill = year), position = "stack")
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


- percentage

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='fill')
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

 
- groups are dodge with `actual number `

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='dodge')

```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

- groups are dodge with `percentage`

<!-- generated-learning-note -->
**Purpose.** This block prepares the analysis data. Follow each transformation carefully because filtering, recoding, grouping, and reshaping determine the observations used downstream.

```{r}
gop2 <- gop %>% 
  group_by(growth ) %>% 
  mutate(Sepal.Length.prop=Sepal.Length.mean/sum(Sepal.Length.mean))

ggplot(gop2) + 
  geom_bar(aes(x=growth, y=Sepal.Length.prop, fill=Species), 
           stat='identity', 
           position='dodge') + 
  ylab("Votes (%)")
```

<!-- generated-learning-note -->
**How to read the output.** Compare the data before and after the transformation. Confirm that row counts, group sizes, missingness, and variable values changed as intended.


 
#### Line charts

<!-- generated-learning-note -->
This subsection explains **Line charts** and shows how it contributes to the overall Data visualization workflow.

 

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.

 
#### Grouped by colour variable

<!-- generated-learning-note -->
This section prepares the information needed for **Grouped by colour variable**. Check variable definitions, units, coding, and missing values before moving to analysis.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length, colour = Species))
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


 
- grouped by state then set how many rows or columns

<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length) ) + 
  
  facet_wrap(~Species, nrow = 1) +      #set how many rows
  coord_flip()  
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


#### Multiple aesthetics

<!-- generated-learning-note -->
This subsection explains **Multiple aesthetics** and shows how it contributes to the overall Data visualization workflow.


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) + 
  geom_line(size=2,color="purple")+  
  # number format
  scale_x_log10(labels = scales::label_number())+
  geom_point( aes(size =  Sepal.Length,colour = as.factor(growth)),show.legend = F)+
  facet_wrap(~ Species) 

```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.

## Interpreting the Results

Use summaries and visual checks to confirm ranges, distributions, duplicates, and missing-data patterns. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
