---
title: "Confounding and effect measure modification"
output: html_document
date: '2022-11-04'
author: Daniel
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to analyze confounding and effect measure modification 
- If there is only confounding: The stratum-specific measures of association will be similar to one another, but they will be different from the overall crude estimate by 10% or more. In this situation, one can use Mantel-Haenszel methods to calculate a pooled estimate (RR or OR) and p-value.

- If there is neither confounding nor effect modification: The crude estimate of association and the stratum-specific estimates will be similar. They don't have to be identical, just similar.

- If there is only effect modification: The stratum-specific estimates will differ from one another significantly. Whether they are "significantly different" can be tested by using a chi-square test of homogeneity, as described in the Aschengrau & Seage textbook.

- If there is both effect modification and confounding: Here, you need to consider two possibilities:

  - 1) If the stratum-specific estimates differ from one another, and they are both less than the crude estimate or if they are both greater than the crude estimate, then there is both confounding and effect modification.

  - 2) If the stratum-specific estimates differ from one another, and the crude estimate is between the two stratum-specific estimates, then you need to pool the stratum-specific estimates (with a Mantel-Haenszel equation) to determine whether the pooled estimate is more than 10% different from the crude estimate.

  - Note that in this situation you are only pooling the stratum-specific estimates in order to make a decision about whether confounding is present; you should not report the pooled estimate as an "adjusted" measure of association if there is effect modification
  
  - confouding factor is not related to emm directly. confounding is adjusted covariate; effect modification is interaction effect in the regression model. One should adjust for confounders, but report the different effects seen for effect modifers.

please see the detail [here](https://sphweb.bumc.bu.edu/otlt/mph-modules/bs/bs704-ep713_confounding-em/bs704-ep713_confounding-em_print.html). 

### Practice in assessing confounding and modification 

- Load package and data set
```{r, message=FALSE}
## Importing required packages
library(tidyverse)   
library(skimr)      
## Read in clean data set
df_dia <- read_csv("C:\\Users\\hed2\\Downloads\\code-storage\\code\\diabetes-data.csv")

```
- Skim the dataset
```{r}
## Skim the dataset
skim(df_dia)
```

- Check the structure of the data
```{r}
## Explore data set
glimpse(df_dia)
```
- Create (three-way) interaction terms
```{r }
## Create interaction terms
df_dia <- df_dia %>% 
  mutate(
    padmit = patient_visits*admit_source,
    hadmit = HbA1c*admit_source,
    hpatient = HbA1c*patient_visits,
    hpadmit = HbA1c*patient_visits*admit_source)%>% 
  mutate_at(vars(diabetesMed, HbA1c, admit_source, hadmit), as.factor)
## Check the structure of the data
glimpse(df_dia)
```

#### Explore whether it is a confounder

if the covariate is not sig, it must be not confounder. in other words, the regresion model can adjust confoudners. 
```{r}
## patient_visits is   confounder  
model7 <- glm(diabetesMed ~ HbA1c + admit_source + patient_visits ,
              df_dia, family = "binomial")

## Print the coefficients of the model
coef(summary(model7))
```
patient_visits is not  confounder but independent predictor, because it is sig but there is no change in the coefficient of hba1c.  

# Before and after adjusting  
```{r}
##  
model8 <- glm(diabetesMed ~ HbA1c + admit_source   ,
              df_dia, family = "binomial")

## Print the coefficients of the model
coef(summary(model8))
```

- Plot confounding in log odds

examine the primary association of interest at different levels of a potential confounding factor.

```{r}
curve((1.03209664  +0.81957667*0      +0.10412221       +0.03830031 *x)   , from=1, to=80, n=300, xlab="patient_visits", ylab="log odds", 
             col="red", lwd=2, main="confounding of exposure and patient_visits"  )

curve((1.03209664  +0.81957667*1      +0.10412221       +0.03830031 *x )  , from=1, to=80, n=300, add=T, xlab="patient_visits", ylab="log odds", 
             col="blue", lwd=2, main="confounding of exposure and patient_visits"  )
```
- Plot confounding in odds
```{r}
curve( exp(1.03209664  +0.81957667*0      +0.10412221       +0.03830031 *x)   , from=1, to=80, n=300, xlab="patient_visits", ylab="odds", 
             col="red", lwd=2, main="confounding of exposure and patient_visits"  )

curve( exp(1.03209664  +0.81957667*1      +0.10412221       +0.03830031 *x )  , from=1, to=80, n=300, add=T, xlab="patient_visits", ylab="odds", 
             col="blue", lwd=2, main="confounding of exposure and patient_visits"  )
```
- Plot confounding in odds ratio
```{r}
curve( exp(1.03209664  +0.81957667*0      +0.10412221       +0.03830031 *x) / exp(1.03209664  +0.81957667*1      +0.10412221       +0.03830031 *x )  , from=1, to=80, n=300,   xlab="patient_visits", ylab="OR", 
             col="blue", lwd=2, main="confounding of exposure and patient_visits"  )
```

#### Explore whether it is a effect measure modification  

although the covariate is not sig, it may be emm as long as the interaction is sig. This occurs when an association between an exposure X and an outcome Y can depend on another variable Z, and Z does not have to be directly related to X and/or Y.

- Step 1: Full model
```{r}
## Step 1: Full model
full_model <- glm(as.factor(diabetesMed) ~ HbA1c + admit_source + patient_visits 
                  + padmit + hadmit + hpatient + hpadmit, 
                  df_dia , family = "binomial")
## Print the summary of the model
coef(summary(full_model)) 
```

- Step 2: Remove no sig variables
```{r}
## Step 3:  
model5 <- glm(diabetesMed ~ HbA1c + patient_visits + hpatient,
              df_dia, family = "binomial")

## Print the coefficients of the model
coef(summary(model5))
```
`equation with exposure, outcome and interaction`
```{r}
# 
# y=0.9940305 +0.9267593*HbA1c1           +0.1340291*patient_visits  - 0.1322058* HbA1c1*patient_visits
# y0=0.9940305 +0.9267593*0           +0.1340291*patient_visits  - 0.1322058* 0*patient_visits
# y1=0.9940305 +0.9267593*1           +0.1340291*patient_visits  - 0.1322058* 1*patient_visits

```

- Plot interaction of exposure and patient_visits in log odds
```{r}
curve( (0.9940305 +0.9267593*0           +0.1340291*x  - 0.1322058* 0*x), from=1, to=20, n=300, xlab="patient_visits", ylab="log odds", 
             col="red", lwd=2, main="interaction of exposure and patient_visits"  )

curve( (0.9940305 +0.9267593*1           +0.1340291*x  - 0.1322058* 1*x), from=1, to=20, n=300, add=TRUE, xlab="patient_visits", ylab="log odds", 
             col="blue", lwd=2, main="interaction of exposure and patient_visits"  )
```

- Plot interaction of exposure and patient_visits in odds
```{r}
curve( exp(0.9940305 +0.9267593*0           +0.1340291*x  - 0.1322058* 0*x), from=1, to=20, n=300, xlab="patient_visits", ylab="odds", 
             col="red", lwd=2, main="interaction of exposure and patient_visits"  )

curve( exp(0.9940305 +0.9267593*1           +0.1340291*x  - 0.1322058* 1*x), from=1, to=20, n=300, add=TRUE, xlab="patient_visits", ylab="odds", 
             col="blue", lwd=2, main="interaction of exposure and patient_visits"  )
```

- Plot interaction of exposure and patient_visits in odds ratio
```{r}
curve(  exp(0.9940305 +0.9267593*0           +0.1340291*x  - 0.1322058* 0*x)/ exp(0.9940305 +0.9267593*1           +0.1340291*x  - 0.1322058* 1*x), from=1, to=20, n=300,  xlab="patient_visits", ylab="OR", 
             col="blue", lwd=2, main="interaction of exposure and patient_visits"  )
```

This suggests that patient visits modify the outcome-exposure relationship, while others do not modify the relationship between HbA1c levels (exposure) and diabetic medication change (outcome). a covariate cannot confound and modify the exposure-outcome relationship simultaneously. 

- Get the confidence interval and OR
```{r}
## Get the confidence interval
CI <- confint(model5)
CI

## Find the odds ratio
Odds_ratios <- exp(model5$coefficients)
Odds_ratios

```

 




