---
title: "Poisson regression"
description: "A practical R learning guide to Poisson regression, with concise explanations, reproducible examples, and interpretation guidance."
date: 2026-08-31
section: "Biostatistics"
category: "Statistical Models"
topic: "Poisson regression"
type: article
tags: ["Statistical Models", "Biostatistics", "Poisson", "regression"]
status: published
featured: true
youtube: []
pdf: []
downloads: []
validate-yaml: false
execute:
  echo: true
  warning: false
  error: true
---

## Why This Topic Matters

A statistical model is a structured approximation that links observed data to an estimand or prediction target.

## Learning Objectives

After working through this material, you should be able to:

- explain the main ideas behind **Poisson regression**;
- interpret the R workflow and its output; and
- adapt the examples to a related analysis.

## Lesson Roadmap

The original material is organized around the following topics:

1. Poisson regression
2. Dependent variable is count
3. Dependent variable is rate (count/population)

## How to Use This Lesson

Read the explanation before each workflow, then run the R chunks in sequence. Code and output are shown together so that each numerical or graphical result can be connected to the command that produced it. Some source examples require optional packages or external data. Those setup-dependent chunks remain visible for reference but are marked `eval: false` when they cannot be reproduced safely in a clean website build.

## Learning Material

The following material develops **Poisson regression** from core ideas to practical implementation. Definitions and assumptions come first; worked code then shows how the ideas are represented computationally.

### Conceptual Foundation

The material below closely follows the original R Markdown source, with its examples retained for study and reuse.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Poisson regression

<!-- generated-learning-note -->
This section develops **Poisson regression**. Read the model specification together with its assumptions and the scale of each reported effect.



$$
\mathrm{P}(Y=y \mid \lambda)=\frac{e^{-\lambda} \lambda^y}{y !}
$$

**Poisson distribution not equal Poisson regression**

### R Workflow and Examples

Run the examples in order because later blocks may reuse objects created earlier. Each block includes a short purpose statement and guidance for reading its output.


<!-- generated-learning-note -->
**Purpose.** This block creates a graphical summary. The plot should be read together with its axes, scales, groups, and any representation of uncertainty.

```{r}
# numbers from  
x<-seq(0,30,1)

# calculate the distribution function
# based on the parameters
pdf1<- dpois(x, 1)
pdf2<- dpois(x, 7)
pdf3<- dpois(x, 12)

# Plotting the PDF 
plot(x,pdf1,type = "l")
lines(x,pdf2 ,col=2)
lines(x,pdf3 ,col=3)
```

<!-- generated-learning-note -->
**How to read the output.** Look for the intended pattern, but also inspect sparse groups, outliers, scale choices, and uncertainty before generalizing from the figure.


Negative binomial regression is a popular generalization of Poisson regression because it loosens the highly restrictive assumption that the variance is equal to the mean made by the Poisson model. The traditional negative binomial regression model is based on the Poisson-gamma mixture distribution. This model is popular because it models the Poisson heterogeneity with a gamma distribution.

Poisson regression is typically used to model count data. But, sometimes, it is more relevant to model rates instead of counts. This is relevant when individuals are not followed the same amount of time. For example, six cases over 1 year should not amount to the same as six cases over 10 years.





In Poisson regression the dependent variable $(Y)$ is an observed count that follows the Poisson distribution. The rate $\lambda$ is determined by a set of $k$ predictors $\mathbf{X}=\left(X_1, \ldots, X_k\right)$. 


since

$$
\mu_x=\lambda=\exp \{\mathbf{X} \beta\} .
$$

$$
\log \mu_x=\beta_0+\beta_1 x
$$

(where $\mu_x$ is the expected count for those with covariate $x$ )


for outcome is rate not count

$$
\log \frac{\mu_x}{t_x}=\beta_0^{\prime}+\beta_1^{\prime} x
$$

(where $t_x$ is the exposure time for those with covariate $x$ ). Now, the last equation could be rewritten

$$
\log \mu_x=\log t_x+\beta_0^{\prime}+\beta_1^{\prime} x
$$

and $\log t_x$ plays the role of an offset.





#### Dependent variable is count

<!-- generated-learning-note -->
This section prepares the information needed for **Dependent variable is count**. Check variable definitions, units, coding, and missing values before moving to analysis.




<!-- generated-learning-note -->
**Purpose.** This source block depends on optional software, external data, or a machine-specific path. It is retained as a reference and is not executed during the website build.

```{r}
#| eval: false
#| code-fold: true
# example
p <- read.csv("https://stats.idre.ucla.edu/stat/data/poisson_sim.csv")
p <- within(p, {
  prog <- factor(prog, levels=1:3, labels=c("General", "Academic", 
                                            "Vocational"))
  id <- factor(id)
})

head(p)

hist(p$num_awards,probability = T)
lines(density(p$num_awards))

summary(m1 <- glm(num_awards ~ prog + math, family="poisson", data=p))
```

<!-- generated-learning-note -->
**How to read the output.** Because this block is not executed, it produces no website output. Enable it only after its dependencies and input paths have been validated.



#### Dependent variable is rate (count/population)

<!-- generated-learning-note -->
This section prepares the information needed for **Dependent variable is rate (count/population)**. Check variable definitions, units, coding, and missing values before moving to analysis.


rate data can be modeled by including the log(n) term with coefficient of 1. This is called an offset. 


<!-- generated-learning-note -->
**Purpose.** This block loads the packages and supporting tools used by the following analysis. Package versions can affect numerical results and printed output.

```{r}
library(ISwR)
data(eba1977)
cancer.data = eba1977
head(cancer.data)

# histogram count
hist(cancer.data$case,probability = T)
lines(density(cancer.data$case))

# histogram rate
hist(cancer.data$case/cancer.data$pop,probability = T)
lines(density(cancer.data$case/cancer.data$pop))

# using offset
poisson.model.rate <- glm(cases ~ city + age+ offset(log(pop)), family = poisson(link = "log"), data = cancer.data)

#display summary
summary(poisson.model.rate)

```

<!-- generated-learning-note -->
**How to read the output.** A successful setup block is usually silent. Warnings about missing packages or version conflicts should be resolved before trusting later output.


- Interpretation 

Compare to 40-54, age55-59 group increases the number of cases by exp (1.1)=3, while controlling for the effect of city.

## Interpreting the Results

Interpret coefficients with their coding and scale, examine uncertainty, and use diagnostics to evaluate assumptions. Treat the output as evidence generated under the stated data, coding choices, and assumptions. Before using the workflow in practice, replace example inputs with validated data and confirm that the method targets the intended scientific question.

## Key Takeaways

- **Connect method to purpose.** Identify the question and target quantity before choosing code or a model.
- **Read code and output together.** Confirm what each object, estimate, interval, table, or plot represents.
- **Check assumptions and diagnostics.** A successful computation is not by itself evidence that the result is reliable.
- **Adapt deliberately.** When applying the example elsewhere, document changes to data definitions, parameters, and analytical choices.
